---
title: |
  | Practical assignments during the class
subtitle: "LGN 5822 - Statistical Genetics I: Foundations"
author:
  - "Bruna Marques Moreno"
institute: |
  | University of Sao Paulo
  | "Luiz de Quieroz" College of Agriculture
  | Departament of Genetics
  | Graduate Program in Genetics and Plant Breeding
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 2
    number_sections: true
    fig_width: 8
    fig_height: 8
    fig_retina: 2
    fig_caption: true
    df_print: kable
    keep_md: true
---

## Activity of the week (3^rd^ class)

Using your USP number as seed, simulate phenotypic values (don’t forget, P=G+E) using the normal distribution (`rnorm`). Then, compute the mean and the variance of P, G and E, and build a histogram, a density plot, a box plot and a scatter plot depicting the relationship between P and G. Do it for three different sample sizes: 10, 100 and 1000.

### Sample size 10

```{r}
#| warning: false
library(dplyr)
set.seed(8970011)

gen_10 <- rnorm(n=10, mean = 2000, sd = 276)
env_10 <- rnorm(n=10, mean = 859, sd = 165)
pheno_10 <- gen_10 + env_10

dados <- c(gen_10, env_10, pheno_10)
dat_10 <- data.frame("class" = c(rep("Gen", 10), rep("Env", 10), rep("Pheno", 10)),
                     "valores" = dados)

df_10 <- data.frame("Gen" = gen_10,
                    "Env" = env_10,
                    "Pheno" = pheno_10)

mean <- tapply(dat_10$valores, dat_10$class, mean)

var <- tapply(dat_10$valores, dat_10$class, var)
var
```

### Graphs sample size 10

```{r}
#| warning: false
library(ggplot2)

# Histogram
ggplot(data = dat_10, aes(x = valores, fill = class)) +
  geom_histogram()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotypic value", color = "Mean Genotypic value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environmental value", color = "Mean Environmental value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotypic value", color = "Mean Phenotypic value"), linewidth = 1)+
  labs(linetype = "Parameter", color = "Parameter")

# Density
ggplot(data = dat_10, aes(x = valores, fill = class)) +
  geom_density()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotype", color = "Mean Genotype"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environment", color = "Mean Environment"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotype", color = "Mean Phenotype"), linewidth = 1)

# Boxplot
ggplot(data = dat_10, aes(x = class, y= valores, fill = class)) +
  geom_boxplot() 

ggplot(data = dat_10, aes(x = class, y= valores, fill = class)) +
  geom_violin()

# Scatter plot
ggplot(data = df_10, aes(x = Gen, y = Pheno)) +
  geom_jitter(size = 3, alpha = 0.7, width = 0.2, color = "orange") +
  labs(x = "Genotype",
       y = "Phenotype",
       subtitle = paste("Correlação de Pearson =", round(cor(df_10$Gen, df_10$Pheno), 4)))
```

### Sample size 100

```{r}
#| warning: false
library(dplyr)
set.seed(8970011)

gen_100 <- rnorm(n=100, mean = 2000, sd = 276)
env_100 <- rnorm(n=100, mean = 859, sd = 165)
pheno_100 <- gen_100 + env_100

dados <- c(gen_100, env_100, pheno_100)
dat_100 <- data.frame("class" = c(rep("Gen", 100), rep("Env", 100), rep("Pheno", 100)),
                     "valores" = dados)

df_100 <- data.frame("Gen" = gen_100,
                    "Env" = env_100,
                    "Pheno" = pheno_100)

mean <- tapply(dat_100$valores, dat_100$class, mean)

var <- tapply(dat_100$valores, dat_100$class, var)
var
```

### Graphs sample size 100

```{r}
#| warning: false
library(ggplot2)

# Histogram
ggplot(data = dat_100, aes(x = valores, fill = class)) +
  geom_histogram()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotypic value", color = "Mean Genotypic value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environmental value", color = "Mean Environmental value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotypic value", color = "Mean Phenotypic value"), linewidth = 1)+
  labs(linetype = "Parameter", color = "Parameter")

# Density
ggplot(data = dat_100, aes(x = valores, fill = class)) +
  geom_density()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotype", color = "Mean Genotype"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environment", color = "Mean Environment"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotype", color = "Mean Phenotype"), linewidth = 1)

# Boxplot
ggplot(data = dat_100, aes(x = class, y= valores, fill = class)) +
  geom_boxplot() 

ggplot(data = dat_100, aes(x = class, y= valores, fill = class)) +
  geom_violin()

# Scatter plot
ggplot(data = df_100, aes(x = Gen, y = Pheno)) +
  geom_jitter(size = 3, alpha = 0.7, width = 0.2, color = "orange") +
  labs(x = "Genotype",
       y = "Phenotype",
       subtitle = paste("Correlação de Pearson =", round(cor(df_100$Gen, df_100$Pheno), 4)))
```

### Sample size 1000

```{r}
#| warning: false
library(dplyr)
set.seed(8970011)

gen_1000 <- rnorm(n=1000, mean = 2000, sd = 276)
env_1000 <- rnorm(n=1000, mean = 859, sd = 165)
pheno_1000 <- gen_1000 + env_1000

dados <- c(gen_1000, env_1000, pheno_1000)
dat_1000 <- data.frame("class" = c(rep("Gen", 1000), rep("Env", 1000), rep("Pheno", 1000)),
                     "valores" = dados)

df_1000 <- data.frame("Gen" = gen_1000,
                    "Env" = env_1000,
                    "Pheno" = pheno_1000)

mean <- tapply(dat_1000$valores, dat_1000$class, mean)

var <- tapply(dat_1000$valores, dat_1000$class, var)
var
```

### Graphs sample size 100

```{r}
#| warning: false
library(ggplot2)

# Histogram
ggplot(data = dat_1000, aes(x = valores, fill = class)) +
  geom_histogram()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotypic value", color = "Mean Genotypic value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environmental value", color = "Mean Environmental value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotypic value", color = "Mean Phenotypic value"), linewidth = 1)

ggplot(data = dat_1000, aes(x = valores, fill = class)) +
  geom_histogram()+
  facet_grid(class ~ . , scale = "free_y")+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotypic value", color = "Mean Genotypic value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environmental value", color = "Mean Environmental value"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotypic value", color = "Mean Phenotypic value"), linewidth = 1)

# Density
ggplot(data = dat_1000, aes(x = valores, fill = class)) +
  geom_density()+
  geom_vline(aes(xintercept = mean["Gen"], linetype = "Mean Genotype", color = "Mean Genotype"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Env"],  linetype = "Mean Environment", color = "Mean Environment"), linewidth = 1) +
  geom_vline(aes(xintercept = mean["Pheno"],  linetype = "Mean Phenotype", color = "Mean Phenotype"), linewidth = 1)

# Boxplot
ggplot(data = dat_1000, aes(x = class, y= valores, fill = class)) +
  geom_boxplot() 

ggplot(data = dat_1000, aes(x = class, y= valores, fill = class)) +
  geom_violin()

# Scatter plot
ggplot(data = df_1000, aes(x = Gen, y = Pheno)) +
  geom_jitter(size = 3, alpha = 0.7, width = 0.2, color = "orange") +
  labs(x = "Genotype",
       y = "Phenotype",
       subtitle = paste("Correlação de Pearson =", round(cor(df_1000$Gen, df_1000$Pheno), 4)))
```
