---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true

crossref: {}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)
library(car)

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

# Exercise 
Choose two traits. Perform data analysis of all trials in your dataset using a RCBD model. Justify your choice in using the treatment and the block effects as fixed or random. Report the variance component estimates, the hypothesis test results, the broad-sense heritability, the coefficient of experimental variation, and the selection candidates' **BLUPs and reliabilities**. Finally, compare the rankings between trials and speculate the reasons why they are different/similar. This task must be carried out within groups. 


# Introduction


# Trait \#1 - GY



## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation


# Trait \#2 - PH
## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

# Selection of candidates


# Discussion


A fazer

- filtrar os outliers do dataframe total para as duas características
- ajustar o modelo para todos os environments (modelo = $$ y = env + env/bloco + gen + gen*env + irrigação + \epsilon$$) - sendo env, gen e gen*env como aleatório.
- tabela para mostrar os locais nos anos que escolhemos para fazer (2) - estabilidade/adaptabilidade
- ajustar o modelo para cada local ao longo dos anos.
- ajustar o modelo para um ano qual é o melhor local.



se der tempo: estrutura de covariância entre anos/locais para identificar as correlações entre anos. Quanto mais próximos, maior. Tem covariância entre os anos.