---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml")
invisible(lapply(packages, library, character.only = TRUE))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: outliers_PH

# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE) + 
               c(-3, 3) * sd(data$PH_R8, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = PH_R8, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))
```

```{r}
#| label: outliers_GY

# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removendo outlier
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Ajustando colunas
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

USAR O DF COMO BASE DE DADOS!!! ELE JÁ ESTÁ FILTRADO

# Exercise

Choose two traits. Perform data analysis of all trials in your dataset
using a RCBD model. Justify your choice in using the treatment and the
block effects as fixed or random. Report the variance component
estimates, the hypothesis test results, the broad-sense heritability,
the coefficient of experimental variation, and the selection candidates'
**BLUPs and reliabilities**. Finally, compare the rankings between
trials and speculate the reasons why they are different/similar. This
task must be carried out within groups.

# Introduction

This report provides a data analysis of a multi-environment soybean breeding in Zambia dataset, focusing on two agronomic traits **Grain Yield** and **Plant Height**. Utilizing a Randomized Complete Block Design (RCBD) framework, all of the 33 individual trials were evaluated through mixed model methodology. Beyond estimation of  variance components, statistical hypothesis tests, and experimental quality were audited using broad-sense heritability ($H^{2}$) and the coefficient of variation (CV%). Furthermore, selected candidates were evaluated based on their Best Linear Unbiased Predictions (BLUPs) and respective reliability values. Finally,  a cross-trial ranking comparison is presented, providing a critical speculation on the underlying genetic and environmental factors that govern the differences or similarities in genotype performance across locations.

# Trait 1 - GY

Our dataset comprised 33 trials conducted in Zambia. The experiments were arranged in a Randomized Complete Block Design (RCBD) with 3 blocks and approximately 30 genotypes per trial. Not all genotypes were present across all trials. Furthermore, because some experiments had missing plots over the course of the evaluation, we implemented the following mixed model:
$$y_{ij}=b_{j}+g_{i}+e_{ij}$$
Where:
$b_{j}$ was considered a fixed effect because there are only 3 blocks within each experiment.
$g_{i}$ was considered a random effect, where $g_i \sim N(0, \sigma^2_g)$, since the genotypes are samples from a larger breeding population. Additionally, modeling genotypes as random allowed us to recover information and account for the unbalanced nature of the data.

```{r}
#| label: GY_analysis

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")


# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt.asreml(model, m_red_GY, boundary = T)
  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

# Summary of Grain Yield Model Analysis 

```{r}
#| label: tbl-gy
#| tbl-cap: Summary 33 soybean breeding trials in Zambia evaluating Grain Yield (GY)

tabela_geral_GY |> 
  mutate(significance = case_when(
    `pvalue` < 0.01 ~ "**",
    `pvalue` < 0.05 ~ "*",
    TRUE             ~ "ns" 
  )) |> 
  kable(
    align = "c",
    col.names = c(
      "Trial", 
      "LRT statistic", 
      "p-value", 
      "Significance", 
      "Broad-sense heritability (H²)", 
      "Coefficient of Variation (%)"
    )
  ) |> 
  kable_styling(bootstrap_options = "striped", full_width = TRUE) |> 
  footnote(
    general = c("* - p < 0.05 (5%);", "** - p < 0.01 (1%);", "ns - non-significant."),
    general_title = "Significance levels: ")


```

For Grain Yield (GY) it is visible that the trials `E0108`, `E0174`, `E0175`, `E0224`, `E0280`, and `E0285` did not present variance difference, that is, the test did not rejected $H_0$ which means the genetic variance is equal to zero.

# BLUPs and Reability

In this section we can observe the predicted value of each genotype in each trial individually. There are some environments that have more genotypes or different genotypes. Also, observing the standard errors it is possible to identify what trials are unbalanced.

```{r}
#| label: tbl-GY-blup-table
#| tbl-cap: Predicted values of Grain Yield (kg/ha) and standard errors for grain yield of soybean breeding genotypes evaluated across 33 trials in Zambia.

selected_trials <- names(singleTrial_blup_GY)
tabela_grande <- singleTrial_blup_GY[selected_trials] |> 
  map(~ .x |> select(gen, predicted.value, std.error)) |>
  reduce(full_join, by = "gen")

colunas_valores <- rep(c("Predicted value", "Stand. error"), length(selected_trials))

tabela_grande |> 
  kable(
    align = "c",
    col.names = c("Genotype", colunas_valores)
  ) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, setNames(rep(2, length(selected_trials)), selected_trials))) |> 
  scroll_box(width = "100%", height = "600px")

```

As shown in @top-genotypes-gy, filtering for trials with a CV under 30% and positive genetic variance resulted in the selection of 20 out of 33 trials. This high retention rate reflects the overall experimental quality of the breeding program.

```{r}
#| label: tbl-top-genotypes-gy
#| tbl-cap-location: top
#| tbl-cap: Predicted value of the top five soybean breeding genotypes ranked (1 to 5) in each trial in Zambia 

low_cv_envs <- tabela_geral_GY |> 
  select(env, h2, cv, significance) |> 
  filter(significance == "TRUE" & cv < 30) |> 
  arrange(cv)

envs <- low_cv_envs$env

genotipos_repetidos <- singleTrial_blup_GY[envs] |> 
  map_dfr(~ .x |> arrange(desc(predicted.value)) |> head(5) |> select(gen)) |> 
  count(gen) |> 
  filter(n > 1) |> 
  pull(gen) |> 
  as.character()

set.seed(42) 
cores_paleta <- hcl.colors(length(genotipos_repetidos), palette = "Earth") |> 
  paste0("80") 
mapa_cores <- setNames(cores_paleta, genotipos_repetidos)

tabela_longa <- envs |> 
  set_names() |> 
  map_dfr(~ {
    singleTrial_blup_GY[[.x]] |>  
      arrange(desc(predicted.value)) |>   
      head(5) |> 
      select(gen, predicted.value) |> 
      mutate(
        rank = c("1st", "2nd", "3rd", "4th", "5th"),
        gen = as.character(gen)
      )
  }, .id = "env")

tabela_larga <- tabela_longa |> 
  pivot_wider(
    id_cols = env, 
    names_from = rank, 
    values_from = c(gen, predicted.value),
    names_glue = "{rank}_{.value}" 
  ) |> 
  select(
    env,
    `1st_gen`, `1st_predicted.value`,
    `2nd_gen`, `2nd_predicted.value`,
    `3rd_gen`, `3rd_predicted.value`,
    `4th_gen`, `4th_predicted.value`,
    `5th_gen`, `5th_predicted.value`
  )

sub_colunas <- c("Trial", rep(c("Genotype", "Predicted value"), 5))

tabela_final <- kbl(tabela_larga, align = "c", col.names = sub_colunas) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, "1st" = 2, "2nd" = 2, "3rd" = 2, "4th" = 2, "5th" = 2))

colunas_indices <- c(2, 4, 6, 8, 10)

for (col in colunas_indices) {
  valores_coluna <- as.character(tabela_larga[[col]])

  marcar_negrito <- valores_coluna %in% genotipos_repetidos
  
  cores_fundo <- ifelse(marcar_negrito, mapa_cores[valores_coluna], "transparent")
  cores_fundo[is.na(cores_fundo)] <- "transparent"
  
  tabela_final <- tabela_final |> 
    column_spec(
      column = col, 
      bold = marcar_negrito,
      background = cores_fundo
    )
}

tabela_final
```

```{r}
#| label: tbl-freq-top-five
#| tbl-cap: Frequency of each genotype in the top five genotypes in each trial

# How many times a genotype appears in the table
frequencia_geral <- tabela_larga |>
  select(env, ends_with("_gen")) |> 
  pivot_longer(cols = ends_with("_gen"), values_to = "genotipo") |> 
  # Conta apenas quantas vezes o genótipo aparece no total
  count(genotipo, name = "total_ambientes") |> 
  arrange(desc(total_ambientes))

kable(frequencia_geral,
      col.names = c("Genotype", "Frequency")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The top two genotypes with the highest frequency of occurrence behave differently across the trials. Genotype `G092` exhibits a constant predicted value, whereas genotype `G223` shows variable performance. Despite both being top-ranked, their distinct predicted values across trials indicate genotype-by-environment (GxE) interactions that were not captured in this report.

```{r}
G233 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G233" ) |> arrange(rank)
kable(G233,
      align = "c",
      col.names = c("Genotype","Environment", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

G092 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G092" ) |> arrange(rank)
kable(G092,
      align = "c",
      col.names = c("Genotype","Environment", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Trait 2 - Plant Height (PH)  

```{r}
#| label: PH_analysis
#| output: false

# Removendo ambientes com nenhuma observação
df <- df |> filter(!env %in% c("E0171","E0172","E0225"))
envs_list <- envs_list[!envs_list %in% c("E0171","E0172","E0225")]

# Verificando dados faltantes para PH em cada ENV
info_env_PH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(PH_R8)),
    .groups = "drop")

# Rodando para cada trial isoladamente
singleTrial_models_PH <- list()
singleTrial_varcomp_PH <- list()
singleTrial_blup_PH <- list()
singleTrial_reliability_PH <- list()
tabela_geral_PH <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = PH_R8 ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_PH <- asreml(fixed = PH_R8 ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt.asreml(model, m_red_PH, boundary = T)
  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$PH, na.rm=T)
  
  tabela_geral_PH <- rbind(tabela_geral_PH,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_PH[[i]] <- model
  singleTrial_varcomp_PH[[i]] <- varcomp
  singleTrial_blup_PH[[i]] <- blups
  singleTrial_reliability_PH[[i]] <- reliability
  }
```

# Summary of Plant Height Model Analysis 

```{r}
#| label: tbl-ph
#| tbl-cap: Summary of the main genetic and statistical parameters for Plant Height (PL) across 33 trials of the soybean breeding program in Zambia.

tabela_geral_PH |> 
  mutate(significance = case_when(
    `pvalue` < 0.01 ~ "**",
    `pvalue` < 0.05 ~ "*",
    TRUE             ~ "ns" 
  )) |> 
  kable(
    align = "c",
    col.names = c(
    "Trial", 
    "LRT statistic", 
    "p-value", 
    "Significance", 
    "Broad-sense heritability (H²)", 
    "Coefficient of Variation (%)"
  )) |> 
  kable_styling(bootstrap_options = "striped", full_width = TRUE)

```

Regarding Plant Height (PH), trials `E0108`, `E0170`, and `E0230` were the only ones that did not present genetic variance above zero, meaning the Likelihood Ratio Test failed to reject the null hypothesis ($H_{0}$). This same behavior was observed for broad-sense heritability ($H^2$); in environments with no significant genetic variance, heritability was consequently zero or extremely close to it. Overall, heritability values varied from zero to 0.87. Furthermore, the coefficient of variation (CV) was low in general, indicating that the experiments were conducted with high experimental precision.

# BLUPs and Reability

In this section we can observe the predicted value of each genotype in each trial individually. There are some environments that have more genotypes or different genotypes. Also, observing the standard errors it is possible to identify what trials are unbalanced.

```{r}
#| label: tbl-ph-blup
#| tbl-cap: Predicted values (BLUPs) and standard errors for Plant Height (PH) of soybean breeding genotypes evaluated across 33 trials in Zambia.

selected_trials <- names(singleTrial_blup_PH)
tabela_grande <- singleTrial_blup_PH[selected_trials] |> 
  map(~ .x |> select(gen, predicted.value, std.error)) |>
  reduce(full_join, by = "gen")

colunas_valores <- rep(c("Predicted value", "Stand. error"), length(selected_trials))

tabela_grande |> 
  kable(
    align = "c",
    col.names = c("Genotype", colunas_valores)
  ) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, setNames(rep(2, length(selected_trials)), selected_trials))) |> 
  scroll_box(width = "100%", height = "600px")

```


```{r}
#| label: tbl-common-geno-ph
#| tbl-cap: Top five ranked soybean breeding genotype in Zambia 
low_cv_envs <- tabela_geral_PH |> 
  select(env, h2, cv, significance) |> 
  filter(significance == "TRUE" & cv < 30) |> 
  arrange(cv)

envs <- low_cv_envs$env

genotipos_repetidos <- singleTrial_blup_PH[envs] |> 
  map_dfr(~ .x |> arrange(desc(predicted.value)) |> head(5) |> select(gen)) |> 
  count(gen) |> 
  filter(n > 1) |> 
  pull(gen) |> 
  as.character()

set.seed(42) 
cores_paleta <- hcl.colors(length(genotipos_repetidos), palette = "Earth") |> 
  paste0("80") 
mapa_cores <- setNames(cores_paleta, genotipos_repetidos)

tabela_longa <- envs |> 
  set_names() |> 
  map_dfr(~ {
    singleTrial_blup_PH[[.x]] |>  
      arrange(desc(predicted.value)) |>   
      head(5) |> 
      select(gen, predicted.value) |> 
      mutate(
        rank = c("1st", "2nd", "3rd", "4th", "5th"),
        gen = as.character(gen)
      )
  }, .id = "env")

tabela_larga <- tabela_longa |> 
  pivot_wider(
    id_cols = env, 
    names_from = rank, 
    values_from = c(gen, predicted.value),
    names_glue = "{rank}_{.value}" 
  ) |> 
  select(
    env,
    `1st_gen`, `1st_predicted.value`,
    `2nd_gen`, `2nd_predicted.value`,
    `3rd_gen`, `3rd_predicted.value`,
    `4th_gen`, `4th_predicted.value`,
    `5th_gen`, `5th_predicted.value`
  )

sub_colunas <- c("Trial", rep(c("Genotype", "Predicted value"), 5))

tabela_final <- kbl(tabela_larga, align = "c", col.names = sub_colunas) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, "1st" = 2, "2nd" = 2, "3rd" = 2, "4th" = 2, "5th" = 2))

colunas_indices <- c(2, 4, 6, 8, 10)

for (col in colunas_indices) {
  valores_coluna <- as.character(tabela_larga[[col]])

  marcar_negrito <- valores_coluna %in% genotipos_repetidos
  
  cores_fundo <- ifelse(marcar_negrito, mapa_cores[valores_coluna], "transparent")
  cores_fundo[is.na(cores_fundo)] <- "transparent"
  
  tabela_final <- tabela_final |> 
    column_spec(
      column = col, 
      bold = marcar_negrito,
      background = cores_fundo
    )
}

tabela_final
```

In the table above the genotypes in bold and colored are repeated at least once in this top five rank in each trial. It is interesting to observe that for Plant Height, the same genotypes were repeated more frequently compared to Grain Yield. Genotypes `G092` and `G233` seem to be promising, since they appear frequently across both traits. Probably, these materials can be considered stable, meaning they present a consistent behavior across different environments.

```{r}
#| label: tbl-freq-top-five-ph
#| tbl-cap: Frequency of each soybean breeding genotype in the top five genotypes in each trial in Zambia.

frequencia_geral <- tabela_larga |>
  select(env, ends_with("_gen")) |> 
  pivot_longer(cols = ends_with("_gen"), values_to = "genotipo") |> 
  count(genotipo, name = "total_ambientes") |> 
  arrange(desc(total_ambientes))

kable(frequencia_geral,
      col.names = c("Genotype", "Frequency")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
#| label: tbl-common-geno
#| tbl-cap: Common genotypes

G146 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G146" ) |> arrange(rank)
kable(G146,
      align = "c",
      col.names = c("Genotype","Trial", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

G244 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G244" ) |> arrange(rank)
kable(G244,
      align = "c",
      col.names = c("Genotype","Trial", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

As observed for the previous trait, Plant Height (PH) values varied considerably among trials. For example, genotype `G244` reached a height of 60 cm and was top-ranked in trial`E0283`. Genotype `G146`had similar results. These results highlight the need to include genotype-by-environment (GxE) interactions in the model. This is evidenced by the fact that a single genotype can exhibit superior performance in specific environments while performing poorly in others, despite its high overall ranking.

# Conclusion

