---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml")
invisible(lapply(packages, library, character.only = TRUE))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: outliers_PH

# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE) + 
               c(-3, 3) * sd(data$PH_R8, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = PH_R8, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))
```

```{r}
#| label: outliers_GY

# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removendo outlier
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Ajustando colunas
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

USAR O DF COMO BASE DE DADOS!!! ELE JÁ ESTÁ FILTRADO

# Exercise

Choose two traits. Perform data analysis of all trials in your dataset
using a RCBD model. Justify your choice in using the treatment and the
block effects as fixed or random. Report the variance component
estimates, the hypothesis test results, the broad-sense heritability,
the coefficient of experimental variation, and the selection candidates'
**BLUPs and reliabilities**. Finally, compare the rankings between
trials and speculate the reasons why they are different/similar. This
task must be carried out within groups.

# Introduction

Blocos como fixo (poucos blocos) e genotipos como aleatorio (amostras de
uma possivel populacao dos cruzamentos, recuperacao da info em alguns
casos de desbalanceamento.

# Trait 1 - GY

```{r}
#| label: GY_analysis

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")


# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt.asreml(model, m_red_GY, boundary = T)
  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

# Trait 2 - PH  

```{r}
#| label: PH_analysis

# Removendo ambientes com nenhuma observação
df <- df |> filter(!env %in% c("E0171","E0172","E0225"))
envs_list <- envs_list[!envs_list %in% c("E0171","E0172","E0225")]

# Verificando dados faltantes para PH em cada ENV
info_env_PH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(PH_R8)),
    .groups = "drop")

# Rodando para cada trial isoladamente
singleTrial_models_PH <- list()
singleTrial_varcomp_PH <- list()
singleTrial_blup_PH <- list()
singleTrial_reliability_PH <- list()
tabela_geral_PH <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = PH_R8 ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_PH <- asreml(fixed = PH_R8 ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt.asreml(model, m_red_PH, boundary = T)
  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$PH, na.rm=T)
  
  tabela_geral_PH <- rbind(tabela_geral_PH,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_PH[[i]] <- model
  singleTrial_varcomp_PH[[i]] <- varcomp
  singleTrial_blup_PH[[i]] <- blups
  singleTrial_reliability_PH[[i]] <- reliability
  }
```

## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

# Discussion

\

\
