---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

Choose two traits. Perform data analysis of all trials in your dataset
using a RCBD model. Justify your choice in using the treatment and the
block effects as fixed or random. Report the variance component
estimates, the hypothesis test results, the broad-sense heritability,
the coefficient of experimental variation, and the selection candidates'
BLUPs and reliabilities. Finally, compare the rankings between trials
and speculate the reasons why they are different/similar. This task must
be carried out within groups.

# Introduction

This report provides a data analysis of a multi-environment soybean
breeding in Zambia dataset, focusing on two agronomic traits **Grain
Yield** and **Plant Height**. Utilizing a Randomized Complete Block
Design (RCBD) framework, all of the 33 individual trials were evaluated
through mixed model methodology. Beyond estimation of variance
components, statistical hypothesis tests, and experimental quality were
audited using broad-sense heritability ($H^{2}$) and the coefficient of
variation ($CV%$). Furthermore, selected candidates were evaluated based
on their Best Linear Unbiased Predictions (BLUPs) and respective
reliability values. Finally, a cross-trial ranking comparison is
presented, providing a critical speculation on the underlying genetic
and environmental factors that govern the differences or similarities in
genotype performance across locations.

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: fig-outliers-PH
#| fig-cap: Exploratory data analysis of Plant Height (PH) showcasing distribution behavior and outlier detection. The vertical dotted orange lines and the boxplot's red markers identify the extreme values candidate for filtering or transformation prior to subsequent statistical modeling.

# Normal Distribution Sd's
ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE) + 
               c(-3, 3) * sd(data$PH_R8, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = PH_R8, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))
```

```{r}
#| label: fig-outliers-GY
#| fig-cap: Exploratory data analysis of Grain Yield (GY) showcasing distribution behavior and outlier detection. The vertical dotted orange lines and the boxplot's red markers identify the extreme values candidate for filtering or transformation prior to subsequent statistical modeling.

# Normal Distribution Sd's
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removing outliers
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Converting columns
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

# Statistical Model

Our dataset comprised 33 trials conducted in Zambia. The experiments
were arranged in a Randomized Complete Block Design (RCBD) with 3 blocks
and approximately 30 genotypes per trial. Not all genotypes were present
across all trials.

For both traits analysed, the statistical model used is describe as
follow:

$$y_{ij}=b_{j}+g_{i}+e_{ij}$$

Where:

$b_{j}$ was considered a fixed effect because there are only 3 blocks
within each experiment.

$g_{i}$ was considered a random effect, where
$g_i \sim N(0, \sigma^2_g)$, since genotypes are samples from a larger
breeding population. Additionally, modeling genotypes as random allowed
us to recover information and account for the unbalanced nature of the
data, as some trials had missing plots.

$e_{ij}$ was considered a random effect, where
$e_{ij} \sim N(0, \sigma^2_e)$.

As most trials are completely balanced, the broad-sense heritability was
estimated using the simple variance components ratio:

$$H² = \frac{\sigma²_g}{( \sigma²_g + \sigma²_e)}$$

# Trait 1 - GY

```{r}
#| label: GY_analysis
#| output: false

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")

library(asreml)

# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_GY, boundary = T)  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

# Summary of Grain Yield Model Analysis

```{r}
#| label: tbl-gy
#| tbl-cap: Agronomic summary of 33 soybean breeding trials for Grain Yield (GY) in Zambia

tabela_geral_GY |> 
  mutate(significance = case_when(
    `pvalue` < 0.01 ~ "**",
    `pvalue` < 0.05 ~ "*",
    TRUE             ~ "ns" 
  )) |> 
  kable(
    align = "c",
    col.names = c(
      "Trial", 
      "LRT statistic", 
      "p-value", 
      "Significance", 
      "Broad-sense heritability (H²)", 
      "Coefficient of Variation (%)"
    )
  ) |> 
  kable_styling(bootstrap_options = "striped", full_width = TRUE) |> 
  footnote(
    general = c("* - p < 0.05 (5%);", "** - p < 0.01 (1%);", "ns - non-significant."),
    general_title = "Significance levels: ")


```

For Grain Yield (GY) it is visible that the trials `E0108`, `E0174`,
`E0175`, `E0224`, `E0280`, and `E0285` did not present significant
genetic variance, that is, the test did not rejected $H_0$, indicating
that the genetic variance was not significantly different from zero.

# BLUPs and Reability

In this section we can observe the predicted value of each genotype in
each trial individually. There are some environments that have more
genotypes or different genotypes. Also, observing the standard errors it
is possible to identify what trials are unbalanced.

```{r}
#| label: fig-create-gy-plots
#| fig-cap: Predicted values for Grain Yield (GY) and respective genotype rankings across individual trials. Each bar chart displays the performance rank within a specific trial.
#| echo: false

graficos_GY <- list()

for (ambiente in names(singleTrial_blup_GY)) {
  
  dados_blup <- singleTrial_blup_GY[[ambiente]]
  
  p <- ggplot(data = dados_blup, aes(x = reorder(gen, -predicted.value), y = predicted.value)) +
    geom_col(color = "white", fill = "#10342d") + 
    
    geom_errorbar(aes(ymin = predicted.value - std.error, ymax = predicted.value + std.error), width = 0.2) +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          text = element_text(size = 8)) +  
    labs(title = paste("BLUPs for Grain Yield (GY) - Trial:", ambiente),
         x = "Genotype", 
         y = "BLUP (Predicted Value)")
  
  graficos_GY[[ambiente]] <- p
}
```

Here, the BLUPs for Grain Yield (GY) across the trials are presented.
The bar charts (see the panel tabset below) illustrate the genotype
ranking within each environment. The error bars represent the standard
error of the prediction, providing a visual assessment of the
reliability of the estimates in each specific trial.

```{r}
#| label: fig-graficos-gy
#| results: asis
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: Predicted values for Grain Yield (GY) and respective genotype rankings across individual trials. Each bar chart displays the performance rank within a specific trial.


# Imprimindo gráficos
grupos <- split(graficos_GY,ceiling(seq_along(graficos_GY) / 4))
for(i in seq_along(grupos)){
  painel <- wrap_plots(grupos[[i]],ncol = 2,nrow = 2)
  print(painel)
}
```



```{r}
#| label: fig-blup_by_envs_plots_GY
#| fig-cap: Genotypic performance and temporal stability. The line graph illustrates the BLUP-predicted values (Y-axis) for the grain yield (GY) trait across the 2019 to 2023 period. The horizontal axis discriminates the different testing environments exibiting the genotype-by-environment (G×E) interaction of the soybean breeding genotypes along the years.

blup_all_GY <- bind_rows(
  lapply(names(singleTrial_blup_GY), function(env){
    df <- singleTrial_blup_GY[[env]]
    df$env <- env
    return(df)}))

# subset_1 <- blup_all_GY |> filter(env %in% c("E0252","E0253","E0254","E0255","E0278","E0279","E0280","E0281","E0282","E0283","E0284","E0285"))
#   
# subset_2 <- blup_all_GY |> filter(!env %in% c("E0252","E0253","E0254","E0255","E0278","E0279","E0280","E0281","E0282","E0283","E0284","E0285"))
# 
# ggplot(subset_1,
#        aes(x = env, y = predicted.value, group = gen, color = gen)) +
#       geom_line(alpha = 0.3) +
#       geom_point(size = 1) +
#       theme_minimal() +
#       labs(x = "Environment", y = "Predicted BLUP (GY)",
#         title = "2022 - 2023 Genotype's Predicted Values Across Diferent Environments") +
#       theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
# 
# ggplot(subset_2,
#        aes(x = env, y = predicted.value, group = gen, color = gen)) +
#       geom_line(alpha = 0.3) +
#       geom_point(size = 1) +
#       theme_minimal() +
#       labs(x = "Environment", y = "Predicted BLUP (GY)",
#         title = "2019 - 2021 Genotype's Predicted Values Across Diferent Environments") +
#       theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

ggplot(blup_all_GY,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (GY)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```

The visual analysis of the BLUPs in each individual trials reveals a
dynamic performance behavior among the genotypes. The fluctuation in the
top-ranking positions highlights the presence of a strong Genotype by
Environment (GxE) interaction. Also, this can be seen in
@fig-blup_by_envs_plots_GY, by the cross-over patterns in the different
lines.

While the ranking order shifts depending on the trial, the consistent
appearance of certain materials, such as genotypes `G092` and `G233`, at
the top of the distribution suggests that these lines possess a degree
of wide adaptation and stability across different Zambian environments.

The variation in the magnitude of the error bars across the charts
reflects the distinct experimental precisions and levels of
unbalanceness among the trials. However, by relying on mixed-model
methodology, the information recovery can provide a more precise
estimate of these genetic values.

For a comprehensive numerical reference, the predicted values and
standard errors for all genotypes across all environments are detailed
in @tbl-GY-blup-table.

```{r}
#| label: tbl-GY-blup-table
#| tbl-cap: Predicted values of Grain Yield (kg/ha) and standard errors for grain yield of soybean breeding genotypes evaluated across 33 trials in Zambia.

selected_trials <- names(singleTrial_blup_GY)
tabela_grande <- singleTrial_blup_GY[selected_trials] |> 
  map(~ .x |> select(gen, predicted.value, std.error)) |>
  reduce(full_join, by = "gen")

colunas_valores <- rep(c("Predicted value", "Stand. error"), length(selected_trials))

tabela_grande |> 
  kable(
    align = "c",
    col.names = c("Genotype", colunas_valores)
  ) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, setNames(rep(2, length(selected_trials)), selected_trials))) |> 
  scroll_box(width = "100%", height = "600px")

```

As shown in @tbl-top-genotypes-gy, filtering for trials with a CV under 30% and positive genetic variance resulted in the selection of 20 out of 33 trials. This high retention rate reflects the solid experimental quality of the breeding program. Furthermore, the highlighted genotypes indicate those that recur across different trials or maintain top positions within the ranking, showcasing their high yield potential and stability across environments in Zambia.

```{r}
#| label: tbl-top-genotypes-gy
#| tbl-cap-location: top
#| tbl-cap:  Predicted value of Grain Yield (GY) of the top five soybean breeding genotypes ranked (1 to 5) in each trial in Zambia 

low_cv_envs <- tabela_geral_GY |> 
  select(env, h2, cv, significance) |> 
  filter(significance == "TRUE" & cv < 30) |> 
  arrange(cv)

envs <- low_cv_envs$env

genotipos_repetidos <- singleTrial_blup_GY[envs] |> 
  map_dfr(~ .x |> arrange(desc(predicted.value)) |> head(5) |> select(gen)) |> 
  count(gen) |> 
  filter(n > 1) |> 
  pull(gen) |> 
  as.character()

set.seed(42) 
cores_paleta <- hcl.colors(length(genotipos_repetidos), palette = "Greens 2") |> 
  paste0("80") 
mapa_cores <- setNames(cores_paleta, genotipos_repetidos)

tabela_longa <- envs |> 
  set_names() |> 
  map_dfr(~ {
    singleTrial_blup_GY[[.x]] |>  
      arrange(desc(predicted.value)) |>   
      head(5) |> 
      select(gen, predicted.value) |> 
      mutate(
        rank = c("1st", "2nd", "3rd", "4th", "5th"),
        gen = as.character(gen)
      )
  }, .id = "env")

tabela_larga <- tabela_longa |> 
  pivot_wider(
    id_cols = env, 
    names_from = rank, 
    values_from = c(gen, predicted.value),
    names_glue = "{rank}_{.value}" 
  ) |> 
  select(
    env,
    `1st_gen`, `1st_predicted.value`,
    `2nd_gen`, `2nd_predicted.value`,
    `3rd_gen`, `3rd_predicted.value`,
    `4th_gen`, `4th_predicted.value`,
    `5th_gen`, `5th_predicted.value`
  )

sub_colunas <- c("Trial", rep(c("Genotype", "Predicted value"), 5))

tabela_final <- kbl(tabela_larga, align = "c", col.names = sub_colunas) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, "1st" = 2, "2nd" = 2, "3rd" = 2, "4th" = 2, "5th" = 2))

colunas_indices <- c(2, 4, 6, 8, 10)

for (col in colunas_indices) {
  valores_coluna <- as.character(tabela_larga[[col]])

  marcar_negrito <- valores_coluna %in% genotipos_repetidos
  
  cores_fundo <- ifelse(marcar_negrito, mapa_cores[valores_coluna], "transparent")
  cores_fundo[is.na(cores_fundo)] <- "transparent"
  
  tabela_final <- tabela_final |> 
    column_spec(
      column = col, 
      bold = marcar_negrito,
      background = cores_fundo
    )
}

tabela_final
```

```{r}
#| label: fig-freq-top-five
#| fig-cap: Frequency of each genotype in the top five genotypes in each trial

# How many times a genotype appears in the table
frequencia_geral <- tabela_larga |>
  select(env, ends_with("_gen")) |> 
  pivot_longer(cols = ends_with("_gen"), values_to = "genotipo") |> 
  count(genotipo, name = "total_ambientes") |> 
  arrange(desc(total_ambientes))

ggplot(data = frequencia_geral, aes(x = total_ambientes, y = reorder(genotipo, total_ambientes))) +
  geom_col(color = "white", fill = "#10342d") +  
  theme_minimal() +
  labs(
    x = "Frequency (Number of Trials in Top 5)", 
    y = "Genotype"
  ) +
  theme(
    text = element_text(size = 08),
    panel.grid.minor = element_blank() 
  )
```

The top two genotypes with the highest frequency of occurrence behave
differently across the trials. Genotype `G092` exhibits a constant
predicted value, whereas genotype `G233` shows variable performance.
Despite both being top-ranked, their distinct predicted values across
trials indicate genotype-by-environment (GxE) interactions that were not
captured in this report.

```{r}
#| label: tbl-top-two-behavior
#| tbl-cap: Behavior of the two most frequent genotypes across ranked environments


G233 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G233" ) |> arrange(rank)
kable(G233,
      align = "c",
      col.names = c("Genotype","Environment", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

G092 <- tabela_longa |> select (gen, env, rank, predicted.value) |> filter(gen == "G092" ) |> arrange(rank)
kable(G092,
      align = "c",
      col.names = c("Genotype","Environment", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

A closer examination of the two most frequent genotypes highlights
contrasting adaptive strategies. Genotype `G092` emerges as a highly
stable material with wide adaptation; it consistently secures top-tier
rankings (predominantly 1st and 2nd) and maintains reliable predicted
values across diverse conditions. Conversely, genotype `G233` exhibits
phenotypic plasticity. It is highly responsive to environments, eaching
a peak predicted yield of over 4,800 kg/ha in trial `E0229`, but
experiences significant yield penalties in more restrictive environments
(dropping to approximately 1,935 kg/ha in `E0283`).

From a practical breeding perspective, `G092` is highly recommended for
broad geographical deployment to ensure yield stability. Meanwhile,
`G233` represents a valuable asset for targeted selection, strategically
suited for high-yielding Target Population of Environments (TPEs).
Ultimately, these distinct behavioral patterns definitively validate the
necessity of characterizing Genotype-by-Environment (GxE) interactions
in multi-environment trial evaluations.

# Trait 2 - Plant Height (PH)

```{r}
#| label: PH_analysis
#| output: false

# Removendo ambientes com nenhuma observação
df <- df |> filter(!env %in% c("E0171","E0172","E0225"))
envs_list <- envs_list[!envs_list %in% c("E0171","E0172","E0225")]

# Verificando dados faltantes para PH em cada ENV
info_env_PH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(PH_R8)),
    .groups = "drop")

# Rodando para cada trial isoladamente
singleTrial_models_PH <- list()
singleTrial_varcomp_PH <- list()
singleTrial_blup_PH <- list()
singleTrial_reliability_PH <- list()
tabela_geral_PH <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2 = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = PH_R8 ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen")$pvals
  
  blups <- blups[order(blups$predicted.value, decreasing = TRUE),]
  
  reliability <- 1 - (blups$std.error^2/varcomp[1,1])
  
  
  m_red_PH <- asreml(fixed = PH_R8 ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_PH, boundary = T)  
  
  
  h2 <- vpredict(model, H2 ~ V1/(V1+V2))
  cv <- sqrt(varcomp[2,1])/mean(df_env$PH, na.rm=T)
  
  tabela_geral_PH <- rbind(tabela_geral_PH,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2 = round(h2$Estimate,digits=4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_PH[[i]] <- model
  singleTrial_varcomp_PH[[i]] <- varcomp
  singleTrial_blup_PH[[i]] <- blups
  singleTrial_reliability_PH[[i]] <- reliability
  }
```

# Summary of Plant Height Model Analysis

```{r}
#| label: tbl-ph
#| tbl-cap: Summary of the main genetic and statistical parameters for Plant Height (PL) across 33 trials of the soybean breeding program in Zambia.

tabela_geral_PH |> 
  mutate(significance = case_when(
    `pvalue` < 0.01 ~ "**",
    `pvalue` < 0.05 ~ "*",
    TRUE             ~ "ns" 
  )) |> 
  kable(
    align = "c",
    col.names = c(
    "Trial", 
    "LRT statistic", 
    "p-value", 
    "Significance", 
    "Broad-sense heritability (H²)", 
    "Coefficient of Variation (%)"
  )) |> 
  kable_styling(bootstrap_options = "striped", full_width = TRUE)|> 
  footnote(
    general = paste("* - p < 0.05 (5%);", "** - p < 0.01 (1%);", "ns - non-significant."),
    general_title = "Significance levels: ",
    threeparttable = TRUE)

```

Regarding Plant Height (PH), trials `E0108`, `E0170`, and `E0230` were
the only ones that did not present genetic variance above zero, meaning
the Likelihood Ratio Test failed to reject the null hypothesis
($H_{0}$). This same behavior was observed for broad-sense heritability
($H^2$); in environments with no significant genetic variance,
heritability was consequently zero or extremely close to it. Overall,
heritability values varied from zero to 0.87. Furthermore, the
coefficient of variation (CV) was low in general, indicating that the
experiments were conducted with high experimental precision.

# BLUPs and Reability

In this section we can observe the predicted value of each genotype in
each trial individually. There are some environments that have more
genotypes or different genotypes. Also, observing the standard errors it
is possible to identify what trials are unbalanced.

```{r}
#| label: create-ph-plots
#| echo: false

# Criando a lista de gráficos para PH
graficos_PH <- list()

for (ambiente in names(singleTrial_blup_PH)) {
  
  dados_blup <- singleTrial_blup_PH[[ambiente]]
  
  p <- ggplot(data = dados_blup, aes(x = reorder(gen, -predicted.value), y = predicted.value)) +
    geom_col(color = "white", fill = "#10342d") + 
    geom_errorbar(aes(ymin = predicted.value - std.error, ymax = predicted.value + std.error), width = 0.2) +
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          text = element_text(size = 8)) +  
    labs(title = paste("BLUPs for Plant Height (PH) - Trial:", ambiente),
         x = "Genotype", 
         y = "BLUP (Predicted Value)")
  
  graficos_PH[[ambiente]] <- p
}
```

Here, we present the BLUPs for Plant Height (PH) across the trials. The
bar charts (see the panel tabset below) illustrate the genotype ranking
within each environment. We used error bars to represent the standard
error of the prediction, providing a visual assessment of the
reliability of our predictions in each specific trial.

```{r}
#| label: graficos-ph
#| results: asis
#| fig-width: 10
#| fig-height: 6
#| echo: false

# Imprimindo gráficos
grupos <- split(graficos_PH,ceiling(seq_along(graficos_PH) / 4))
for(i in seq_along(grupos)){
  painel <- wrap_plots(grupos[[i]],ncol = 2,nrow = 2)
  print(painel)
}
```

```{r}
#| label: fig-blup_by_envs_plots_PH
#| fig-cap: Ppredicted values for Plant Height (PH) and respective genotype rankings across individual trials. Each bar chart displays the performance rank within a specific trial.

blup_all_PH <- bind_rows(
  lapply(names(singleTrial_blup_PH), function(env){
    df <- singleTrial_blup_PH[[env]]
    df$env <- env
    return(df)}))

# subset_1 <- blup_all_PH |> filter(env %in% c("E0252","E0253","E0254","E0255","E0278","E0279","E0280","E0281","E0282","E0283","E0284","E0285"))
#   
# subset_2 <- blup_all_PH |> filter(!env %in% c("E0252","E0253","E0254","E0255","E0278","E0279","E0280","E0281","E0282","E0283","E0284","E0285"))
# 
# ggplot(subset_1,
#        aes(x = env, y = predicted.value, group = gen, color = gen)) +
#       geom_line(alpha = 0.3) +
#       geom_point(size = 1) +
#       theme_minimal() +
#       labs(x = "Environment", y = "Predicted BLUP (PH)",
#         title = "2022 - 2023 Genotype's Predicted Values Across Diferent Environments") +
#       theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
# 
# ggplot(subset_2,
#        aes(x = env, y = predicted.value, group = gen, color = gen)) +
#       geom_line(alpha = 0.3) +
#       geom_point(size = 1) +
#       theme_minimal() +
#       labs(x = "Environment", y = "Predicted BLUP (PH)",
#         title = "2019 - 2021 Genotype's Predicted Values Across Diferent Environments") +
#       theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")

ggplot(blup_all_PH,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (PH)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```

Our visual analysis of the BLUPs for Plant Height (PH) across the
individual trials reveals substantial variation among environments, as
observed in @fig-blup_by_envs_plots_PH. In trials with non-significant
genetic variance (such as `E0108`, `E0170`, and `E0230`), we observe
that the genotype distributions appear flat, reflecting the lack of
genetic differentiation. Conversely, in highly heritable environments,
we can see a clear genetic gradient. The generally low standard errors
and tight error bars visually confirm the high experimental precision
(low CV%) of our breeding trials.

For a comprehensive numerical reference, we detailed the predicted
values and standard errors for all genotypes across all environments in
@tbl-ph-blup.

```{r}
#| label: tbl-ph-blup
#| tbl-cap: Predicted values (BLUPs) and standard errors for Plant Height (PH) of soybean breeding genotypes evaluated across 33 trials in Zambia.

selected_trials <- names(singleTrial_blup_PH)
tabela_grande <- singleTrial_blup_PH[selected_trials] |> 
  map(~ .x |> select(gen, predicted.value, std.error)) |>
  reduce(full_join, by = "gen")

colunas_valores <- rep(c("Predicted value", "Stand. error"), length(selected_trials))

tabela_grande |> 
  kable(
    align = "c",
    col.names = c("Genotype", colunas_valores)
  ) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, setNames(rep(2, length(selected_trials)), selected_trials))) |> 
  scroll_box(width = "100%", height = "600px")

```

```{r}
#| label: tbl-common-geno-ph
#| tbl-cap: Predicted value of Plant Height (PH) of the top five soybean breeding genotypes ranked (1 to 5) in each trial in Zambia 

low_cv_envs <- tabela_geral_PH |> 
  select(env, h2, cv, significance) |> 
  filter(significance == "TRUE" & cv < 30) |> 
  arrange(cv)

envs <- low_cv_envs$env

genotipos_repetidos <- singleTrial_blup_PH[envs] |> 
  map_dfr(~ .x |> arrange(desc(predicted.value)) |> head(5) |> select(gen)) |> 
  count(gen) |> 
  filter(n > 1) |> 
  pull(gen) |> 
  as.character()

set.seed(42) 
cores_paleta <- hcl.colors(length(genotipos_repetidos), palette = "Greens 2") |> 
  paste0("80") 
mapa_cores <- setNames(cores_paleta, genotipos_repetidos)

tabela_longa <- envs |> 
  set_names() |> 
  map_dfr(~ {
    singleTrial_blup_PH[[.x]] |>  
      arrange(desc(predicted.value)) |>   
      head(5) |> 
      select(gen, predicted.value) |> 
      mutate(
        rank = c("1st", "2nd", "3rd", "4th", "5th"),
        gen = as.character(gen)
      )
  }, .id = "env")

tabela_larga <- tabela_longa |> 
  pivot_wider(
    id_cols = env, 
    names_from = rank, 
    values_from = c(gen, predicted.value),
    names_glue = "{rank}_{.value}" 
  ) |> 
  select(
    env,
    `1st_gen`, `1st_predicted.value`,
    `2nd_gen`, `2nd_predicted.value`,
    `3rd_gen`, `3rd_predicted.value`,
    `4th_gen`, `4th_predicted.value`,
    `5th_gen`, `5th_predicted.value`
  )

sub_colunas <- c("Trial", rep(c("Genotype", "Predicted value"), 5))

tabela_final <- kbl(tabela_larga, align = "c", col.names = sub_colunas) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |> 
  add_header_above(c(" " = 1, "1st" = 2, "2nd" = 2, "3rd" = 2, "4th" = 2, "5th" = 2))

colunas_indices <- c(2, 4, 6, 8, 10)

for (col in colunas_indices) {
  valores_coluna <- as.character(tabela_larga[[col]])

  marcar_negrito <- valores_coluna %in% genotipos_repetidos
  
  cores_fundo <- ifelse(marcar_negrito, mapa_cores[valores_coluna], "transparent")
  cores_fundo[is.na(cores_fundo)] <- "transparent"
  
  tabela_final <- tabela_final |> 
    column_spec(
      column = col, 
      bold = marcar_negrito,
      background = cores_fundo
    )
}

tabela_final
```

In the table above the genotypes in bold and colored are repeated at
least once in this top five rank in each trial. It is interesting to
observe that for Plant Height, the same genotypes were repeated more
frequently compared to Grain Yield. Genotypes `G092` and `G233` seem to
be promising, since they appear frequently across both traits. Probably,
these materials can be considered stable, meaning they present a
consistent behavior across different environments.

```{r}
#| label: fig-freq-top-five-ph
#| fig-cap: Frequency of each soybean breeding genotype in the top five genotypes in each trial for Plant Height in Zambia.

frequencia_geral_ph <- tabela_larga |>
  select(env, ends_with("_gen")) |> 
  pivot_longer(cols = ends_with("_gen"), values_to = "genotipo") |> 
  count(genotipo, name = "total_ambientes") |> 
  arrange(desc(total_ambientes))

ggplot(data = frequencia_geral_ph, aes(x = total_ambientes, y = reorder(genotipo, total_ambientes))) +
  geom_col(color = "white", fill = "#10342d") +  
  theme_minimal() +
  labs(
    x = "Frequency (Number of Trials in Top 5)", 
    y = "Genotype"
  ) +
  theme(
    text = element_text(size = 8),
    panel.grid.minor = element_blank() 
  )
```

As observed for the previous trait, Plant Height (PH) values varied
considerably among trials. This is further evidenced by tracking
individual high-ranking lines across different environmental conditions
in @tbl-common-geno-specific.

```{r}
#| label: tbl-common-geno-specific
#| tbl-cap: Behavior of high-ranking genotypes across specific environmental trials.

G146 <- tabela_longa |> select(gen, env, rank, predicted.value) |> filter(gen == "G146") |> arrange(rank)
kable(G146,
      align = "c",
      col.names = c("Genotype","Trial", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

G244 <- tabela_longa |> select(gen, env, rank, predicted.value) |> filter(gen == "G244") |> arrange(rank)
kable(G244,
      align = "c",
      col.names = c("Genotype","Trial", "Rank position", "Predicted value")) |> 
kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

When we take a closer analysis of the specific behavior of genotypes
`G146` and `G244`, the environmental effect on plant height becomes
clearer. We can observe that genotype `G146` appears exceptionally
frequently at the top of the ranking, occupying the 1st or 2nd position
in different new trials. However, its predicted height is highly
variable; it reaches approximately 109 cm in highly designed
environments such as `E0284` and `E0255`, but suffers a drastic
reduction to only 28.4 cm in trial `E0107`.

Similarly, we note that genotype `G244` secures the 1st position in both
`E0278` and `E0283`, but its absolute height in these two environments
shows a large contrast (101.5 cm versus 60.4 cm). This indicates that,
although these genotypes are consistently among the tallest in their
various trials, their absolute growth is greatly influenced by the
specific conditions of each location.

From a practical agronomic point of view, we emphasize that this level
of phenotypic plasticity is a critical factor for selection. A plant
height of 28.4 cm (as observed for `G146`) is generally too low for
efficient mechanized harvesting, leading to potential yield losses in
the field. On the other hand, heights exceeding 100 cm can increase the
risk of lodging, depending on stem architecture. Therefore, we conclude
that, although these lines exhibit strong genetic potential for growth,
identifying and understanding the environmental stresses that cause
these drastic height reductions is essential before making any safe
varietal recommendations for Zambian environments.

# Cross-Trait Ranking Comparison: GY vs PH

To formally assess whether genotype rankings for Grain Yield (GY) and
Plant Height (PH) are associated within each trial, we computed the
Spearman rank correlation coefficient ($\rho$) between the BLUPs of both
traits for each environment independently. This non-parametric approach
is appropriate given that it does not assume a linear relationship
between the traits, relying solely on the rank order of genotypes. A
positive ρ indicates that taller genotypes tend to also yield more
within that trial, whereas a negative ρ suggests a trade-off between the
two traits.

```{r}
#| label: fig-spearman-gy-ph
#| fig-cap: Spearman rank correlation between GY and PH BLUPs within each trial.

# Trials presentes em ambos os traits
trials_comuns <- intersect(names(singleTrial_blup_GY), names(singleTrial_blup_PH))

# Correlação de Spearman em loop
spearman_results <- map_dfr(trials_comuns, function(env) {
  
  blup_gy <- singleTrial_blup_GY[[env]] |>
    select(gen, GY = predicted.value)
  
  blup_ph <- singleTrial_blup_PH[[env]] |>
    select(gen, PH = predicted.value)
  
  merged <- inner_join(blup_gy, blup_ph, by = "gen")
  
  if (nrow(merged) < 5) return(NULL)
  
  test <- cor.test(merged$GY, merged$PH, method = "spearman", exact = FALSE)
  
  data.frame(
    env      = env,
    rho      = round(test$estimate, 3),
    pvalue   = round(test$p.value, 4),
    sig      = case_when(
      test$p.value < 0.01 ~ "**",
      test$p.value < 0.05 ~ "*",
      TRUE                ~ "ns"
    ),
    n_gen    = nrow(merged)
  )
})

# Heatmap
ggplot(spearman_results, aes(x = 1, y = reorder(env, rho), fill = rho)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(rho, " ", sig)), size = 3.2) +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "white",
    high     = "#1a9850",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman ρ"
  ) +
  labs(
    title = "Spearman Correlation (GY vs PH) per Trial",
    x     = NULL,
    y     = "Trial"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid   = element_blank(),
    text         = element_text(size = 11)
  )
```

The Spearman correlation coefficients varied considerably across trials,
ranging from $\rho$ = −0.81 (`E0278`) to $\rho$ = 0.65 (`E0229`), which
itself reflects the magnitude of the GxE interaction previously
discussed. Among the 30 trials evaluated, 14 presented a statistically
significant correlation (p \< 0.05), of which 12 were positive and two
were negative (`E0278`, $\rho$ = −0.81; `E0284`, $\rho$ = −0.47). The
remaining 18 trials showed non-significant correlations, suggesting that
in most environments GY and PH rankings are largely independent.

The predominantly positive correlations may reflect a shared dependence
on biomass accumulation, in favorable environments, genotypes with
greater vegetative growth tend to simultaneously achieve higher grain
yields. Conversely, the strong negative correlations observed in `E0278`
and `E0284` suggest that in those specific conditions, taller genotypes
may have allocated more photoassimilates to vegetative structures at the
expense of grain filling, a classical source-sink competition pattern.
Or possibly, that taller genotypes had problems with lodging, which
decreases the grain yield. Overall, these results reinforce that
multi-trait evaluation across environments is essential, as the GY–PH
relationship is highly context-dependent and cannot be assumed to be
consistent across the Zambian target population of environments.

# Conclusion

This report evaluated 33 soybean trials in Zambia using a mixed
randomized block design (CRD) model to estimate genetic merit (BLUPs)
for Grain Yield (GY) and Plant Height (PH). Quality control retained
environments with a coefficient of variation below 30% and significant
genetic variance (LRT, p \< 0.05), ensuring that only experimentally
reliable trials were used for selection decisions. Broad-sense
heritability values ​​reached up to 0.87, reflecting a high genetic signal
in the most favorable environments.

Significant changes in genotype ranking across locations definitively
established Genotype-Environment (GxE) interaction as a primary factor
in phenotypic expression across the target population of Zambian
environments. This was illustrated by the contrasting adaptive
strategies of the best-performing lines: for GY, the `G092` genotype
demonstrated remarkable stability across trials, while `G233` exhibited
high phenotypic plasticity, ranging from a peak of 4,825.4 kg/ha under
highly favorable conditions to 1,935.7 kg/ha in restrictive
environments. A similar pattern was observed for PH, where the `G146`
genotype ranged from an optimal value of 109.1 cm to a severely stunted
value of 28.4 cm, with direct implications for mechanical harvesting
efficiency and lodging risk.

To formally quantify the relationship between trait ratings, we
calculated Spearman correlations between the BLUPs of GY and PH in each
trial. The coefficients ranged from $\rho$ = −0.81 (`E0278`) to $\rho$ =
0.65 (`E0229`), with 14 of the 30 trials showing statistically
significant associations (p \< 0.05). The predominantly positive
correlations in favorable environments suggest a shared dependence on
biomass accumulation, while the strong negative values ​​observed in
`E0278` and `E0284` point to a source-sink competition between
vegetative growth and grain filling under specific conditions. Overall,
the GY–PH relationship proved to be highly context-dependent,
reinforcing that selection for a single trait would be insufficient for
this target population.

Trait consensus analysis identified genotypes `G092` and `G233` as the
most promising candidates for advancement in Zambia's breeding program.
The `G092` line consistently ranked among the best performers for both
grain yield (GY) and plant height (PH) in multiple environments,
suggesting broad adaptation and stability. The `G233` line, despite its
plasticity, achieved peak performance, making it a strong candidate for
targeted deployment in high-yielding Target Populations (TPEs).
Together, these results highlight the value of multi-environmental and
multi-trait assessment under a mixed model framework as a basis for
informed and reliable selection decisions in soybean breeding programs.
