---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot")
invisible(lapply(packages, library, character.only = TRUE))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$PH_R8, na.rm = TRUE) + 
               c(-3, 3) * sd(data$PH_R8, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = PH_R8)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = PH_R8, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))
```

```{r}
## GY 
# Para datasets grandes, ajuste o número de bins
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histograma com limites de outliers (±3σ)",
       x = "Valor", y = "Frequência") +
  theme_minimal()

# Criar histograma com boxplot abaixo
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removendo outlier
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Ajustando colunas
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

Add isso

```{r}
library(asreml)
```

USAR O DF COMO BASE DE DADOS!!! ELE JÁ ESTÁ FILTRADO

# Exercise

Choose two traits. Perform data analysis of all trials in your dataset
using a RCBD model. Justify your choice in using the treatment and the
block effects as fixed or random. Report the variance component
estimates, the hypothesis test results, the broad-sense heritability,
the coefficient of experimental variation, and the selection candidates'
**BLUPs and reliabilities**. Finally, compare the rankings between
trials and speculate the reasons why they are different/similar. This
task must be carried out within groups.

# Introduction

USAR O DF COMO BASE DE DADOS!!! ELE JÁ ESTÁ FILTRADO \# Trait #1 - GY

#Modelo completo - João \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

#Modelo local 1 todos os anos - Bruna \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

----

#Modelo local 2 todos os anos - Renan \## Variance component estimates

Before diving into the mixed models, we selected location L099 for this
location-specific temporal analysis. Based on the cross-tabulation of
the trial network, L099 stands out as one of the most consistent
environments, possessing robust phenotypic data across almost all
evaluated years, with a single gap in the 2022 season.

```{r}
#| label: model-local2-gy
#| warning: false
#| message: false

# Local 2 Model (All Years) - L099 | Trait: GY (Grain Yield)

# 1. Data Preparation
df_L099_GY <- df |> 
  filter(loc == "L099") |> 
  droplevels()

# 2. Mixed Model Fitting (by ASReml)
mod_L099_GY <- asreml(
  fixed = GY ~ 1 + RAINFED,
  random = ~ YEAR + YEAR:rep + gen + gen:YEAR,
  residual = ~ idv(units),
  data = df_L099_GY,
  trace = FALSE #
)
mod_L099_GY <- update(mod_L099_GY)
```

#RAINFED do modelo do L099?? será que o Rainfed é o melhor como efeito
fixo???

## Hypothesis test

```{r}
#| label: var-test-local2-gy

## Variance component estimates
print("Variance Components (Location L099 - GY):")
summary(mod_L099_GY)$varcomp

## Hypothesis test
print("Wald Test for Fixed Effect (RAINFED):")
wald.asreml(mod_L099_GY)

# LRT for Genotypic Effect
mod_L099_GY_reduced <- asreml(
  fixed = GY ~ 1 + RAINFED,
  random = ~ YEAR + YEAR:rep + gen:YEAR,
  residual = ~ idv(units),
  data = df_L099_GY,
  trace = FALSE
)
print("LRT for Genotypic Variance:")
lrt(mod_L099_GY, mod_L099_GY_reduced)
```

The Wald test indicated that the water regime (RAINFED) had no
significant effect (p = 0.25) on grain yield at the L099 location.
Furthermore, the variance components revealed a massive Genotype
$\times$ Year interaction variance (174,271), which was over five times
larger than the genotypic variance (31,247). Consequently, the main
genotypic effect was not significant in the LRT (p = 0.24),
demonstrating that climatic fluctuations across seasons heavily dictate
cultivar performance and cause severe rank changes over time.

## Broad-sense heritability estimation & Coefficient of experimental variation

```{r}
#| label: metrics-local2-gy

## Broad-sense heritability estimation
n_years_L099 <- length(levels(df_L099_GY$YEAR))

reps_per_year_L099 <- df_L099_GY |> 
  group_by(YEAR) |> 
  summarise(n = n_distinct(rep))
n_rep_mean_L099 <- mean(reps_per_year_L099$n)

vc_L099 <- summary(mod_L099_GY)$varcomp
vg_L099  <- vc_L099["gen", "component"]
vga_L099 <- vc_L099["gen:YEAR", "component"]
ve_L099  <- vc_L099["units!units", "component"]

# Heritability 
h2_L099_manual <- vg_L099 / (vg_L099 + (vga_L099 / n_years_L099) + (ve_L099 / (n_years_L099 * n_rep_mean_L099)))

print(paste("Broad-sense Heritability (H2):", round(h2_L099_manual, 4)))

## Coefficient of experimental variation
var_res_L099_real <- vc_L099["units!units", "component"]
mean_GY_L099 <- mean(df_L099_GY$GY, na.rm = TRUE)

cve_L099 <- (sqrt(var_res_L099_real) / mean_GY_L099) * 100
cat("Coefficient of Experimental Variation (CVe):", round(cve_L099, 2), "%\n")
```

The coefficient of experimental variation (CVe) was 18.57%,
demonstrating excellent experimental precision for grain yield trials.
Broad-sense heritability was estimated at 0.31, a moderate-to-low value
that directly reflects the strong influence of the Genotype $\times$
Year interaction, confirming that climatic variation across seasons at
L099 heavily masks the pure genetic expression of the cultivars.

```{r}
#| label: blups-local2-gy

## BLUPs and Predicted Means Extraction
pred_L099 <- predict(mod_L099_GY, classify = "gen")
predicted_means <- pred_L099$pvals

## Extracting Pure Random Effects (To get the correct BLUP Standard Error)
random_effects <- summary(mod_L099_GY, coef = TRUE)$coef.random |> 
  as.data.frame() |> 
  tibble::rownames_to_column("Term") |> 
  filter(grepl("^gen_", Term)) |> 
  mutate(gen = gsub("gen_", "", Term)) |> 
  rename(SE_BLUP = std.error) |> 
  select(gen, SE_BLUP)

## Genetic Variance
vg_L099 <- summary(mod_L099_GY)$varcomp["gen", "component"]

## Merging and calculating True Reliability
blups_L099_Final <- predicted_means |> 
  left_join(random_effects, by = "gen") |> 
  mutate(Reliability = 1 - (SE_BLUP^2 / vg_L099)) |> 
  select(gen, predicted.value, SE_BLUP, status, Reliability) |> 
  arrange(desc(predicted.value))

print("Top 10 Genotypes (Predicted Means and Corrected Reliability):")
head(blups_L099_Final, 10) |> 
  knitr::kable(digits = 3)
```

Genotypic predictions (BLUPs) revealed that materials G068, G319, and
G092 showed the highest yield potential for the L099 location. The
reliabilities of these predictions ranged from 0.10 to 0.25. These
values reflect the high Genotype $\times$ Year interaction variance,
indicating that the recommendation of these cultivars must account for
strong climatic fluctuations between seasons (across years)

-----

#Modelo ano 1 todos os locais - Bruna \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

-----

#Modelo ano 1 todos os locais - Renan \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

To evaluate spatial stability, we selected the 2023, which presented the
broadest geographical coverage in the dataset (8 active trial
locations). Prior to fitting the spatial model, we assessed the trial
network's connectivity to avoid highly unbalanced estimations. The
dataset demonstrated balance: out of 30 evaluated genotypes, 28 were
concurrently tested across all 8 locations.

```{r}
#| label: model-year1-gy
#| warning: false
#| message: false

# 1. Data Preparation
# Filtering the main dataset for 2023 and dropping unused levels.
df_2023_GY <- df |> 
  filter(YEAR == "2023") |> 
  droplevels()

# 2. Mixed Model Fitting 
mod_2023_GY <- asreml(
  fixed = GY ~ 1 + RAINFED,
  random = ~ loc + loc:rep + gen + gen:loc,
  residual = ~ idv(units),
  data = df_2023_GY,
  trace = FALSE
)

# Ensuring variance convergence
mod_2023_GY <- update(mod_2023_GY)
```

For this single-year spatial analysis, the trial locations represent the
distinct environmental conditions. We fitted a mixed model treating the
management regime (RAINFED) as a fixed effect, while the 8 locations,
experimental repetitions, and genotypes were treated as random effects.
This structure allows us to model the spatial environmental noise and
accurately isolate the Genotype $\times$ Location variance to identify
the most geographically stable cultivars.

## Hypothesis test

```{r}
#| label: var-test-year1-gy

## Variance component estimates
print("Variance Components (Year 2023 - GY):")
summary(mod_2023_GY)$varcomp

## Hypothesis test
print("Wald Test for Fixed Effect (RAINFED):")
wald.asreml(mod_2023_GY)

# LRT for Genotypic Effect
mod_2023_GY_reduced <- asreml(
  fixed = GY ~ 1 + RAINFED,
  random = ~ loc + loc:rep + gen:loc,
  residual = ~ idv(units),
  data = df_2023_GY,
  trace = FALSE
)
print("LRT for Genotypic Variance:")
lrt(mod_2023_GY, mod_2023_GY_reduced)
```

The Wald test indicated that the water management regime (RAINFED) had a
highly significant effect (p \< 0.001) on grain yield across the
evaluated locations in 2023. Furthermore, the variance components
revealed a massive Genotype $\times$ Location interaction variance
(176,858.86), nearly four times larger than the pure genotypic variance
(47,547.28). Despite this spatial environmental influence causing rank
changes among sites, the Likelihood Ratio Test (LRT) confirmed that the
main genotypic effect remained highly significant (p \< 0.001),
indicating a strong and selectable genetic contribution to yield across
the trial network.

## Broad-sense heritability estimation & Coefficient of experimental variation

```{r}
#| label: metrics-year1-gy

## Broad-sense heritability estimation
n_locs_2023 <- length(levels(df_2023_GY$loc))

reps_per_loc_2023 <- df_2023_GY |> 
  group_by(loc) |> 
  summarise(n = n_distinct(rep))
n_rep_mean_2023 <- mean(reps_per_loc_2023$n)

# Extracting components by name
vc_2023 <- summary(mod_2023_GY)$varcomp
vg_2023  <- vc_2023["gen", "component"]
vgl_2023 <- vc_2023["gen:loc", "component"]
ve_2023  <- vc_2023["units!units", "component"]

h2_2023_manual <- vg_2023 / (vg_2023 + (vgl_2023 / n_locs_2023) + (ve_2023 / (n_locs_2023 * n_rep_mean_2023)))

print(paste("Broad-sense Heritability (H2) for 2023:", round(h2_2023_manual, 4)))

## Coefficient of experimental variation
var_res_2023_real <- vc_2023["units!units", "component"]
mean_GY_2023 <- mean(df_2023_GY$GY, na.rm = TRUE)

cve_2023 <- (sqrt(var_res_2023_real) / mean_GY_2023) * 100
cat("Coefficient of Experimental Variation (CVe):", round(cve_2023, 2), "%\n")
```

The coefficient of experimental variation (CVe) was 20.36%, indicating
an acceptable and typical level of experimental precision for
multi-environment field trials evaluating grain yield. Broad-sense
heritability was estimated at 0.56. This metric corroborates the highly
significant genotypic effect observed in the Likelihood Ratio Test,
demonstrating that despite the considerable Genotype $\times$ Location
interaction, the pure genetic variance is robust enough to allow for
reliable selection of superior and spatially stable cultivars across
trial network.

```{r}
#| label: blups-year1-gy

## BLUPs and Predicted Means Extraction
pred_2023 <- predict(mod_2023_GY, classify = "gen")
predicted_means_2023 <- pred_2023$pvals

random_effects_2023 <- summary(mod_2023_GY, coef = TRUE)$coef.random |> 
  as.data.frame() |> 
  tibble::rownames_to_column("Term") |> 
  filter(grepl("^gen_", Term)) |> 
  mutate(gen = gsub("gen_", "", Term)) |> 
  rename(SE_BLUP = std.error) |> 
  select(gen, SE_BLUP)

## Merging and calculating True Reliability
blups_2023_Final <- predicted_means_2023 |> 
  left_join(random_effects_2023, by = "gen") |> 
  mutate(Reliability = 1 - (SE_BLUP^2 / vg_2023)) |> 
  select(gen, predicted.value, SE_BLUP, status, Reliability) |> 
  arrange(desc(predicted.value))

print("Top 10 Genotypes (2023 Spatial Stability):")
head(blups_2023_Final, 10) |> 
  knitr::kable(digits = 3)
```

Genotypic predictions (BLUPs) identified G137 as the standout cultivar,
demonstrating the highest spatial stability and yield potential across
the 2023 trial network, followed by G335 and G359. The reliabilities of
these predictions were robust, stabilizing around 0.53 to 0.54. These
values are substantially higher than those observed in the
single-location temporal model, reflecting the strong broad-sense
heritability (0.56) and confirming that these specific genotypes
maintain superior, reliable performance across diverse geographical
environments despite the spatial Genotype $\times$ Location interaction.

is:

```{r}
1 + 1
```

------

USAR O DF COMO BASE DE DADOS!!! ELE JÁ ESTÁ FILTRADO \# Trait #2 - PH
#Modelo completo - João \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

#Modelo local 1 todos os anos - Bruna \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

#Modelo local 2 todos os anos - Renan \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

#Modelo ano 1 todos os locais - Bruna \## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

#Modelo ano 1 todos os locais - Renan \## Variance component estimates

# Discussion

A fazer

-   filtrar os outliers do dataframe total para as duas características
-   ajustar o modelo para todos os environments (modelo =
    $$ y = env + env/bloco + gen + gen*env + irrigação + \epsilon$$) -
    sendo env, gen e gen\*env como aleatório.
-   tabela para mostrar os locais nos anos que escolhemos para
    fazer (2) - estabilidade/adaptabilidade
-   ajustar o modelo para cada local ao longo dos anos.
-   ajustar o modelo para um ano qual é o melhor local.

se der tempo: estrutura de covariância entre anos/locais para
identificar as correlações entre anos. Quanto mais próximos, maior. Tem
covariância entre os anos.

