---
title: "Data Analysis RCBD in Zambia"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true

crossref: {}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)
library(car)

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial)) |> na.omit()

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1") |> na.omit()
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2") |> na.omit()
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3") |> na.omit()
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4") |> na.omit()
genotypes_4 <- trial_4$gen
```

# Exercise 
Choose two traits. Perform data analysis of all trials in your dataset using a RCBD model. Justify your choice in using the treatment and the block effects as fixed or random. Report the variance component estimates, the hypothesis test results, the broad-sense heritability, the coefficient of experimental variation, and the selection candidates' BLUPs and reliabilities. Finally, compare the rankings between trials and speculate the reasons why they are different/similar. This task must be carried out within groups. 


# Introduction


# Trait \#1 
## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation


# Trait \#2
## Variance component estimates

## Hypothesis test

## Broad-sense heritability estimation

## Coefficient of experimental variation

# Selection of candidates


# Discussion