---
title: "Multi Trial EDA"
author: 
  - name: João Vitor Sanches Calvo
    orcid: 0009-0002-5827-6109
    email: joaovitor.calvo@usp.br
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
---

# Exercise

Using the dataset you will use in the project, perform a comprehensive exploratory data analysis. Choose four trials and perform EDA for five traits (if available). Check the variation, detect outliers and possible errors, build histograms, density plots and boxplots; compute measures of central tendency and dispersion, and investigate the relationship between traits. Finally, perform an across trials EDA: check if there is any major changes within traits across trials.

# Database selection - Zambia

```{r}
#| warning: false

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4")
genotypes_4 <- trial_4$gen
```

# Functions

```{r}
#| warning: false

# Function 1 - Single Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  # Correlation between traits
  cor_mat <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")
  
  p1 <-ggcorrplot(cor_mat, 
           hc.order = TRUE,           
           type = "upper",               
           lab = TRUE,      
           lab_size = 3,           
           colors = c("red", "white", "blue"),      
           title = title)   
  
  # Traits Boxplots
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p4 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = p1,boxplot = p2, density = p3, histogram = p4))
}


# Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (scatter = p1)
}

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18)) +
        theme(axis.text.x = element_text(size = 10))
  
  return (boxplot_genotypes = p1)
}

# Function 4 - Single Trials Analysis Together
single_trials_together <- function(df, title, traits) {
  
  # Dataframe long
  df_long <- df |> 
    mutate(Trial = as.factor(trial)) |> 
    pivot_longer(
      cols = all_of(traits),
      names_to = "Trait",
      values_to = "Value") |> 
    filter(!is.na(Value))
  
  # Boxplots 
  p1 <- ggplot(df_long, aes(x = Trial, y = Value, fill = Trial)) +
          geom_boxplot(outlier.colour = "black", outlier.size = 0.5) +
          facet_wrap(~Trait, scales = "free") +
          theme_minimal() +
          labs(title = title)

  # Density Plot
  p2 <- ggplot(df_long, aes(x = Value, fill = Trial, group = Trial)) +
          geom_density(alpha = 0.4) +
          facet_wrap(~Trait, scales = "free") +
          theme_light() +
          labs(title = title)

return (list(boxplots = p1, density = p2))
}

# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}


# Function 6 - Across Trials Correlation
across_trials_correlation <- function(df, title, traits) {

  # Correlation Plot with all trials
  cor_total <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")

  p1 <-ggcorrplot(cor_total, 
         hc.order = TRUE,           
         type = "upper",               
         lab = TRUE,      
         lab_size = 3,           
         colors = c("red", "white", "blue"),      
         title = title)
  
  return(all_cor = p1)
}
```

# Trials Overview

Explicar genotipos do experimento, e cada trait analisado.

# Trial 1 - Local "L0118" / Year "2023"

## A FAZER

1\) Listar os genotipos do experimento, e explicar cada trait analisado.

2\) Fazer um exemplo manual para verificar as medidas de posição, dispersão, correlação e funcções

3\) Definir critério de avaliação para os outliers (Trial 1 tem 1 outlier na altura)

4\) Rodar as funções para cada Trial e depois entre Trials

5\) Organizar documentos e fazer as interpretações

## Uso das Funções

```{r}
#| warning: false

## Function 1 - Single Trial Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 30)
}

# Example - Trial 1
all_trials_summary[[1]]$stats
all_trials_summary[[1]]$correlation
all_trials_summary[[1]]$density
all_trials_summary[[1]]$boxplot
all_trials_summary[[1]]$histogram

#######################################################################

## Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter(trial_2,"Título", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")

######################################################################

## Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Título", "GY", "GY (kg/ha)")

######################################################################

## Function 4 - Single Trials Analysis Together
all_trials_together <- single_trials_together(selected_data, "Título", traits)

# Outputs
all_trials_together$density
all_trials_together$boxplot

#######################################################################

# Function 5 - All Trials Boxplot By Genotype
all_genotype_GY <- boxplot_genotype_all_in_one(selected_data,"GY")
all_genotype_GY$boxplot_gt_trial
all_genotype_GY$boxplot_gt

#######################################################################

# Function 6 - Across Trials Correlation
across_trials <- across_trials_correlation(selected_data, "Título", traits)
across_trials
```

## Outras Anotações

```{r}
#| label: fig-genotype-comparison
#| fig-width: 22
#| fig-height: 8
#| warning: false

#| label: fig-corr-indiv
#| layout-ncol: 2
#| fig-height: 8
#| warning: false

#| label: fig-corr-all
#| fig-height: 10
#| fig-cap: "Correlation"
#| warning: false

#indiviual plots
boxplot_genotype_all_in_one(selected_data, "FLW_DAYS")
boxplot_genotype_all_in_one(selected_data, "GY")
boxplot_genotype_all_in_one(selected_data, "PL_EMERG_COUNT")
boxplot_genotype_all_in_one(selected_data, "Pod_Shattering")

#plot PH_R8, filtring outlier (16  Zambia 2023 E0283 G321 L0118 // 308.33)
all_trials_cleanPH_R8 <- selected_data %>%
  filter(PH_R8 > 10 & PH_R8 < 160) 
boxplot_genotype_all_in_one(all_trials_cleanPH_R8 , "PH_R8")

across_trials_correlation(selected_data, "Titulo", traits)
```
