---
title: "Multi Trial EDA"
author: 
  - name: João Vitor Sanches Calvo
    orcid: 0009-0002-5827-6109
    email: joaovitor.calvo@usp.br
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

# Exercise

Using the dataset you will use in the project, perform a comprehensive exploratory data analysis. Choose four trials and perform EDA for five traits (if available). Check the variation, detect outliers and possible errors, build histograms, density plots and boxplots; compute measures of central tendency and dispersion, and investigate the relationship between traits. Finally, perform an across trials EDA: check if there is any major changes within traits across trials.

# Database selection - Zambia

```{r}

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4")
genotypes_4 <- trial_4$gen
```

# Functions

```{r}

# Function 1 - Single Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  # Correlation between traits
  cor_mat <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")
  
  p1 <-ggcorrplot(cor_mat, 
           hc.order = TRUE,           
           type = "upper",               
           lab = TRUE,      
           lab_size = 3,           
           colors = c("red", "white", "blue"),      
           title = title)   
  
  # Traits Boxplots
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p4 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = p1,boxplot = p2, density = p3, histogram = p4))
}


# Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (scatter = p1)
}

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10)) +
        theme(axis.text.x = element_text(size = 8))
  
  return (boxplot_genotypes = p1)
}

# Function 4 - Single Trials Analysis Together
single_trials_together <- function(df, title, traits) {
  
  # Dataframe long
  df_long <- df |> 
    mutate(Trial = as.factor(trial)) |> 
    pivot_longer(
      cols = all_of(traits),
      names_to = "Trait",
      values_to = "Value") |> 
    filter(!is.na(Value))
  
  # Boxplots 
  p1 <- ggplot(df_long, aes(x = Trial, y = Value, fill = Trial)) +
          geom_boxplot(outlier.colour = "black", outlier.size = 0.5) +
          facet_wrap(~Trait, scales = "free") +
          theme_minimal() +
          labs(title = title)

  # Density Plot
  p2 <- ggplot(df_long, aes(x = Value, fill = Trial, group = Trial)) +
          geom_density(alpha = 0.4) +
          facet_wrap(~Trait, scales = "free") +
          theme_light() +
          labs(title = title)

return (list(boxplots = p1, density = p2))
}

# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}


# Function 6 - Across Trials Correlation
across_trials_correlation <- function(df, title, traits) {

  # Correlation Plot with all trials
  cor_total <- df |>
    select(all_of(traits)) |>
    cor(use = "pairwise.complete.obs")

  p1 <-ggcorrplot(cor_total, 
         hc.order = TRUE,           
         type = "upper",               
         lab = TRUE,      
         lab_size = 3,           
         colors = c("red", "white", "blue"),      
         title = title)
  
  return(all_cor = p1)
}
```

# Trials Overview

Explicar genotipos do experimento, e cada trait analisado.

Revisar tabela

```{r}
#| label: tabela-trait-local
#| tbl-cap: "Measures of central tendency and dispersion of trials conducted in Zambia"

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}

# Table 
envs <- c("T1", "T2", "T3", "T4")
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
df_empilhado <- map_df(seq_along(all_trials_summary), function(i) {
  all_trials_summary[[i]]$stats |> 
    mutate(Local = envs[i]) 
})

df_para_tabela <- df_empilhado |>
    pivot_longer(
    cols = c(mean, median, variance, sd, min, max, range), 
    names_to = "Estatistica", 
    values_to = "Valor"
  ) |>
   mutate(Estatistica = case_match(Estatistica,
    "mean" ~ "Mean",
    "median" ~ "Median",
    "variance" ~ "Variance",
    "sd" ~ "Sd",
    "min" ~ "Min",
    "max" ~ "Max",
    "range" ~ "Range"
  )) |> 
  pivot_wider(names_from = c(Variable, Local), values_from = Valor)

n_traits <- 5
n_locais <- 4

cabecalho_superior <- c(" " = 1)
cabecalho_superior[traits] <- n_locais 

df_para_tabela |>
  kable(
    digits = 2,
    align = "l",
    col.names = c(" ", rep(envs, n_traits)), 
    caption = "Measures of central tendency and dispersion of trials conducted in Zambia"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE
  ) |>
  add_header_above(cabecalho_superior) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```

```{r}

# Removing outlier
trial_1 <- trial_1 |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8))
selected_data <- selected_data |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8))

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}
```

```{r}
# Function 4 - Single Trials General Overview
all_trials_together <- single_trials_together(selected_data, "All Trials Overview", traits)

all_trials_together$density
all_trials_together$boxplot

# Function 5 - All Trials Boxplot By Genotype
# Grain Yield
all_genotype_GY <- boxplot_genotype_all_in_one(selected_data,"GY")
all_genotype_GY$boxplot_gt_trial
all_genotype_GY$boxplot_gt

# Plant Height
all_genotype_PH <- boxplot_genotype_all_in_one(selected_data,"PH_R8")
all_genotype_PH$boxplot_gt_trial
all_genotype_PH$boxplot_gt

# Pod_Shattering
all_genotype_PS <- boxplot_genotype_all_in_one(selected_data,"Pod_Shattering")
all_genotype_PS$boxplot_gt_trial
all_genotype_PS$boxplot_gt

# Days to Flowering
all_genotype_FL <- boxplot_genotype_all_in_one(selected_data,"FLW_DAYS")
all_genotype_FL$boxplot_gt_trial
all_genotype_FL$boxplot_gt

# Plant Emergence
all_genotype_PL <- boxplot_genotype_all_in_one(selected_data,"PL_EMERG_COUNT")
all_genotype_PL$boxplot_gt_trial
all_genotype_PL$boxplot_gt

# Function 6 - Across Trials Correlation
across_trials_corr <- across_trials_correlation(selected_data, "Across Trials Correlation", traits)
across_trials_corr
```

# Trial 1 - Local "L0118" / Year "2023"

```{r}

# Trial 1
all_trials_summary[[1]]$stats
all_trials_summary[[1]]$correlation
all_trials_summary[[1]]$density
all_trials_summary[[1]]$boxplot
all_trials_summary[[1]]$histogram

# Function 2 - Single Trial Pairwise Scatter Plot
pairwise_scatter(trial_1,"Trial 1", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
pairwise_scatter(trial_1,"Trial 1", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
pairwise_scatter(trial_1,"Trial 1", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
pairwise_scatter(trial_1,"Trial 1", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
pairwise_scatter(trial_1,"Trial 1", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "GY", "GY (kg/ha)")
boxplot_by_genotype(trial_1,"Trial 1", "PH_R8", "PH (cm)")
boxplot_by_genotype(trial_1,"Trial 1", "FLW_DAYS", "Days to Flowering")
boxplot_by_genotype(trial_1,"Trial 1", "Pod_Shattering", "Pod_Shattering score")
boxplot_by_genotype(trial_1,"Trial 1", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 2 - Local "L0133" / Year "2023"

```{r}

# Trial 2
all_trials_summary[[2]]$stats
all_trials_summary[[2]]$correlation
all_trials_summary[[2]]$density
all_trials_summary[[2]]$boxplot
all_trials_summary[[2]]$histogram

# Function 2 - Single Trial Pairwise Scatter Plot
pairwise_scatter(trial_2,"Trial 2", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
pairwise_scatter(trial_2,"Trial 2", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
pairwise_scatter(trial_2,"Trial 2", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
pairwise_scatter(trial_2,"Trial 2", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
pairwise_scatter(trial_2,"Trial 2", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "GY", "GY (kg/ha)")
boxplot_by_genotype(trial_2,"Trial 2", "PH_R8", "PH (cm)")
boxplot_by_genotype(trial_2,"Trial 2", "FLW_DAYS", "Days to Flowering")
boxplot_by_genotype(trial_2,"Trial 2", "Pod_Shattering", "Pod_Shattering score")
boxplot_by_genotype(trial_2,"Trial 2", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 3 - Local "L074" / Year "2023"

```{r}
# Trial 3
all_trials_summary[[3]]$stats
all_trials_summary[[3]]$correlation
all_trials_summary[[3]]$density
all_trials_summary[[3]]$boxplot
all_trials_summary[[3]]$histogram

# Function 2 - Single Trial Pairwise Scatter Plot
pairwise_scatter(trial_3,"Trial 3", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
pairwise_scatter(trial_3,"Trial 3", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
pairwise_scatter(trial_3,"Trial 3", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
pairwise_scatter(trial_3,"Trial 3", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
pairwise_scatter(trial_3,"Trial 3", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "GY", "GY (kg/ha)")
boxplot_by_genotype(trial_3,"Trial 3", "PH_R8", "PH (cm)")
boxplot_by_genotype(trial_3,"Trial 3", "FLW_DAYS", "Days to Flowering")
boxplot_by_genotype(trial_3,"Trial 3", "Pod_Shattering", "Pod_Shattering score")
boxplot_by_genotype(trial_3,"Trial 3", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 4 - Local "L0134" / Year "2023"

```{r}
# Trial 4
all_trials_summary[[4]]$stats
all_trials_summary[[4]]$correlation
all_trials_summary[[4]]$density
all_trials_summary[[4]]$boxplot
all_trials_summary[[4]]$histogram

# Function 2 - Single Trial Pairwise Scatter Plot
pairwise_scatter(trial_4,"Trial 4", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
pairwise_scatter(trial_4,"Trial 4", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
pairwise_scatter(trial_4,"Trial 4", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
pairwise_scatter(trial_4,"Trial 4", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
pairwise_scatter(trial_4,"Trial 4", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "GY", "GY (kg/ha)")
boxplot_by_genotype(trial_4,"Trial 4", "PH_R8", "PH (cm)")
boxplot_by_genotype(trial_4,"Trial 4", "FLW_DAYS", "Days to Flowering")
boxplot_by_genotype(trial_4,"Trial 4", "Pod_Shattering", "Pod_Shattering score")
boxplot_by_genotype(trial_4,"Trial 4", "PL_EMERG_COUNT", "Plant Emergence")
```

## A FAZER

1\) Listar os genotipos do experimento, e explicar cada trait analisado.

2\) Revisar tabela

3\) Definir critério de avaliação para os outliers (Trial 1 tem 1 outlier na altura)

4\) Organizar cunks, formatar documentos e fazer as interpretações
