---
title: "Exploratory Data Analysis"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true

crossref: {}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

# Introduction

This report presents a comprehensive exploratory data analysis of a soybean breeding program in **Zambia**, exploring **four trials** and **five key soybean traits**. The main objectives are to assess variation, detect unusual values, compute measures of central tendency and dispersion, and investigate relationships between traits. First, we describe the custom functions developed to perform the EDA. In the following section, we present our insights regarding the experimental database. Finally, we analyze the performance of a trait to support genotype selection across the evaluated trials.

# Data characterization

The dataset is composed of phenotypic records from a soybean breeding program conducted in Zambia during the 2023 growing season. A total of 29 genotypes were evaluated across four distinct locations (L0118, L0133, L074, and L0134), representing different environmental conditions. Each genotype was replicated three times per location, except for "G321", which had six replicates in each trial.

Five agronomic traits were selected based on data availability and breeding importance: Days to Flowering (FLW_DAYS), Grain Yield (GY), Plant Height at R8 stage (PH_R8), Emergence Count (PL_EMERG_COUNT), and Pod Shattering.

```{r}
#| label: trials-data

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4")
genotypes_4 <- trial_4$gen
```

# Used Functions

```{r}
#| label: functions

# Function 1 - Single Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      cv = (sd/mean),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  # Correlation between traits
  cor_mat <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")
  
  p1 <-ggcorrplot(cor_mat, 
           hc.order = TRUE,           
           type = "upper",               
           lab = TRUE,      
           lab_size = 3,           
           colors = c("red", "white", "blue"),      
           title = title)   
  
  # Traits Boxplots
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p4 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = p1,boxplot = p2, density = p3, histogram = p4))
}


# Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (scatter = p1)
}

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10)) +
        theme(axis.text.x = element_text(size = 8))
  
  return (boxplot_genotypes = p1)
}

# Function 4 - Single Trials Analysis Together
single_trials_together <- function(df, title, traits) {
  
  # Dataframe long
  df_long <- df |> 
    mutate(Trial = as.factor(trial)) |> 
    pivot_longer(
      cols = all_of(traits),
      names_to = "Trait",
      values_to = "Value") |> 
    filter(!is.na(Value))
  
  # Boxplots 
  p1 <- ggplot(df_long, aes(x = Trial, y = Value, fill = Trial)) +
          geom_boxplot(outlier.colour = "black", outlier.size = 0.5) +
          facet_wrap(~Trait, scales = "free") +
          theme_minimal() +
          labs(title = title)

  # Density Plot
  p2 <- ggplot(df_long, aes(x = Value, fill = Trial, group = Trial)) +
          geom_density(alpha = 0.4) +
          facet_wrap(~Trait, scales = "free") +
          theme_light() +
          labs(title = title)

return (list(boxplots = p1, density = p2))
}

# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}


# Function 6 - Across Trials Correlation
across_trials_correlation <- function(df, title, traits) {

  # Correlation Plot with all trials
  cor_total <- df |>
    select(all_of(traits)) |>
    cor(use = "pairwise.complete.obs")

  p1 <-ggcorrplot(cor_total, 
         hc.order = TRUE,           
         type = "upper",               
         lab = TRUE,      
         lab_size = 3,           
         colors = c("red", "white", "blue"),      
         title = title)
  
  return(all_cor = p1)
}
```

# Trials Overview

First, some basic statistics were computed for each trial, to facilitate comparison and evaluation of traits:

```{r}
#| label: stats-table1
#| echo: false

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}

# Table 
envs <- c("T1", "T2", "T3", "T4")
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")

df_conc <- map_df(seq_along(all_trials_summary), function(i) {
  all_trials_summary[[i]]$stats |> 
    mutate(Local = envs[i])})

table_df <- df_conc |>
    pivot_longer(
    cols = c(mean, median, variance, sd, cv, min, max, range), 
    names_to = "Stat", 
    values_to = "Value"
  ) |>
   mutate(Stat = case_match(Stat,
    "mean" ~ "Mean",
    "median" ~ "Median",
    "variance" ~ "Variance",
    "sd" ~ "Sd",
    "cv" ~ "Cv",
    "min" ~ "Min",
    "max" ~ "Max",
    "range" ~ "Range"
  )) |> 
  pivot_wider(names_from = c(Variable, Local), values_from = Value)

n_traits <- 5
n_local <- 4

sup_header <- c(" " = 1)
sup_header[traits] <- n_local 

order <- unlist(lapply(traits, function(trait) {
  paste0(trait, "_", envs)}))

table_df <- table_df |>
  select(Stat, all_of(order))

```

```{r}
#| label: tbl-stats-table2
#| tbl-cap: "Measures of central tendency and dispersion of trials conducted in Zambia"
#| echo: false
table_df |>
  kable(
    digits = 2,
    align = "l",
    col.names = c(" ", rep(envs, n_traits))) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE
  ) |>
  add_header_above(sup_header) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```

An outlier was identified for plant height (PH_R8) ( @tbl-stats-table2). This data point was excluded due to its biological anomalousness. It is highly unlikely for a soybean plant to reach 308.33 cm, especially when the trial mean was only 51.34 cm. This inconsistency likely originates from a manual recording error.

```{r}
#| label: outlier

# Removing outliers
trial_1 <- trial_1 |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8))
trial_3 <- trial_3 |> 
  mutate(across(all_of(traits),~ if_else(gen == "G216" & env == "E0282", NA_real_ ,.))) |>
  mutate(FLW_DAYS = ifelse(FLW_DAYS == 0, NA, FLW_DAYS)) 

selected_data <- selected_data |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8)) |> 
  mutate(across(all_of(traits),~ if_else(gen == "G216" & env == "E0282", NA_real_ ,.))) |> 
  mutate(FLW_DAYS = ifelse(FLW_DAYS == 0, NA, FLW_DAYS))

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}
```

Now, some plots analysing all the traits within each trial:

```{r}
#| label: fig-all_trials_density
#| fig.cap: "Density plot of traits within each trial"
#| echo: false

# Function 4 - Single Trials General Overview
all_trials_together <- single_trials_together(selected_data, "All Trials Overview", traits)
all_trials_together$density
```

```{r}
#| label: fig-all_trials_boxplot
#| fig.cap: "Boxplot of traits within each trial"
#| echo: false
all_trials_together$boxplot
```

Regarding the boxplot analyses performed by trait, comparing genotypes and environments, clear trends can be observed. For the Grain Yield (GY) trait, mean values are consistently higher across almost all genotypes in environment L0134 ( @fig-all_trials_boxplot). This location stands out as a high-potential environment, likely providing optimal edaphoclimatic conditions that allowed the germplasm to express its maximum genetic potential.


In contrast, the genotypes did not perform well for this trait in environment L0118. The overall shift toward lower yield values in this specific location suggests a restrictive environment, possibly due to soil limitations or water stress during critical stages of crop development. This marked difference between L0134 and L0118 highlights a strong Environmental Effect, which is a fundamental component in evaluating the adaptability and stability of these genotypes across the Zambian landscape.
Regarding the flowering cycle (FLW_DAYS), it is observed that environment L0133 promotes a longer vegetative period, leading to later flowering for the vast majority of genotypes. This contrasts sharply with the earliness observed in environment L0134. As for plant emergence (PL_EMERG_COUNT), a clear trend of superiority is noted in environment L0133, which exhibited higher mean stands for most genotypes compared to the other locations; the remaining environments showed no significant differentiation among themselves. Finally, regarding plant height (PH_R8), environment L0118 stood out negatively, showing a consistent trend of lower mean values across nearly all genotypes when compared to the other environments.


```{r}
#| label: fig-gt_boxplot_gy
#| fig.cap: "Genotypes Boxplot - Grain Yield"
#| echo: false

# Function 5 - Grain Yield
all_genotype_GY <- boxplot_genotype_all_in_one(selected_data,"GY")
all_genotype_GY$boxplot_gt
#all_genotype_GY$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_ph
#| fig.cap: "Genotypes Boxplot - Plant Height"
#| echo: false

# Function 5 - Plant Height
all_genotype_PH <- boxplot_genotype_all_in_one(selected_data,"PH_R8")
all_genotype_PH$boxplot_gt
#all_genotype_PH$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_ps
#| fig.cap: "Genotypes Boxplot - Pod Shattering"
#| echo: false

# Function 5 - Pod_Shattering
all_genotype_PS <- boxplot_genotype_all_in_one(selected_data,"Pod_Shattering")
all_genotype_PS$boxplot_gt
#all_genotype_PS$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_fl
#| fig.cap: "Genotypes Boxplot - Days to Flowering"
#| echo: false

# Function 5 - Days to Flowering
all_genotype_FL <- boxplot_genotype_all_in_one(selected_data,"FLW_DAYS")
all_genotype_FL$boxplot_gt
#all_genotype_FL$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_pl
#| fig.cap: "Genotypes Boxplot - Plant Emergence"
#| echo: false

# Function 5 - Plant Emergence
all_genotype_PL <- boxplot_genotype_all_in_one(selected_data,"PL_EMERG_COUNT")
all_genotype_PL$boxplot_gt
#all_genotype_PL$boxplot_gt_trial
```

The across-trials analysis revealed that vegetative vigor is a primary driver of yield performance in the Zambia trials, evidenced by the positive correlation of 0.45 between plant height (PH_R8) and grain yield (GY). This relationship indicates that genotypes with greater structural height provide a more robust physiological framework to support reproductive structures, allowing the breeder to use plant height as a stable and easily phenotyped proxy for yield potential. The positive associations between initial stand and development suggest that successful crop establishment is the foundational step in ensuring that the genetic potential for growth is expressed across diverse environmental conditions.


In contrast, the strong negative correlations between pod shattering and grain yield (-0.53) highlight a critical bottleneck in the production system, where pre-harvest seed loss acts as a direct constraint on final output. Additionally, the negative relationship between plant stand (PL) and yield (-0.54) suggests the presence of phenotypic plasticity or environmental stress in high-density scenarios, where increased plant populations did not translate into higher productivity. Taken together, these findings demonstrate that while vegetative vigor dictates the yield ceiling, yield stability in Zambia depends fundamentally on mitigating losses through shattering resistance and optimizing population density management.


```{r}
#| label: fig-across_corr
#| fig.cap: "Traits Correlation Across Trials"
#| echo: false 

# Function 6 - Across Trials Correlation
across_trials_corr <- across_trials_correlation(selected_data, "Across Trials Correlation", traits)
across_trials_corr 
```

# Trial 1 - Local "L0118" / Year "2023"

Trait correlations on trial 1 were generally low ( @fig-t1_corr). Pod shattering exhibited a weak positive correlation with grain yield (GY) (0.26), as did plant height at maturity (PH_R8) with days to flowering (FLW_DAYS) (0.25). A significant finding was the negative correlation between FLW_DAYS and GY (-0.45), indicating that early-maturing genotypes are more likely to achieve higher yields in this environment.


```{r}
#| label: fig-t1_corr
#| fig.cap: "Trial 1 - Traits Correlation"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$correlation
```

Trial 1 (L0118) appears to have faced challenges during the growing season. It presented the lowest mean grain yield (1,402.99 kg/ha), the longest time to flowering (50.34 days), reduced plant growth (51.34 cm), and higher pod shattering (2.27). These results may have been influenced by climatic conditions or specific management practices affecting genotypic performance. Interestingly, the plant emergence count (PL_EMERG_COUNT) was comparable to Trial 4, suggesting that the experiment did not encounter significant issues in the early phases. However, the high Coefficient of Variation (CV%) emphasizes that environmental heterogeneity significantly influenced the trial. This is further supported by the density plot ( @fig-all_trials_density), where Trial 1 appears clearly displaced from the other environments, converging on lower performance values.

```{r}
#| label: fig-t1_dens
#| fig.cap: "Trial 1 - Traits Density Plot"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$density
```

```{r}
#| label: fig-t1_box
#| fig.cap: "Trial 1 - Traits Boxplot"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$boxplot
```

```{r}
#| label: fig-t1_hist
#| fig.cap: "Trial 1 - Traits Histogram"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$boxplot
```

Examining correlations through scatter plots, the association between pod shattering and GY  appears counter intuitive ( @fig-t1_scatter2). Normally, higher-yielding plants are expected to exhibit lower shattering, as seed dispersal in the environment results in direct yield loss. Other correlations followed the trends previously discussed in the correlation matrix.

```{r}
#| label: fig-t1_scatter1
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p1 <-pairwise_scatter(trial_1,"Trial 1", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t1_p2 <-pairwise_scatter(trial_1,"Trial 1", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t1_p1 | t1_p2)
```

```{r}
#| label: fig-t1_scatter2
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p3 <- pairwise_scatter(trial_1,"Trial 1", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t1_p4 <- pairwise_scatter(trial_1,"Trial 1", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t1_p3 | t1_p4)
```

```{r}
#| label: fig-t1_scatter3
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p5 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t1_p6 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t1_p5 | t1_p6)
```

```{r}
#| label: fig-t1_scatter4
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p7 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t1_p8 <- pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t1_p7 | t1_p8)
```

```{r}
#| label: fig-t1_scatter5
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p9 <- pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t1_p10 <- pairwise_scatter(trial_1,"Trial 1", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t1_p9 | t1_p10)
```

Evaluating the trait grain yield (GY), nine genotypes performed above the trial mean (1,402.99 kg/ha). These genotypes showed consistency across replicates, indicating adaptation to these specific edaphoclimatic conditions ( @fig-t1_box_gy). Their performance suggests these are rustic materials suitable for restrictive environments. Notably, genotype G233 stood out for its expressive yield and high stability (consistent values among replicates) despite the other genotypes' performance.

```{r}
#| label: fig-t1_box_gy
#| fig.cap: "Trial 1 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "GY", "GY (kg/ha)")
```

For plant height (PH_R8), no clear pattern emerged among the boxplot graph. Several materials showed large discrepancies between observations, as evidenced by the high CV (58%), which indicates low experimental precision for this trait. Remarkably, genotypes G072, G214, G244, G245, and G311 showed lower intra-genotypic variation, though they were characterized by shorter stature, except genotype G244.

```{r}
#| label: fig-t1_box_ph
#| fig.cap: "Trial 1 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "PH_R8", "PH (cm)")
```

In terms of plant emergence (PL_EMERG_COUNT), genotypes G076 and G319 were prominent. Although they had lower absolute emergence counts, they were more consistent across replicates compared to G188, G198, and G263, which had higher means but greater dispersion. For FLW_DAYS, genotype G057 was a highlight for its precocity, reaching flowering earlier than the rest. The reduced amplitude of the boxplots for this trait suggests lower environmental influence on flowering time. The most promising genotypes for an "escape" strategy (faster flowering to ensure a shorter cycle) are G233, G076, and G143. Conversely, G359, G315, G246, and G214 showed high stability despite being later-maturing. Consequently, two selection strategies could be employed: selecting early-flowering genotypes (e.g., G076 and G233) to escape terminal drought, or selecting longer-cycle genotypes (e.g., G321) to maximize yield potential if rainfall distribution allows.

```{r}
#| label: fig-t1_box_fl
#| fig.cap: "Trial 1 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t1_box_ps
#| fig.cap: "Trial 1 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t1_box_pl
#| fig.cap: "Trial 1 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 2 - Local "L0133" / Year "2023"

The Trial 2 (L0133) stands out as a high-potential environment with strong phenotypic stability for most traits. The mean GY reached 3,676 kg/ha, with a median of 3,712 kg/ha. This symmetry, achieved after the removal of previously identified outliers, indicates a balanced data distribution, allowing for a reliable assessment of genotype performance. The range (1,421 to 5,684 kg/ha) demonstrates that despite being a favorable environment, there is sufficient genetic variability to select superior materials. The density plot for GY confirms this trend, showing a well-defined curve shifted toward high-yield levels.


Regarding the crop cycle, the mean of 56.7 days to flowering (FLW_DAYS) confirms that this environment induces a longer cycle. The low standard deviation (5.64) and the narrow peak in the density plot suggest that the genotypes responded uniformly to local photoperiodic or thermal stimuli. However, PL_EMERG_COUNT showed a lot variability (9 to 219 plants) ( @fig-t2_corr). The density plot for emergence displayed a flat and wide base, visually confirming this lack of uniformity ( @fig-t2_dens). 


```{r}
#| label: fig-t2_corr
#| fig.cap: "Trial 2 - Traits Correlation"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$correlation
```

A correlation of approximatelly -0.30 between FLW_DAYS and GY was observed when contrasted with PL_EMERG_COUNT (FLW_DAYS x PL_EMERG_COUNT; FLW_DAYS x PL_EMERG_COUNT), indicating that while the stand was irregular. Similarly, the -0.35 correlation between PH_R8 and GY suggests that taller plants were not necessarily the most productive at this site; in high-density areas, etiolation for light competition may have occurred, however, these clarification values are not very high. Finally, Pod_Shattering showed a mean of 1.07, indicating that environmental conditions exerted no selection pressure for pod dehiscence. Correlations near zero with other variables confirm that this trait remained "silent" in Trial 2. Therefore, while this site is excellent for evaluating yield potential and plant architecture, it does not provide discriminatory information for shattering resistance, as nearly all materials displayed stable behavior and minimum scores.

```{r}
#| label: fig-t2_dens
#| fig.cap: "Trial 2 - Traits Density Plot"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$density
```

Regarding the boxplots comparing treatments within each genotype, it is visually evident that most flowering data, regardless of the genotype, is concentrated within a range of 50–60 days ( @fig-t2_box_fl), with few individuals falling outside this window. This indicates high phenotypic stability for this trait or a uniform response to the photoperiod and temperature conditions of the specific site and growing season. A similar pattern is observed for pod shattering, where observations across genotypes remained between 0.5 and 1.5 ( @fig-t2_box_ps), suggesting that the evaluated population possesses high mechanical resistance to pod dehiscence and low genetic variability for this in Trial 2.

```{r}
#| label: fig-t2_box
#| fig.cap: "Trial 2 - Traits Boxplot"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$boxplot
```

```{r}
#| label: fig-t2_hist
#| fig.cap: "Trial 2 - Traits Histogram"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$boxplot
```

The scatter plot analysis for Trial 2 clarifies the trade-offs between traits under high-potential conditions. A notable negative trend was observed between plant height and yield, where the highest-yielding genotypes were concentrated within an intermediate height range (80–100 cm). Plants exceeding 110 cm showed a significant decline in productivity, suggesting that excessive vegetative growth at this site was likely a result of etiolation due to light competition rather than high genetic vigor ( @fig-t2_scatter1). This indicates an inefficient energy partitioning, where the plants prioritized stem elongation over grain filling.

```{r}
#| label: fig-t2_scatter1
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p1 <-pairwise_scatter(trial_2,"Trial 2", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t2_p2 <-pairwise_scatter(trial_2,"Trial 2", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t2_p1 | t2_p2)
```

```{r}
#| label: fig-t2_scatter2
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p3 <- pairwise_scatter(trial_2,"Trial 2", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t2_p4 <- pairwise_scatter(trial_2,"Trial 2", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t2_p3 | t2_p4)
```

```{r}
#| label: fig-t2_scatter3
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p5 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t2_p6 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t2_p5 | t2_p6)
```

```{r}
#| label: fig-t2_scatter4
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p7 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t2_p8 <- pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t2_p7 | t2_p8)
```

```{r}
#| label: fig-t2_scatter5
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p9 <- pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t2_p10 <- pairwise_scatter(trial_2,"Trial 2", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t2_p9 | t2_p10)
```

In Trial 2, the most significant contrasts between genotypes are observed in Grain Yield (GY), Plant Height (PH), and plant emergence. Regarding GY, genotypes G072, G076, G137, G143, G194, G216, G219, G246, and G359 stood out with positive performance, while G146, G198, G315, G318, G319, and G321 showed negative highlights, with means settling near 3,000 kg/ha  ( @fig-t2_box_gy).


In contrast, for PH, genotype G076 presented a negative highlight compared to the others, with a mean height near 60 cm. Interestingly, genotypes G244, G311, G315, G318, G319, and G321 obtained the highest means for height ( @fig-t2_box_ph). This suggests a potential inverse relationship or trade-off in certain materials, where excessive investment in vegetative growth and taller stature (as seen in the G315–G321 group) may be occurring at the expense of grain productivity, whereas compact genotypes like G076 demonstrate a more efficient harvest index. These visual observations align with the correlation analysis, confirming a negative association between PH and GY in this trial, discussed previously.

```{r}
#| label: fig-t2_box_gy
#| fig.cap: "Trial 2 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t2_box_ph
#| fig.cap: "Trial 2 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t2_box_fl
#| fig.cap: "Trial 2 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t2_box_ps
#| fig.cap: "Trial 2 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t2_box_pl
#| fig.cap: "Trial 2 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 3 - Local "L074" / Year "2023"

As in the other trials, the correlation values between the traits analysed were also low. The biggest correlation observed was with plant emergence (PL_EMERG_COUNT) and pod shattering, with a value around 0.34. This value is low, and does not suggest a strong relationship between the variables.

```{r}
#| label: fig-t3_corr
#| fig.cap: "Trial 3 - Traits Correlation"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$correlation
```

```{r}
#| label: fig-t3_dens
#| fig.cap: "Trial 3 - Traits Density Plot"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$density
```

Analysing the grain yield (GY) information displayed in @fig-t3_hist, @fig-t3_dens and @fig-t3_box, it can be seen that this trial was more productive than the first one, although the values were not higher than in Trial 4. Regarding days to flowering (FLW_DAYS), the mean value was lower than on the previous trials, indicating that the genotypes in this environment may have an early flowering or shorter cycles, even though it can be seen a genetic variation of this trait in @fig-t3_dens. Looking at plant height (PH_R8), the distribution is similar to a normal, and the mean value is slightly smaller than the observed in trials 2 and 4, indicating that this environment may result in shorter plants. For plant emergence (PL_EMERG_COUNT), the mean value is next to 180, however, there is a large variation in this trait, which can be caused by differences in seed germination and vigor, or operational issues during planting. Finally, analyzing pod shattering, it is noticeable that most of the genotypes were scored as “1”, while only 6 genotypes were scored as “2”, showing that, in this trial, there were no major issues with pod shattering. 
The “unusual values” shown in the @fig-t3_box for grain yield and pod shattering were not removed because they are possible values.

```{r}
#| label: fig-t3_box
#| fig.cap: "Trial 3 - Traits Boxplot"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$boxplot
```

```{r}
#| label: fig-t3_hist
#| fig.cap: "Trial 3 - Traits Histogram"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$boxplot
```

Looking into the pairwise scatter plots, clear patterns can not be detected, as the correlation values in this trial, for the traits selected, are low. 

```{r}
#| label: fig-t3_scatter1
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p1 <-pairwise_scatter(trial_3,"Trial 3", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t3_p2 <-pairwise_scatter(trial_3,"Trial 3", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t3_p1 | t3_p2)
```

```{r}
#| label: fig-t3_scatter2
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p3 <- pairwise_scatter(trial_3,"Trial 3", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t3_p4 <- pairwise_scatter(trial_3,"Trial 3", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t3_p3 | t3_p4)
```

```{r}
#| label: fig-t3_scatter3
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p5 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t3_p6 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t3_p5 | t3_p6)
```

```{r}
#| label: fig-t3_scatter4
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p7 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t3_p8 <- pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t3_p7 | t3_p8)
```

```{r}
#| label: fig-t3_scatter5
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p9 <- pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t3_p10 <- pairwise_scatter(trial_3,"Trial 3", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t3_p9 | t3_p10)
```

In trial 3, the grain yield (GY) information was also analysed at genotype level, as shown in @fig-t3_box_gy. There were genotypes with mean yield around 2,000 kg/ha and others with means superior to 4000 kg/ha, indicating that the performance of genotypes are different in this environment. Besides that, it can be seen that in some cases, there is a considerable variation within genotype level, which can be caused by several agronomical, climate and other environmental effects. Ideally, the experimental design and the statistical analysis should be able to deal with this variation. For other traits, the variation within genotypes and across genotypes can also be seen. For example, in plant emergence ( @fig-t3_box_pl), there were genotypes with very slightly variations, while others had almost 50% variation.

```{r}
#| label: fig-t3_box_gy
#| fig.cap: "Trial 3 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t3_box_ph
#| fig.cap: "Trial 3 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t3_box_fl
#| fig.cap: "Trial 3 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t3_box_ps
#| fig.cap: "Trial 3 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t3_box_pl
#| fig.cap: "Trial 3 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 4 - Local "L0134" / Year "2023"

In Trial 4 (Location L0134), the phenotypic analysis reveals a lack of strong associations between characters, with the highest correlation observed between Grain Yield (GY) and Plant Height (PH) at a weak value of r = 0.19 ( @fig-t4_corr). This independence between traits is further elucidated by the density distributions ( @fig-t4_dens). The flowering data (FLW) shows a bimodal distribution with two prominent peaks, indicating a clear division of the population into two distinct maturation groups under these environmental conditions. 

```{r}
#| label: fig-t4_corr
#| fig.cap: "Trial 4 - Traits Correlation"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$correlation
```

Regarding Grain Yield (GY), the density curve is nearly flat at the top, reflecting high variability and the absence of a dominant performance trend among genotypes. While PH shows a more defined peak compared to other traits, its weak correlation with GY suggests that height was not a limiting or determining factor for productivity in this trial. Overall, the dispersed nature of the yield data and the varied density profiles indicate that at L0134, genotypic performance was highly decentralized, with different traits responding independently to environmental stimuli. 

```{r}
#| label: fig-t4_dens
#| fig.cap: "Trial 4 - Traits Density Plot"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$density
```

Complementing this, the boxplots ( @fig-t4_box) identify potential outliers for PH, Plant Emergence, and Pod Shattering, each represented by a single data point positioned away from the main distribution. In contrast, FLW and GY do not exhibit such atypical values, showing more continuous distributions. Furthermore, the scatterplots confirm the lack of a clear linear trend between GY and the other agronomic traits ( @fig-t4_scatter1; @fig-t4_scatter2). Although the correlation between PH and GY was the highest recorded in this trial, its low magnitude (0.19) reinforces that, in this specific location, traditional yield components did not act as strong predictors for final output.

```{r}
#| label: fig-t4_box
#| fig.cap: "Trial 4 - Traits Boxplot"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$boxplot
```

```{r}
#| label: fig-t4_hist
#| fig.cap: "Trial 4 - Traits Histogram"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$boxplot
```

```{r}
#| label: fig-t4_scatter1
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p1 <-pairwise_scatter(trial_4,"Trial 4", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t4_p2 <-pairwise_scatter(trial_4,"Trial 4", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t4_p1 | t4_p2)
```

```{r}
#| label: fig-t4_scatter2
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p3 <- pairwise_scatter(trial_4,"Trial 4", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t4_p4 <- pairwise_scatter(trial_4,"Trial 4", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t4_p3 | t4_p4)
```

```{r}
#| label: fig-t4_scatter3
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p5 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t4_p6 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t4_p5 | t4_p6)
```

```{r}
#| label: fig-t4_scatter4
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p7 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t4_p8 <- pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t4_p7 | t4_p8)
```

```{r}
#| label: fig-t4_scatter5
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p9 <- pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t4_p10 <- pairwise_scatter(trial_4,"Trial 4", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t4_p9 | t4_p10)
```

In the comparative boxplots among genotypes, a significant amplitude in Grain Yield (GY) was observed within the same environment ( @fig-t4_box_gy). Genotypes G057 and G137 were the top performers, with means exceeding 6000 kg/ha, contrasting sharply with G204 and G215, which fell below 4000 kg/ha. A similar trend was noted for Plant Height (PH), where genotypes G057 and G146 stood out with heights above 100 cm (notably, G057 excelled in both yield and stature) ( @fig-t4_box_ph). In contrast, genotype G076 remained below 60 cm, while the remaining genotypes clustered visually above the 80 cm mark. Furthermore, flowering data remained tightly grouped between 40–50 days, and pod shattering values were consistently below 1.5, with the exception of a single observation for genotype G263 that exceeded 2.0, identifying it as a potential outlier ( @fig-t4_box_ps).

```{r}
#| label: fig-t4_box_gy
#| fig.cap: "Trial 4 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t4_box_ph
#| fig.cap: "Trial 4 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t4_box_fl
#| fig.cap: "Trial 4 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t4_box_ps
#| fig.cap: "Trial 4 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t4_box_pl
#| fig.cap: "Trial 4 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "PL_EMERG_COUNT", "Plant Emergence")
```

# Conclusion
In conclusion, a cross-environmental analysis of Grain Yield (GY) highlights contrasting strategies among the evaluated materials. Genotypes G057 and G137 demonstrated remarkable performance, maintaining a positive standout status in three out of the four evaluated trials. This consistency suggests high yield stability and a robust genetic potential that is less sensitive to environmental fluctuations across different locations in Zambia.

In contrast, genotype G146 exhibited high environmental sensitivity (Genotype x Environment Interaction). While it maintained average performance in two locations, it showed significant negative highlights for GY in the remaining two trials. These results indicate that while G057 and G137 are strong candidates for broad adaptation, G146's productivity is highly dependent on specific local conditions, making it a riskier choice for environments with high unpredictability.

