---
title: "Exploratory Data Analysis"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true

crossref: {}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

# Introduction

This report presents a comprehensive exploratory data analysis of a soybean breeding program in **Zambia**, exploring **four trials** and **five key soybean traits**. The main objectives are to assess variation, detect unusual values, compute measures of central tendency and dispersion, and investigate relationships between traits.

# Data characterization

The dataset is composed of phenotypic records from a soybean breeding program conducted in Zambia during the 2023 growing season. A total of 29 genotypes were evaluated across four distinct locations (L0118, L0133, L074, and L0134), representing different environmental conditions. Each genotype was replicated three times per location, except for "G321", which had six replicates in each trial.

Five agronomic traits were selected based on data availability and breeding importance: Days to Flowering (FLW_DAYS), Grain Yield (GY), Plant Height at R8 stage (PH_R8), Emergence Count (PL_EMERG_COUNT), and Pod Shattering.

```{r}
#| label: trials-data

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4")
genotypes_4 <- trial_4$gen
```

# Used Functions

```{r}
#| label: functions

# Function 1 - Single Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      cv = (sd/mean),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  # Correlation between traits
  cor_mat <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")
  
  p1 <-ggcorrplot(cor_mat, 
           hc.order = TRUE,           
           type = "upper",               
           lab = TRUE,      
           lab_size = 3,           
           colors = c("red", "white", "blue"),      
           title = title)   
  
  # Traits Boxplots
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p4 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = p1,boxplot = p2, density = p3, histogram = p4))
}


# Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (scatter = p1)
}

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_text(size = 10)) +
        theme(axis.text.x = element_text(size = 8))
  
  return (boxplot_genotypes = p1)
}

# Function 4 - Single Trials Analysis Together
single_trials_together <- function(df, title, traits) {
  
  # Dataframe long
  df_long <- df |> 
    mutate(Trial = as.factor(trial)) |> 
    pivot_longer(
      cols = all_of(traits),
      names_to = "Trait",
      values_to = "Value") |> 
    filter(!is.na(Value))
  
  # Boxplots 
  p1 <- ggplot(df_long, aes(x = Trial, y = Value, fill = Trial)) +
          geom_boxplot(outlier.colour = "black", outlier.size = 0.5) +
          facet_wrap(~Trait, scales = "free") +
          theme_minimal() +
          labs(title = title)

  # Density Plot
  p2 <- ggplot(df_long, aes(x = Value, fill = Trial, group = Trial)) +
          geom_density(alpha = 0.4) +
          facet_wrap(~Trait, scales = "free") +
          theme_light() +
          labs(title = title)

return (list(boxplots = p1, density = p2))
}

# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("All Trials Boxplot:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}


# Function 6 - Across Trials Correlation
across_trials_correlation <- function(df, title, traits) {

  # Correlation Plot with all trials
  cor_total <- df |>
    select(all_of(traits)) |>
    cor(use = "pairwise.complete.obs")

  p1 <-ggcorrplot(cor_total, 
         hc.order = TRUE,           
         type = "upper",               
         lab = TRUE,      
         lab_size = 3,           
         colors = c("red", "white", "blue"),      
         title = title)
  
  return(all_cor = p1)
}
```

# Trials Overview

First, some basic statistics were computed for each trial, to facilitate comparison and evaluation of traits:

```{r}
#| label: stats-table1
#| echo: false

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}

# Table 
envs <- c("T1", "T2", "T3", "T4")
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")

df_conc <- map_df(seq_along(all_trials_summary), function(i) {
  all_trials_summary[[i]]$stats |> 
    mutate(Local = envs[i])})

table_df <- df_conc |>
    pivot_longer(
    cols = c(mean, median, variance, sd, cv, min, max, range), 
    names_to = "Stat", 
    values_to = "Value"
  ) |>
   mutate(Stat = case_match(Stat,
    "mean" ~ "Mean",
    "median" ~ "Median",
    "variance" ~ "Variance",
    "sd" ~ "Sd",
    "cv" ~ "Cv",
    "min" ~ "Min",
    "max" ~ "Max",
    "range" ~ "Range"
  )) |> 
  pivot_wider(names_from = c(Variable, Local), values_from = Value)

n_traits <- 5
n_local <- 4

sup_header <- c(" " = 1)
sup_header[traits] <- n_local 

order <- unlist(lapply(traits, function(trait) {
  paste0(trait, "_", envs)}))

table_df <- table_df |>
  select(Stat, all_of(order))

```

```{r}
#| label: tbl-stats-table2
#| tbl-cap: "Measures of central tendency and dispersion of trials conducted in Zambia"
#| echo: false
table_df |>
  kable(
    digits = 2,
    align = "l",
    col.names = c(" ", rep(envs, n_traits))) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE
  ) |>
  add_header_above(sup_header) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```

In @tbl-stats-table2 ... Outlier Trial 1 (PH_R8) -\> removing outlier observation:

```{r}
#| label: outlier

# Removing outliers
trial_1 <- trial_1 |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8))
trial_3 <- trial_3 |> 
  mutate(across(all_of(traits),~ if_else(gen == "G216" & env == "E0282", NA_real_ ,.))) |>
  mutate(FLW_DAYS = ifelse(FLW_DAYS == 0, NA, FLW_DAYS)) 

selected_data <- selected_data |> 
  mutate(PH_R8 = ifelse(PH_R8 > 200, NA, PH_R8)) |> 
  mutate(across(all_of(traits),~ if_else(gen == "G216" & env == "E0282", NA_real_ ,.))) |> 
  mutate(FLW_DAYS = ifelse(FLW_DAYS == 0, NA, FLW_DAYS))

# Function 1 - Single Trials Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)
for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 50)
}
```

Now, some plots analysing all the traits within each trial:

```{r}
#| label: fig-all_trials_density
#| fig.cap: "Density plot of traits within each trial"
#| echo: false

# Function 4 - Single Trials General Overview
all_trials_together <- single_trials_together(selected_data, "All Trials Overview", traits)
all_trials_together$density
```

```{r}
#| label: fig-all_trials_boxplot
#| fig.cap: "Boxplot of traits within each trial"
#| echo: false
all_trials_together$boxplot
```

Analysing the boxplot @fig-all_trials_boxplot aaa

```{r}
#| label: fig-gt_boxplot_gy
#| fig.cap: "Genotypes Boxplot - Grain Yield"
#| echo: false

# Function 5 - Grain Yield
all_genotype_GY <- boxplot_genotype_all_in_one(selected_data,"GY")
all_genotype_GY$boxplot_gt
#all_genotype_GY$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_ph
#| fig.cap: "Genotypes Boxplot - Plant Height"
#| echo: false

# Function 5 - Plant Height
all_genotype_PH <- boxplot_genotype_all_in_one(selected_data,"PH_R8")
all_genotype_PH$boxplot_gt
#all_genotype_PH$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_ps
#| fig.cap: "Genotypes Boxplot - Pod Shattering"
#| echo: false

# Function 5 - Pod_Shattering
all_genotype_PS <- boxplot_genotype_all_in_one(selected_data,"Pod_Shattering")
all_genotype_PS$boxplot_gt
#all_genotype_PS$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_fl
#| fig.cap: "Genotypes Boxplot - Days to Flowering"
#| echo: false

# Function 5 - Days to Flowering
all_genotype_FL <- boxplot_genotype_all_in_one(selected_data,"FLW_DAYS")
all_genotype_FL$boxplot_gt
#all_genotype_FL$boxplot_gt_trial
```

```{r}
#| label: fig-gt_boxplot_pl
#| fig.cap: "Genotypes Boxplot - Plant Emergence"
#| echo: false

# Function 5 - Plant Emergence
all_genotype_PL <- boxplot_genotype_all_in_one(selected_data,"PL_EMERG_COUNT")
all_genotype_PL$boxplot_gt
#all_genotype_PL$boxplot_gt_trial
```

```{r}
#| label: fig-across_corr
#| fig.cap: "Traits Correlation Across Trials"
#| echo: false 

# Function 6 - Across Trials Correlation
across_trials_corr <- across_trials_correlation(selected_data, "Across Trials Correlation", traits)
across_trials_corr 
```

# Trial 1 - Local "L0118" / Year "2023"

```{r}
#| label: fig-t1_corr
#| fig.cap: "Trial 1 - Traits Correlation"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$correlation
```

```{r}
#| label: fig-t1_dens
#| fig.cap: "Trial 1 - Traits Density Plot"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$density
```

```{r}
#| label: fig-t1_box
#| fig.cap: "Trial 1 - Traits Boxplot"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$boxplot
```

```{r}
#| label: fig-t1_hist
#| fig.cap: "Trial 1 - Traits Histogram"
#| echo: false 

# Trial 1
all_trials_summary[[1]]$boxplot
```

```{r}
#| label: fig-t1_scatter1
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p1 <-pairwise_scatter(trial_1,"Trial 1", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t1_p2 <-pairwise_scatter(trial_1,"Trial 1", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t1_p1 | t1_p2)
```

```{r}
#| label: fig-t1_scatter2
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p3 <- pairwise_scatter(trial_1,"Trial 1", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t1_p4 <- pairwise_scatter(trial_1,"Trial 1", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t1_p3 | t1_p4)
```

```{r}
#| label: fig-t1_scatter3
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p5 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t1_p6 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t1_p5 | t1_p6)
```

```{r}
#| label: fig-t1_scatter4
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p7 <- pairwise_scatter(trial_1,"Trial 1", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t1_p8 <- pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t1_p7 | t1_p8)
```

```{r}
#| label: fig-t1_scatter5
#| fig.cap: "Trial 1 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t1_p9 <- pairwise_scatter(trial_1,"Trial 1", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t1_p10 <- pairwise_scatter(trial_1,"Trial 1", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t1_p9 | t1_p10)
```

```{r}
#| label: fig-t1_box_gy
#| fig.cap: "Trial 1 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t1_box_ph
#| fig.cap: "Trial 1 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t1_box_fl
#| fig.cap: "Trial 1 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t1_box_ps
#| fig.cap: "Trial 1 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t1_box_pl
#| fig.cap: "Trial 1 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Trial 1", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 2 - Local "L0133" / Year "2023"

```{r}
#| label: fig-t2_corr
#| fig.cap: "Trial 2 - Traits Correlation"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$correlation
```

```{r}
#| label: fig-t2_dens
#| fig.cap: "Trial 2 - Traits Density Plot"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$density
```

```{r}
#| label: fig-t2_box
#| fig.cap: "Trial 2 - Traits Boxplot"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$boxplot
```

```{r}
#| label: fig-t2_hist
#| fig.cap: "Trial 2 - Traits Histogram"
#| echo: false 

# Trial 2
all_trials_summary[[2]]$boxplot
```

```{r}
#| label: fig-t2_scatter1
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p1 <-pairwise_scatter(trial_2,"Trial 2", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t2_p2 <-pairwise_scatter(trial_2,"Trial 2", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t2_p1 | t2_p2)
```

```{r}
#| label: fig-t2_scatter2
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p3 <- pairwise_scatter(trial_2,"Trial 2", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t2_p4 <- pairwise_scatter(trial_2,"Trial 2", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t2_p3 | t2_p4)
```

```{r}
#| label: fig-t2_scatter3
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p5 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t2_p6 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t2_p5 | t2_p6)
```

```{r}
#| label: fig-t2_scatter4
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p7 <- pairwise_scatter(trial_2,"Trial 2", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t2_p8 <- pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t2_p7 | t2_p8)
```

```{r}
#| label: fig-t2_scatter5
#| fig.cap: "Trial 2 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t2_p9 <- pairwise_scatter(trial_2,"Trial 2", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t2_p10 <- pairwise_scatter(trial_2,"Trial 2", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t2_p9 | t2_p10)
```

```{r}
#| label: fig-t2_box_gy
#| fig.cap: "Trial 2 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t2_box_ph
#| fig.cap: "Trial 2 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t2_box_fl
#| fig.cap: "Trial 2 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t2_box_ps
#| fig.cap: "Trial 2 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t2_box_pl
#| fig.cap: "Trial 2 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_2,"Trial 2", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 3 - Local "L074" / Year "2023"

```{r}
#| label: fig-t3_corr
#| fig.cap: "Trial 3 - Traits Correlation"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$correlation
```

```{r}
#| label: fig-t3_dens
#| fig.cap: "Trial 3 - Traits Density Plot"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$density
```

```{r}
#| label: fig-t3_box
#| fig.cap: "Trial 3 - Traits Boxplot"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$boxplot
```

```{r}
#| label: fig-t3_hist
#| fig.cap: "Trial 3 - Traits Histogram"
#| echo: false 

# Trial 3
all_trials_summary[[3]]$boxplot
```

```{r}
#| label: fig-t3_scatter1
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p1 <-pairwise_scatter(trial_3,"Trial 3", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t3_p2 <-pairwise_scatter(trial_3,"Trial 3", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t3_p1 | t3_p2)
```

```{r}
#| label: fig-t3_scatter2
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p3 <- pairwise_scatter(trial_3,"Trial 3", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t3_p4 <- pairwise_scatter(trial_3,"Trial 3", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t3_p3 | t3_p4)
```

```{r}
#| label: fig-t3_scatter3
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p5 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t3_p6 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t3_p5 | t3_p6)
```

```{r}
#| label: fig-t3_scatter4
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p7 <- pairwise_scatter(trial_3,"Trial 3", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t3_p8 <- pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t3_p7 | t3_p8)
```

```{r}
#| label: fig-t3_scatter5
#| fig.cap: "Trial 3 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t3_p9 <- pairwise_scatter(trial_3,"Trial 3", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t3_p10 <- pairwise_scatter(trial_3,"Trial 3", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t3_p9 | t3_p10)
```

```{r}
#| label: fig-t3_box_gy
#| fig.cap: "Trial 3 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t3_box_ph
#| fig.cap: "Trial 3 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t3_box_fl
#| fig.cap: "Trial 3 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t3_box_ps
#| fig.cap: "Trial 3 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t3_box_pl
#| fig.cap: "Trial 3 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_3,"Trial 3", "PL_EMERG_COUNT", "Plant Emergence")
```

# Trial 4 - Local "L0134" / Year "2023"

```{r}
#| label: fig-t4_corr
#| fig.cap: "Trial 4 - Traits Correlation"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$correlation
```

```{r}
#| label: fig-t4_dens
#| fig.cap: "Trial 4 - Traits Density Plot"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$density
```

```{r}
#| label: fig-t4_box
#| fig.cap: "Trial 4 - Traits Boxplot"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$boxplot
```

```{r}
#| label: fig-t4_hist
#| fig.cap: "Trial 4 - Traits Histogram"
#| echo: false 

# Trial 4
all_trials_summary[[4]]$boxplot
```

```{r}
#| label: fig-t4_scatter1
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p1 <-pairwise_scatter(trial_4,"Trial 4", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")
t4_p2 <-pairwise_scatter(trial_4,"Trial 4", "GY", "FLW_DAYS", "GY (kg/ha)", "Days to Flowering")
(t4_p1 | t4_p2)
```

```{r}
#| label: fig-t4_scatter2
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p3 <- pairwise_scatter(trial_4,"Trial 4", "GY", "PL_EMERG_COUNT", "GY (kg/ha)", "Plant Emergence")
t4_p4 <- pairwise_scatter(trial_4,"Trial 4", "GY", "Pod_Shattering", "GY (kg/ha)", "Pod_Shattering score")
(t4_p3 | t4_p4)
```

```{r}
#| label: fig-t4_scatter3
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p5 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PH_R8", "Pod_Shattering score", "PH (cm)")
t4_p6 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "FLW_DAYS", "Pod_Shattering score", "Days to Flowering")
(t4_p5 | t4_p6)
```

```{r}
#| label: fig-t4_scatter4
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p7 <- pairwise_scatter(trial_4,"Trial 4", "Pod_Shattering", "PL_EMERG_COUNT", "Pod_Shattering score", "Plant Emergence")
t4_p8 <- pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "PH_R8", "Plant Emergence", "PH (cm)")
(t4_p7 | t4_p8)
```

```{r}
#| label: fig-t4_scatter5
#| fig.cap: "Trial 4 - Scatter Plots"
#| echo: false 
# Function 2 - Single Trial Pairwise Scatter Plot
t4_p9 <- pairwise_scatter(trial_4,"Trial 4", "PL_EMERG_COUNT", "FLW_DAYS", "Plant Emergence", "Days to Flowering")
t4_p10 <- pairwise_scatter(trial_4,"Trial 4", "FLW_DAYS", "PH_R8", "Days to Flowering", "PH (cm)")
(t4_p9 | t4_p10)
```

```{r}
#| label: fig-t4_box_gy
#| fig.cap: "Trial 4 Genotypes Boxplot - Grain Yield"
#| echo: false 
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "GY", "GY (kg/ha)")
```

```{r}
#| label: fig-t4_box_ph
#| fig.cap: "Trial 4 Genotypes Boxplot - Plant Height"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "PH_R8", "PH (cm)")
```

```{r}
#| label: fig-t4_box_fl
#| fig.cap: "Trial 4 Genotypes Boxplot - Flowering Days"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "FLW_DAYS", "Days to Flowering")
```

```{r}
#| label: fig-t4_box_ps
#| fig.cap: "Trial 4 Genotypes Boxplot - Pod Shattering"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "Pod_Shattering", "Pod_Shattering score")
```

```{r}
#| label: fig-t4_box_pl
#| fig.cap: "Trial 4 Genotypes Boxplot - Plant Emergence"
#| echo: false
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_4,"Trial 4", "PL_EMERG_COUNT", "Plant Emergence")
```

## A FAZER

3\) Definir critério de avaliação para os outliers (Trial 1 tem 1 outlier na altura)

4\) Organizar cunks, formatar documentos e fazer as interpretações
