---
title: "Multi Trial EDA"
author: 
  - name: João Vitor Sanches Calvo
    orcid: 0009-0002-5827-6109
    email: joaovitor.calvo@usp.br
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
---

# Introduction 

This report examines a  comprehensive exploratory data analysis of a soybean breeding program in Zambia, exploring four trials and five key soybean traits. The main objectives are to assess variation, detect unusual values, compute measures of central tendency and dispersion, and investigate relationships between traits.

# Data characterization
The dataset is composed of phenotypic records from a soybean breeding program conducted in Zambia during the 2023 growing season. . A total of 30 genotypes were evaluated across four distinct locations (L0118, L0133, L074, and L0134), representing different environmental conditions. Each genotype had triplicate per location. Five agronomic traits were selected based on their classic and breeding objectives: Days to Flowering (FLW_DAYS), Grain Yield (GY), Plant Height at R8 stage (PH_R8), Emergence Count (PL_EMERG_COUNT), and Pod Shattering. The data were collected in 2023 and only the trait PL_EMERG_COUNT did not have data. Each genotype is replicated three times in each location.

```{r}
#| warning: false

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)
library(patchwork)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(trial == "1")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(trial == "2")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(trial == "3")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L0134" / Year "2023" 
trial_4 <- selected_data |> filter(trial == "4")
genotypes_4 <- trial_4$gen
```

To quantify the phenotypic variation among the genotypes and experimental precision across the four locations, a summary of descriptive statistics was compiled (@tabela-trait-local). This includes measures of central tendency, such as the mean and median, alongside dispersion metrics such as standard deviation and the Coefficient of Variation (CV%). 


```{r}
#| warning: FALSE
#| label: tabela-trait-local
#| tbl-cap: "Summary Statistics of soybean traits across four locations in Zambia"

# individual trial analysis
trial_summary <- function(df, title, traits, bins) {
  df_long <- df |>
    pivot_longer(cols = all_of(traits), names_to = "Variable", values_to = "value")
  
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE),
      cv = (sd / mean) * 100 
    ) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  cor_mat <- cor(df |> select(all_of(traits)), use = "complete.obs")
  p1 <- ggcorrplot(cor_mat, hc.order = TRUE, type = "upper", lab = TRUE, title = title)
  
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) + facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable), shape = 17) +
        theme_minimal()
  return(list(stats = stats_df, correlation = p1, boxplot = p2))
}

# Loop
traits_list <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
envs_list   <- c("L0118", "L0133", "L074", "L0134")
trials_data <- list(trial_1, trial_2, trial_3, trial_4)

all_trials_summary <- list()

for(i in seq_along(trials_data)) {
  all_trials_summary[[i]] <- trial_summary(trials_data[[i]], envs_list[i], traits_list, 30)
}


df_empilhado <- map_df(seq_along(all_trials_summary), function(i) {
  all_trials_summary[[i]]$stats |> 
    mutate(Local = envs_list[i])
})

# kable
df_para_tabela <- df_empilhado |>
  mutate(Variable = factor(Variable, levels = traits_list)) |> 
  pivot_longer(
    cols = c(mean, median, variance, sd, cv, min, max, range), 
    names_to = "Statistic", 
    values_to = "Value"
  ) |>
  mutate(Statistic = factor(Statistic, 
                            levels = c("mean", "median", "variance", "sd", "cv", "min", "max", "range"),
                            labels = c("Mean", "Median", "Variance", "Std. Deviation", "CV (%)", "Minimum", "Maximum", "Range"))) |>
  pivot_wider(names_from = c(Variable, Local), values_from = Value) |>
  arrange(Statistic)

# table
header_top <- c(" " = 1)
for(t in traits_list) { header_top[t] <- length(envs_list) }

df_para_tabela |>
  kable(digits = 2, align = "l", col.names = c("Statistic", rep(envs_list, length(traits_list)))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) |>
  add_header_above(header_top) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```

```{r}
#| warnings: false
#| fig-cap: "Comparative boxplots of five agronomic traits across four experimental trials"
#| label: box-single-trial
## Function 4 - Single Trials Analysis Together
all_trials_together <- single_trials_together(selected_data, " ", traits)

# Outputs
all_trials_together$boxplot
```

```{r}
#| warnings: false
#| fig-cap: "Comparative density plots of five agronomic traits across four experimental trials"
#| label: dens-single-trial

all_trials_together$density
```


# Analysis by genotype

```{r}
# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18)) +
        theme(axis.text.x = element_text(size = 10))
  
  return (boxplot_genotypes = p1)
}

# Output


```

# Analysis between genotypes

Regarding the boxplot analyses performed by trait, comparing genotypes and environments, clear trends can be observed. For the Grain Yield (GY) trait, mean values are consistently higher across almost all genotypes in environment L0134. This location stands out as a high-potential environment, likely providing optimal edaphoclimatic conditions that allowed the germplasm to express its maximum genetic potential.

```{r}
# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}
#| label: fig-genotype-comparison
#| fig-width: 22
#| fig-height: 8
#| warning: false

#| label: fig-corr-indiv
#| layout-ncol: 2
#| fig-height: 8
#| warning: false

#| label: fig-corr-all
#| fig-height: 10
#| fig-cap: "Correlation"
#| warning: false

#indiviual plots
g1 <- boxplot_genotype_all_in_one(selected_data, "FLW_DAYS")
g2 <- boxplot_genotype_all_in_one(selected_data, "GY")
g3 <- boxplot_genotype_all_in_one(selected_data, "PL_EMERG_COUNT")
g4 <- boxplot_genotype_all_in_one(selected_data, "Pod_Shattering")

g1[[1]]/ g1[[2]]
```

In contrast, the genotypes did not perform well for this trait in environment L0118. The overall shift toward lower yield values in this specific location suggests a restrictive environment, possibly due to soil limitations or water stress during critical stages of crop development. This marked difference between L0134 and L0118 highlights a strong Environmental Effect, which is a fundamental component in evaluating the adaptability and stability of these genotypes across the Zambian landscape.

```{r}
g2[[1]]/ g2[[2]]
```

Regarding the flowering cycle (FLW_DAYS), it is observed that environment L0133 promotes a longer vegetative period, leading to later flowering for the vast majority of genotypes. This contrasts sharply with the earliness observed in environment L0134. As for plant emergence (PL_EMERG_COUNT), a clear trend of superiority is noted in environment L0133, which exhibited higher mean stands for most genotypes compared to the other locations; the remaining environments showed no significant differentiation among themselves. Finally, regarding plant height (PH_R8), environment L0118 stood out negatively, showing a consistent trend of lower mean values across nearly all genotypes when compared to the other environments.

```{r}
g3[[1]]/ g3[[2]]
g4[[1]]/ g4[[2]]
```














# Functions

```{r}
#| warning: false

# Function 1 - Single Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE),)
      cv = (sd/mean)*100 |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  # Correlation between traits
  cor_mat <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")
  
  p1 <-ggcorrplot(cor_mat, 
           hc.order = TRUE,           
           type = "upper",               
           lab = TRUE,      
           lab_size = 3,           
           colors = c("red", "white", "blue"),      
           title = title)   
  
  # Traits Boxplots
  p2 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p4 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = p1,boxplot = p2, density = p3, histogram = p4))
}


# Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (scatter = p1)
}

# Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype <- function(df, title, trait, trait_lab){
    
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(trait),
      names_to = "Variable",
      values_to = "value")
  stats_df <- df_long |>
    group_by(gen) |>
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")
  df_plot <- df_long |> 
    left_join(stats_df, by = "gen")
  
  p1 <- ggplot(df_plot, aes(x = as.factor(gen), y = value, fill = mean)) +
        geom_boxplot(alpha = 0.7) +
        geom_jitter(width = .2, size = 2, alpha = .75, color = "black") +
        scale_fill_viridis_c() +
        labs(title = title,
          x = "Genotypes",
          y = trait_lab,
          fill = "Mean") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18)) +
        theme(axis.text.x = element_text(size = 10))
  
  return (boxplot_genotypes = p1)
}

# Function 4 - Single Trials Analysis Together
single_trials_together <- function(df, title, traits) {
  
  # Dataframe long
  df_long <- df |> 
    mutate(Trial = as.factor(trial)) |> 
    pivot_longer(
      cols = all_of(traits),
      names_to = "Trait",
      values_to = "Value") |> 
    filter(!is.na(Value))
  
  # Boxplots 
  p1 <- ggplot(df_long, aes(x = Trial, y = Value, fill = Trial)) +
          geom_boxplot(outlier.colour = "black", outlier.size = 0.5) +
          facet_wrap(~Trait, scales = "free") +
          theme_minimal() +
          labs(title = title) + 
          scale_x_discrete(labels = c("1" = "L0118", "2" = "L0133", "3" = "L074", "4" = "L0134")) +
          scale_fill_discrete(labels = c("1" = "L0118", "2" = "L0133", "3" = "L074", "4" = "L0134"))

  # Density Plot
  p2 <- ggplot(df_long, aes(x = Value, fill = Trial, group = Trial)) +
          geom_density(alpha = 0.4) +
          facet_wrap(~Trait, scales = "free") +
          theme_light() +
          labs(title = title) +
          scale_x_discrete(labels = c("1" = "L0118", "2" = "L0133", "3" = "L074", "4" = "L0134")) +
          scale_fill_discrete(labels = c("1" = "L0118", "2" = "L0133", "3" = "L074", "4" = "L0134"))


return (list(boxplots = p1, density = p2))
}

# Function 5 - All Trials Boxplot By Genotype 
boxplot_genotype_all_in_one <- function(df, trait) {
  
  # Boxplot by Genotype and Trial
  p1 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait), fill = trial)) +
        geom_boxplot(alpha = 0.7, outlier.shape = NA, na.rm = TRUE) + 
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          fill = "Trial"
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
        
  # Boxplot by Genotype
  p2 <- ggplot(df, aes(x = reorder(gen, !!sym(trait)), y = !!sym(trait))) +
        geom_boxplot(fill = "grey80", color = "black", outlier.shape = NA,na.rm = TRUE) +
        geom_jitter(aes(color = trial),width = 0.2,size = 1.2,alpha = 0.8, na.rm = TRUE) +
        labs(
          title = paste("Performance Across Trials:", trait),
          x = "Genotypes",
          y = trait,
          color = "Trial") +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
          legend.position = "top",
          panel.grid.minor = element_blank())
   
  return (list(boxplot_gt_trial = p1, boxplot_gt = p2))
}


# Function 6 - Across Trials Correlation
across_trials_correlation <- function(df, title, traits) {

  # Correlation Plot with all trials
  cor_total <- df |>
    select(all_of(traits)) |>
    cor(use = "complete.obs")

  p1 <-ggcorrplot(cor_total, 
         hc.order = TRUE,           
         type = "upper",               
         lab = TRUE,      
         lab_size = 3,           
         colors = c("red", "white", "blue"),      
         title = title)
  
  return(all_cor = p1)
}


```

# Trials Overview

Explicar genotipos do experimento, e cada trait analisado.

# Trial 1 - Local "L0118" / Year "2023"

## A FAZER

1\) Listar os genotipos do experimento, e explicar cada trait analisado.

2\) Fazer um exemplo manual para verificar as medidas de posição, dispersão, correlação e funcções

3\) Definir critério de avaliação para os outliers (Trial 1 tem 1 outlier na altura)

4\) Rodar as funções para cada Trial e depois entre Trials

5\) Organizar documentos e fazer as interpretações

## Uso das Funções

```{r}
#| warning: false

## Function 1 - Single Trial Summary
all_trials_summary <- list()
trials <- list(trial_1, trial_2, trial_3, trial_4)

for(i in seq_along(trials)){
  title <- paste("Trial", i)
  all_trials_summary[[i]] <- trial_summary(trials[[i]],title, traits, 30)
}

# Example - Trial 1
all_trials_summary[[1]]$stats
all_trials_summary[[1]]$correlation
all_trials_summary[[1]]$density
all_trials_summary[[1]]$boxplot
all_trials_summary[[1]]$histogram

#######################################################################

## Function 2 - Single Trial Pairwise Scatter plot
pairwise_scatter(trial_2,"Título", "GY", "PH_R8", "GY (kg/ha)", "PH (cm)")

######################################################################

## Function 3 - Single Trial Boxplot by Genotype
boxplot_by_genotype(trial_1,"Título", "GY", "GY (kg/ha)")

######################################################################



#######################################################################

# Function 5 - All Trials Boxplot By Genotype
all_genotype_GY <- boxplot_genotype_all_in_one(selected_data,"GY")
all_genotype_GY$boxplot_gt_trial
all_genotype_GY$boxplot_gt

#######################################################################

# Function 6 - Across Trials Correlation
across_trials <- across_trials_correlation(selected_data, "Título", traits)
across_trials
```

## Outras Anotações

```{r}
#| label: fig-genotype-comparison
#| fig-width: 22
#| fig-height: 8
#| warning: false

#| label: fig-corr-indiv
#| layout-ncol: 2
#| fig-height: 8
#| warning: false

#| label: fig-corr-all
#| fig-height: 10
#| fig-cap: "Correlation"
#| warning: false

#indiviual plots
boxplot_genotype_all_in_one(selected_data, "FLW_DAYS")
boxplot_genotype_all_in_one(selected_data, "GY")
boxplot_genotype_all_in_one(selected_data, "PL_EMERG_COUNT")
boxplot_genotype_all_in_one(selected_data, "Pod_Shattering")

#plot PH_R8, filtring outlier (16  Zambia 2023 E0283 G321 L0118 // 308.33)
all_trials_cleanPH_R8 <- selected_data %>%
  filter(PH_R8 > 10 & PH_R8 < 160) 
boxplot_genotype_all_in_one(all_trials_cleanPH_R8 , "PH_R8")

across_trials_correlation(selected_data, " ", traits)
```

```{r}
#| warning: FALSE
#| label: tabela-trait-local
#| tbl-cap: "Medidas de tendência central e dispersão dos experimentos performados na Zâmbia"
envs <- c("L0118", "L0133", "L074", "L099")
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
df_empilhado <- map_df(seq_along(results), function(i) {
  results[[i]]$stats |> 
    mutate(Local = envs[i]) 
})

df_para_tabela <- df_empilhado |>
    pivot_longer(
    cols = c(mean, median, variance, sd, min, max, range), 
    names_to = "Estatistica", 
    values_to = "Valor"
  ) |>
   mutate(Estatistica = case_match(Estatistica,
    "mean" ~ "Média",
    "median" ~ "Mediana",
    "variance" ~ "Variância",
    "sd" ~ "Desvio Padrão",
    "min" ~ "Mínimo",
    "max" ~ "Máximo",
    "range" ~ "Amplitude"
  )) |>
  pivot_wider(names_from = c(Variable, Local), values_from = Valor)

n_traits <- 5
n_locais <- 4

cabecalho_superior <- c(" " = 1)
cabecalho_superior[traits] <- n_locais 

df_para_tabela |>
  kable(
    digits = 2,
    align = "l",
    col.names = c(" ", rep(envs, n_traits)), 
    caption = "Measures of central tendency and dispersion of the five traits and four locations"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), 
    full_width = FALSE
  ) |>
  add_header_above(cabecalho_superior) |>
  column_spec(1, bold = TRUE, border_right = TRUE)
```
