---
title: "Multi Trial EDA"
author: 
  - name: João Vitor Sanches Calvo
    orcid: 0009-0002-5827-6109
    email: joaovitor.calvo@usp.br
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
---

# Exercise

Using the dataset you will use in the project, perform a comprehensive exploratory data analysis. Choose four trials and perform EDA for five traits (if available). Check the variation, detect outliers and possible errors, build histograms, density plots and boxplots; compute measures of central tendency and dispersion, and investigate the relationship between traits. Finally, perform an across trials EDA: check if there is any major changes within traits across trials.

# Database selection - Zambia

```{r}
#| warning: false

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> filter(loc %in% c("L0118", "L0133", "L074", "L099") & YEAR == "2023")

# Trial 1 - Local "L0118" / Year "2023" 
trial_1 <- selected_data |> filter(loc == "L0118", YEAR == "2023")
genotypes_1 <- trial_1$gen

# Trial 2 - Local "L0133" / Year "2023" 
trial_2 <- selected_data |> filter(loc == "L0133", YEAR == "2023")
genotypes_2 <- trial_2$gen

# Trial 3 - Local "L074" / Year "2023" 
trial_3 <- selected_data |> filter(loc == "L074", YEAR == "2023")
genotypes_3 <- trial_3$gen

# Trial 4 - Local "L099" / Year "2023" 
trial_4 <- selected_data |> filter(loc == "L099", YEAR == "2023")
genotypes_4 <- trial_4$gen
```

# Trials Overview

Explicar genotipos do experimento, e cada trait analisado.

# Functions

```{r}
# Function - Trial Summary
trial_summary <- function(df, title, traits, bins){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(traits),
      names_to = "Variable",
      values_to = "value")
  
  # Measures of Position and Dispersion - Quantitative Variable
  stats_df <- df_long |>
    group_by(Variable) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      variance = var(value, na.rm = TRUE),
      sd = sd(value, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      range = max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) |> 
    mutate(across(where(is.numeric), ~round(.x, 3)))
  
  
  # Traits Boxplots
  p1 <- ggplot(df_long, aes(x = value, y = Variable, fill = Variable)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~Variable, scales = "free") +
        geom_point(data = stats_df, aes(x = mean, y = Variable),
          shape = 17, color = "#10342d", size = 2) +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Values",
          y = "",
          fill = "Trait") +
        theme_minimal() +
        theme(axis.text.y = element_blank(),text = element_text(size = 18))
  
  # Traits Density Plot
  p2 <- ggplot(df_long, aes(x = value, fill = Variable)) +
        geom_density(alpha = 0.6, show.legend = FALSE) +
        geom_vline(data = stats_df, aes(xintercept = mean),
          linetype = "dashed", linewidth = 0.8, show.legend = FALSE) +
        geom_text(data = stats_df,aes(x = mean, y = Inf, label = round(mean, 2)),
          vjust = 1.5,hjust = -0.4 ,size = 3.5, color = "black") +
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(
          title = paste(title),
          x = "Value",
          y = "Density") +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  # Traits Histogram
  p3 <- ggplot(df_long, aes(x = value, fill = Variable)) + 
        geom_histogram(bins = 30, color = "white", 
          alpha = 0.7, show.legend = FALSE) + 
        facet_wrap(~Variable, scales = "free") +
        scale_fill_brewer(palette = "Set2") +
        labs(title = title,
          x = "Value", 
          y = "Count") + 
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return(list(stats = stats_df,correlation = cor_mat,boxplot = p1, density = p2, histogram = p3))
}



# Correlation Analysis
# Standardized Plot Function
plot_std_corr <- function(df, title_name, type = "upper") {
  
  # 1. Select traits
  df_sub <- df %>% 
    select(all_of(traits)) %>% 
    mutate(across(everything(), as.numeric))
  
  # 2. Filter out columns with SD = 0
  cols_to_keep <- sapply(df_sub, function(x) sd(x, na.rm = TRUE) > 0)
  df_filtered <- df_sub[, cols_to_keep, drop = FALSE]
  
  # 3. Correlation Matrix
  m <- cor(df_filtered, use = "complete.obs")
  
  # 4. Plotting with standardized visual style
  corrplot(m, 
           method = "color",           
           type = type,               
           addCoef.col = "black",      
           tl.col = "black",           
           diag = TRUE,                
           number.cex = 0.7,           
           tl.srt = 45,                
           title = paste("\nLocation:", title_name),
           mar = c(0,0,2,0))          
}

#| label: fig-corr-indiv
#| layout-ncol: 2
#| fig-height: 8
#| warning: false

# individual plots 
plot_std_corr(trial_1, "L0118")
plot_std_corr(trial_2, "L0133")
plot_std_corr(trial_3, "L074")
plot_std_corr(trial_4, "L0134")


#| label: fig-corr-all
#| fig-height: 10
#| fig-cap: "Correlation"
#| warning: false

#  all trial data into a single dataframe for across-trial analysis
all_trials <- bind_rows(trial_1, trial_2, trial_3, trial_4) %>%
  mutate(across(all_of(traits), as.numeric))

# correlation matrix
cor_total <- cor(all_trials %>% select(all_of(traits)), use = "complete.obs")
corrplot(cor_total, 
         method = "color", 
         type = "full", 
         addCoef.col = "black", 
         tl.col = "black", 
         diag = TRUE,
         title = "\nCORRELATION_ALL TRIALS",
         mar = c(0,0,2,0))





# Function - Pairwise Scatter plot
pairwise_scatter <- function(df, title, trait1, trait2, t1_lab, t2_lab){
  
  # Dataframe long
  df_long <- df |>
    pivot_longer(cols = all_of(c(trait1, trait2)),
      names_to = "variable",
      values_to = "value") |>
    mutate(Variable = variable)
  
  # Trait1 and Trait2 - Scatter plot
  p1 <- ggplot(df, aes(x = .data[[trait1]], y = .data[[trait2]])) +
        geom_point(color = "olivedrab4", alpha = 0.7) +
        labs(
          title = paste(title),
          x = t1_lab,
          y = t2_lab,
          caption = paste("Correlation: ",
          round(cor(df[[trait1]], df[[trait2]], use = "complete.obs"), 2))) +
        theme_minimal() +
        theme(text = element_text(size = 18))
  
  return (p1)
}
```

# Trial 1 - Local "L0118" / Year "2023"

```{r}

# Trial Summary
T1_summary <- trial_summary(trial_1, "Trial 1", traits, 30)
T1_summary$stats
T1_summary$correlation
T1_summary$boxplot
T1_summary$density
T1_summary$histogram
```

## A FAZER

1\) Boxplot do trait destrinchando os genótipos

2\) Calcular correlacao (matriz) e comparar com valor do scatter plot

3\) Verificar as medidas de posicao e dispersao

4\) Explicar genotipos do experimento, e cada trait analisado.

5\) Critério para outliers (Trial 1 tem 1 outlier na altura)

6\) Fazer funcao para comparar caracteristicas de trials de diferentes anos

7\) Interpretações
