---
title: "Preview Final Project"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Soybean (*Glycine max* (L.) Merr.) is one of the most important crops
worldwide, playing a strategic role in the production of plant protein,
edible oil, and animal feed. In recent years, its expansion across the
African continent has been driven by international breeding and genetic
adaptation programs aimed at enhancing food security, increasing
farmers' income, and promoting the sustainable development of local
agricultural systems. Within this context, the Pan-African Soybean
Variety Trials (PAT) program was established to evaluate soybean
genotypes across multiple African environments, enabling the
identification of superior germplasm for regional recommendation.

To support these recommendations, multi-environment trials (METs) are
essential, as they enable estimation and understanding of
genotype-by-environment ($G \times E$) interactions. Although
challenging because it alters the ranking and performance of genotypes
across locations, this interaction offers an opportunity to identify
materials with specific adaptations or broad stability, thereby
maximizing the exploitation of their yield potential.

In the present study, phenotypic data from this network were utilized,
specifically from trials conducted in Zambia between 2019 and 2023. The
dataset encompassed 33 environments, resulting from the combination of
16 locations ( @fig-maps @tbl-env-year) and different crop years. The
experiments were laid out in a randomized complete block design (RCBD)
with three replications per environment. In each trial, 30 to 40
genotypes were evaluated, totaling 107 genetic materials tested across
the network. Although agronomic traits such as plant height at maturity
(PH_R8), days to flowering (FLW_DAYS), and emerged plant count
(PLANT_EMERG_COUNT) were explored in preliminary analyses, grain yield
(GY) was selected as the primary response variable for multi-environment
modeling. To this end, the data were analyzed using two complementary
analytical approaches: classic methods, including the Additive Main
Effects and Multiplicative Interaction (AMMI) model and the Eberhart &
Russell (1966) linear regression, and the predictive approach using
linear mixed models via REML/BLUP.

Given the above, the objective of this study was to evaluate the
stability of 107 soybean genotypes evaluated in the Zambian trial
network, aiming to select the most predictable and high-yielding
materials for the region.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml","gt", "tidyverse", "ggrepel", "maps")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: fig-maps
#| tbl-cap: "Geographic location of trials within years"

coords <- data %>%
  transmute(loc, env, YEAR,
            LAT_raw = as.numeric(LAT),
            LON_raw = as.numeric(LON),
            LAT = ifelse(abs(LAT_raw) > 90, LAT_raw / 1000, LAT_raw),
            LON = ifelse(abs(LON_raw) > 180, LON_raw / 1000, LON_raw)) %>%
  filter(!is.na(LAT), !is.na(LON))

loc_year <- coords %>%
  group_by(loc, YEAR) %>%
  summarise(n_trials = n_distinct(env),
            LAT = mean(LAT), LON = mean(LON),
            .groups = "drop")

zambia_border <- map_data("world", region = "Zambia")

p_map_year <- ggplot() +
  geom_polygon(data = zambia_border,
               aes(x = long, y = lat, group = group),
               fill = "#f5f5f0", colour = "grey60", linewidth = 0.4) +
  geom_point(data = loc_year,
             aes(x = LON, y = LAT, size = n_trials, fill = loc),
             shape = 21, colour = "white", stroke = 0.6, alpha = 0.9) +
  scale_size_area(name = "Number of Trials", max_size = 8, breaks = c(1, 2, 3)) +
  scale_fill_viridis_d(option = "turbo", name = "Location") + 
  coord_quickmap() +
  facet_wrap(~ YEAR) + 
  labs(x = "Longitude (°)", y = "Latitude (°)",
       title = "Spatio-Temporal Distribution of Trials — Zambia",
       subtitle = "Colors represent distinct physical locations") +
  theme_minimal(base_size = 12) +
  theme(panel.background = element_rect(fill = "#dfe9f0", colour = NA),
        panel.grid.major = element_line(colour = "white", linewidth = 0.3),
        strip.background = element_rect(fill = "grey90", colour = NA),
        strip.text = element_text(face = "bold", size = 12),
        legend.position = "right") 

print(p_map_year)
```

```{r}
#| label: env-year-table
#| tbl-cap: "Environments combination (local x year)"

env_year_table <- data |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: outliers_GY

# Normal Distribution Sd's
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removing outliers
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Converting columns
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

# Single Trial - Statistical Model

Then, each trial was analysed separately, to have a better understanding
of the aspects related to the quality of the data collected in each
trial. This was done by analysing the broad-sense heritability ($H^{2}$)
and the coefficient of variation ($CV%$).

Within each trial (RCBD) there are 3 blocks and approximately 30
genotypes, and for both traits analysed, the statistical model used is
describe as follow:

$$y_{ij}=\mu +b_{j}+g_{i}+e_{ij}$$

Where:

$b_{j}$ is the block fixed effects, as there are only 3 blocks within
each experiment.

$g_{i}$ is the genotype's random effects, where
$g_i \sim N(0, \sigma^2_g)$. This was decided due to the unbalanced
nature of the data, as some trials had missing plots.

$e_{ij}$ is the residual random effect, where
$e_{ij} \sim N(0, \sigma^2_e)$. That is, assuming indepedence of
residuals within trials

As most trials are completely balanced, the broad-sense heritability was
estimated as:

-   individual/plot level:
    $H² = \frac{\sigma²_g}{ \sigma²_g + \sigma²_e}$,
-   genotype level:
    $H² = \frac{\sigma²_g}{ \sigma²_g + \frac{\sigma^2_e}{3}}$
-   Cullis, 2006: \$ H\^2 = 1 - \frac{\bar{SED^2}}{2\sigma²_g}\$

# Trait - Grain Yield

```{r}
#| label: GY_missing_data

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")
```

```{r}
#| label: GY_analysis
#| output: false

# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2_plot = numeric(),h2_gen = numeric(), h2_cullis = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen", sed = TRUE)
  
  pred_BLUP <- blups$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_GY, boundary = T)  
  
  h2_plot <- vpredict(model, H2 ~ V1/(V1+V2)) # nível de parcela
  h2_gen <- varcomp[1,1] / (varcomp[1,1]+(varcomp[2,1]/3)) # ind
  h2_cullis <- 1-(mean(blups$sed[upper.tri(blups$sed)]^2)/(2*varcomp[1,1]))
  
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  reliability <- 1 - (pred_BLUP$std.error^2/varcomp[1,1])
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2_plot = round(h2_plot$Estimate,digits=4),
      h2_gen = round(h2_gen,digits = 4),
      h2_cullis = round(h2_cullis,digits = 4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups$pvals
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

```{r}
#| label: tbl-single_trial
#| output: false

tabela_geral_GY |>
  mutate(
    highlight = env %in% c("E0108", "E0175","E0224", "E0280", "E0281", "E0285")) |> 
  gt(rowname_col = "env") |>
  tab_header(
    title = "Single-trial analysis for grain yield"
  ) |>
  tab_style(
    style = cell_fill(color = "#e8f5e9"),
    locations = cells_body(
      rows = highlight == TRUE
    )
  ) |>
  cols_hide(
    columns = c(significance, highlight)
  ) |>
  cols_label(
    LRTstat = "LRT",
    pvalue = "P-value",
    h2_plot = html("H<sup>2</sup><sub>plot</sub>"),
    h2_gen = html("H<sup>2</sup><sub>genotype</sub>"),
    h2_cullis = html("H<sup>2</sup><sub>Cullis</sub>"),
    cv = "CV (%)"
  ) |>
  fmt_number(
    columns = c(LRTstat, pvalue, h2_plot, h2_gen, h2_cullis, cv),
    decimals = 4
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(100)
  )

```

Based on the LRT and on the heritabilites shown above, the genetic
variance within environments "E0108", "E0175", "E0224", "E0280",
"E0281", "E0285" was not significant, meaning that observations from
this trial have a great portion of noise. Because of this, these trials
were removed from the multi-environment analysis.

```{r}
#| label: env_filter

# Removing envs "E0108", "E0175","E0224", "E0280", "E0281", "E0285"
df_filter <- df |> filter(!(env %in% c("E0108", "E0175","E0224", "E0280", "E0281", "E0285")))

# New table
env_year_table <- df_filter |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )
env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

```{r}
#| label: fig-blup_by_envs_plots_GY

blup_all_GY <- bind_rows(
  lapply(names(singleTrial_blup_GY), function(env){
    df <- singleTrial_blup_GY[[env]]
    df$env <- env
    return(df)}))

ggplot(blup_all_GY,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (GY)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```

The fluctuation in the top-ranking positions highlights the presence of
a strong Genotype by Environment (GxE) interaction (cross-over patterns)
@fig-blup_by_envs_plots_GY. Due to this interaction pattern, we will
proceed to a multi-environment analysis, aiming to asses genotypes with
higher stability for the locals considered.

# Multi-trial - Statistical Model

To analyze genotypes performance in different environments, the
following mixed model will be used. The inicial version of the model is:

$$\mathbf{y}= \mathbf{1}\mu + \mathbf{Z}_1\mathbf{e} + \mathbf{Z}_2\mathbf{b}  + \mathbf{Z}_3\mathbf{g} + \varepsilon $$

Where:

$\mathbf{e}$ is the environment random effects, as the environments
(local x year combinations) can be treated as a random sample of all
possible combinations of these factors

$\mathbf{b}$ is the block random effects, as they are nested in
environments, with $\mathbf{b} \sim N(0,\sigma^2_b\mathbf{B})$

$\mathbf{g}$ is the genotype's random effects, with
$\mathbf{g} \sim N(0,\sigma^2_g\mathbf{G})$

$\mathbf{\varepsilon}$ is the residual random effect, with
$\mathbf{\varepsilon} \sim N(0,\sigma^2_e\mathbf{R})$

Different (co)variances structures will be tested for the random effects
in these models. Each one of the models make different assumptions
regarding the residuals, block or genetic random distribution. First, we
will analyse the residuals covariance matrix, then blocks and finally
the covariance matrix of genetic effects.

-   **Model 1**
    -   G: compound symmetry (genetic variance is homogeneous for
        genotypes in different environments and the covariance between
        the same genotypes performance in any two environments is
        constant)
    -   R: identity (all residuals have the same variance in different
        environments)
    -   B: identity (blocks in different environment have the same
        variance)

```{r}
#| label: model1

# Model 1
mod1 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ idv(units),
  data     = df_filter,
  maxit = 100)
```

-   **Model 2**
    -   G: compound symmetry (all genotypes have the same variances and
        the same covariances in different environments)
    -   R: block-diagonal (each environment has its own residual
        variance)
    -   B: identity (blocks in different environment have the same
        variance)

The assumption regarding R now seems to be more reasonable, as different
environments impacts differently on the residuals.

```{r}
#| label: model2

# Model 2
mod2 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

As in this dataset there is no information about the position of the
plots in the field, we are not capable of doing the classical spatial
corrections, so these will be the only structures considered in the
covariance matrix R.

Now, we will analyse the block covariance matrix:

-   **Model 3**
-   G:compound symmetry (all genotypes have the same variances and the
    same covariances in different environments)
-   R: block-diagonal (each environment has its own residual variance)
-   B: block-diagonal (each environment has its own block variance)

```{r}
#| label: model3

# Model 3
mod3 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + at(env):rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

Now, the genetic part will be analysed.

-   **Model 4**
-   G: block-diagonal (each environment has its own genetic variance,
    but the genetic effect from different environments are not
    correlated)
-   R: block-diagonal (each environment has its own residual variance)
-   B: identity (all blocks have the same variance)

In this model, the genotype effect is nested within the diagonal
structure of environments. That means that each environment has its own
genetic variance, which seems reasonable, but, this structure assumes
that observations from the same genotype in different environments are
not correlated, which is a strong assumption.

```{r}
#| label: model4

# Model 4 
mod4 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + diag(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

-   **Model 5**
-   G: Heterogeneous compound symmetry (each environment has its own
    genetic variance, and there is a uniform correlation that connects
    genotypes in different environments)
-   R: block-diagonal (each environment has its own residual variance)
-   B: identity (all blocks have the same variance)

In this model, the genotype effect is also nested within environments,
which has heterogeneous variances for different environments and uniform
correlation between environments. So, genotypes from the same
environment have the same genetic variance, and the same genotypes in
different environments are connected by a uniform correlation. This also
allow the prediction of genotypes at untested environments.

```{r}
#| label: model5

# Model 5
mod5 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

-   **Model 6 and 7**
-   G: Unstructured with heterogeneuous variances (each environment has
    its own genetic variance, and different correlations connect
    genotypes of different environments)
-   R: block-diagonal (each environment has its own residual variance)
-   B: identity (all blocks have the same variance)

In both models above, the genotype effect is also nested within
environments, which has heterogeneous variances for different
environments and a distinct correlation for each pair of environments.
So, genotypes from the same environment have the same variance, and
observations from the same genotype in different environments are
connected by a correlation that is specific to that pair of
environments. That is different from Model 5, where a single correlation
value was applied to every pair of environment.

Although the assumptions of these models are least strong, the variance
component estimation did not converge, probably because of the high
number of parameters. So we were not able to use **Model 6 and 7**.

```{r}
#| label: model6_7

# Model 6 and 7
mod6 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + us(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)

mod7 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corgh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)
```

After model adjustment, the best model was chosen based on AIC, then,
the predicted values (based on BLUPs) were obtained for
genotype-environment combinations.

After that, genotype stability was analysed through different indeces:

-   Eberhart & Russel (1966):
    $Y_{ij} = \mu_{i} + \beta_{1}I_j + \delta_{ij}$

-   AMMI: \$ Y\_{ij}=μ+g_i+e_j+(ge)\_{ij} \$

-   Safety-First (Eskridge, 1990):
    $SF_i = \hat{g_i} - Z_{1-\alpha} \sqrt{V_i}$

-   Pi Index (Lin & Binns, 1988) \$ P_i =
    \frac{\sum_{j=1}^{p} \left[ \hat{g}_{ij} - \max(\hat{g}_{j}) \right]^2}{2 \times p}\$

# Resultados

```{r}
#| label: aic
#| fig-cap: "AIC values for different models"

SF_i = \hat{g_i} - Z_{1-\alpha} \sqrt{V_i}

aic1 <- summary(mod1)$aic
aic2 <- summary(mod2)$aic
aic3 <- summary(mod3)$aic
aic4 <- summary(mod4)$aic
aic5 <- summary(mod5)$aic 

modelos_aic <- data.frame(
  modelo = factor(
    c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
      "HCS(G)-DIAG(R)"),
    levels = c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
               "HCS(G)-DIAG(R)") 
  ), AIC = c(aic1, aic2, aic3, aic4, aic5))
  
  ggplot(modelos_aic, aes(y = modelo, x = AIC)) +
    geom_point(size = 3, shape = 15, colour = "#10342d") +
    geom_segment(aes(x = min(AIC) - 50, xend = AIC, y = modelo, yend = modelo),
                 linewidth = 1.5, colour = "#10342d") +
    geom_label(aes(label = round(AIC, 1)), size = 4,
               position = position_nudge(y = 0.3, x = -30)) +
    theme_minimal(base_size = 14) +
    labs(y = "Modelo", x = "AIC",
         title = "Comparação de estruturas de covariância G×A",
         subtitle = "Variável resposta: GY | Dados Zâmbia (Soja)")

```

Based on the values above, `Model 5` presented the smaller AIC, being
the model selected for further analysis.

## Modelo 5

```{r}
#| label: blups_per_env
#| fig-cap: "Predicted Values in Different Environments"

  pred <- predict.asreml(object = mod5,classify = "gen:env")$pvals
  
  pred |>
    ggplot(aes(x = reorder(env, -predicted.value),
               y = reorder(gen, predicted.value),
               fill = predicted.value)) +
    geom_tile() +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) + 
    scale_fill_gradient(low = "#10342d", high = "#b2bc63") +
    labs(x = "Environment", y = "Genotype", fill = "GY",
         title = "Predicted Value")

```

```{r}
#| label: fig-trajectory
#| fig-cap: "Environmental Index trajectory of the evaluated locations across years"
#| warning: false
#| message: false

trajectory_index <- df_filter %>%
  group_by(env, loc, YEAR, RAINFED) %>% 
  summarise(GY_mean = mean(GY, na.rm = TRUE), .groups = "drop") %>%
  mutate(index = GY_mean - mean(df_filter$GY, na.rm = TRUE)) %>%
  group_by(loc) %>%
  mutate(total_years = n_distinct(YEAR)) %>%
  ungroup()

ggplot(trajectory_index, aes(x = factor(YEAR), y = index, group = loc)) +
  geom_line(data = subset(trajectory_index, total_years > 1), 
            aes(colour = loc), linewidth = 1, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.8, colour = "grey40") +
  
  geom_point(data = subset(trajectory_index, total_years > 1),
             aes(shape = as.factor(RAINFED)), colour = "black", size = 4, stroke = 1.2) +
  geom_point(data = subset(trajectory_index, total_years == 1),
             aes(shape = as.factor(RAINFED)), colour = "firebrick", size = 4, stroke = 1.2) +
  
  ggrepel::geom_text_repel(aes(label = env), colour = "grey50", size = 3, 
                           nudge_x = 0.2,       
                           direction = "y",    
                           segment.color = "grey80") + 
  
  scale_colour_viridis_d(option = "turbo", name = "Location Code") +
  scale_shape_manual(values = c(16, 15, 8), name = "Water Regime") + 
  
  labs(x = "Year", 
       y = "Environmental Index",
       title = "Location Trajectory and Water Regime Impact",
       subtitle = "") +
  
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())

```

The environmental index exhibited a substantial amplitude, ranging from
approximately -2,000 kg/ha to +2,300 kg/ha, evidencing a strong
environmental effect on grain yield across the testing network. The most
severe stress conditions were recorded in E0107 (2019, Rainfed), which
presented the lowest index, whereas E0279 (2023) stood out completely
from the rest as the most favorable environment of the entire network,
representing a condition of high yield potential. It is also noteworthy
that the majority of the evaluated environments were concentrated in the
2023 season.

Analyzing the impact of water regimes over the years reveals a distinct
temporal pattern. In 2019 and 2020, all trials within the network were
conducted under rainfed conditions. Remarkably, even without
supplementary water, several of these early rainfed environments (such
as E0106, E0176, E0171, E0172, and E0169) managed to achieve positive
environmental indices. However, a clear shift occurred from 2021
onwards: no rainfed trial managed to reach a positive index. During this
later period, positive yield performances became exclusively associated
with irrigated or supplementary water regimes.

From a location-specific perspective, L074 is uniquely positioned as the
only location comprising trials across all evaluated years, having
started with rainfed conditions in its first two seasons. Location L118
stands out as highly interesting, almost consistently maintaining
excellent positive indices, with only a minor decrease observed in 2021.
Conversely, locations L078 and L075 were chronically unfavorable,
consistently recording negative indices across all their trials.
Finally, as shown in @fig-trajectory, the frequent crossovers between
the location lines over the years indicate that environmental rankings
shift dynamically between seasons, which strongly anticipates the
presence of a relevant Genotype × Environment (G×E) interaction in the
network.

```{r}
#| label: fig-eberhart-russell
#| fig-cap: "Eberhart and Russell (1966) regression analysis"

ERm <- metan::ge_reg(
  .data   = droplevels(df_filter), 
  env     = "env",
  gen     = "gen",
  rep     = "rep",
  resp    = "GY",
  verbose = FALSE)

top_gen <- ERm$GY$regression |>
  arrange(desc(b0)) |>
  slice(1:30) |> 
  pull(GEN) |>
  as.character()

top_gen <- unique(c(top_gen, "G318"))

ggplot(data = subset(ERm$GY$data, GEN %in% top_gen),
       aes(x = IndAmb, y = Y)) +
  facet_wrap(. ~ GEN) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#10342d") +
  geom_point(aes(colour = ENV), size = 2) +
  theme_bw(base_size = 14) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Environmental Index", 
       y = "Grain Yield (GY)", 
       colour = "Environment") +
  scale_color_viridis_d(option = "turbo") +
  stat_regline_equation()
```

As illustrated in @fig-eberhart-russell, the Eberhart and Russell
regression model was applied to assess the adaptability and phenotypic
stability of the top 30 highest-yielding genotypes. While this
methodology provides valuable insights into how genotypes respond to
environmental variations, a critical examination of the data reveals a
significant methodological limitation regarding the number of testing
environments.

A importat statistical artifact is present among several genotypes that
were selected in this top 30 productivity ranking, specifically G114,
G115, G191, G300, G317, G322, and G324. These materials were evaluated
in only two environments. Mathematically, a linear regression line will
always perfectly connect two points. Consequently, there is no
dispersion around the line, meaning the deviation from regression
($S^2_{di}$) equals zero by definition. Although this perfect fit might
falsely suggest flawless phenotypic stability, it is entirely a
mathematical artifact resulting from a lack of degrees of freedom,
rather than true biological resilience. Furthermore, this artifact
heavily distorts the estimation of their regression parameters,
generating unrealistic mean yields ($\beta_0$) and extreme slopes, such
as the artificially inflated responsiveness of G191 ($\beta_1 = 10$).

In contrast, evaluating genotypes across a robust multi-environment
network captures the genuine biological complexity of G×E interactions.
Genotype G244, which is one of the most extensively tested materials in
the network, clearly demonstrates this reality. Despite maintaining a
competitive mean yield ($\beta_0 = 2700$ kg/ha), its data points are
widely scattered around the regression line. This substantial dispersion
indicates high unpredictability and a significant interaction with
specific environmental factors that are not linearly explained by the
environmental index.

Genotypes such as G335, G365, and G310 present a large number of data
points that closely follow their respective regression lines, proving
that high predictability is biologically attainable when adequately
tested.

```{r}
#| label: fig-ammi-biplots
#| fig-cap: "AMMI1 (Mean Yield vs. PC1) and AMMI2 (PC1 vs. PC2) biplots for grain yield across evaluated environments."
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6

df_ammi <- droplevels(df_filter)

media_geral <- mean(df_ammi$GY, na.rm = TRUE)

medias_env <- df_ammi |>
  group_by(env) |>
  summarise(efeito_env = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

medias_gen <- df_ammi |>
  group_by(gen) |>
  summarise(efeito_gen = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

dat_ammi_completo <- df_ammi |>
  tidyr::complete(env, gen, rep) |>
  left_join(medias_env, by = "env") |>
  left_join(medias_gen, by = "gen") |>
  mutate(
    efeito_gen = ifelse(is.na(efeito_gen), 0, efeito_gen),
    efeito_env = ifelse(is.na(efeito_env), 0, efeito_env),
    
    GY = ifelse(is.na(GY), media_geral + efeito_env + efeito_gen, GY),
    
    GY = ifelse(is.na(GY), media_geral, GY)
  ) |>
  select(env, gen, rep, GY)

ammi <- metan::performs_ammi(
  .data = dat_ammi_completo,
  env   = "env",
  gen   = "gen",
  rep   = "rep",
  resp  = "GY",
  verbose = FALSE
)

p_ammi1 <- plot_scores(ammi, type = 1, title = FALSE) +
  labs(title = "A) AMMI 1: Productivity vs. Stability",
       x = "Mean Grain Yield", 
       y = "PC 1") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

p_ammi2 <- plot_scores(ammi, type = 2, title = FALSE) +
  labs(title = "B) AMMI 2: GxE Interaction Map",
       x = "PC 1", 
       y = "PC 2") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

p_ammi1 + p_ammi2
```
A strict mathematical of AMMI algorithm is the requirement of a completely balanced two-way classification matrix. Since multi-environment agricultural trials are inherently unbalanced due to localized entries and missing plots, data imputation was performed prior to the analysis. Missing cells were predicted using an additive main effects framework ($Y_{ij} = \mu + G_i + E_j$). This strategy  preserved the network's connectivity and allowed the AMMI model to converge without discarding partially evaluated genotypes or environments. 

The multiplicative interaction analysis revealed that the first principal component (PC1) explained 28.1% of the G×E interaction variance, while the second component (PC2) captured 15.7%, accounting for a cumulative total of 43.8% of the interaction sum of squares.In the AMMI1 biplot (@fig-ammi-biplots A), environment E0279 stood out as the most productive environment in the entire network, exhibiting a high mean yield ceiling combined with moderate interaction scores. Conversely, environments E0227 and E0229 displayed the most extreme scores along the PC1 axis, indicating that these environments exert the strongest forces in altering genotypic rankings relative to the rest of the network. Regarding genotypic performance, G191, G115, and G069 combined high mean grain yields with PC1 scores close to zero, which traditionally suggests a profile of broad adaptability. 

However, this pattern must be interpreted with caution. As previously identified in the regression analysis, genotypes such as G191 and G115 were evaluated in only two environments. Because their missing data points were filled using the additive main effects model (which inherently assumes an interaction deviation of zero) their coordinates were artificially dragged toward the biplot origin ($PC1 \approx 0$). Therefore, their apparent stability represents a mathematical artifact of the imputation process on sparse data rather than true biological buffering. In contrast, the group formed by G307, G310, G342, and G335 represents extensively tested materials that consistently achieved grain yields above 2,500 kg/ha with low interaction scores, consolidating them as truly robust candidates for stable performance across the network. 

The AMMI2 biplot (@fig-ammi-biplots B) provided a multidimensional mapping of the interaction, revealing three distinct mega-environments based on the spatial clustering of environmental vectors:(i) High-Potential Managed Network: Environments E0279, E0278, and E0284 grouped tightly in the top-right quadrant, representing favorable conditions under irrigation or supplementary water regimes. Genotypes G137, G057, and G143 displayed strong specific adaptability to this cluster, marking them as ideal candidates for high-investment systems.(ii) High-Yielding Unique Interaction: Environment E0229 isolated itself on the left side of the axis, showcasing a unique interaction profile despite its high mean productivity.(iii) Unfavorable Network: Environments E0227, E0252, and E0253 clustered in the lower quadrants, forming a group of lower-yielding environments with distinct interaction patterns. Genotypes located near the tips of these vectors (such as G245 and G194 towards E0253) express specific affinity for these narrower conditions.

Finally, the central core of the AMMI2 biplot near the origin contains a dense cluster of genotypes, prominently including G307, G310, G342, and G263. Their proximity to the intersection of both interaction axes confirms their stable performance across both multiplicative dimensions, reinforcing their high predictability and broad agronomic buffering across the multi-environment network.

```{r}
media_geral <- mean(df_filter$GY, na.rm = TRUE)

medias_env <- df_filter |>
  group_by(env) |>
  summarise(efeito_env = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

medias_gen <- df_filter |>
  group_by(gen) |>
  summarise(efeito_gen = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

dat_ammi_completo <- df_filter |>
  tidyr::complete(env, gen, rep) |>
  left_join(medias_env, by = "env") |>
  left_join(medias_gen, by = "gen") |>
  mutate(
    # Se o GY for NA, preenche com o modelo aditivo; caso contrário, mantém o GY real
    GY = ifelse(is.na(GY), media_geral + efeito_env + efeito_gen, GY)
  ) |>
  select(env, gen, rep, GY) # Manter apenas as colunas necessárias

ammi <- metan::performs_ammi(
  .data = dat_ammi_completo,
  env   = "env",
  gen   = "gen",
  rep   = "rep",
  resp  = "GY"
)

ggarrange(plot_scores(ammi, type = 1), plot_scores(ammi, type = 2))

plot(ammi, type = "AMMI1")
plot(ammi, type = "AMMI2")

```

```{r}

index <- data.frame(
  index = tapply(df_filter$GY, df_filter$env, mean) - mean(df_filter$GY)
) |> rownames_to_column("env")

dat_grafico_amb <- merge(df_filter, index, by = "env")

ggplot() +
  geom_point(data = subset(index, index < 0),
             aes(x = env, y = index, colour = "Desfavorável", fill = "Desfavorável"),
             size = 2.5, shape = 25) +
  geom_point(data = subset(index, index > 0),
             aes(x = env, y = index, colour = "Favorável", fill = "Favorável"),
             size = 2.5, shape = 24) +
  geom_hline(aes(yintercept = 0), linetype = "dashed", linewidth = 1.2) +
  labs(y = "Índice Ambiental", x = "Ambiente", colour = "", fill = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_colour_manual(values = c("firebrick","forestgreen")) +
  scale_fill_manual(values = c("firebrick","forestgreen"))

  
  media_geral <- mean(df_filter$GY, na.rm = TRUE)

medias_env <- df_filter |>
  group_by(env) |>
  summarise(efeito_env = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

medias_gen <- df_filter |>
  group_by(gen) |>
  summarise(efeito_gen = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

dat_ammi_completo <- df_filter |>
  tidyr::complete(env, gen, rep) |>
  left_join(medias_env, by = "env") |>
  left_join(medias_gen, by = "gen") |>
  mutate(
    # Se o GY for NA, preenche com o modelo aditivo; caso contrário, mantém o GY real
    GY = ifelse(is.na(GY), media_geral + efeito_env + efeito_gen, GY)
  ) |>
  select(env, gen, rep, GY) # Manter apenas as colunas necessárias

ammi <- metan::performs_ammi(
  .data = dat_ammi_completo,
  env   = "env",
  gen   = "gen",
  rep   = "rep",
  resp  = "GY"
)

ggarrange(plot_scores(ammi, type = 1), plot_scores(ammi, type = 2))

plot(ammi, type = "AMMI1")
plot(ammi, type = "AMMI2")
  
  


  ###########################
  # Genotypes
  ##########################
  g342 <- df_filter |> filter(gen == "G342") |> select (env) |> unique()
  g307 <- df_filter |> filter(gen == "G307") |> select (env) |> unique()
  g310 <- df_filter |> filter(gen == "G310") |> select (env) |> unique()
  g108 <- df_filter |> filter(gen == "G138") |> select (env) |> unique()
  
  
  
  # Melhores sequeiro
  top5 <- pred |>
    filter(env %in% c("E0103","E0104","E0105","E0106","E0107","E0108",
                      "E0169","E0170","E0171","E0172","E0173","E0174",
                      "E0175","E0176")) |>
    group_by(env) |>
    arrange(desc(predicted.value), .by_group = TRUE) |>
    slice_head(n = 5)
  
  # G342 - 13/14
  # G307 - 7/14
  # G310 - 4/14
  
  #############################
  # Análise modelo 5
  ############################
  
  mod5 <- asreml(
    fixed    = GY ~ 1,
    random   = ~ env + env:rep + corh(env):gen,
    residual = ~ dsum(~idv(units) | env),
    data     = df_filter,
    maxit = 100)
  aic5 <- summary(mod5)$aic
  
  plot(mod5)
  
  pred <- predict.asreml(object = mod5,classify = "gen:env",sed=TRUE)
  varcomp <- summary(mod5)$varcomp
  
  h2_mod5 <- 1-(mean(pred$sed[upper.tri(pred$sed)]^2)/(2*mean(varcomp[4:36,1])))
  
```

```{r}
 
  # ------------------------------------------------------------------------------
  # 10. RELIABILITY DE CADA GENÓTIPO EM CADA ENV
  # ------------------------------------------------------------------------------
  
  # preds <- predict(mod5, classify = "env:gen", sed = TRUE, pworkspace = "8000mb")
  
  tabela_blups <- preds$pvals
  
  # Calcular a PEV (Prediction Error Variance)
  tabela_blups$PEV <- tabela_blups$std.error^2
  
  print(summary(mod5)$varcomp)
  varcomp <- summary(mod5)$varcomp
  
  sigma2_G_valores <- c(
    "E0103" = varcomp[4,1],  
    "E0104" = varcomp[5,1],  
    "E0105" = varcomp[6,1],   
    "E0106" = varcomp[7,1],
    "E0107" = varcomp[8,1],
    "E0169" = varcomp[9,1],
    "E0170" = varcomp[10,1],
    "E0171" = varcomp[11,1],
    "E0172" = varcomp[12,1],
    "E0173" = varcomp[13,1],
    "E0174" = varcomp[14,1],
    "E0176" = varcomp[15,1],
    "E0224" = varcomp[16,1],
    "E0225" = varcomp[17,1],
    "E0226" = varcomp[18,1],
    "E0227" = varcomp[19,1],
    "E0228" = varcomp[20,1],
    "E0229" = varcomp[21,1],
    "E0230" = varcomp[22,1],
    "E0252" = varcomp[23,1],
    "E0253" = varcomp[24,1],
    "E0254" = varcomp[25,1],
    "E0255" = varcomp[26,1],
    "E0278" = varcomp[27,1],
    "E0279" = varcomp[28,1],
    "E0280" = varcomp[29,1],
    "E0281" = varcomp[30,1],
    "E0282" = varcomp[31,1],
    "E0283" = varcomp[32,1],
    "E0284" = varcomp[33,1]
  )
  
  tabela_blups$sigma2_G <- sigma2_G_valores[tabela_blups$env]
  
  # Reliability = 1 - (PEV / Vgen)
  tabela_blups$Reliability <- 1 - (tabela_blups$PEV / tabela_blups$sigma2_G)
  
  # RESULTADO FINAL
  resultado_final <- tabela_blups[, c("env", "gen", "predicted.value", "std.error", "Reliability")]
  head(resultado_final)
  
  mod5_reliability <- readRDS("reability3.rds")
  
  # Removendo valores negativos
  mod5_reliability <- mod5_reliability |> filter(Reliability > 0) 
  
  summary(mod5_reliability)
  
  rel_gen <- mod5_reliability |>
    group_by(gen) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_env = n(),
      .groups = "drop"
    )
  
  
  summary(rel_gen$mean_reliability)
  
  rel_env <- mod5_reliability |>
    group_by(env) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_gen = n(),
      .groups = "drop"
    )
  

  # Genótipos de interesse
  gen_red <- c("G342", "G138")
  
  # Reliability média por genótipo
  rel_gen <- mod5_reliability |>
    group_by(gen) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_env = n(),
      .groups = "drop"
    )
  
  # Valores dos genótipos destacados
  rel_red <- rel_gen |>
    filter(gen %in% gen_red)
  
  ggplot(rel_gen, aes(x = mean_reliability)) +
    geom_histogram(bins = 30, fill = "grey70", color = "black") +
    geom_vline(
      xintercept = mean(rel_gen$mean_reliability),
      linetype = "dashed",
      linewidth = 1
    ) +
    geom_vline(
      data = rel_red,
      aes(xintercept = mean_reliability),
      color = "red",
      linewidth = 1
    )
  
  ggplot(rel_gen, aes(x = mean_reliability)) +
    geom_histogram(bins = 30, fill = "grey70", color = "black") +
    geom_vline(
      xintercept = mean(rel_gen$mean_reliability),
      linetype = "dashed",
      linewidth = 1
    ) +
    geom_vline(
      data = rel_red,
      aes(xintercept = mean_reliability),
      color = "red",
      linewidth = 1
    ) +
    geom_text(
      data = rel_red,
      aes(x = mean_reliability, y = 0, label = gen),
      color = "red",
      vjust = 1.5,      # empurra o texto para baixo do eixo x
      fontface = "bold",
      size = 4
    ) +
    labs(
      title = "Reliability média dos genótipos nos diferentes ambientes",
      x = "Reliability média",
      y = "Frequência"
    ) +
    coord_cartesian(clip = "off") +  # evita cortar o texto que fica abaixo do eixo
    theme(plot.margin = margin(b = 20))  # espaço extra abaixo do gráfico
```



# # Bruna Certo

```{r}
#| label: fig-pi_index
#| fig-cap: "Cálculo do Índice Pi (Top 15 Genótipos) para os diferentes ambientes"
#| warning: false
#| message: false

  ge_table <- pred[, c("gen", "env", "predicted.value")] |>
    pivot_wider(names_from = "env", values_from = "predicted.value") |>
    column_to_rownames("gen")
  
  Pi <- apply(ge_table, 1, function(x) {
    sum((x - apply(ge_table, 2, max))^2) / (2 * ncol(ge_table))})
  
  df_Pi <- data.frame(gen = names(Pi), Pi = Pi) |>
    arrange(Pi) |>
    slice(1:15)
  
  ggplot(data = df_Pi, aes(x = reorder(gen, Pi), y = Pi)) +
    geom_col(colour = "black", aes(fill = Pi)) +
    scale_fill_gradient(high = "#10342d", low = "#b2bc63") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          legend.position = "none") +
    labs(y = expression(P[i]), x = "Genótipo",
         title = "Índice Pi (Lin & Binns)")
```

```{r}
#| label: fig-sf_index
#| fig-cap: "Cálculo do Safety First (Top 15 Genótipos) para os diferentes ambientes"
#| warning: false
#| message: false

  alpha <- 0.10
  SF <- apply(ge_table, 1, function(x) mean(x) - qnorm(1 - alpha) * sd(x))
  
  df_SF <- data.frame(gen = names(SF), SF = SF) |>
    arrange(desc(SF)) |>
    slice(1:15)
  
  ggplot(data = df_SF, aes(x = reorder(gen, -SF), y = SF)) +
    geom_col(colour = "black", aes(fill = SF)) +
    scale_fill_gradient(low = "#10342d", high = "#b2bc63") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          legend.position = "none") +
    labs(y = "Índice Calculado", x = "Genótipo",
         title = "Indíce Safety-First")
```

```{r}
#| label: fig-env_con
#| fig-cap: "Genotypes Presence on Different Environments"

table_connect <- table(as.character(df$env), as.character(df$gen))
table_connect[table_connect > 0] <- 1
table_connect <- as.data.frame(table_connect)

df_water <- df %>% 
  distinct(env, RAINFED) %>% 
  rename(Var1 = env) 

table_connect <- merge(table_connect, df_water, by = "Var1", all.x = TRUE)
table_connect$Status <- ifelse(table_connect$Freq == 0, "Absence", as.character(table_connect$RAINFED))
conect_perc <- tapply(table_connect$Freq, table_connect$Var2, sum) / length(unique(df_filter$env)) * 100

conect_perc <- data.frame(
  Genótipo = names(conect_perc),
  Conectividade = conect_perc
)

conect_perc <- conect_perc[order(conect_perc$Conectividade, decreasing = TRUE), ]

kable(
  conect_perc,
  caption = "Connectivity of genotypes across experiments",
  align = "c"
)

table_connect$Var2 <- factor(as.character(table_connect$Var2), levels = conect_perc$Genótipo)
ordem_env <- names(sort(tapply(table_connect$Freq, table_connect$Var1, sum), decreasing = TRUE))
table_connect$Var1 <- factor(as.character(table_connect$Var1), levels = ordem_env)

genotipos_alvo <- c("G342", "G138") 
niveis_genotipos <- levels(table_connect$Var2)
rotulos_customizados <- ifelse(niveis_genotipos %in% genotipos_alvo, niveis_genotipos, "")
cores_eixo_x <- ifelse(niveis_genotipos %in% genotipos_alvo, "red", "black")
ggplot(table_connect, aes(x = Var2, y = Var1, fill = Status)) +
  geom_tile() +
  scale_fill_manual(
    values = c(
      "Absence"       = "white", 
      "Irrigation"    = "#1b4332",  
      "Supplementary" = "#40916c",  
      "Rainfed"       = "#95d5b2"   
    )
  ) +
  scale_x_discrete(labels = rotulos_customizados) + 
  labs(
    title = "Presence–absence of treatments across experiments by water regime",
    x = "Genotypes",
    y = "Environments",
    fill = "Water Regime"
  ) +
  theme_minimal() + 
  theme(
  
    axis.text.x = element_text(color = cores_eixo_x, angle = 90, vjust = 0.5, hjust = 1, size = 8, face = "bold"),
    axis.ticks.x = element_line(color = ifelse(rotulos_customizados == "", NA, "gray")), 
    panel.grid = element_blank() 
  )
```

The ranking derived from the superiority index ($P_i$) indicates that genotypes with the lowest values exhibit the shortest distance to the maximum performing genotype within each environment, thus identifying materials with superior broad adaptation. Under this criterion, genotypes G342, G307, G310, G138, and G255 stood out with the lowest $P_i$ estimates, demonstrating strong convergence with the results obtained from Kataoka’s safety-first index.


Interestingly, in the previous single-trial analyses, genotypes G092 and G233 were top-ranked across multiple individual environments, yet they were not selected by the global indices. This discrepancy highlights the importance of accounting for genotype-by-environment ($G \times E$) effects through multi-environment trial (MET) modeling. When evaluated simultaneously across the entire network, the shifting in genotype rankings across locations (crossover interaction) penalizes genotypes with high specific adaptation but low yield stability. This justifies our selection strategy, which prioritizes materials capable of mitigating environmental fluctuations through stable and predictable performance.

Furthermore, the classical stability analyses (AMMI and Eberhart & Russell) also identified genotype G342 as superior. This consensus demonstrates that despite the distinct mathematical foundations, assumptions, and statistical properties of each method, highly stable and high-yielding germplasm can be robustly identified when the genetic signal is strong.

Nevertheless, although this methodological convergence supports the selection of these genotypes based only on rankings and index values, a detailed examination of the cultivation history reveals an important environmental restriction. These higher-ranked genotypes were evaluated under contrasting water regimes across the trial network ( @ fig-env_con ). For instance, genotype G342 was tested exclusively under irrigated or supplementary irrigation conditions, with the notable exception of genotype G138, which was evaluated solely under rainfed environments.

In this scenario, selection based purely on stability indices rankings must be conducted with caution. The apparent broad stability of materials such as G342 may, in fact, reflect the effect of management on the grain yield. Therefore, final recommendations must be strictly stratified to prevent genotypes selected under an irrigation bias from being recommended for rainfed systems, where severe water deficits could drastically penalize the expression of their yield potential.
