---
title: "Preview Final Project"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

Using provided datasets, you are expected to analyse the data and
interpret the results by applying the concepts and methodologies covered
throughout the course.

Please develop a report using R Markdown or Quarto. The final document
must include the following sections:

Introduction: A brief overview of the research problem. Materials and
Methods: A detailed description of the statistical approach. Results: A
clear presentation of the findings (with Figures and Tables).
Discussion: A concise interpretation of the results in the context of
the study.

# Introduction

This report provides a data analysis of a multi-environment soybean
breeding in Zambia dataset, focusing on two agronomic traits **Grain
Yield** and **Plant Height**. The experimental design of the 33 trials
in Zambia dataset is Randomized Complete Block Design (RCBD), in which
all trials are located in different environment, being the **environment
a combination of year x local**.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml","gt")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: env-year-table
#| tbl-cap: "Environment distribution"

env_year_table <- data |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: outliers_GY

# Normal Distribution Sd's
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removing outliers
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Converting columns
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

# Single Trial - Statistical Model

Then, each trial was analysed separately, to have a better understanding
of the aspects related to the quality of the data collected in each
trial. This was done by analysing the broad-sense heritability ($H^{2}$)
and the coefficient of variation ($CV%$).

Within each trial (RCBD) there are 3 blocks and approximately 30
genotypes, and for both traits analysed, the statistical model used is
describe as follow:

$$y_{ij}=b_{j}+g_{i}+e_{ij}$$

Where:

$b_{j}$ is the block fixed effects, as there are only 3 blocks within
each experiment.

$g_{i}$ is the genotype's random effects, where
$g_i \sim N(0, \sigma^2_g)$. This was decided due to the unbalanced
nature of the data, as some trials had missing plots.

$e_{ij}$ is the residual random effect, where
$e_{ij} \sim N(0, \sigma^2_e)$. That is, assuming indepedence of
residuals within trials

As most trials are completely balanced, the broad-sense heritability was estimated as: 

- individual/plot level: $H² = \frac{\sigma²_g}{ \sigma²_g + \sigma²_e}$, 
- genotype level: $H² = \frac{\sigma²_g}{ \sigma²_g + \frac{\sigma^2_e}{3}}$ 
- Cullis, 2006:  $ H^2 =  1 - \frac{\bar{SED^2}}{2\sigma²_g}$

# Trait 1 - GY

```{r}
#| label: GY_missing_data

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")
```

```{r}
#| label: GY_analysis
#| output: false

# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2_plot = numeric(),h2_gen = numeric(), h2_cullis = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen", sed = TRUE)
  
  pred_BLUP <- blups$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_GY, boundary = T)  
  
  h2_plot <- vpredict(model, H2 ~ V1/(V1+V2)) # nível de parcela
  h2_gen <- varcomp[1,1] / (varcomp[1,1]+(varcomp[2,1]/3)) # ind
  h2_cullis <- 1-(mean(blups$sed[upper.tri(blups$sed)]^2)/(2*varcomp[1,1]))
  
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  reliability <- 1 - (pred_BLUP$std.error^2/varcomp[1,1])
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2_plot = round(h2_plot$Estimate,digits=4),
      h2_gen = round(h2_gen,digits = 4),
      h2_cullis = round(h2_cullis,digits = 4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups$pvals
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

```{r}

tabela_geral_GY |>
  mutate(
    highlight = env %in% c("E0108", "E0175", "E0285")) |> 
  gt(rowname_col = "env") |>
  tab_header(
    title = "Single-trial analysis for grain yield"
  ) |>
  tab_style(
    style = cell_fill(color = "#e8f5e9"),
    locations = cells_body(
      rows = highlight == TRUE
    )
  ) |>
  cols_hide(
    columns = c(significance, highlight)
  ) |>
  cols_label(
    LRTstat = "LRT",
    pvalue = "P-value",
    h2_plot = html("H<sup>2</sup><sub>plot</sub>"),
    h2_gen = html("H<sup>2</sup><sub>genotype</sub>"),
    h2_cullis = html("H<sup>2</sup><sub>Cullis</sub>"),
    cv = "CV (%)"
  ) |>
  fmt_number(
    columns = c(LRTstat, pvalue, h2_plot, h2_gen, h2_cullis, cv),
    decimals = 4
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(100)
  )

```

Com base nas herdabilidades observadas na tabela acima, nota-se que as observações dos ambientes "E0108", "E0175" e "E0285" tem uma herdabildiade nula ou muito pequena. Isso mostra que somente uma pequena parte da variação das observações desses experimentos são causadas por fatores genético, e grande parte é causada pelo efeito do ambiente e fatores não controlados. 

Analisar se o modelo fica melhor removendo esses ambientes ou não

```{r}
# Removing envs "E0108", "E0175" e "E0285"
df_filter <- df |> filter(!(env %in% c("E0108", "E0175", "E0285")))

# New table
env_year_table <- df_filter |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

```{r}
#| label: fig-blup_by_envs_plots_GY

blup_all_GY <- bind_rows(
  lapply(names(singleTrial_blup_GY), function(env){
    df <- singleTrial_blup_GY[[env]]
    df$env <- env
    return(df)}))

ggplot(blup_all_GY,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (GY)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```



The fluctuation in the top-ranking positions highlights the presence of a strong Genotype by Environment (GxE) interaction (cross-over patterns) @fig-blup_by_envs_plots_GY

Devido à interação, analisaremos todos os ambientes juntos (estabilidade e GxE)

# Multi trial - Statistical Model