---
title: "Preview Final Project"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

Using provided datasets, you are expected to analyse the data and
interpret the results by applying the concepts and methodologies covered
throughout the course.

Please develop a report using R Markdown or Quarto. The final document
must include the following sections:

Introduction: A brief overview of the research problem. Materials and
Methods: A detailed description of the statistical approach. Results: A
clear presentation of the findings (with Figures and Tables).
Discussion: A concise interpretation of the results in the context of
the study.

# Introduction

This report provides a data analysis of a multi-environment soybean
breeding in Zambia dataset, focusing on two agronomic traits **Grain
Yield** and **Plant Height**. The experimental design of the 33 trials
in Zambia dataset is Randomized Complete Block Design (RCBD), in which
all trials are located in different environment, being the **environment a combination of year x local**.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml","gt", "tidyverse")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: env-year-table
#| tbl-cap: "Environment distribution"

env_year_table <- data |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: outliers_GY

# Normal Distribution Sd's
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removing outliers
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Converting columns
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

# Single Trial - Statistical Model

Then, each trial was analysed separately, to have a better understanding
of the aspects related to the quality of the data collected in each
trial. This was done by analysing the broad-sense heritability ($H^{2}$)
and the coefficient of variation ($CV%$).

Within each trial (RCBD) there are 3 blocks and approximately 30
genotypes, and for both traits analysed, the statistical model used is
describe as follow:

$$y_{ij}=\mu +b_{j}+g_{i}+e_{ij}$$

Where:

$b_{j}$ is the block fixed effects, as there are only 3 blocks within
each experiment.

$g_{i}$ is the genotype's random effects, where
$g_i \sim N(0, \sigma^2_g)$. This was decided due to the unbalanced
nature of the data, as some trials had missing plots.

$e_{ij}$ is the residual random effect, where $e_{ij} \sim N(0, \sigma^2_e)$. That is, assuming indepedence of residuals within trials

As most trials are completely balanced, the broad-sense heritability was estimated as: 

- individual/plot level: $H² = \frac{\sigma²_g}{ \sigma²_g + \sigma²_e}$, 
- genotype level: $H² = \frac{\sigma²_g}{ \sigma²_g + \frac{\sigma^2_e}{3}}$ 
- Cullis, 2006:  $ H^2 =  1 - \frac{\bar{SED^2}}{2\sigma²_g}$

# Trait - Grain Yield

```{r}
#| label: GY_missing_data

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")
```

```{r}
#| label: GY_analysis
#| output: false

# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2_plot = numeric(),h2_gen = numeric(), h2_cullis = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen", sed = TRUE)
  
  pred_BLUP <- blups$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_GY, boundary = T)  
  
  h2_plot <- vpredict(model, H2 ~ V1/(V1+V2)) # nível de parcela
  h2_gen <- varcomp[1,1] / (varcomp[1,1]+(varcomp[2,1]/3)) # ind
  h2_cullis <- 1-(mean(blups$sed[upper.tri(blups$sed)]^2)/(2*varcomp[1,1]))
  
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  reliability <- 1 - (pred_BLUP$std.error^2/varcomp[1,1])
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2_plot = round(h2_plot$Estimate,digits=4),
      h2_gen = round(h2_gen,digits = 4),
      h2_cullis = round(h2_cullis,digits = 4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups$pvals
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

```{r}
#| label: tbl-single_trial
#| output: false

tabela_geral_GY |>
  mutate(
    highlight = env %in% c("E0108", "E0175","E0224", "E0280", "E0281", "E0285")) |> 
  gt(rowname_col = "env") |>
  tab_header(
    title = "Single-trial analysis for grain yield"
  ) |>
  tab_style(
    style = cell_fill(color = "#e8f5e9"),
    locations = cells_body(
      rows = highlight == TRUE
    )
  ) |>
  cols_hide(
    columns = c(significance, highlight)
  ) |>
  cols_label(
    LRTstat = "LRT",
    pvalue = "P-value",
    h2_plot = html("H<sup>2</sup><sub>plot</sub>"),
    h2_gen = html("H<sup>2</sup><sub>genotype</sub>"),
    h2_cullis = html("H<sup>2</sup><sub>Cullis</sub>"),
    cv = "CV (%)"
  ) |>
  fmt_number(
    columns = c(LRTstat, pvalue, h2_plot, h2_gen, h2_cullis, cv),
    decimals = 4
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(100)
  )

```

Based on the LRT and on the heritabilites shown above, the genetic variance within environments "E0108", "E0175", "E0224", "E0280", "E0281", "E0285" was not significant, meaning that observations from this trial have a great portion of noise. Because of this, these trials were removed from the multi-environment analysis.


```{r}
#| label: env_filter

# Removing envs "E0108", "E0175","E0224", "E0280", "E0281", "E0285"
df_filter <- df |> filter(!(env %in% c("E0108", "E0175","E0224", "E0280", "E0281", "E0285")))

# New table
env_year_table <- df_filter |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )
env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

```{r}
#| label: fig-blup_by_envs_plots_GY

blup_all_GY <- bind_rows(
  lapply(names(singleTrial_blup_GY), function(env){
    df <- singleTrial_blup_GY[[env]]
    df$env <- env
    return(df)}))

ggplot(blup_all_GY,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (GY)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```



The fluctuation in the top-ranking positions highlights the presence of a strong Genotype by Environment (GxE) interaction (cross-over patterns) @fig-blup_by_envs_plots_GY. Due to this interaction pattern, we will proceed to a multi-environment analysis, aiming to asses genotypes with higher stability for the locals considered.


# Multi-trial - Statistical Model

To analyze genotypes performance in different environments, the following mixed model will be used. The inicial version of the model is:

$$\mathbf{y}= \mathbf{1}\mu + \mathbf{Z}_1\mathbf{e} + \mathbf{Z}_2\mathbf{b}  + \mathbf{Z}_3\mathbf{g} + \varepsilon $$

Where:

$\mathbf{e}$ is the environment random effects, as the environments (local x year combinations) can be treated as a random sample of all possible combinations of these factors

$\mathbf{b}$ is the block random effects, as they are nested in environments, with $\mathbf{b} \sim N(0,\sigma^2_b\mathbf{B})$

$\mathbf{g}$ is the genotype's random effects, with $\mathbf{g} \sim N(0,\sigma^2_g\mathbf{G})$

$\mathbf{\varepsilon}$ is the residual random effect, with $\mathbf{\varepsilon} \sim N(0,\sigma^2_e\mathbf{R})$

Different (co)variances structures will be tested for the random effects in these models. Each one of the models make different assumptions regarding the residuals, block or genetic random distribution. First, we will analyse the residuals covariance matrix, then blocks and finally the covariance matrix of genetic effects.

- **Model 1**: G - compound symmetry (genetic variance is homogeneous for genotypes in different environments and the covariance between the same genotypes performance in any two environments is constant)
           R - identity (all residuals have the same variance in different environments)
           B - identity (blocks in different environment have the same variance)
            
```{r}
#| label: model1

# Model 1
mod1 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ idv(units),
  data     = df_filter,
  maxit = 100)
```

- **Model 2**: G - compound symmetry (all genotypes have the same variances and the same covariances in different environments)
           R - block-diagonal (each environment has its own residual variance)
           B - identity (blocks in different environment have the same variance)
           
The assumption regarding R now seems to be more reasonable, as different environments impacts differently on the residuals.
```{r}
#| label: model2

# Model 2
mod2 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

As in this dataset there is no information about the position of the plots in the field, we are not capable of doing the classical spatial corrections, so these will be the only structures considered in the covariance matrix R. 

Now, we will analyse the block covariance matrix:

- **Model 3**: G - compound symmetry (all genotypes have the same variances and the same covariances in different environments)
           R - block-diagonal (each environment has its own residual variance)
           B - block-diagonal (each environment has its own block variance)

```{r}
#| label: model3

# Model 3
mod3 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + at(env):rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

Now, the genetic part will be analysed.

- **Model 4**: G - block-diagonal (each environment has its own genetic variance, but the genetic effect from different environments are not correlated) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)
           
In this model, the genotype effect is nested within the diagonal structure of environments. That means that each environment has its own genetic variance, which seems reasonable, but, this structure assumes that observations from the same genotype in different environments are not correlated, which is a strong assumption.  

```{r}
#| label: model4

# Model 4 
mod4 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + diag(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```

- **Model 5**: G - Heterogeneous compound symmetry (each environment has its own genetic variance, and there is a uniform correlation that connects genotypes in different environments) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)

In this model, the genotype effect is also nested within environments, which has heterogeneous variances for different environments and uniform correlation between environments. So, genotypes from the same environment have the same genetic variance, and the same genotypes in different environments are connected by a uniform correlation. This also allow the prediction of genotypes at untested environments.


```{r}
#| label: model5

# Model 5
mod5 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
```


- **Model 6 and 7**: G - Unstructured with heterogeneuous variances (each environment has its own genetic variance, and different correlations connect genotypes of different environments) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)

In both models above, the genotype effect is also nested within environments, which has heterogeneous variances for different environments and a distinct correlation for each pair of environments. So, genotypes from the same environment have the same variance, and observations from the same genotype in different environments are connected by a correlation that is specific to that pair of environments. That is different from Model 5, where a single correlation value was applied to every pair of environment.

Although the assumptions of these models are least strong, the variance component estimation did not converge, probably because of the high number of parameters. So we were not able to use **Model 6 and 7**.

```{r}
#| label: model6_7

# Model 6 and 7
mod6 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + us(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)

mod7 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corgh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)
```




# Resultados

```{r}
#| label: aic

aic1 <- summary(mod1)$aic
aic2 <- summary(mod2)$aic
aic3 <- summary(mod3)$aic
aic4 <- summary(mod4)$aic
aic5 <- summary(mod5)$aic 

modelos_aic <- data.frame(
  modelo = factor(
    c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
      "HCS(G)-DIAG(R)"),
    levels = c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
               "HCS(G)-DIAG(R)") 
  ), AIC = c(aic1, aic2, aic3, aic4, aic5))
  
  ggplot(modelos_aic, aes(y = modelo, x = AIC)) +
    geom_point(size = 3, shape = 15, colour = "#10342d") +
    geom_segment(aes(x = min(AIC) - 50, xend = AIC, y = modelo, yend = modelo),
                 linewidth = 1.5, colour = "#10342d") +
    geom_label(aes(label = round(AIC, 1)), size = 4,
               position = position_nudge(y = 0.3, x = -30)) +
    theme_minimal(base_size = 14) +
    labs(y = "Modelo", x = "AIC",
         title = "Comparação de estruturas de covariância G×A",
         subtitle = "Variável resposta: GY | Dados Zâmbia (Soja)")

```


```{r}
#| label: fig-trajectory
#| fig-cap: "Environmental Index trajectory of the evaluated locations across years"
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(ggrepel)

trajectory_index <- df_filter %>%
  group_by(env, loc, YEAR, RAINFED) %>% 
  summarise(GY_mean = mean(GY, na.rm = TRUE), .groups = "drop") %>%
  mutate(index = GY_mean - mean(df_filter$GY, na.rm = TRUE)) %>%
  group_by(loc) %>%
  mutate(total_years = n_distinct(YEAR)) %>%
  ungroup()

ggplot(trajectory_index, aes(x = factor(YEAR), y = index, group = loc)) +
  geom_line(data = subset(trajectory_index, total_years > 1), 
            aes(colour = loc), linewidth = 1, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.8, colour = "grey40") +
  
  geom_point(data = subset(trajectory_index, total_years > 1),
             aes(shape = as.factor(RAINFED)), colour = "black", size = 4, stroke = 1.2) +
  geom_point(data = subset(trajectory_index, total_years == 1),
             aes(shape = as.factor(RAINFED)), colour = "firebrick", size = 4, stroke = 1.2) +
  
  ggrepel::geom_text_repel(aes(label = env), colour = "grey50", size = 3, 
                           nudge_x = 0.2,       
                           direction = "y",    
                           segment.color = "grey80") + 
  
  scale_colour_viridis_d(option = "turbo", name = "Location Code") +
  scale_shape_manual(values = c(16, 15, 8), name = "Water Regime") + 
  
  labs(x = "Year", 
       y = "Environmental Index",
       title = "Location Trajectory and Water Regime Impact",
       subtitle = "") +
  
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())

```

correto #########################################################


```{r}
#| label: fig-eberhart-russell
#| fig-cap: "Eberhart and Russell (1966) regression analysis"
#| 
# 1. Run the Eberhart-Russell model
ERm <- metan::ge_reg(
  .data   = df_filter,          
  env     = "env",
  gen     = "gen",
  rep     = "rep",
  resp    = "GY",
  verbose = FALSE
)

# 2. Select the top 30 genotypes with the highest overall mean (b0)
top_gen <- ERm$GY$regression |>
  arrange(desc(b0)) |>
  slice(1:30) |> 
  pull(GEN) |>
  as.character()

# 3. Force genotype G318 into the plot (if not already in the top 30)
top_gen <- unique(c(top_gen, "G318"))

# 4. Plot the regression lines
ggplot(data = subset(ERm$GY$data, GEN %in% top_gen),
       aes(x = IndAmb, y = Y)) +
  facet_wrap(. ~ GEN) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#10342d") +
  geom_point(aes(colour = ENV), size = 2) +
  theme_bw(base_size = 14) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Environmental Index", 
       y = "Grain Yield (GY)", 
       colour = "Environment") +
  scale_color_viridis_d(option = "turbo") +
  stat_regline_equation()
```


```{r}

index <- data.frame(
  index = tapply(df_filter$GY, df_filter$env, mean) - mean(df_filter$GY)
) |> rownames_to_column("env")

dat_grafico_amb <- merge(df_filter, index, by = "env")

ggplot() +
  geom_point(data = subset(index, index < 0),
             aes(x = env, y = index, colour = "Desfavorável", fill = "Desfavorável"),
             size = 2.5, shape = 25) +
  geom_point(data = subset(index, index > 0),
             aes(x = env, y = index, colour = "Favorável", fill = "Favorável"),
             size = 2.5, shape = 24) +
  geom_hline(aes(yintercept = 0), linetype = "dashed", linewidth = 1.2) +
  labs(y = "Índice Ambiental", x = "Ambiente", colour = "", fill = "") +
  theme_bw(base_size = 14) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_colour_manual(values = c("firebrick","forestgreen")) +
  scale_fill_manual(values = c("firebrick","forestgreen"))

# ------------------------------------------------------------------------------
# 3. REGRESSÃO DE EBERHART-RUSSELL
# ------------------------------------------------------------------------------

media_geral <- mean(df_filter$GY, na.rm = TRUE)

medias_env <- df_filter |>
  group_by(env) |>
  summarise(efeito_env = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

medias_gen <- df_filter |>
  group_by(gen) |>
  summarise(efeito_gen = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

dat_ammi_completo <- df_filter |>
  tidyr::complete(env, gen, rep) |>
  left_join(medias_env, by = "env") |>
  left_join(medias_gen, by = "gen") |>
  mutate(
    # Se o GY for NA, preenche com o modelo aditivo; caso contrário, mantém o GY real
    GY = ifelse(is.na(GY), media_geral + efeito_env + efeito_gen, GY)
  ) |>
  select(env, gen, rep, GY) # Manter apenas as colunas necessárias


ERm <- metan::ge_reg(
  .data   = dat_ammi_completo,          
  env     = "env",
  gen     = "gen",
  rep     = "rep",
  resp    = "GY",
  verbose = FALSE
)

# Selecionar os genótipos com maior média geral (b0)
top_gen <- ERm$GY$regression |>
  arrange(desc(b0)) |>
  slice(1:30) |> #dá para mudar os 30
  pull(GEN) |>
  as.character()

#mantém os genótipos mesmo que ná esteja em todos os ambientes?

ggplot(data = subset(ERm$GY$data, GEN %in% top_gen),
       aes(x = IndAmb, y = Y)) +
  facet_wrap(. ~ GEN) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#10342d") +
  geom_point(aes(colour = ENV), size = 2) +
  theme_bw(base_size = 14) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Índice Ambiental", y = "Produtividade (GY)", colour = "Ambiente") +
  scale_color_viridis_d(option = "turbo") +
  stat_regline_equation()



melhor_mod <- mod5 # o model 5 teve o menor AIC
  
  pred <- predict.asreml(object = melhor_mod,classify = "gen:env")$pvals
  
  pred |>
    ggplot(aes(x = reorder(env, -predicted.value),
               y = reorder(gen, predicted.value),
               fill = predicted.value)) +
    geom_tile() +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right",
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) + 
    scale_fill_gradient(low = "#10342d", high = "#b2bc63") +
    labs(x = "Ambiente", y = "Genótipo", fill = "GY",
         title = "Valores preditos — Genótipo × Ambiente")
  
  
  media_geral <- mean(df_filter$GY, na.rm = TRUE)

medias_env <- df_filter |>
  group_by(env) |>
  summarise(efeito_env = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

medias_gen <- df_filter |>
  group_by(gen) |>
  summarise(efeito_gen = mean(GY, na.rm = TRUE) - media_geral, .groups = "drop")

dat_ammi_completo <- df_filter |>
  tidyr::complete(env, gen, rep) |>
  left_join(medias_env, by = "env") |>
  left_join(medias_gen, by = "gen") |>
  mutate(
    # Se o GY for NA, preenche com o modelo aditivo; caso contrário, mantém o GY real
    GY = ifelse(is.na(GY), media_geral + efeito_env + efeito_gen, GY)
  ) |>
  select(env, gen, rep, GY) # Manter apenas as colunas necessárias

ammi <- metan::performs_ammi(
  .data = dat_ammi_completo,
  env   = "env",
  gen   = "gen",
  rep   = "rep",
  resp  = "GY"
)

ggarrange(plot_scores(ammi, type = 1), plot_scores(ammi, type = 2))

plot(ammi, type = "AMMI1")
plot(ammi, type = "AMMI2")
  
  # ------------------------------------------------------------------------------
  # 8. ÍNDICES DE ESTABILIDADE 
  # ------------------------------------------------------------------------------
  
  ge_table <- pred[, c("gen", "env", "predicted.value")] |>
    pivot_wider(names_from = "env", values_from = "predicted.value") |>
    column_to_rownames("gen")
  
  # -- Índice Pi (Lin & Binns) --
  Pi <- apply(ge_table, 1, function(x) {
    sum((x - apply(ge_table, 2, max))^2) / (2 * ncol(ge_table))
  })
  
  df_Pi <- data.frame(gen = names(Pi), Pi = Pi) |>
    arrange(Pi) |>
    slice(1:15) #top 15
  
  ggplot(data = df_Pi, aes(x = reorder(gen, Pi), y = Pi)) +
    geom_col(colour = "black", aes(fill = Pi)) +
    scale_fill_gradient(high = "#10342d", low = "#b2bc63") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          legend.position = "none") +
    labs(y = expression(P[i]), x = "Genótipo",
         title = "Top 15 - Índice Pi (Lin & Binns)")
  
  # -- Safety-First Index--
  alpha <- 0.10
  SF <- apply(ge_table, 1, function(x) mean(x) - qnorm(1 - alpha) * sd(x))
  
  df_SF <- data.frame(gen = names(SF), SF = SF) |>
    arrange(desc(SF)) |>
    slice(1:15)
  
  ggplot(data = df_SF, aes(x = reorder(gen, -SF), y = SF)) +
    geom_col(colour = "black", aes(fill = SF)) +
    scale_fill_gradient(low = "#10342d", high = "#b2bc63") +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          legend.position = "none") +
    labs(y = "Safety-First Index", x = "Genótipo",
         title = "Top 15 - Safety-First (Kataoka)")
  

```

```{r}

  ###########################
  # Genotypes
  ##########################
  g342 <- df_filter |> filter(gen == "G342") |> select (env) |> unique()
  g307 <- df_filter |> filter(gen == "G307") |> select (env) |> unique()
  g310 <- df_filter |> filter(gen == "G310") |> select (env) |> unique()
  g108 <- df_filter |> filter(gen == "G138") |> select (env) |> unique()
  
  
  
  # Melhores sequeiro
  top5 <- pred |>
    filter(env %in% c("E0103","E0104","E0105","E0106","E0107","E0108",
                      "E0169","E0170","E0171","E0172","E0173","E0174",
                      "E0175","E0176")) |>
    group_by(env) |>
    arrange(desc(predicted.value), .by_group = TRUE) |>
    slice_head(n = 5)
  
  # G342 - 13/14
  # G307 - 7/14
  # G310 - 4/14
  
  #############################
  # Análise modelo 5
  ############################
  
  mod5 <- asreml(
    fixed    = GY ~ 1,
    random   = ~ env + env:rep + corh(env):gen,
    residual = ~ dsum(~idv(units) | env),
    data     = df_filter,
    maxit = 100)
  aic5 <- summary(mod5)$aic
  
  plot(mod5)
  
  pred <- predict.asreml(object = mod5,classify = "gen:env",sed=TRUE)
  varcomp <- summary(mod5)$varcomp
  
  h2_mod5 <- 1-(mean(pred$sed[upper.tri(pred$sed)]^2)/(2*mean(varcomp[4:36,1])))
  
```



```{r}
 
  # ------------------------------------------------------------------------------
  # 10. RELIABILITY DE CADA GENÓTIPO EM CADA ENV
  # ------------------------------------------------------------------------------
  library(dplyr)
  library(asreml)
  
  
  # preds <- predict(mod5, classify = "env:gen", sed = TRUE, pworkspace = "8000mb")
  
  tabela_blups <- preds$pvals
  
  # Calcular a PEV (Prediction Error Variance)
  tabela_blups$PEV <- tabela_blups$std.error^2
  
  print(summary(mod5)$varcomp)
  varcomp <- summary(mod5)$varcomp
  
  sigma2_G_valores <- c(
    "E0103" = varcomp[4,1],  
    "E0104" = varcomp[5,1],  
    "E0105" = varcomp[6,1],   
    "E0106" = varcomp[7,1],
    "E0107" = varcomp[8,1],
    "E0169" = varcomp[9,1],
    "E0170" = varcomp[10,1],
    "E0171" = varcomp[11,1],
    "E0172" = varcomp[12,1],
    "E0173" = varcomp[13,1],
    "E0174" = varcomp[14,1],
    "E0176" = varcomp[15,1],
    "E0224" = varcomp[16,1],
    "E0225" = varcomp[17,1],
    "E0226" = varcomp[18,1],
    "E0227" = varcomp[19,1],
    "E0228" = varcomp[20,1],
    "E0229" = varcomp[21,1],
    "E0230" = varcomp[22,1],
    "E0252" = varcomp[23,1],
    "E0253" = varcomp[24,1],
    "E0254" = varcomp[25,1],
    "E0255" = varcomp[26,1],
    "E0278" = varcomp[27,1],
    "E0279" = varcomp[28,1],
    "E0280" = varcomp[29,1],
    "E0281" = varcomp[30,1],
    "E0282" = varcomp[31,1],
    "E0283" = varcomp[32,1],
    "E0284" = varcomp[33,1]
  )
  
  tabela_blups$sigma2_G <- sigma2_G_valores[tabela_blups$env]
  
  # Reliability = 1 - (PEV / Vgen)
  tabela_blups$Reliability <- 1 - (tabela_blups$PEV / tabela_blups$sigma2_G)
  
  # RESULTADO FINAL
  resultado_final <- tabela_blups[, c("env", "gen", "predicted.value", "std.error", "Reliability")]
  head(resultado_final)
  
  mod5_reliability <- readRDS("reability3.rds")
  
  # Removendo valores negativos
  mod5_reliability <- mod5_reliability |> filter(Reliability > 0) 
  
  summary(mod5_reliability)
  
  rel_gen <- mod5_reliability |>
    group_by(gen) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_env = n(),
      .groups = "drop"
    )
  
  
  summary(rel_gen$mean_reliability)
  
  rel_env <- mod5_reliability |>
    group_by(env) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_gen = n(),
      .groups = "drop"
    )
  

  library(dplyr)
  library(ggplot2)
  
  # Genótipos de interesse
  gen_red <- c("G342", "G138")
  
  # Reliability média por genótipo
  rel_gen <- mod5_reliability |>
    group_by(gen) |>
    summarise(
      mean_reliability = mean(Reliability, na.rm = TRUE),
      n_env = n(),
      .groups = "drop"
    )
  
  # Valores dos genótipos destacados
  rel_red <- rel_gen |>
    filter(gen %in% gen_red)
  
  ggplot(rel_gen, aes(x = mean_reliability)) +
    geom_histogram(bins = 30, fill = "grey70", color = "black") +
    geom_vline(
      xintercept = mean(rel_gen$mean_reliability),
      linetype = "dashed",
      linewidth = 1
    ) +
    geom_vline(
      data = rel_red,
      aes(xintercept = mean_reliability),
      color = "red",
      linewidth = 1
    )
  
  ggplot(rel_gen, aes(x = mean_reliability)) +
    geom_histogram(bins = 30, fill = "grey70", color = "black") +
    geom_vline(
      xintercept = mean(rel_gen$mean_reliability),
      linetype = "dashed",
      linewidth = 1
    ) +
    geom_vline(
      data = rel_red,
      aes(xintercept = mean_reliability),
      color = "red",
      linewidth = 1
    ) +
    geom_text(
      data = rel_red,
      aes(x = mean_reliability, y = 0, label = gen),
      color = "red",
      vjust = 1.5,      # empurra o texto para baixo do eixo x
      fontface = "bold",
      size = 4
    ) +
    labs(
      title = "Reliability média dos genótipos nos diferentes ambientes",
      x = "Reliability média",
      y = "Frequência"
    ) +
    coord_cartesian(clip = "off") +  # evita cortar o texto que fica abaixo do eixo
    theme(plot.margin = margin(b = 20))  # espaço extra abaixo do gráfico

```


```{r}

  #################
  # Sequeiro
  library(tidyverse)
  
  seq_pred <- pred |>
    filter(env %in% c("E0103","E0104","E0105","E0106","E0107","E0108",
                      "E0169","E0170","E0171","E0172","E0173","E0174",
                      "E0175","E0176"))
  
  
  ge_table <- seq_pred[, c("gen", "env", "predicted.value")] |>
    pivot_wider(names_from = "env", values_from = "predicted.value") |>
    column_to_rownames("gen")
  
  # -- Índice Pi (Lin & Binns) --
  Pi <- apply(ge_table, 1, function(x) {
    sum((x - apply(ge_table, 2, max))^2) / (2 * ncol(ge_table))
  })
  
  df_Pi <- data.frame(gen = names(Pi), Pi = Pi) |>
    arrange(Pi) |>
    slice(1:15) #top 15
  
  ggplot(data = df_Pi, aes(x = reorder(gen, Pi), y = Pi)) +
    geom_col(colour = "black", aes(fill = Pi)) +
    scale_fill_gradient(high = "#10342d", low = "#b2bc63") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          legend.position = "none") +
    labs(y = expression(P[i]), x = "Genótipo",
         title = "Top 15 - Índice Pi (Lin & Binns)")
  
  # -- Safety-First Index (Kataoka) --
  alpha <- 0.10
  SF <- apply(ge_table, 1, function(x) mean(x) - qnorm(1 - alpha) * sd(x))
  
  df_SF <- data.frame(gen = names(SF), SF = SF) |>
    arrange(desc(SF)) |>
    slice(1:15)
```

