---
title: "Preview Final Project"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

Using provided datasets, you are expected to analyse the data and
interpret the results by applying the concepts and methodologies covered
throughout the course.

Please develop a report using R Markdown or Quarto. The final document
must include the following sections:

Introduction: A brief overview of the research problem. Materials and
Methods: A detailed description of the statistical approach. Results: A
clear presentation of the findings (with Figures and Tables).
Discussion: A concise interpretation of the results in the context of
the study.

# Introduction

This report provides a data analysis of a multi-environment soybean
breeding in Zambia dataset, focusing on two agronomic traits **Grain
Yield** and **Plant Height**. The experimental design of the 33 trials
in Zambia dataset is Randomized Complete Block Design (RCBD), in which
all trials are located in different environment, being the **environment
a combination of year x local**.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml","gt")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia")
```

```{r}
#| label: env-year-table
#| tbl-cap: "Environment distribution"

env_year_table <- data |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: outliers_GY

# Normal Distribution Sd's
ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean(data$GY, na.rm = TRUE) + 
               c(-3, 3) * sd(data$GY, na.rm = TRUE), 
             color = "orange", linetype = "dotted", size = 1) +
  labs(title = "Histogram with outlier limits (±3σ)",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Boxplot outliers
p_hist <- ggplot(data, aes(x = GY)) +
  geom_histogram(bins = 100, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.title.x = element_blank())

p_box <- ggplot(data, aes(y = GY, x = "")) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.color = "red", 
               outlier.size = 2, outlier.alpha = 0.5) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

p_hist / p_box + plot_layout(heights = c(3, 1))

## Removing outliers
df <- data|> 
  mutate (GY = case_when(
    GY == 10650.85 ~ NA,
    TRUE ~ GY
  )) |> 
  mutate(PH_R8 = case_when(
    PH_R8 == 0 ~ NA,
    PH_R8 == 0.0 ~ NA,
    PH_R8 == 308.33 ~ NA,
    TRUE ~ PH_R8
  ))

## Converting columns
df$YEAR <- as.factor(df$YEAR)
df$loc <- as.factor(df$loc)
df$env <- as.factor(df$env)
df$check <- as.factor(df$check)
df$rep <- as.factor(df$rep)
df$gen <- as.factor(df$gen)
df$RAINFED <- as.factor(df$RAINFED)
```

# Single Trial - Statistical Model

Then, each trial was analysed separately, to have a better understanding
of the aspects related to the quality of the data collected in each
trial. This was done by analysing the broad-sense heritability ($H^{2}$)
and the coefficient of variation ($CV%$).

Within each trial (RCBD) there are 3 blocks and approximately 30
genotypes, and for both traits analysed, the statistical model used is
describe as follow:

$$y_{ij}=\mu +b_{j}+g_{i}+e_{ij}$$

Where:

$b_{j}$ is the block fixed effects, as there are only 3 blocks within
each experiment.

$g_{i}$ is the genotype's random effects, where
$g_i \sim N(0, \sigma^2_g)$. This was decided due to the unbalanced
nature of the data, as some trials had missing plots.

$e_{ij}$ is the residual random effect, where $e_{ij} \sim N(0, \sigma^2_e)$. That is, assuming indepedence of residuals within trials

As most trials are completely balanced, the broad-sense heritability was estimated as: 

- individual/plot level: $H² = \frac{\sigma²_g}{ \sigma²_g + \sigma²_e}$, 
- genotype level: $H² = \frac{\sigma²_g}{ \sigma²_g + \frac{\sigma^2_e}{3}}$ 
- Cullis, 2006:  $ H^2 =  1 - \frac{\bar{SED^2}}{2\sigma²_g}$

# Trait 1 - GY

```{r}
#| label: GY_missing_data

# Lista de ambientes
envs_list <- unique(as.character(df$env))

# Verificando dados faltantes para GY em cada ENV
info_env_GH <- df |>
  group_by(env) |>
  summarise(
    obs = n(),
    NAs = sum(is.na(GY)),
    .groups = "drop")
```

```{r}
#| label: GY_analysis
#| output: false

# Rodando para cada trial isoladamente
singleTrial_models_GY <- list()
singleTrial_varcomp_GY <- list()
singleTrial_blup_GY <- list()
singleTrial_reliability_GY <- list()
tabela_geral_GY <- data.frame(env = character(),LRTstat = numeric(),pvalue = numeric(),significance = logical(), h2_plot = numeric(),h2_gen = numeric(), h2_cullis = numeric(), cv = numeric(), stringsAsFactors = FALSE)

for (i in envs_list){
  
  df_env <- df |> filter(env == i)
  df_env$gen <- as.factor(df_env$gen)
  df_env$rep <- as.factor(df_env$rep)
  i <- as.character(i)
  
  model <- asreml(fixed = GY ~ rep,
                               random = ~ gen ,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))

  varcomp <- summary(model)$varcomp
  
  blups <- predict.asreml(model, classify = "gen", sed = TRUE)
  
  pred_BLUP <- blups$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))
  
  m_red_GY <- asreml(fixed = GY ~ rep,
                               data = df_env,
                               maxit = 100,
                               na.action = na.method(y=c("omit")))
  
  pval <- lrt(model, m_red_GY, boundary = T)  
  
  h2_plot <- vpredict(model, H2 ~ V1/(V1+V2)) # nível de parcela
  h2_gen <- varcomp[1,1] / (varcomp[1,1]+(varcomp[2,1]/3)) # ind
  h2_cullis <- 1-(mean(blups$sed[upper.tri(blups$sed)]^2)/(2*varcomp[1,1]))
  
  cv <- sqrt(varcomp[2,1])/mean(df_env$GY, na.rm=T)
  
  reliability <- 1 - (pred_BLUP$std.error^2/varcomp[1,1])
  
  tabela_geral_GY <- rbind(tabela_geral_GY,
    data.frame(
      env = i,
      LRTstat = round(pval$`LR-statistic`,digits=3),
      pvalue = round(pval$`Pr(Chisq)`,digits=4) ,
      significance = pval$`Pr(Chisq)` < 0.05,
      h2_plot = round(h2_plot$Estimate,digits=4),
      h2_gen = round(h2_gen,digits = 4),
      h2_cullis = round(h2_cullis,digits = 4),
      cv = round((cv*100),digits=2)))
  
  singleTrial_models_GY[[i]] <- model
  singleTrial_varcomp_GY[[i]] <- varcomp
  singleTrial_blup_GY[[i]] <- blups$pvals
  singleTrial_reliability_GY[[i]] <- reliability
  }
```

```{r}

tabela_geral_GY |>
  mutate(
    highlight = env %in% c("E0108", "E0175", "E0285")) |> 
  gt(rowname_col = "env") |>
  tab_header(
    title = "Single-trial analysis for grain yield"
  ) |>
  tab_style(
    style = cell_fill(color = "#e8f5e9"),
    locations = cells_body(
      rows = highlight == TRUE
    )
  ) |>
  cols_hide(
    columns = c(significance, highlight)
  ) |>
  cols_label(
    LRTstat = "LRT",
    pvalue = "P-value",
    h2_plot = html("H<sup>2</sup><sub>plot</sub>"),
    h2_gen = html("H<sup>2</sup><sub>genotype</sub>"),
    h2_cullis = html("H<sup>2</sup><sub>Cullis</sub>"),
    cv = "CV (%)"
  ) |>
  fmt_number(
    columns = c(LRTstat, pvalue, h2_plot, h2_gen, h2_cullis, cv),
    decimals = 4
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_options(
    table.font.size = px(12),
    table.width = pct(100)
  )

```

Com base nas herdabilidades observadas na tabela acima, nota-se que as observações dos ambientes "E0108", "E0175" e "E0285" tem uma herdabildiade nula ou muito pequena. Isso mostra que somente uma pequena parte da variação das observações desses experimentos são causadas por fatores genético, e grande parte é causada pelo efeito do ambiente e fatores não controlados. 

Analisar se o modelo fica melhor removendo esses ambientes ou não

```{r}
# Removing envs "E0108", "E0175" e "E0285"
df_filter <- df |> filter(!(env %in% c("E0108", "E0175", "E0285")))

# New table
env_year_table <- df_filter |>
  distinct(loc, YEAR, env) |>
  pivot_wider(
    names_from = YEAR,
    values_from = env,
    values_fn = list(env = paste),
    values_fill = "-"
  )

env_year_table |>
  gt(rowname_col = "loc") |>
  tab_header(
    title = "Environment distribution"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = 12
  )
```

```{r}
#| label: fig-blup_by_envs_plots_GY

blup_all_GY <- bind_rows(
  lapply(names(singleTrial_blup_GY), function(env){
    df <- singleTrial_blup_GY[[env]]
    df$env <- env
    return(df)}))

ggplot(blup_all_GY,
       aes(x = env, y = predicted.value, group = gen, color = gen)) +
      geom_line(alpha = 0.3) +
      geom_point(size = 1) +
      theme_minimal() +
      labs(x = "Environment", y = "Predicted BLUP (GY)",
        title = "2019 - 2023 Genotype's Predicted Values Across Diferent Environments") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```



The fluctuation in the top-ranking positions highlights the presence of a strong Genotype by Environment (GxE) interaction (cross-over patterns) @fig-blup_by_envs_plots_GY

Devido à interação, analisaremos todos os ambientes juntos (estabilidade e GxE)

# Multi trial - Statistical Model

# Conectividade de genótipos entre os ambientes



```{r}
#| label: gen_connectivity

# 
table_connect <- table(as.character(df_filter$env), as.character(df_filter$gen))
  
# Convert to presence/absence
table_connect[table_connect > 0] <- 1
table_connect <- as.data.frame(table_connect)
  
# Percentage of experiments in which each treatment is present
conect_perc <- tapply(table_connect$Freq, table_connect$Var2, sum) /length(unique(df_filter$env)) * 100
  
conect_perc <- data.frame(Genótipo = names(conect_perc),
                          Conectividade = conect_perc)

table_connect

conect_perc <- conect_perc[order(conect_perc$Conectividade, decreasing = TRUE),]

kable(conect_perc, caption = "Connectivity of genotypes across experiments",
align = "c")

#Presence 
pres <- mean(table_connect$Freq) * 100

#Absence
abs <- 100 - pres

  
  table_connect$Var2 <- factor(as.character(table_connect$Var2), levels = conect_perc$Genótipo)
  ordem_env <- names(sort(tapply(table_connect$Freq, table_connect$Var1, sum), decreasing = TRUE))
  table_connect$Var1 <- factor(as.character(table_connect$Var1), levels = ordem_env)
  
  ggplot(table_connect, aes(x = Var2, y = Var1, fill = factor(Freq))) +
    geom_tile() +
    scale_fill_manual(
      values = c("white", "#A7C7E7"),
      labels = c("Absence", "Presence")
    ) +
    labs(
      title = "Presence–absence of treatments across experiments",
      x = "Genotypes (Sorted by Connectivity)",
      y = "Environments",
      fill = "Presence"
    ) +
    theme_minimal() + 
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid = element_blank() 
    )
  
  # Matrix
  M <- xtabs(Freq ~ Var1 + Var2, data = table_connect)
  
  # Shared treatments
  shared <- M %*% t(M)
  
  # Number of treatments per experiment
  n_trat <- rowSums(M)
  
  # Jaccard
  jaccard <- shared
  
  for(i in rownames(shared)) {
    for(j in colnames(shared)) {
      jaccard[i, j] <- shared[i, j] / (n_trat[i] + n_trat[j] - shared[i, j])
    }
  }
  
  diag(jaccard) <- 1
  
  pheatmap(
    jaccard,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    clustering_method = "average",
    color = rev(hcl.colors(100, "Blues")),
    fontsize_row = 10,
    fontsize_col = 10,
    angle_col = 90,
    main = "Experiment connectivity based on shared genotypes"
  )
  
  ## Heatmap com os genótipos
  
  # 1. Matriz de contingência original (Linhas = Ambientes, Colunas = Genótipos)
  M <- xtabs(Freq ~ Var1 + Var2, data = table_connect)
  
  # 2. Shared environments per genotype (INVERTIDO: t(M) %*% M gera Genótipo x Genótipo)
  shared_gen <- t(M) %*% M
  
  # 3. Número de ambientes em que cada genótipo está presente (Agora sim usamos colSums)
  n_env_per_gen <- colSums(M)
  
  # 4. Inicializa a matriz de Jaccard para os Genótipos
  jaccard_gen <- shared_gen
  
  # 5. Loop do Jaccard adaptado para os genótipos (com proteção contra divisões por zero)
  for(i in rownames(shared_gen)) {
    for(j in colnames(shared_gen)) {
      denominador <- n_env_per_gen[i] + n_env_per_gen[j] - shared_gen[i, j]
      
      if (denominador == 0) {
        jaccard_gen[i, j] <- 0
      } else {
        jaccard_gen[i, j] <- shared_gen[i, j] / denominador
      }
    }
  }
  
  # Garante a diagonal unitária (similaridade de um genótipo com ele mesmo é 1)
  diag(jaccard_gen) <- 1
  
  # 6. Plot do Heatmap dos Genótipos
  library(pheatmap)
  pheatmap(
    jaccard_gen,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    clustering_method = "average",
    color = rev(hcl.colors(100, "Blues")),
    fontsize_row = 6,  
    fontsize_col = 6,
    angle_col = 90,
    main = "Genotype connectivity based on shared environments"
  )
  
```

Para analisar o desempenho dos genótipos nos diferentes locais conjuntamente, será também utilizado modelos mistos, que é mais apropriado para o desbalanceamento de dados.

Our inicial model is:

**REVER MODELO (GxE)**

$$y_{ijk}=l_{i} +b_{j}+g_{k}+e_{ijk}$$

Where:

$l_{i}$ is the environment random effects, as the environments (local x year combinations) can be treated as a random sample of all possible combinations of these factors

$b_{j}$ is the block random effects, as they are nested in environments $b \sim N(0,\sigma^2_bB)$

$g_{k}$ is the genotype's random effects, where $g \sim N(0,\sigma^2_gG)$

$e_{ijk}$ is the residual random effect $e \sim N(0,\sigma^2_eR)$

Different (co)variances structures will be tested for the random effects in these models. Each one of the models make different assumptions regarding the residuals, block or genetic random distribution. First, we will analyse the residuals covariance matrix, then blocks and genetics.

- Model 1: G - compound symmetry (genetic variance is homogeneous for genotypes in different environments and the covariance between the same genotypes performance in any two environments is constant)
           R - identity (all residuals have the same variance in different environments)
           B - identity (blocks in different environment have the same variance)
            
```{r}
# Model 1
mod1 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ idv(units),
  data     = df_filter,
  maxit = 100)
aic1 <- summary(mod1)$aic ; aic1

```

- Model 2: G - compound symmetry (all genotypes have the same variances and the same covariances in different environments)
           R - block-diagonal (each environment has its own residual variance)
           B - identity (blocks in different environment have the same variance)
           
The assumption regarding R now seems to be more reasonable, as different environments impacts differently on the residuals. By analysing the AIC value, we observe that the model fitted better, compensating the parameters that were added.

```{r}
# Model 2
mod2 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
aic2 <- summary(mod2)$aic; aic2
```

As in this trial there is no information about the position of the plots in the field, we are not capable of doing the classical spatial corrections, so these will be the structure adopted to the covariance matrix R. Now, we will analyse the block covariance matrix:

- Model 3: G - compound symmetry (all genotypes have the same variances and the same covariances in different environments)
           R - block-diagonal (each environment has its own residual variance)
           B - block-diagonal (each environment has its own block variance)

```{r}
# Model 3
mod3 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + at(env):rep + gen + gen:env,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
aic3 <- summary(mod3)$aic; aic3
```

Based on the AIC value, the third model did not fitted better than the second one. This occurred probably because in each environment, there are only 3 blocks to estimate the variance component. Based on that, we will assume that the previous structure for covariance matrix B (identity - Model 2) is better. Now, the genetic part will be analysed.

- Model 4: G - block-diagonal (each environment has its own genetic variance, but the genetic effect from different environments are not correlated) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)
           
In this model, the genotype effect is nested within the diagonal structure of environments. That means that each environment has its own genetic variance, which seems reasonable, but, this structure assumes that observations from the same genotype in different environments are not correlated, which is a strong assumption.  

```{r}
# Model 4 
mod4 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + diag(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
aic4 <- summary(mod4)$aic; aic4

```

- Model 5: G - Heterogeneous compound symmetry (each environment has its own genetic variance, and there is a uniform correlation that connects genotypes in different environments) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)

In this model, the genotype effect is also nested within environments, which has heterogeneous variances for different environments and uniform correlation between environments. So, genotypes from the same environment have the same genetic variance, and the same genotypes in different environments are connected by a uniform correlation. This also allow the prediction of genotypes at untested environments.


```{r}
# Model 5
mod5 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 100)
aic5 <- summary(mod5)$aic ; aic5

```


- Model 6 and 7: G - Unstructured with heterogeneuous variances (each environment has its own genetic variance, and different correlations connect genotypes of different environments) 
           R - block-diagonal (each environment has its own residual variance)
           B - identity (all blocks have the same variance)

In both models below, the genotype effect is also nested within environments, which has heterogeneous variances for different environments and a distinct correlation for each pair of environments. So, genotypes from the same environment have the same variance, and observations from the same genotype in different environments are connected by a correlation that is specific to that pair of environments. That is different from Model 5, where a single correlation value was applied to every pair of environment.

Although the assumptions of these models are least strong, the variance component estimation did not converge, probably because of the high number of parameters. So we were not able to use this models below.

```{r}
# Model 6 and 7
mod6 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + us(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)

mod7 <- asreml(
  fixed    = GY ~ 1,
  random   = ~ env + env:rep + corgh(env):gen,
  residual = ~ dsum(~idv(units) | env),
  data     = df_filter,
  maxit = 300)
```


```{r}
modelos_aic <- data.frame(
  modelo = factor(
    c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
      "HCS(G)-DIAG(R)"),
    levels = c("CS(G)-ID(R)", "CS(G)-DIAG(R)", "CS(G)-DIAG(B)-DIAG(R)", "DIAG(G)-DIAG(R)",
               "HCS(G)-DIAG(R)") 
  ), AIC = c(aic1, aic2, aic3, aic4, aic5))
  
  ggplot(modelos_aic, aes(y = modelo, x = AIC)) +
    geom_point(size = 3, shape = 15, colour = "#10342d") +
    geom_segment(aes(x = min(AIC) - 50, xend = AIC, y = modelo, yend = modelo),
                 linewidth = 1.5, colour = "#10342d") +
    geom_label(aes(label = round(AIC, 1)), size = 4,
               position = position_nudge(y = 0.3, x = -30)) +
    theme_minimal(base_size = 14) +
    labs(y = "Modelo", x = "AIC",
         title = "Comparação de estruturas de covariância G×A",
         subtitle = "Variável resposta: GY | Dados Zâmbia (Soja)")

```




