---
title: "Regression"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    fig-width: 10
    theme: 
      light: journal
      dark: darkly
    embed-resources: true

crossref: {}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```


```{r}
#| label: trials-data

# Loading packages
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
library(grid)
library(ggcorrplot)
library(car)

# Loading data (Zambia)
data = read.csv("https://raw.githubusercontent.com/mauricioaraujj/Pan_African_Trials_Network/refs/heads/main/data/data.csv", sep = ';') |> filter(COUNTRY == "Zambia") 

# Selecting traits
traits <- c("FLW_DAYS", "GY", "PH_R8", "PL_EMERG_COUNT", "Pod_Shattering")
selected_data <- data |> 
  select(COUNTRY, YEAR, env, gen, loc, SEASON, rep, check, all_of(traits)) |> 
  filter(loc %in% c("L0118", "L0133", "L074", "L0134") & YEAR == "2023") |> 
  mutate(trial = case_when(
    loc == "L0118" & YEAR == "2023" ~ 1,
    loc == "L0133" & YEAR == "2023" ~ 2,
    loc == "L074"  & YEAR == "2023" ~ 3,
    loc == "L0134" & YEAR == "2023" ~ 4)) |> 
  mutate(trial = as.factor(trial))
```


# Multiple Regression

Now, a multiple regression model with the 3 independent variables will be considered: $y_i =\beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \beta_3x_{i3} + \varepsilon_{i}$:

```{=latex}
\begin{equation*}
\begin{bmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
\vdots \\
y_{354} \\
\end{bmatrix}
=
\begin{bmatrix}
1 & x_{1\mathbf{1}} & x_{1\mathbf{2}} & x_{1\mathbf{3}} \\
1 & x_{2\mathbf{1}} & x_{2\mathbf{2}} & x_{2\mathbf{3}} \\
\vdots & \vdots & \vdots & \vdots \\
1 & x_{354\mathbf{1}} & x_{354\mathbf{2}} & x_{354\mathbf{3}}
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\
\beta_1 \\
\beta_2 \\
\beta_3 \\
\end{bmatrix}
+
\begin{bmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
\vdots \\
\varepsilon_{354}
\end{bmatrix}
\end{equation*}
```

As it is a full rank model, the parameters estimation is:

$$
\underset{\sim}{\hat\beta} = (X'X)^{-1}X'\underset{\sim}{y}
$$

The estimated values are:
```{r}
# Removing outliers and other variables
selected_data <- selected_data |> 
  slice(-c(91, 109, 110, 190, 270, 286)) |> 
  select(GY, PL_EMERG_COUNT, PH_R8, FLW_DAYS)

# Observations vector
y <- as.matrix(selected_data$GY)

# Design matrix
X <- as.matrix(cbind(1, selected_data[,-1]))

# Parameters estimation
(beta <- solve(t(X)%*% X) %*% t(X) %*% y)

# R check
(modelo_completo <- lm(selected_data$GY ~ selected_data$PL_EMERG_COUNT + selected_data$PH_R8 + selected_data$FLW_DAYS)) 

```

So, considering the 3 variables, the model adjusted for all observations (without considering effects associated with their specific local/trial) is:

$$
y_i = 4014.965 - 2.638x_{i1} + 42.728x_{i2} - 71.994x_{i3}  + \varepsilon_i
$$

Each partial regression coefficient represents the **isolated effect** of that variable in the response variable (considering the presence of the others independent variables in the model).

For exemple, if all independent variables remain constant, with the increase of one unit (1cm) in plant heigth (variable $x_2$), an increase of 42.728 kg/ha in grain yield is expected. The same reasoning follows for the other variables.

First, we will test the hypothesis $H_0: \beta_1 = \beta_2 = \beta_3 = 0$, to verify if at least one variable is significant to explain grain yield.

Assuming $y \sim N_n(XB, \sigma^2I)$, we can calculate:

$$
F = \frac{QM_{REG}}{QM_{RES}}
$$

$$
SQ_{REG} = \underset{\sim}{y}'(X(X'X)^{-1}X' - \frac{1}{n}J)\underset{\sim}{y}
$$

$$
SQ_{RES} = \underset{\sim}{y}'(I-X(X'X)^{-1}X')\underset{\sim}{y}
$$

Since they are quadratic forms, the rank of the associated matrix determines the degrees of freedom of sums of squares. As the matrices are idempotent, their traces are equal to their ranks.

```{r}
# Declaring matrices
n <- nrow(X)
I <- diag(n) # Identity
J <- matrix(1, nrow=n, ncol=n) # 1´s matrix

# Sums os squares
nucleo_A <- X %*% solve(t(X) %*% X) %*% t(X) - 1/n * J
SQ_reg <- t(y) %*% nucleo_A %*% y 

nucleo_B <- I - X %*% solve(t(X) %*% X) %*% t(X)
SQ_res <- t(y) %*% nucleo_B %*% y

# Degrees of freedom
gl_reg <- sum(diag(nucleo_A)) 
gl_res <- sum(diag(nucleo_B)) 

# Mean squares
QM_reg <- SQ_reg / gl_reg
QM_res <- SQ_res / gl_res

# F
F_calculado <- QM_reg / QM_res 

# p-value
pvalor <- pf(F_calculado, gl_reg, gl_res, lower.tail = FALSE)

# ANOVA table
tabela_anova <- data.frame(
  "Variation Causes" = c("Regression", "Residuals"),
  SQ = c(round(SQ_reg, digits = 3),round(SQ_res, digits =3)),
  gl = c(gl_reg,gl_res),
  QM = c(round(QM_reg,digits=3), round(QM_res, digits = 3)),
  Fcalc = c(round(F_calculado, digits = 3), ""), 
  Pvalor = c(pvalor,""), check.names = FALSE
)
kable(tabela_anova, align = c("l", "r", "r", "r","r"))
```

To test the same hypothesis with R, we will compare the full and the null-model:

```{r}
# R
modelo_nulo <- lm(selected_data$GY~1, data=selected_data) # only intercept
anova(modelo_nulo, modelo_completo)
```

The results are the same. We conclude that at least one of the variables are important to explain grain yield, considering the model explained above.

Now, each variable will be tested separately, in the presence of the other variables in the model. The hypothesis are $H_0: \beta_i = 0$ e $H_a : \beta_i \ne 0$, for i = 1, 2, 3, and the test used is: $H_0 : C\underset{\sim}{\beta} = 0$ . 


$$
SQ_{HIP} = (C\underset{\sim}{\hat\beta})'[C (X'X)^{-1}C']^{-1}(C\underset{\sim}{\hat\beta})
$$

```{r}
# Vectors C
c_1 <- matrix(data=c(0,1,0,0),ncol=4)
c_2 <- matrix(data=c(0,0,1,0),ncol=4)
c_3 <- matrix(data=c(0,0,0,1),ncol=4)

# Hypothesis 1 - Sums of squares
Cbeta <- c_1 %*% beta 
nucleo_C <- solve(c_1 %*% solve(t(X) %*% X) %*% t(c_1))
SQ_hip_1 <- t(Cbeta) %*% nucleo_C %*% Cbeta

# Hypothesis 2 - Sums of squares
Cbeta <- c_2 %*% beta 
nucleo_C <- solve(c_2 %*% solve(t(X) %*% X) %*% t(c_2))
SQ_hip_2 <- t(Cbeta) %*% nucleo_C %*% Cbeta

# Hypothesis 3 - Sums of squares
Cbeta <- c_3 %*% beta 
nucleo_C <- solve(c_3 %*% solve(t(X) %*% X) %*% t(c_3))
SQ_hip_3 <- t(Cbeta) %*% nucleo_C %*% Cbeta

# F calculados de cada hipótese
F_calculado_1 <- SQ_hip_1 / QM_res
F_calculado_2 <- SQ_hip_2 / QM_res
F_calculado_3 <- SQ_hip_3 / QM_res

# P valor de cada hipótese
pvalor_1 <- pf(F_calculado_1, 1 , gl_res, lower.tail = FALSE)
pvalor_2 <- pf(F_calculado_2, 1 , gl_res, lower.tail = FALSE)
pvalor_3 <- pf(F_calculado_3, 1 , gl_res, lower.tail = FALSE)
```


```{r, echo=FALSE}
# Imprimindo resultados 
cat("calculated F - Beta1 = 0:", F_calculado_1,"\tp-value - Beta1 = 0:", pvalor_1, "\ncalculated F - Beta 2 = 0:", F_calculado_2,"\tp-value - Beta2 = 0:", pvalor_2,"\ncalculated F - Beta3 = 0:", F_calculado_3,"\tp-value - Beta3 = 0:", pvalor_3) 
```

```{r}
Anova(modelo_completo, type = 3)
```

The result is equal to the one produced by Anova() function

Considering $\alpha_f = 0.05$ , without correcting the p-values, we **reject** the hypothesis $H_0 : \beta_1 = 0$ ,$H_0 : \beta_2 = 0$ and $H_3: \beta_3 = 0$ , meaning that variables $x_1$ - "PL_EMERG_COUNT", $x_2$ - "PH_R8" e $x_3$ - "FLW_DAYS" are significant in the **full model**.

However, when doing this, the gobalwise level of significance is $\alpha_f=1-(1-0.05)^3= 0.142$ . To adress this problem, Bonferroni's Correction could be used, so that $\alpha_c = 0.05 / 3= 0.016$. Now, $H_0 : \beta_2 =0$ and $H_0: \beta_3 = 0$ are **rejected**, which means that only $x_2$ - "PH_R8" and $x_3$ - "FLW_DAYS" are significant in the **full model**, when correcting the significance level.


## Stepwise

Other method that could be used to select variables in the model is stepwise:

1.  In each step is added a variable in the model (usually the one more correlated with the response variable);

2.  The added variable is avaluated in the model, using a criteria such as  p-value, AIC, BIC..

3.  The full model is obtained when the addition of no remaining variable is significant


```{r}
# Verifing multicolinearity
vif(modelo_completo)
```

No multicolinearity is observed. The order of variable inclusion will be: $x_2 \rightarrow x_3 \rightarrow x_1$ , which was decided using the correlation between these variables and y.

### First Step - Adding X2 (Plant Height)

Model 1: $y_i = \beta_0 + \beta_2x_{i2} + \varepsilon_i$

Hip: $H_0: \beta_2 = 0$


```{r, results='hold'}
# Design matrix
X_1 <- X[, c(1,3)]

# Model 1 - SQReg
nucleo_A_1 <- X_1 %*% solve(t(X_1) %*% X_1) %*% t(X_1) - 1/n * J
SQ_reg_1 <- t(y) %*% nucleo_A_1 %*% y 

# Model 1 - SQRes
nucleo_B_1 <- I - X_1 %*% solve(t(X_1) %*% X_1) %*% t(X_1)
SQ_res_1 <- t(y) %*% nucleo_B_1 %*% y 

# Model 1 - df
gl_reg_1 <- sum(diag(nucleo_A_1)) 
gl_res_1 <- sum(diag(nucleo_B_1)) 

# Model 1 - Mean Squares
QM_reg_1 <- SQ_reg_1 / gl_reg_1
QM_res_1 <- SQ_res_1 / gl_res_1

# Model 1 - F and p-value
F_calculado_1 <- QM_reg_1 / QM_res_1 
pvalor_1 <- pf(F_calculado_1, gl_reg_1, gl_res_1, lower.tail = FALSE)

```

```{r, echo=FALSE}
# Model 1 - ANOVA Table
tabela_anova <- data.frame(
  "Variation Causes" = c("Regression", "Residuals"),
  SQ = c(round(SQ_reg_1, digits = 3),round(SQ_res_1, digits =3)),
  gl = c(gl_reg_1,gl_res_1),
  QM = c(round(QM_reg_1,digits=3), round(QM_res_1, digits = 3)),
  Fcalc = c(round(F_calculado_1, digits = 3), ""), 
  Pvalor = c(pvalor_1,""), check.names = FALSE
)
kable(tabela_anova, align = c("l", "r", "r", "r","r"))
```

As this variable was significant, it will be mantained in the model.

### Second Step - Adding X3 (Days to Flowering)

Model 2: $y_i = \beta_0 + \beta_2x_{i2} + \beta_1x_{i3} +\varepsilon_i$

Hip: $H_0: \beta_3 = 0$ . 

If the hypothesis is rejected, we will test $H_0 : \beta_2 = 0$ , to verify if this variable remains significant after adding X3.
We can compare both models using: 
$$
F = \frac{SQ_{REG \ (CURRENT)} - SQ_{REG \ (PREVIOUS)}}{QM_{RES\ (CURRENT)}}
$$

```{r, results='hold'}
# Design Matrix
X_2 <- X[, c(1,3,4)]

# Model 2 - SQReg
nucleo_A_2 <- X_2 %*% solve(t(X_2) %*% X_2) %*% t(X_2) - 1/n * J
SQ_reg_2 <- t(y) %*% nucleo_A_2 %*% y 

# Model 2 - SQRes
nucleo_B_2 <- I - X_2 %*% solve(t(X_2) %*% X_2) %*% t(X_2)
SQ_res_2 <- t(y) %*% nucleo_B_2 %*% y 

# Model 2 - df
gl_res_2 <- sum(diag(nucleo_B_2)) 

# Model 2 - Mean Squares
QM_hip_2 <- SQ_reg_2 - SQ_reg_1
QM_res_2 <- SQ_res_2 / gl_res_2

# Model 2 - F and p-value
F_calculado_2 <- QM_hip_2 / QM_res_2 
pvalor_2 <- pf(F_calculado_2, 1 , gl_res_2, lower.tail = FALSE)

cat("calculated F - Model 2:", F_calculado_2,"\tp-value - Model 2:", pvalor_2)
```

As variable X3 is significant, it stays in the model. Now, we will calculate if variable X2 remains significant after the addition of X3, by calculating $SQ(x_2 \mid x_3)$.

```{r, results='hold'}
# Design matrix
X_2_2 <- X[, c(1,4)] 

# SQReg after removing X2
nucleo_A_2_2 <- X_2_2 %*% solve(t(X_2_2) %*% X_2_2) %*% t(X_2_2) - 1/n * J
SQ_reg_2_2 <- t(y) %*% nucleo_A_2_2 %*% y 

# F and p-value
F_calculado_2_2 <- (SQ_reg_2 - SQ_reg_2_2) / QM_res_2 
pvalor_2_2 <- pf(F_calculado_2_2, 1, gl_res_2, lower.tail = FALSE)

cat("calculated F - X2 removed:", F_calculado_2_2,"\tp-value - X2 removed:", pvalor_2_2)
```

It can be seen that $SQ(x_2 \mid x_3)$ is significant, meaning that $x_2$ is still important after the addition of $x_1$. So, both variables stay in the model

### Third 3 - Adding X1 (Plant Emergence Count)

Model 3: $y_i = \beta_0 + \beta_5x_{i2} + \beta_1x_{i3} + \beta_2 x_{i1} +\varepsilon_i$

Hip: $H_0: \beta_1 = 0$ . 

The tests are done as in the previous steps.

```{r, results='hold'}
# Design matrix
X_3 <- X  # full model

# Model 3 - SQReg
nucleo_A_3 <- X_3 %*% solve(t(X_3) %*% X_3) %*% t(X_3) - 1/n * J
SQ_reg_3 <- t(y) %*% nucleo_A_3 %*% y 

# Model 3 - SQRes
nucleo_B_3 <- I - X_3 %*% solve(t(X_3) %*% X_3) %*% t(X_3)
SQ_res_3 <- t(y) %*% nucleo_B_3 %*% y 

# Model 3 - df
gl_res_3 <- sum(diag(nucleo_B_3)) 

# Model 3 - Mean Squares
QM_hip_3 <- SQ_reg_3 - SQ_reg_2
QM_res_3 <- SQ_res_3 / gl_res_3

# Model 3 - F and p-value
F_calculado_3 <- QM_hip_3 / QM_res_3 
pvalor_3 <- pf(F_calculado_3, 1 , gl_res_3, lower.tail = FALSE)

cat("calculated F - Model 3:", F_calculado_3,"\tp-value - Modelo 3:", pvalor_3)
```

We verified that $SQ(x_1 \mid x_2 ,x_3)$ is significant, considering $\alpha_c = 0.05$. Now, we will verify if the other variables remain significant.

```{r, results='hold'}
# Removing X2

# Design matrix
X_3_1 <- X[, c(1,2,4)] 

# SQReg after removing X2
nucleo_A_3_1 <- X_3_1 %*% solve(t(X_3_1) %*% X_3_1) %*% t(X_3_1) - 1/n * J
SQ_reg_3_1 <- t(y) %*% nucleo_A_3_1 %*% y 

# F 
F_calculado_3_1 <- (SQ_reg_3 - SQ_reg_3_1) / QM_res_3 
pvalor_3_1 <- pf(F_calculado_3_1, 1, gl_res_3, lower.tail = FALSE)

# Removing X3
X_3_2 <- X[, c(1,2,3)]

# SQReg after removing X3
nucleo_A_3_2 <- X_3_2 %*% solve(t(X_3_2) %*% X_3_2) %*% t(X_3_2) - 1/n * J
SQ_reg_3_2 <- t(y) %*% nucleo_A_3_2 %*% y 

# F
F_calculado_3_2 <- (SQ_reg_3 - SQ_reg_3_2) / QM_res_3 
pvalor_3_2 <- pf(F_calculado_3_2, 1, gl_res_3, lower.tail = FALSE)

# Imprimindo resultados 
cat("calculated F - Remoção X2:", F_calculado_3_1,"\tp-value - Removing X2:", pvalor_3_1, "\ncalculated F - Remoção X3:", F_calculado_3_2,"\tp-value - Removing X3:", pvalor_3_2)
```

The same model is obtained using stepwise() function, which uses AIC:

```{r}
# Utilizando função stepwise
modelo_stepwise <- step(lm(GY ~ 1, data =as.data.frame(selected_data)),
scope = list(lower = ~1, upper = ~ PL_EMERG_COUNT + PH_R8 + FLW_DAYS),
direction = "forward", trace=0)

summary(modelo_stepwise)
```

Considering the full model, the parameters can be estimated as $\underset{\sim}{\hat\beta} = (X'X)^{-1}X'\underset{\sim}{y}$

```{r}
# Design matrix
X_final <- X 

# beta estimation
beta <- solve(t(X_final)%*%X_final) %*% t(X_final) %*% y

# Total SQ
nucleo_total <- I - (1/n*J)
SQ_total <- t(y) %*% nucleo_total %*% y 

# Regression SQ
nucleo_reg <- X_final %*% solve(t(X_final)%*% X_final) %*% t(X_final) - (1/n * J)
SQreg_final <- t(y) %*% nucleo_reg %*% y

# R²
R_2 <- SQreg_final / SQ_total

cat("R²: ",R_2, "\nIntercept:\t\t     ", beta[1], "\nX1 coefficient:   ", beta[2], "\nX2 coefficient:   ", beta[3], "\nX3 coefficient:   ", beta[4])

# R verification
summary(lm(GY ~ PL_EMERG_COUNT + PH_R8 + FLW_DAYS, data=selected_data))
```

The final model is:

$$ y_i = 4014.965 - 2.638x_{i1} + 42.728x_{i2} - 71.994x_{i3} + \varepsilon_i $$

$R^2 \approx 0.41$ indicates that the model explains approximately 41% of the observed grain yield variation. Note that we are considering data from different locations, without including an effect associated with locations or the interaction of genotypes x environment. 

Regarding the interpretation of each partial regression coefficient, as explained, they represent the **isolated effect** of that variable in the response (when the other variables are included in the model). For example, in this model, if "Plant Emergence Count (X1)" and "Plant Height (X2)" are constant, when we increase 1 day in X3 (Days to Flowering), the productivity is expected to reduce in 71,9 kg/ha, indicating that cultivars with faster cycles may be better in this environments.

# SIMPLE LINEAR REGRESSION BY TRIAL

While the across-trials multiple regression provided a comprehensive overview of the main drivers of grain yield in Zambia, it is essential to recognize that phenotypic expression is highly dependent on specific environmental conditions.

In this section, we perform a **Simple Linear Regression** for each of the four locations (Trials 1 to 4) independently. The primary objective is to verify if the positive correlation between plant height ([PH_R8) and yield (GY)]{.underline} observed in the global model holds true for each specific site, while also exploring the behavior of other relevant agronomic traits.

To ensure mathematical transparency *(as proposed by the exercise)*, we calculate the parameters using **Manual Matrix Algebra** through the Ordinary Least Squares (OLS) estimator:

$$ \hat{\beta} = (X'X)^{-1}X'y $$

This step-by-step matrix approach, followed by a validation using the native `lm()` function, allows us to pinpoint environments where the biological response deviates from the general trend—providing a more granular and reliable recommendation for localized crop management.

```{r}

# --- Trial 1: L0118 ---
cat("\n\n=== MATRICES FOR TRIAL 1 (L0118) ===\n")
df_T1 <- selected_data %>% 
  filter(trial == 1, PH_R8 <= 200) %>% 
  na.omit() #O na.omit() garante que apenas linhas completas entrem no cálculo matricial, os NA estavam "quebrando" a matriz

y_1 <- as.matrix(df_T1$GY)
X_1 <- as.matrix(cbind(Intercept = 1, PH_R8 = df_T1$PH_R8))

# Álgebra Manual
beta_1 <- solve(t(X_1) %*% X_1) %*% t(X_1) %*% y_1
print("1. Matrix (X'X) for Trial 1:")
print(round(t(X_1) %*% X_1, 2))
print("2. Manual Coefficients (beta):")
print(beta_1)

# Prova Real
cat("--- Validation with lm() ---\n")
print(coef(lm(GY ~ PH_R8, data = df_T1)))


# --- Trial 2: L0133 ---
cat("\n\n=== MATRICES FOR TRIAL 2 (L0133) ===\n")
df_T2 <- selected_data %>% filter(trial == 2) %>% na.omit()

y_2 <- as.matrix(df_T2$GY)
X_2 <- as.matrix(cbind(Intercept = 1, PH_R8 = df_T2$PH_R8))

beta_2 <- solve(t(X_2) %*% X_2) %*% t(X_2) %*% y_2
print("1. Matrix (X'X) for Trial 2:")
print(round(t(X_2) %*% X_2, 2))
print("2. Manual Coefficients (beta):")
print(beta_2)

cat("--- Validation with lm() ---\n")
print(coef(lm(GY ~ PH_R8, data = df_T2)))


# --- Trial 3: L074 ---
cat("\n\n=== MATRICES FOR TRIAL 3 (L074) ===\n")
df_T3 <- selected_data %>% filter(trial == 3) %>% na.omit()

y_3 <- as.matrix(df_T3$GY)
X_3 <- as.matrix(cbind(Intercept = 1, PH_R8 = df_T3$PH_R8))

beta_3 <- solve(t(X_3) %*% X_3) %*% t(X_3) %*% y_3
print("1. Matrix (X'X) for Trial 3:")
print(round(t(X_3) %*% X_3, 2))
print("2. Manual Coefficients (beta):")
print(beta_3)

cat("--- Validation with lm() ---\n")
print(coef(lm(GY ~ PH_R8, data = df_T3)))


# --- Trial 4: L0134 ---
cat("\n\n=== MATRICES FOR TRIAL 4 (L0134) ===\n")
df_T4 <- selected_data %>% filter(trial == 4) %>% na.omit()

y_4 <- as.matrix(df_T4$GY)
X_4 <- as.matrix(cbind(Intercept = 1, PH_R8 = df_T4$PH_R8))

beta_4 <- solve(t(X_4) %*% X_4) %*% t(X_4) %*% y_4
print("1. Matrix (X'X) for Trial 4:")
print(round(t(X_4) %*% X_4, 2))
print("2. Manual Coefficients (beta):")
print(beta_4)

cat("--- Validation with lm() ---\n")
print(coef(lm(GY ~ PH_R8, data = df_T4)))
```

To facilitate the interpretation of the calculated parameters, we present below the individual regression plots for each trial. Each panel displays the observed data points, the fitted linear regression line (calculated via Ordinary Least Squares), and the 95% confidence interval. The annotations highlight the Coefficient of Determination ($R^2$) and the p-value, providing a visual confirmation of the goodness-of-fit and the statistical significance of the plant height ($x_2$) as a predictor for grain yield ($y$) in each specific environment.

```{r}

# Função para extrair métricas de ajuste do modelo
get_metrics <- function(model) {
  r2 <- round(summary(model)$r.squared, 3)
  p_val <- summary(model)$coefficients[2,4]
  p_txt <- ifelse(p_val < 0.001, "< 0.001", paste("=", round(p_val, 3)))
  return(paste0("R² = ", r2, "\np ", p_txt))
}

# plots individuais
p1_plot <- ggplot(df_T1, aes(x = PH_R8, y = GY)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  annotate("text", x = -Inf, y = Inf, label = get_metrics(mod1), 
           hjust = -0.2, vjust = 1.5, color = "darkred", fontface = "bold") +
  labs(title = "Trial 1: L0118", x = "PH_R8 (cm)", y = "GY (kg/ha)") +
  theme_bw()

p2_plot <- ggplot(df_T2, aes(x = PH_R8, y = GY)) +
  geom_point(color = "forestgreen", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  annotate("text", x = -Inf, y = Inf, label = get_metrics(mod2), 
           hjust = -0.2, vjust = 1.5, color = "darkred", fontface = "bold") +
  labs(title = "Trial 2: L0133", x = "PH_R8 (cm)", y = "GY (kg/ha)") +
  theme_bw()

p3_plot <- ggplot(df_T3, aes(x = PH_R8, y = GY)) +
  geom_point(color = "darkorange", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  annotate("text", x = -Inf, y = Inf, label = get_metrics(mod3), 
           hjust = -0.2, vjust = 1.5, color = "darkred", fontface = "bold") +
  labs(title = "Trial 3: L074", x = "PH_R8 (cm)", y = "GY (kg/ha)") +
  theme_bw()

p4_plot <- ggplot(df_T4, aes(x = PH_R8, y = GY)) +
  geom_point(color = "purple", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  annotate("text", x = -Inf, y = Inf, label = get_metrics(mod4), 
           hjust = -0.2, vjust = 1.5, color = "darkred", fontface = "bold") +
  labs(title = "Trial 4: L0134", x = "PH_R8 (cm)", y = "GY (kg/ha)") +
  theme_bw()



painel <- (p1_plot + p2_plot) / (p3_plot + p4_plot) +
  plot_annotation(title = "Regression Analysis by Trial",
                  subtitle = "Linear models comparing Yield (GY) and Plant Height (PH_R8)",
                  theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5)))

print(painel)
```

The individual trial analysis reveals a significant departure from the across-trials global model, emphasizing that phenotypic expression is highly dependent on specific environmental conditions. While the general trend in Zambia suggested a positive correlation between plant height and yield ($R^2 \approx 0.41$), the dissected data shows that this relationship is strictly environment-specific. In trials such as L0118 and L074, the relationship between plant height ($PH\_R8$) and grain yield ($GY$) proved to be statistically non-significant, as indicated by $p$-values of $0.141$ and $0.116$, respectively. This lack of significance suggests that in these specific sites, vegetative vigor is not a reliable proxy for yield potential, and other environmental variables likely exerted more influence on the final output.

The most striking evidence of GxE interaction is found in Trial L0133, which displays a significant negative correlation ($R^2 = 0.122, p < 0.001$). In this particular environment, genotypes with greater structural height were severely penalized in terms of productivity, a phenomenon often associated with lodging or transpiration stress, where the water demand of a larger leaf area cannot be met by local soil moisture.

In conclusion, the disparity between the global regression model and the individual trial slopes confirms that a "one-size-fits-all" recommendation based on plant height is biologically risky for soybean breeding in Zambia. While height drives yield at a national scale, it can act as a direct constraint in specific environments like L0133

\-

After dissecting the regression models for each individual trial, we conclude this analysis with a consolidated overview of all environments. This "Across-Trials" perspective is fundamental to understanding the general trend of soybean performance in Zambia during the 2023 season. By aggregating the data we can estimate a global slope that represents the average biological response of the crop to variations in plant height ($PH\_R8$). This final step serves as a validation of our initial findings, the goal here is to measure the statistical power of plant height as a primary driver of grain yield on a national scale. Through the same manual matrix algebra and Ordinary Least Squares (OLS) estimation, we aim to confirm if the positive correlation observed in the multiple regression

```{r}

df_across <- selected_data %>% 
  filter(!(trial == 1 & PH_R8 > 200)) %>% 
  na.omit()

cat("\n--- ACROSS-TRIALS MATRIX CALCULATION ---\n")
y_total <- as.matrix(df_across$GY)
X_total <- as.matrix(cbind(Intercept = 1, PH_R8 = df_across$PH_R8))

# Cálculo matricial
beta_total <- solve(t(X_total) %*% X_total, t(X_total) %*% y_total)
print(beta_total)

mod_total <- lm(GY ~ PH_R8, data = df_across)
r2_all <- round(summary(mod_total)$r.squared, 3)

ggplot(df_across, aes(x = PH_R8, y = GY)) +
  geom_point(alpha = 0.3, color = "gray40") +
  geom_smooth(method = "lm", color = "red", fill = "indianred1", size = 1.2) +
  annotate("label", x = -Inf, y = Inf, 
           label = paste0("y = ", round(beta_total[2], 2), "x + ", round(beta_total[1], 2), 
                          "\nR² = ", r2_all, "\np < 0.001"),
           hjust = -0.1, vjust = 1.2, color = "darkred", fontface = "bold", size = 5) +
  labs(title = "Across-Trials Overview (Zambia 2023)", 
       subtitle = "Consolidated relationship between Plant Height and Yield",
       x = "Plant Height - PH_R8 (cm)", 
       y = "Grain Yield - GY (kg/ha)") +
  theme_minimal()
```

The results from the matrix calculation confirm a positive relationship, with a slope coefficient of $42.71$, aligning perfectly with our initial global multiple regression model. This indicates that, on average, every 1 cm increase in plant height corresponds to an additional $42.71$ kg/ha in yield across the country. Furthermore, the coefficient of determination ($R^2 = 0.333$) suggests that approximately 33% of the yield variation is explained solely by height, while the high statistical significance ($p < 0.001$) provides a solid foundation for breeding selections that favor taller, more vigorous genotypes under general conditions.
