---
title: "Spatial analysis of augmented designs in sweet potato"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

The dataset available here has six trials laid out in an augmented row-column design (more details about this design here). Analyse each trial individually with increasing order of spatial adjustment. Report the improvement of the modelling (if any) when spatial trends are dealt with. Your report should have differences in the variance components, broad-sense heritabilities, and rankings. For each trial, determine the best model using semivariograms and the AIC. Don't forget that models with different fixed effects must be compared using the adjusted version of the AIC available on the lmmtools package.

# Introduction

This report provides a data analysis of a spacial correction of a multi-environment sweet potato experiment in Africa. Each one of the six environments were analyzed with multiple models testing different approaches to spacial correction. It was estimated  variance components, broad-sense heritabilities, and rankings based on BLUP predictions.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data
#| warning: false

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml", "lmmtools")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data 
data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data$row_number <- as.factor(data$row_number)
data$col_number <- as.factor(data$col_number)
data$rowgroup <- as.factor(data$rowgroup)
data$colgroup <- as.factor(data$colgroup)
data$check <- as.factor(data$check)
data$geno <- as.factor(data$geno)
data$check.cod = as.factor(data$check.cod)
data$geno.cod  = as.factor(data$geno.cod)
```

The trials were conducted across three countries (Rwanda, Namibia, and Senegal) during the 2021 and 2022 growing seasons. To account for spatial variation, each trial was analyzed individually to implement single-trial spatial corrections.
```{r}
# Trial identification
unique(data$env)
```
This dataset is derived from a multi-environment sweet potato trial. The experimental layout follows an augmented row-column design, where each environment comprises a grid of 50 rows and 30 columns. These rows are grouped into 5 super-rows, and the columns into 3 super-columns. Each resulting combination (hereafter referred to as a miniblock) consists of 10 rows and 10 columns, encompassing a total of 8 checks and 1,178 test genotypes. 


The model used in this report was the same described in the article of [Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325), which can be represented as the model below: 


$$ y = \mu + rowgroup + colgroup + rowgroup.colgroup + rowgroup.row + colgroup.col + genotype + check + \epsilon$$


Where, $\mu$ is the intercept, 


$rowgroup$ is the random effect of the super-row, 


$colgroup$ is the random effect of the super-column, 


$rowgroup.colgroup$ is the random effect of the intersection of super-row and super-column (the miniblock), 


$rowgroup.row$ the random effect of the row number nested in the super-row,


$colgroup.col$ the random effect of the column number nested in the super-column,


$genotype$ the random effect of the genotype,

$check$ the fixed effect of the check and


$\epsilon$ the random effect of the error.


# Namibia season 2022 (Trial 1 - OT1NAM22)

```{r}
#| label: fig-t1-prod

t1 <- data |> filter (env == "OT1NAM22")

unique(data$row_number)
unique(data$col_number)
unique(data$rowgroup)
unique(data$colgroup)
unique(data$check.cod)
length(unique(data$geno.cod))
unique(data$rc.group)

# Conta os genótipos únicos por grupo de linha-coluna
tapply(data$geno, data$rc.group, function(x) length(unique(x)))

# Modifying the geno.cod and check.cod columns
t1 <- t1 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn)
  )
```

In order to test the spatial correction, in this season and country was analyzed six models. The most basic model is the same described in [Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325). 

```{r}
#| label: models-t1
#| output: false

# Piepho & Williams (2016) model as a starter

model_1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)


# Correcting col and row number residuals

model_2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Adding units to random effects (nuggets) + row and colgroups as fixed effects

model_3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + check.cod,
             random = ~ rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

#Adding the linearization to try to reduce the noise in the lines and columns 

model_4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Adding units to random effects (nuggets)

model_5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

#Reduction of model complexity

model_6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

```

### Wald test

In every model, the genetic variance was tested with the LRT test and it was significant. This result indicates that at least two genotypes are statistically different from each other. The same tendency is observed in the checks, that were tested using the Wald test.

:::{.panel-tabset}

### Model 1 
```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1)
```

### Model 2
```{r}
wald.asreml(model_2)
```

### Model 3

```{r}
wald.asreml(model_3)
```

### Model 4

```{r}
wald.asreml(model_4)
```

### Model 5

```{r}
wald.asreml(model_5)
```

### Model 6

```{r}
wald.asreml(model_6)
```

:::

### LRT

```{r}
#| output: false
#Random effects

red_model1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)

red_model2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + check.cod,
             random = ~ rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

lrt(model_1, red_model1)
lrt(model_2, red_model2)
lrt(model_3, red_model3)
lrt(model_4, red_model4)
lrt(model_5, red_model5)
lrt(model_6, red_model6)
```

::: {.panel-tabset}

### Model 1

```{r}
lrt(model_1, red_model1)

```

### Model 2

```{r}
lrt(model_2, red_model2)

```

### Model 3

```{r}
lrt(model_3, red_model3)

```

### Model 4

```{r}
lrt(model_4, red_model4)

```

### Model 5

```{r}
lrt(model_5, red_model5)

```

### Model 6

```{r}
lrt(model_6, red_model6)
```

:::

### Semivariograms

Evaluating the semivariograms and the heatmap of root yield, a clear spatial trend can be observed around the center of the superblock. In both visualizations, the central portion of the block exhibits a distinct pattern characterized by lower yields, which corresponds to an almost flat trend in the semivariogram. Even though spatial corrections were applied, this behavior persists. This may indicate a major source of micro-environmental heterogeneity in the field, potentially corresponding to the centerline of the irrigation system or, conversely, a specific zone that did not receive adequate water from the sprinkler irrigation.

```{r}
#| label: fig-variograms-t1
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (AR1 × AR1 Residual Structure)"
#|   - "Semivariogram for Model 3 (Fixed Row/Column Effects + Nugget Effect)"
#|   - "Semivariogram for Model 4 (Linear Trend + Spatial Adjustment)"
#|   - "Semivariogram for Model 5 (Adding units to random effects (nuggets))"
#|   - "Semivariogram for Model 6 (Reduction of model complexity)"
#|   - "Heatmap of the distribution the root yield in tonnes per ha (rytha) in trial 1 - Namibia season 2022"


plot(varioGram(model_1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

ggplot(data = t1, aes(x = col_number, y = row_number, fill = rytha)) + 
  geom_tile() + 
  theme_minimal() + 
  theme(text = element_text(size = 12)) + 
  labs(x = "Column", y = "Row") +
  scale_fill_viridis_c(option = "magma")
```


### AIC 

This analysis presented results similar to those obtained from the semivariograms. The baseline model (Model 1) exhibited the highest Akaike Information Criterion (AIC) value (8,723.64), indicating the poorest fit. Conversely, Model 6, which incorporated the units term into the random effects to account for the nugget effect, was the best-performing model, yielding an AIC value of 7,982.90—a substantial reduction of 740.74 units.

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel = icREML(list(model_1, model_2, model_3, model_4, model_5, model_6)); modsel
modsel <- as.data.frame(modsel)

```

:::{.panel-tabset}

### Model 1

```{r}
icREML(list(model_1))
```

### Model 2

```{r}
icREML(list(model_2))
```

### Model 3

```{r}
icREML(list(model_3))
```

### Model 4

```{r}
icREML(list(model_4))
```

### Model 5

```{r}
icREML(list(model_5))
```

### Model 6

```{r}
icREML(list(model_6))
```

:::

### Variance components

:::{.panel-tabset}

### Model 1
```{r}
# Variance components

varcomp_m1 <- summary(model_1)$varcomp
varcomp_m1 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2 

```{r}
varcomp_m2 <- summary(model_2)$varcomp
varcomp_m2|> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3 <- summary(model_3)$varcomp
varcomp_m3|> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4 <- summary(model_4)$varcomp
varcomp_m4|> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5 <- summary(model_5)$varcomp 
varcomp_m5|> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 6

```{r}
varcomp_m6 <- summary(model_6)$varcomp
varcomp_m6|> 
  kbl(caption = "Variance components for Model 6.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

:::

### Ranking 

```{r}
#| label: t1-blups

#BLUPs
blup1 <- predict.asreml(model_1, classify = "geno.cod", sed = TRUE)
blup2 <- predict.asreml(model_2, classify = "geno.cod", sed = TRUE)
blup3 <- predict.asreml(model_3, classify = "geno.cod", sed = TRUE)
blup4 <- predict.asreml(model_4, classify = "geno.cod", sed = TRUE)
blup5 <- predict.asreml(model_5, classify = "geno.cod", sed = TRUE)
blup6 <- predict.asreml(model_6, classify = "geno.cod", sed = TRUE)

blup <- list(blup1, blup2, blup3, blup4, blup5, blup6)
predic_value_1 <- list()

for (i in 1:length(blup)) {
  
  predic_value_1[[i]] <- blup[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1 <- predict(model_1, classify = "check.cod")$pvals
blues_checks2 <- predict(model_2, classify = "check.cod")$pvals
blues_checks3 <- predict(model_3, classify = "check.cod")$pvals
blues_checks4 <- predict(model_4, classify = "check.cod")$pvals
blues_checks5 <- predict(model_5, classify = "check.cod")$pvals
blues_checks6 <- predict(model_6, classify = "check.cod")$pvals

blue <- list(blues_checks1, blues_checks2, blues_checks3, blues_checks4, blues_checks5, blues_checks6)
bpredic_value_1 <- list()

for (i in 1:length(blue)) {
  
  bpredic_value_1[[i]] <- blue[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

## Graph
aux <- rbind(
  blup1$pvals |>
    mutate(
      model = 'model 1'
    ),
  blup5$pvals |> mutate(
    model = 'model 5'
  )
)

ggplot(data = aux, aes(x = model, y = predicted.value)) +
  geom_point(size = 3, alpha = .5, color = "green") +
  geom_line(aes(group = geno.cod), linewidth = 1.3, alpha = .4, color = "#10342d") +
  theme_minimal() + 
  labs(x = "Model", y = "Yield (BLUP)") + 
  theme(text = element_text(size = 20))
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t1
#| tbl-cap: Broad-sense Heritability of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1 <- 1-(mean(blup1$sed[upper.tri(blup1$sed)]^2)/(2*varcomp_m1[6,1])); h2_mod1
h2_mod2 <- 1-(mean(blup2$sed[upper.tri(blup2$sed)]^2)/(2*varcomp_m2[6,1])); h2_mod2
h2_mod3 <- 1-(mean(blup3$sed[upper.tri(blup3$sed)]^2)/(2*varcomp_m3[4,1])); h2_mod3
h2_mod4 <- 1-(mean(blup4$sed[upper.tri(blup4$sed)]^2)/(2*varcomp_m4[6,1])); h2_mod4
h2_mod5 <- 1-(mean(blup5$sed[upper.tri(blup5$sed)]^2)/(2*varcomp_m5[6,1])); h2_mod5
h2_mod6 <- 1-(mean(blup6$sed[upper.tri(blup6$sed)]^2)/(2*varcomp_m6[3,1])); h2_mod6

h2 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6"),
  H2 = c(h2_mod1, h2_mod2, h2_mod3, h2_mod4, h2_mod5, h2_mod6)
)

options(knitr.table.format = "html")

h2 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Trial 2 - OT1RWE21

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 3 - OT1RWE22

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 4 - OT1SER22

variance components, broad-sense heritabilities, and rankings

## Best model

Encaixar com o da Bruna no começo
```{r}

library(asreml)
library(lmmtools)
library(tidyverse)
library(knitr)     

data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data <- data |>
  mutate(
    env = as.factor(env),
    row_number = as.factor(row_number),
    col_number = as.factor(col_number),
    rowgroup = as.factor(rowgroup),
    colgroup = as.factor(colgroup),
    check = as.factor(check),
    geno = as.factor(geno)
  )
```









# Trial 5 - OT2NAM21

```{r}
# ====================================================================
# Trial 5 - OT2NAM21
# ====================================================================
#| label: trial5-analysis
#| message: false
#| warning: false

t5 <- data |> filter(env == "OT2NAM21") |> arrange(row_number, col_number)

# -----------------------------------------------------------------
# 1. Model Fit
# -----------------------------------------------------------------
M1_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno,
  residual = ~ idv(units),
  na.action = na.method("include"),
  data = t5, workspace = 128e6
)

M2_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"),
  data = t5, workspace = 128e6
)

M3_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"),
  data = t5, workspace = 128e6
)


# -----------------------------------------------------------------
# 2. Model Selection (Adjusted AIC / icREML)
# -----------------------------------------------------------------
cat("### Model Comparison (icREML) - OT2NAM21\n")

lista_modelos_t5 <- list(
  Baseline = M1_t5, 
  Spatial_AR1 = M2_t5, 
  Spatial_Nugget = M3_t5
)

aic_table_t5 <- lmmtools::icREML(lista_modelos_t5)

aic_table_t5 <- aic_table_t5 |> 
  arrange(AIC)

print(aic_table_t5)

best_model_name_t5 <- aic_table_t5$model[1]
cat("\nThe best model for OT2NAM21 is:", best_model_name_t5, "\n\n")

best_model_t5 <- if(best_model_name_t5 == "Baseline") M1_t5 else if(best_model_name_t5 == "Spatial_AR1") M2_t5 else M3_t5

# -----------------------------------------------------------------
# 3. Semivariograms Comparison
# -----------------------------------------------------------------
cat("### Semivariograms Comparison\n")
plot(varioGram(M1_t5, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Baseline (M1)")
plot(varioGram(M2_t5, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Spatial AR1 (M2)")
plot(varioGram(M3_t5, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Spatial Nugget (M3)")

# -----------------------------------------------------------------
# 4. Variance Components & Heritability 
# -----------------------------------------------------------------
cat("### Variance Components\n")
print(summary(best_model_t5)$varcomp)

calc_cullis_corrigido_t5 <- function(model) {
  vc_temp <- summary(model)$varcomp
  gen_var_name <- grep("Candidate.*geno", rownames(vc_temp), value = TRUE)[1]
  gen_var <- vc_temp[gen_var_name, "component"]
  
  pred <- predict(model, classify = "at(check, 'Candidate'):geno", sed = TRUE)
  
  sed_mat <- pred$sed
  vdBLUP_avg <- mean(sed_mat[upper.tri(sed_mat, diag = FALSE)]^2, na.rm = TRUE)
  
  H2 <- 1 - (vdBLUP_avg / (2 * gen_var))
  return(H2)
}

cat("\nBroad-sense Heritability (Cullis):", round(calc_cullis_corrigido_t5(best_model_t5), 3), "\n")

# -----------------------------------------------------------------
# 5. Rankings
# -----------------------------------------------------------------
blups_t5 <- predict(best_model_t5, classify = "at(check, 'Candidate'):geno")$pvals

ranking_t5 <- blups_t5 |>
  filter(check == "Candidate") |>
  mutate(predicted.value = as.numeric(predicted.value)) |>
  arrange(desc(predicted.value)) |>
  select(geno, predicted.value, std.error) |>
  head(15)

cat("\n### Top 15 Candidates\n")
print(ranking_t5)
```
## Best model
For Trial 5 (OT2NAM21), model selection based on the adjusted AIC (icREML) identified the Spatial_Nugget model (M3) as the most adequate fit (icREML = 5474.04). The inclusion of a two-dimensional autoregressive structure along with an independent nugget effect successfully accounted for the spatial heterogeneity in the field, which was visually confirmed by the flattening of the semivariance curve. Applying the rigorous Cullis et al. (2006) method—based on the mean squared standard error of a difference (SED) between all pairs of genotypes—the broad-sense heritability was estimated at 0.516. This moderate heritability indicates a reliable degree of genetic control over the phenotypic expression in this environment, ensuring accuracy in the selection of top-ranking candidates, most notably UGP202000272-5.


# Trial 6 - OT2SER21

# ====================================================================
# Trial 6 - OT2SER21
# ====================================================================
#| label: trial6-analysis
#| message: false
#| warning: false

t6 <- data |> filter(env == "OT2SER21") |> arrange(row_number, col_number)

# -----------------------------------------------------------------
# 1. Model Fit
# -----------------------------------------------------------------
M1_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno,
  residual = ~ idv(units),
  na.action = na.method("include"),
  data = t6, workspace = 128e6
)

M2_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"),
  data = t6, workspace = 128e6
)

M3_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + row_number + col_number + at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"),
  data = t6, workspace = 128e6
)

# -----------------------------------------------------------------
# 2. Model Selection
# -----------------------------------------------------------------
cat("### Model Comparison (icREML) - OT2SER21\n")

lista_modelos_t6 <- list(
  Baseline = M1_t6, 
  Spatial_AR1 = M2_t6, 
  Spatial_Nugget = M3_t6
)

aic_table_t6 <- lmmtools::icREML(lista_modelos_t6)

aic_table_t6 <- aic_table_t6 |> 
  arrange(AIC)

print(aic_table_t6)

best_model_name_t6 <- aic_table_t6$model[1]
cat("\nThe best model for OT2SER21 is:", best_model_name_t6, "\n\n")

best_model_t6 <- if(best_model_name_t6 == "Baseline") M1_t6 else if(best_model_name_t6 == "Spatial_AR1") M2_t6 else M3_t6

# -----------------------------------------------------------------
# 3. Semivariograms Comparison
# -----------------------------------------------------------------
cat("### Semivariograms Comparison\n")
plot(varioGram(M1_t6, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Baseline (M1)")
plot(varioGram(M2_t6, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Spatial AR1 (M2)")
plot(varioGram(M3_t6, type = "default", spatial = "trend", formula = ~NULL, composite = TRUE, model = c("empirical"), metric = c("euclidean"), angle = 0, angle.tol = 180, nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE), main = "Spatial Nugget (M3)")

# -----------------------------------------------------------------
# 4. Variance Components & Heritability
# -----------------------------------------------------------------
cat("### Variance Components\n")
print(summary(best_model_t6)$varcomp)

calc_cullis_corrigido_t6 <- function(model) {
  vc_temp <- summary(model)$varcomp
  gen_var_name <- grep("Candidate.*geno", rownames(vc_temp), value = TRUE)[1]
  gen_var <- vc_temp[gen_var_name, "component"]
  
  pred <- predict(model, classify = "at(check, 'Candidate'):geno", sed = TRUE)
  
  sed_mat <- pred$sed
  vdBLUP_avg <- mean(sed_mat[upper.tri(sed_mat, diag = FALSE)]^2, na.rm = TRUE)
  
  H2 <- 1 - (vdBLUP_avg / (2 * gen_var))
  return(H2)
}

cat("\nBroad-sense Heritability (Cullis):", round(calc_cullis_corrigido_t6(best_model_t6), 3), "\n")

# -----------------------------------------------------------------
# 5. Rankings
# -----------------------------------------------------------------
blups_t6 <- predict(best_model_t6, classify = "at(check, 'Candidate'):geno")$pvals

ranking_t6 <- blups_t6 |>
  filter(check == "Candidate") |>
  mutate(predicted.value = as.numeric(predicted.value)) |>
  arrange(desc(predicted.value)) |>
  select(geno, predicted.value, std.error) |>
  head(15)

cat("\n### Top 15 Candidates\n")
print(ranking_t6)
```
## Best model
For Trial 6 (OT2SER21), model selection via icREML revealed an identical fit between the Spatial_AR1 and Spatial_Nugget models (icREML = 3624.78). Inspection of the variance components explained this convergence: the independent error term (`units`) in the Spatial_Nugget model hit the parameter boundary (Bound 'B'), effectively shrinking its variance to zero. This indicates that the two-dimensional autoregressive structure (AR1xAR1) alone was sufficient to capture all the residual spatial trends in this specific field, making the addition of a nugget effect redundant. Using the SED-based formulation by Cullis et al. (2006), the broad-sense heritability was estimated at 0.146. This relatively low heritability suggests that phenotypic expression in this trial was heavily masked by environmental micro-variations, warranting a more cautious approach when selecting candidates based on these specific BLUPs.

