---
title: "Spatial analysis of augmented designs in sweet potato"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

The dataset available here has six trials laid out in an augmented row-column design (more details about this design here). Analyse each trial individually with increasing order of spatial adjustment. Report the improvement of the modelling (if any) when spatial trends are dealt with. Your report should have differences in the variance components, broad-sense heritabilities, and rankings. For each trial, determine the best model using semivariograms and the AIC. Don't forget that models with different fixed effects must be compared using the adjusted version of the AIC available on the lmmtools package.

# Introduction

This report provides a data analysis of a spacial correction of a multi-environment sweet potato experiment in Africa. Each one of the six environments were analyzed with multiple models testing different approaches to spacial correction. It was estimated  variance components, broad-sense heritabilities, and rankings based on BLUP predictions.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data
#| warning: false

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml", "lmmtools")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data 
data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data$row_number <- as.factor(data$row_number)
data$col_number <- as.factor(data$col_number)
data$rowgroup <- as.factor(data$rowgroup)
data$colgroup <- as.factor(data$colgroup)
data$check <- as.factor(data$check)
data$geno <- as.factor(data$geno)
data$check.cod = as.factor(data$check.cod)
data$geno.cod  = as.factor(data$geno.cod)
```

The trials were conducted across three countries (Rwanda, Namibia, and Senegal) during the 2021 and 2022 growing seasons. To account for spatial variation, each trial was analyzed individually to implement single-trial spatial corrections.
```{r}
# Trial identification
unique(data$env)
```
This dataset is derived from a multi-environment sweet potato trial. The experimental layout follows an augmented row-column design, where each environment comprises a grid of 50 rows and 30 columns. These rows are grouped into 5 super-rows, and the columns into 3 super-columns. Each resulting combination (hereafter referred to as a miniblock) consists of 10 rows and 10 columns, encompassing a total of 8 checks and 1,178 test genotypes. 


The model used in this report was the same described in the article of [Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325), which can be represented as the model below: 


$$ y = \mu + rowgroup + colgroup + rowgroup.colgroup + rowgroup.row + colgroup.col + genotype + check + \epsilon$$


Where, $\mu$ is the intercept, 


$rowgroup$ is the random effect of the super-row, 


$colgroup$ is the random effect of the super-column, 


$rowgroup.colgroup$ is the random effect of the intersection of super-row and super-column (the miniblock), 


$rowgroup.row$ the random effect of the row number nested in the super-row,


$colgroup.col$ the random effect of the column number nested in the super-column,


$genotype$ the random effect of the genotype,

$check$ the fixed effect of the check and


$\epsilon$ the random effect of the error.


# Namibia season 2022 (Trial 1 - OT1NAM22)

```{r}
#| label: fig-t1-prod

t1 <- data |> filter (env == "OT1NAM22")

unique(data$row_number)
unique(data$col_number)
unique(data$rowgroup)
unique(data$colgroup)
unique(data$check.cod)
length(unique(data$geno.cod))
unique(data$rc.group)

# Conta os genótipos únicos por grupo de linha-coluna
tapply(data$geno, data$rc.group, function(x) length(unique(x)))

# Modifying the geno.cod and check.cod columns
t1 <- t1 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn)
  )
```

In order to test the spatial correction, in this season and country was analyzed six models. The most basic model is the same described in [Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325). 

```{r}
#| label: models-t1
#| output: false

# Piepho & Williams (2016) model as a starter

model_1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)


# Correcting col and row number residuals

model_2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Adding units to random effects (nuggets) + row and colgroups as fixed effects

model_3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + check.cod,
             random = ~ rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

#Adding the linearization to try to reduce the noise in the lines and columns 

model_4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Adding units to random effects (nuggets)

model_5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

#Reduction of model complexity

model_6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

```

### Wald test

In every model, the genetic variance was tested with the LRT test and it was significant. This result indicates that at least two genotypes are statistically different from each other. The same tendency is observed in the checks, that were tested using the Wald test.

:::{.panel-tabset}

### Model 1 
```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1)
```

### Model 2
```{r}
wald.asreml(model_2)
```

### Model 3

```{r}
wald.asreml(model_3)
```

### Model 4

```{r}
wald.asreml(model_4)
```

### Model 5

```{r}
wald.asreml(model_5)
```

### Model 6

```{r}
wald.asreml(model_6)
```

:::

### LRT

```{r}
#| output: false
#Random effects

red_model1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)

red_model2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + check.cod,
             random = ~ rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(rn):ar1(cn),
             na.action = na.method("include"),
             data = t1, maxint=50)

lrt(model_1, red_model1)
lrt(model_2, red_model2)
lrt(model_3, red_model3)
lrt(model_4, red_model4)
lrt(model_5, red_model5)
lrt(model_6, red_model6)
```

::: {.panel-tabset}

### Model 1

```{r}
lrt(model_1, red_model1)

```

### Model 2

```{r}
lrt(model_2, red_model2)

```

### Model 3

```{r}
lrt(model_3, red_model3)

```

### Model 4

```{r}
lrt(model_4, red_model4)

```

### Model 5

```{r}
lrt(model_5, red_model5)

```

### Model 6

```{r}
lrt(model_6, red_model6)
```

:::

### Semivariograms

Evaluating the semivariograms and the heatmap of root yield, a clear spatial trend can be observed around the center of the superblock. In both visualizations, the central portion of the block exhibits a distinct pattern characterized by lower yields, which corresponds to an almost flat trend in the semivariogram. Even though spatial corrections were applied, this behavior persists. This may indicate a major source of micro-environmental heterogeneity in the field, potentially corresponding to the centerline of the irrigation system or, conversely, a specific zone that did not receive adequate water from the sprinkler irrigation.

```{r}
#| label: fig-variograms-t1
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (AR1 × AR1 Residual Structure)"
#|   - "Semivariogram for Model 3 (Fixed Row/Column Effects + Nugget Effect)"
#|   - "Semivariogram for Model 4 (Linear Trend + Spatial Adjustment)"
#|   - "Semivariogram for Model 5 (Adding units to random effects (nuggets))"
#|   - "Semivariogram for Model 6 (Reduction of model complexity)"
#|   - "Heatmap of the distribution the root yield in tonnes per ha (rytha) in trial 1 - Namibia season 2022"


plot(varioGram(model_1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

ggplot(data = t1, aes(x = col_number, y = row_number, fill = rytha)) + 
  geom_tile() + 
  theme_minimal() + 
  theme(text = element_text(size = 12)) + 
  labs(x = "Column", y = "Row") +
  scale_fill_viridis_c(option = "magma")
```


### AIC 

This analysis presented results similar to those obtained from the semivariograms. The baseline model (Model 1) exhibited the highest Akaike Information Criterion (AIC) value (8,723.64), indicating the poorest fit. Conversely, Model 6, which incorporated the units term into the random effects to account for the nugget effect, was the best-performing model, yielding an AIC value of 7,982.90—a substantial reduction of 740.74 units.

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel = icREML(list(model_1, model_2, model_3, model_4, model_5, model_6)); modsel
modsel <- as.data.frame(modsel)

```

:::{.panel-tabset}

### Model 1

```{r}
icREML(list(model_1))
```

### Model 2

```{r}
icREML(list(model_2))
```

### Model 3

```{r}
icREML(list(model_3))
```

### Model 4

```{r}
icREML(list(model_4))
```

### Model 5

```{r}
icREML(list(model_5))
```

### Model 6

```{r}
icREML(list(model_6))
```

:::

### Variance components

:::{.panel-tabset}

### Model 1
```{r}
# Variance components

varcomp_m1 <- summary(model_1)$varcomp
varcomp_m1 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2 

```{r}
varcomp_m2 <- summary(model_2)$varcomp
varcomp_m2|> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3 <- summary(model_3)$varcomp
varcomp_m3|> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4 <- summary(model_4)$varcomp
varcomp_m4|> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5 <- summary(model_5)$varcomp 
varcomp_m5|> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 6

```{r}
varcomp_m6 <- summary(model_6)$varcomp
varcomp_m6|> 
  kbl(caption = "Variance components for Model 6.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

:::

### Ranking 

```{r}
#| label: t1-blups

#BLUPs
blup1 <- predict.asreml(model_1, classify = "geno.cod", sed = TRUE)
blup2 <- predict.asreml(model_2, classify = "geno.cod", sed = TRUE)
blup3 <- predict.asreml(model_3, classify = "geno.cod", sed = TRUE)
blup4 <- predict.asreml(model_4, classify = "geno.cod", sed = TRUE)
blup5 <- predict.asreml(model_5, classify = "geno.cod", sed = TRUE)
blup6 <- predict.asreml(model_6, classify = "geno.cod", sed = TRUE)

blup <- list(blup1, blup2, blup3, blup4, blup5, blup6)
predic_value_1 <- list()

for (i in 1:length(blup)) {
  
  predic_value_1[[i]] <- blup[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1 <- predict(model_1, classify = "check.cod")$pvals
blues_checks2 <- predict(model_2, classify = "check.cod")$pvals
blues_checks3 <- predict(model_3, classify = "check.cod")$pvals
blues_checks4 <- predict(model_4, classify = "check.cod")$pvals
blues_checks5 <- predict(model_5, classify = "check.cod")$pvals
blues_checks6 <- predict(model_6, classify = "check.cod")$pvals

blue <- list(blues_checks1, blues_checks2, blues_checks3, blues_checks4, blues_checks5, blues_checks6)
bpredic_value_1 <- list()

for (i in 1:length(blue)) {
  
  bpredic_value_1[[i]] <- blue[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

## Graph
aux <- rbind(
  blup1$pvals |>
    mutate(
      model = 'model 1'
    ),
  blup5$pvals |> mutate(
    model = 'model 5'
  )
)

ggplot(data = aux, aes(x = model, y = predicted.value)) +
  geom_point(size = 3, alpha = .5, color = "green") +
  geom_line(aes(group = geno.cod), linewidth = 1.3, alpha = .4, color = "#10342d") +
  theme_minimal() + 
  labs(x = "Model", y = "Yield (BLUP)") + 
  theme(text = element_text(size = 20))
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t1
#| tbl-cap: Broad-sense Heritability of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1 <- 1-(mean(blup1$sed[upper.tri(blup1$sed)]^2)/(2*varcomp_m1[6,1])); h2_mod1
h2_mod2 <- 1-(mean(blup2$sed[upper.tri(blup2$sed)]^2)/(2*varcomp_m2[6,1])); h2_mod2
h2_mod3 <- 1-(mean(blup3$sed[upper.tri(blup3$sed)]^2)/(2*varcomp_m3[4,1])); h2_mod3
h2_mod4 <- 1-(mean(blup4$sed[upper.tri(blup4$sed)]^2)/(2*varcomp_m4[6,1])); h2_mod4
h2_mod5 <- 1-(mean(blup5$sed[upper.tri(blup5$sed)]^2)/(2*varcomp_m5[6,1])); h2_mod5
h2_mod6 <- 1-(mean(blup6$sed[upper.tri(blup6$sed)]^2)/(2*varcomp_m6[3,1])); h2_mod6

h2 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6"),
  H2 = c(h2_mod1, h2_mod2, h2_mod3, h2_mod4, h2_mod5, h2_mod6)
)

options(knitr.table.format = "html")

h2 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Trial 2 - OT1RWE21

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 3 - OT1RWE22

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 4 - OT1SER22

variance components, broad-sense heritabilities, and rankings

## Best model

Encaixar com o da Bruna no começo

```{r setup-data}
knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold"
)

library(asreml)
library(lmmtools)
library(tidyverse)
library(knitr)     

data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data <- data |>
  mutate(
    env = as.factor(env),
    row_number = as.factor(row_number),
    col_number = as.factor(col_number),
    rowgroup = as.factor(rowgroup),
    colgroup = as.factor(colgroup),
    check = as.factor(check),
    geno = as.factor(geno),
    rn = as.numeric(row_number),
    cn = as.numeric(col_number)
  )
```
# Rwanda season 2021 (Trial 2 - OT1RWE21)

## Modelling

```{r}
#| label: data-t2

t2 <- data |> filter (env == "OT1RWE21")

# Modifying the geno.cod and check.cod columns
t2 <- t2 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn)
  )
```

In order to test the spatial correction, in this season and country was analyzed six models. The most basic model is the same described in [Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325). 

```{r}
#| label: models-t2
#| output: false

# Piepho & Williams (2016) model as a starter

model_1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t2, maxinter=50)


# Correcting col and row number residuals - autoregression

model_2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

# Adding units to random effects (nuggets) + variant autoregression

model_3 <-  asreml(fixed = rytha ~  check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1(row_number):ar1v(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

model_4 <-  asreml(fixed = rytha ~  check.cod + rowgroup + colgroup,
             random = ~  rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1(row_number):ar1v(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

```

### Wald test

In every model, the genetic variance was tested with the LRT test and it was significant. This result indicates that at least two genotypes are statistically different from each other. The same tendency is observed in the checks, that were tested using the Wald test.

:::{.panel-tabset}

### Model 1 
```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1)
```

### Model 2
```{r}
wald.asreml(model_2)
```

### Model 3

```{r}
wald.asreml(model_3)
```

### Model 4

```{r}
wald.asreml(model_4)
```

### Model 5

```{r}
#wald.asreml(model_5)
```

### Model 6

```{r}
#wald.asreml(model_6)
```
:::

### LRT

```{r}
#| output: false
#| warning: false
#Random effects

red_model1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t2, maxinter=50)

red_model2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

red_model3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + check.cod,
             random = ~ rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

red_model4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

#red_model5 <-  asreml(fixed = rytha ~ check.cod,
             #random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             #residual = ~ ar1v(rn):ar1(cn),
             #na.action = na.method("include"),
             #data = t2, maxint=50)

#red_model6 <-  asreml(fixed = rytha ~ check.cod,
             #random = ~ rowgroup:rn + colgroup:cn + units,
             #residual = ~ ar1v(rn):ar1(cn),
             #na.action = na.method("include"),
             #data = t2, maxint=50)

lrt(model_1, red_model1)
lrt(model_2, red_model2)
lrt(model_3, red_model3)
lrt(model_4, red_model4)
#lrt(model_5, red_model5)
#lrt(model_6, red_model6)
```

::: {.panel-tabset}

### Model 1

```{r}
lrt(model_1, red_model1)

```

### Model 2

```{r}
lrt(model_2, red_model2)

```

### Model 3

```{r}
lrt(model_3, red_model3)

```

### Model 4

```{r}
lrt(model_4, red_model4)

```

### Model 5

```{r}
#lrt(model_5, red_model5)

```

### Model 6

```{r}
#lrt(model_6, red_model6)
```

:::

### Semivariograms

Evaluating the semivariograms and the heatmap of root yield, a clear spatial trend can be observed around the center of the superblock. In both visualizations, the central portion of the block exhibits a distinct pattern characterized by lower yields, which corresponds to an almost flat trend in the semivariogram. Even though spatial corrections were applied, this behavior persists. This may indicate a major source of micro-environmental heterogeneity in the field, potentially corresponding to the centerline of the irrigation system or, conversely, a specific zone that did not receive adequate water from the sprinkler irrigation.

```{r}
#| label: fig-variograms-t1
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (AR1 × AR1 Residual Structure)"
#|   - "Semivariogram for Model 3 (Fixed Row/Column Effects + Nugget Effect)"
#|   - "Semivariogram for Model 4 (Linear Trend + Spatial Adjustment)"
#|   - "Semivariogram for Model 5 (Adding units to random effects (nuggets))"
#|   - "Semivariogram for Model 6 (Reduction of model complexity)"
#|   - "Heatmap of the distribution the root yield in tonnes per ha (rytha) in trial 2 - Rwanda season 2021"


plot(varioGram(model_1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

#plot(varioGram(model_5, type = "default", spatial = "trend",
#formula = ~NULL, composite = TRUE, model = c("empirical"),
#metric = c("euclidean"), angle = 0, angle.tol = 180,
#nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

#plot(varioGram(model_6, type = "default", spatial = "trend",
#formula = ~NULL, composite = TRUE, model = c("empirical"),
#metric = c("euclidean"), angle = 0, angle.tol = 180,
#nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

ggplot(data = t2, aes(x = col_number, y = row_number, fill = rytha)) + 
  geom_tile() + 
  theme_minimal() + 
  theme(text = element_text(size = 12)) + 
  labs(x = "Column", y = "Row") +
  scale_fill_viridis_c(option = "magma")
```


### AIC 

This analysis presented results similar to those obtained from the semivariograms. The baseline model (Model 1) exhibited the highest Akaike Information Criterion (AIC) value (8,723.64), indicating the poorest fit. Conversely, Model 6, which incorporated the units term into the random effects to account for the nugget effect, was the best-performing model, yielding an AIC value of 7,982.90—a substantial reduction of 740.74 units.

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel = icREML(list(model_1, model_2, model_3, model_4)) #, model_5, model_6)); modsel
```

:::{.panel-tabset}

### Model 1

```{r}
icREML(list(model_1))
```

### Model 2

```{r}
icREML(list(model_2))
```

### Model 3

```{r}
icREML(list(model_3))
```

### Model 4

```{r}
icREML(list(model_4))
```

### Model 5

```{r}
icREML(list(model_5))
```

### Model 6

```{r}
icREML(list(model_6))
```

:::

### Variance components

:::{.panel-tabset}

### Model 1
```{r}
# Variance components

varcomp_m1 <- summary(model_1)$varcomp
varcomp_m1 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2 

```{r}
varcomp_m2 <- summary(model_2)$varcomp
varcomp_m2|> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3 <- summary(model_3)$varcomp
varcomp_m3|> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4 <- summary(model_4)$varcomp
varcomp_m4|> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5 <- summary(model_5)$varcomp 
varcomp_m5|> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 6

```{r}
varcomp_m6 <- summary(model_6)$varcomp
varcomp_m6|> 
  kbl(caption = "Variance components for Model 6.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

:::

### Ranking 

```{r}
#| label: t1-blups

#BLUPs
blup1 <- predict.asreml(model_1, classify = "geno.cod", sed = TRUE)
blup2 <- predict.asreml(model_2, classify = "geno.cod", sed = TRUE)
blup3 <- predict.asreml(model_3, classify = "geno.cod", sed = TRUE)
blup4 <- predict.asreml(model_4, classify = "geno.cod", sed = TRUE)
blup5 <- predict.asreml(model_5, classify = "geno.cod", sed = TRUE)
blup6 <- predict.asreml(model_6, classify = "geno.cod", sed = TRUE)

blup <- list(blup1, blup2, blup3, blup4, blup5, blup6)
predic_value_1 <- list()

for (i in 1:length(blup)) {
  
  predic_value_1[[i]] <- blup[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1 <- predict(model_1, classify = "check.cod")$pvals
blues_checks2 <- predict(model_2, classify = "check.cod")$pvals
blues_checks3 <- predict(model_3, classify = "check.cod")$pvals
blues_checks4 <- predict(model_4, classify = "check.cod")$pvals
blues_checks5 <- predict(model_5, classify = "check.cod")$pvals
blues_checks6 <- predict(model_6, classify = "check.cod")$pvals

blue <- list(blues_checks1, blues_checks2, blues_checks3, blues_checks4, blues_checks5, blues_checks6)
bpredic_value_1 <- list()

for (i in 1:length(blue)) {
  
  bpredic_value_1[[i]] <- blue[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

## Graph
aux <- rbind(
  blup1$pvals |>
    mutate(
      model = 'model 1'
    ),
  blup5$pvals |> mutate(
    model = 'model 5'
  )
)

ggplot(data = aux, aes(x = model, y = predicted.value)) +
  geom_point(size = 3, alpha = .5, color = "green") +
  geom_line(aes(group = geno.cod), linewidth = 1.3, alpha = .4, color = "#10342d") +
  theme_minimal() + 
  labs(x = "Model", y = "Yield (BLUP)") + 
  theme(text = element_text(size = 20))
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t1
#| tbl-cap: Broad-sense Heritability of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1 <- 1-(mean(blup1$sed[upper.tri(blup1$sed)]^2)/(2*varcomp_m1[6,1])); h2_mod1
h2_mod2 <- 1-(mean(blup2$sed[upper.tri(blup2$sed)]^2)/(2*varcomp_m2[6,1])); h2_mod2
h2_mod3 <- 1-(mean(blup3$sed[upper.tri(blup3$sed)]^2)/(2*varcomp_m3[4,1])); h2_mod3
h2_mod4 <- 1-(mean(blup4$sed[upper.tri(blup4$sed)]^2)/(2*varcomp_m4[6,1])); h2_mod4
h2_mod5 <- 1-(mean(blup5$sed[upper.tri(blup5$sed)]^2)/(2*varcomp_m5[6,1])); h2_mod5
h2_mod6 <- 1-(mean(blup6$sed[upper.tri(blup6$sed)]^2)/(2*varcomp_m6[3,1])); h2_mod6

h2 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6"),
  H2 = c(h2_mod1, h2_mod2, h2_mod3, h2_mod4, h2_mod5, h2_mod6)
)

options(knitr.table.format = "html")

h2 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```







## Trial 5: OT2NAM21

### 1. Data Preparation
```{r t5-prep}
t5 <- data |>
  filter(env == "OT2NAM21") |>
  arrange(row_number, col_number) |>
  mutate(
    # Reclassifica Parentais Elite como Testemunhas (Efeito Fixo)
    check = ifelse(status == "Parent", "Check", as.character(check))
  ) |>
  group_by(check.cod) |>
  mutate(
    is_all_na = (check == "Check" & all(is.na(rytha))),
    check = ifelse(is_all_na, "Candidate", as.character(check)),
    check = as.factor(check)
  ) |>
  ungroup() |>
  select(-is_all_na) |>
  mutate(
    row_num_cov = as.numeric(as.character(row_number)),
    col_num_cov = as.numeric(as.character(col_number))
  )
```

### 2. Model Fit Progression

::: {.panel-tabset}

## Model 1
```{r}
#| code-fold: true
# M1: Baseline — Structural blocks of the augmented design
M1_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno,
  residual = ~ idv(units),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 1.
```{r}
#| echo: false
knitr::kable(summary(M1_t5)$varcomp, digits = 4)
```

## Model 2
```{r}
#| code-fold: true
# M2: AR1xAR1 — Two-dimensional spatial correlation
M2_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 2.
```{r}
#| echo: false
knitr::kable(summary(M2_t5)$varcomp, digits = 4)
```

## Model 3
```{r}
#| code-fold: true
# M3: AR1xAR1 + Nugget — Addition of independent random error
M3_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 3.
```{r}
#| echo: false
knitr::kable(summary(M3_t5)$varcomp, digits = 4)
```

## Model 4
```{r}
#| code-fold: true
# M4: M3 + Nested row/column random effects (Piepho & Williams, 2016)
M4_t5 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 4.
```{r}
#| echo: false
knitr::kable(summary(M4_t5)$varcomp, digits = 4)
```

## Model 5
```{r}
#| code-fold: true
# M5: M4 + Global fixed linear trend
M5_t5 <- asreml(
  fixed = rytha ~ at(check, "Check") + lin(row_num_cov) + lin(col_num_cov),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 5.
```{r}
#| echo: false
knitr::kable(summary(M5_t5)$varcomp, digits = 4)
```

## Model 6
```{r}
#| code-fold: true
# M6: M5 + Splines — Smooth non-linear trend
M6_t5 <- asreml(
  fixed = rytha ~ at(check, "Check") + lin(row_num_cov) + lin(col_num_cov),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              spl(row_num_cov) + spl(col_num_cov) +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t5, workspace = 128e6, maxit = 50
)
```
Variance components for Model 6.
```{r}
#| echo: false
knitr::kable(summary(M6_t5)$varcomp, digits = 4)
```

:::

### 3. Model Selection
```{r t5-selection}
lista_modelos_t5 <- list(
  Mod1_Baseline    = M1_t5,
  Mod2_AR1         = M2_t5,
  Mod3_Nugget      = M3_t5,
  Mod4_RowColGroup = M4_t5,
  Mod5_LinearTrend = M5_t5,
  Mod6_Splines     = M6_t5
)

aic_table_t5 <- lmmtools::icREML(lista_modelos_t5) |> arrange(AIC)
knitr::kable(aic_table_t5, digits = 3)

best_model_name_t5 <- aic_table_t5$model[1]
best_model_t5 <- lista_modelos_t5[[best_model_name_t5]]
```

### 4. Semivariograms
```{r t5-semivario, fig.width=10, fig.height=6}
par(mfrow = c(2, 3))
plot(varioGram(M1_t5), main = "Mod1: Baseline")
plot(varioGram(M2_t5), main = "Mod2: AR1xAR1")
plot(varioGram(M3_t5), main = "Mod3: AR1xAR1 + Nugget")
plot(varioGram(M4_t5), main = "Mod4: + RowColGroup")
plot(varioGram(M5_t5), main = "Mod5: + Linear Trend")
plot(varioGram(M6_t5), main = "Mod6: + Splines")
par(mfrow = c(1, 1))
```

### 5. Heritability & Ranking (Regular Progenies Only)
```{r t5-h2-ranking}
calc_cullis_regular <- function(model, data_trial, geno_var_pattern = "Candidate.*geno") {
  vc_temp <- summary(model)$varcomp
  gen_var_name <- grep(geno_var_pattern, rownames(vc_temp), value = TRUE)[1]
  if (is.na(gen_var_name)) {
    stop("error")
  }
  sigma2_g <- vc_temp[gen_var_name, "component"]

  pred <- predict(model, classify = "geno", sed = TRUE)

  regular_genos <- unique(as.character(data_trial$geno[data_trial$status == "Regular"]))
  keep_idx <- as.character(pred$pvals$geno) %in% regular_genos & pred$pvals$status != "Aliased"

  pvals_r <- pred$pvals[keep_idx, ]
  sed_r   <- pred$sed[keep_idx, keep_idx]

  vdBLUP_avg <- mean(sed_r[upper.tri(sed_r, diag = FALSE)]^2, na.rm = TRUE)
  H2 <- 1 - (vdBLUP_avg / (2 * sigma2_g))

  list(H2 = H2, sigma2_g = sigma2_g, pvals = pvals_r)
}

h2_all_t5 <- lapply(lista_modelos_t5, calc_cullis_regular, data_trial = t5)

pred1_pvals_t5     <- h2_all_t5[["Mod1_Baseline"]]$pvals
pred_best_pvals_t5 <- h2_all_t5[[best_model_name_t5]]$pvals

rank_compare_t5 <- pred1_pvals_t5 |>
  rename(blup_mod1 = predicted.value) |>
  select(geno, blup_mod1) |>
  inner_join(
    pred_best_pvals_t5 |> rename(blup_best = predicted.value) |> select(geno, blup_best),
    by = "geno"
  )

rho_t5 <- cor(rank_compare_t5$blup_mod1, rank_compare_t5$blup_best, method = "spearman", use = "complete.obs")

top15_t5 <- pred_best_pvals_t5 |>
  arrange(desc(predicted.value)) |>
  slice_head(n = 15) |>
  select(geno, predicted.value, std.error)
```

**Spearman rank correlation (Mod1 vs Best Model):** `r round(rho_t5, 4)`  
**Broad-sense Heritability (Cullis) for Best Model:** `r round(h2_all_t5[[best_model_name_t5]]$H2, 3)`

#### Top 15 Candidates (Regulars Only)
```{r t5-top15, echo=FALSE}
knitr::kable(top15_t5, digits = 3)
```

### Discussion: OT2NAM21
The analysis of the OT2NAM21 trial showed the presence of continuous environmental gradients, with the global linear trend model (Mod5_LinearTrend) presenting the best fit according to the penalized icREML criterion (5471.75). The inclusion of two-dimensional row and column covariates captured the spatial variation, making the additional complexity of splines unnecessary. Furthermore, the variance components of the structural blocks of the augmented design collapsed to zero, indicating that the directional linear trends explained most of the field's heterogeneity.

To ensure the accuracy of the biometric parameters, 39 elite parents were reclassified as fixed-effect checks, restricting the genetic random effect exclusively to the 1,138 regular segregating progenies. This filtering prevented the artificial inflation of the genetic variance and resulted in a highly reliable broad-sense heritability estimate (Cullis et al., 2006) of 0.877 in the optimized model. This value reflects a strong genetic signal after the spatial correction of the environmental noise.

The Spearman rank correlation between the BLUPs of the baseline model and the best spatial model was exceptionally high ($\rho = 0.9495$), demonstrating that the original trial data already possessed good ranking quality, which was shaped by the spatial modeling. The final ranking, cleared of the influence of the reference cultivars, highlighted the promising lines of the program. Genotype UGP202000272-5 isolated itself at the top of the selection with a quite precise estimate (BLUP = 16.27), confirming the robustness and efficiency of the trial for early-stage clonal selection.


---

## Trial 6: OT2SER21

### 1. Data Preparation
```{r t6-prep}
t6 <- data |>
  filter(env == "OT2SER21") |>
  arrange(row_number, col_number) |>
  mutate(
    # Reclassifica Parentais Elite como Testemunhas (Efeito Fixo)
    check = ifelse(status == "Parent", "Check", as.character(check))
  ) |>
  group_by(check.cod) |>
  mutate(
    is_all_na = (check == "Check" & all(is.na(rytha))),
    check = ifelse(is_all_na, "Candidate", as.character(check)),
    check = as.factor(check)
  ) |>
  ungroup() |>
  select(-is_all_na) |>
  mutate(
    row_num_cov = as.numeric(as.character(row_number)),
    col_num_cov = as.numeric(as.character(col_number))
  )
```

### 2. Model Fit Progression

::: {.panel-tabset}

## Model 1
```{r}
#| code-fold: true
# M1: Baseline — Structural blocks of the augmented design
M1_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno,
  residual = ~ idv(units),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 1.
```{r}
#| echo: false
knitr::kable(summary(M1_t6)$varcomp, digits = 4)
```

## Model 2
```{r}
#| code-fold: true
# M2: AR1xAR1 — Two-dimensional spatial correlation
M2_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 2.
```{r}
#| echo: false
knitr::kable(summary(M2_t6)$varcomp, digits = 4)
```

## Model 3
```{r}
#| code-fold: true
# M3: AR1xAR1 + Nugget — Addition of independent random error
M3_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup + at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 3.
```{r}
#| echo: false
knitr::kable(summary(M3_t6)$varcomp, digits = 4)
```

## Model 4
```{r}
#| code-fold: true
# M4: M3 + Nested row/column random effects
M4_t6 <- asreml(
  fixed = rytha ~ at(check, "Check"),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 4.
```{r}
#| echo: false
knitr::kable(summary(M4_t6)$varcomp, digits = 4)
```

## Model 5
```{r}
#| code-fold: true
# M5: M4 + Global fixed linear trend
M5_t6 <- asreml(
  fixed = rytha ~ at(check, "Check") + lin(row_num_cov) + lin(col_num_cov),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 5.
```{r}
#| echo: false
knitr::kable(summary(M5_t6)$varcomp, digits = 4)
```

## Model 6
```{r}
#| code-fold: true
# M6: M5 + Splines — Smooth non-linear trend
M6_t6 <- asreml(
  fixed = rytha ~ at(check, "Check") + lin(row_num_cov) + lin(col_num_cov),
  random = ~ rowgroup + colgroup + rowgroup:colgroup +
              rowgroup:row_number + colgroup:col_number +
              spl(row_num_cov) + spl(col_num_cov) +
              at(check, "Candidate"):geno + units,
  residual = ~ ar1(row_number):ar1(col_number),
  na.action = na.method("include"), data = t6, workspace = 128e6, maxit = 50
)
```
Variance components for Model 6.
```{r}
#| echo: false
knitr::kable(summary(M6_t6)$varcomp, digits = 4)
```

:::

### 3. Model Selection
```{r t6-selection}
lista_modelos_t6 <- list(
  Mod1_Baseline    = M1_t6,
  Mod2_AR1         = M2_t6,
  Mod3_Nugget      = M3_t6,
  Mod4_RowColGroup = M4_t6,
  Mod5_LinearTrend = M5_t6,
  Mod6_Splines     = M6_t6
)

aic_table_t6 <- lmmtools::icREML(lista_modelos_t6) |> arrange(AIC)
knitr::kable(aic_table_t6, digits = 3)

best_model_name_t6 <- aic_table_t6$model[1]
best_model_t6 <- lista_modelos_t6[[best_model_name_t6]]
```

### 4. Semivariograms
```{r t6-semivario, fig.width=10, fig.height=6}
par(mfrow = c(2, 3))
plot(varioGram(M1_t6), main = "Mod1: Baseline")
plot(varioGram(M2_t6), main = "Mod2: AR1xAR1")
plot(varioGram(M3_t6), main = "Mod3: AR1xAR1 + Nugget")
plot(varioGram(M4_t6), main = "Mod4: + RowColGroup")
plot(varioGram(M5_t6), main = "Mod5: + Linear Trend")
plot(varioGram(M6_t6), main = "Mod6: + Splines")
par(mfrow = c(1, 1))
```

### 5. Heritability & Ranking (Regular Progenies Only)
```{r t6-h2-ranking}
h2_all_t6 <- lapply(lista_modelos_t6, calc_cullis_regular, data_trial = t6)

pred1_pvals_t6     <- h2_all_t6[["Mod1_Baseline"]]$pvals
pred_best_pvals_t6 <- h2_all_t6[[best_model_name_t6]]$pvals

rank_compare_t6 <- pred1_pvals_t6 |>
  rename(blup_mod1 = predicted.value) |>
  select(geno, blup_mod1) |>
  inner_join(
    pred_best_pvals_t6 |> rename(blup_best = predicted.value) |> select(geno, blup_best),
    by = "geno"
  )

rho_t6 <- cor(rank_compare_t6$blup_mod1, rank_compare_t6$blup_best, method = "spearman", use = "complete.obs")

top15_t6 <- pred_best_pvals_t6 |>
  arrange(desc(predicted.value)) |>
  slice_head(n = 15) |>
  select(geno, predicted.value, std.error)
```

**Spearman rank correlation (Mod1 vs Best Model):** `r round(rho_t6, 4)`  
**Broad-sense Heritability (Cullis) for Best Model:** `r round(h2_all_t6[[best_model_name_t6]]$H2, 3)`

#### Top 15 Candidates (Regulars Only)
```{r t6-top15, echo=FALSE}
knitr::kable(top15_t6, digits = 3)
```

### Discussion: OT2SER21
The analysis of the OT2SER21 trial shows a spatial architecture based on local patches, with the autoregressive model containing independent error (Mod3_Nugget) presenting the best fit according to the penalized icREML criterion (3611.23). The high independent error variance (units = 7.14) combined with the two-dimensional AR1 structure successfully captured the spatial variation, rendering the models with linear trends and splines penalized for unnecessary complexity. This indicates that field heterogeneity was driven by unpredictable, localized micro-variations rather than continuous directional gradients.

To ensure the accuracy of the biometric parameters, the 39 elite parents were reclassified as fixed-effect checks, restricting the genetic random effect exclusively to the 1,138 regular segregating progenies. This filtering prevented the artificial inflation of the genetic variance and resulted in a reliable broad-sense heritability estimate (Cullis et al., 2006) of 0.786 in the optimized model. This value reflects that, even in the presence of strong localized environmental noise, the spatial correction successfully isolated a robust genetic signal.

The Spearman rank correlation between the BLUPs of the baseline model and the best spatial model was high ($\rho = 0.9414$), demonstrating that the spatial adjustment consistently refined the original trial ranking. The final ranking, cleared of the influence of the reference cultivars, highlighted the promising lines of the program. Genotype UGP202000158-7 took the lead in the selection (BLUP = 5.56), closely followed by UGP202000217-17, confirming the efficiency of the trial for early-stage clonal selection.


