---
title: "Spatial analysis of augmented designs in sweet potato"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

The dataset available here has six trials laid out in an augmented
row-column design (more details about this design here). Analyse each
trial individually with increasing order of spatial adjustment. Report
the improvement of the modelling (if any) when spatial trends are dealt
with. Your report should have differences in the variance components,
broad-sense heritabilities, and rankings. For each trial, determine the
best model using semivariograms and the AIC. Don't forget that models
with different fixed effects must be compared using the adjusted version of the AIC available on the lmmtools package.

# Introduction

This report provides a data analysis of a spatial correction of a
multi-environment sweet potato experiment in Africa. Each one of the six environments were analyzed with multiple models testing different
approaches to spacial correction. It was estimated variance components,
broad-sense heritabilities, and rankings based on BLUP predictions.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data
#| warning: false

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml", "lmmtools")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data 
data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data$row_number <- as.factor(data$row_number)
data$col_number <- as.factor(data$col_number)
data$rowgroup <- as.factor(data$rowgroup)
data$colgroup <- as.factor(data$colgroup)
data$check <- as.factor(data$check)
data$geno <- as.factor(data$geno)
data$check.cod = as.factor(data$check.cod)
data$geno.cod  = as.factor(data$geno.cod)
```

The trials were conducted across Uganda in three locations (Rwebitaba - RWE, Namulonge - NAM and Serere - SER) during the 2021 and 2022 growing seasons. To account for spatial variation, each trial was analyzed individually to implement single-trial spatial corrections.

```{r}
# Trial identification
unique(data$env)
```

This dataset is derived from a multi-environment sweet potato trial. The experimental layout follows an augmented row-column design, where each environment comprises a grid of 50 rows and 30 columns. These rows are grouped into 5 super-rows, and the columns into 3 super-columns. Each resulting combination (hereafter referred to as a miniblock) consists of 10 rows and 10 columns, encompassing a total of 8 checks and 1,178 test genotypes.

The model used in this report was the same described in the article of
[Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325),
which can be represented as the model below:

$$ y = \mu + rowgroup + colgroup + rowgroup.colgroup + rowgroup.row + colgroup.col + genotype + check + \epsilon$$

Where, $\mu$ is the intercept,

$rowgroup$ is the random effect of the super-row,

$colgroup$ is the random effect of the super-column,

$rowgroup.colgroup$ is the random effect of the intersection of
super-row and super-column (the miniblock),

$rowgroup.row$ the random effect of the row number nested in the
super-row,

$colgroup.col$ the random effect of the column number nested in the
super-column,

$genotype$ the random effect of the genotype,

$check$ the fixed effect of the check and

$\epsilon$ the random effect of the error.

# Rwebitaba Season 2022/1 (Trial 3 - OT1RWE22)

```{r}
#| label: fig-t3-prod

t3 <- data |> filter (env == "OT1RWE21")

# Modifying the geno.cod and check.cod columns
t3 <- t3 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
t3 |> 
  group_by(check.cod) |>
  summarise(n_total = n(),
            n_na = sum(is.na(rytha)),
            n_obs = sum(!is.na(rytha))) |>
  arrange(n_obs)

t3 <- t3 |> mutate(
  check.cod = as.character(check.cod),
  check.cod = ifelse(check.cod == "RAK808", "Candidato", check.cod),
  check.cod = as.factor(check.cod))

unique(t3$row_number)
unique(t3$col_number)
unique(t3$rowgroup)
unique(t3$colgroup)
unique(t3$check.cod)
length(unique(t3$geno.cod))
unique(t3$rc.group)

```

In order to test the spatial correction, in this season and country was
analyzed six models. The most basic model is the same described in
[Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325).

```{r}
#| label: models-t3
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t3 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t3, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 3 - AR1 in residuals
model_3_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t3 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

```

### Variance components

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t3 <- summary(model_1_t3)$varcomp
varcomp_m1_t3 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t3 <- summary(model_2_t3)$varcomp
varcomp_m2_t3 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t3 <- summary(model_3_t3)$varcomp
varcomp_m3_t3 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t3 <- summary(model_4_t3)$varcomp
varcomp_m4_t3 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t3 <- summary(model_5_t3)$varcomp
varcomp_m5_t3 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t3)
```

### Model 2

```{r}
wald.asreml(model_2_t3)
```

### Model 3

```{r}
wald.asreml(model_3_t3)
```

### Model 4

```{r}
wald.asreml(model_4_t3)
```

### Model 5

```{r}
wald.asreml(model_5_t3)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t3 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t3, maxinter=50)

red_model2_t3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model3_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model4_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model5_t3 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t3, red_model1_t3)

```

### Model 2

```{r}
lrt(model_2_t3, red_model2_t3)

```

### Model 3

```{r}
lrt(model_3_t3, red_model3_t3)
```

### Model 4

```{r}
lrt(model_4_t3, red_model4_t3)

```

### Model 5

```{r}
lrt(model_5_t3, red_model5_t3)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t3
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t3 = icREML(list(model_1_t3, model_2_t3, model_3_t3, model_4_t3, model_5_t3)); modsel_t3
modsel_t3 <- as.data.frame(modsel_t3)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t3))
```

### Model 2

```{r}
icREML(list(model_2_t3))
```

### Model 3

```{r}
icREML(list(model_3_t3))
```

### Model 4

```{r}
icREML(list(model_4_t3))
```

### Model 5

```{r}
icREML(list(model_5_t3))
```
:::

### Spatial Distribution of Residuals

```{r}
resid_long <- t3 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t3, type = "response"),
    Model_2 = residuals(model_2_t3, type = "response"),
    Model_3 = residuals(model_3_t3, type = "response"),
    Model_4 = residuals(model_4_t3, type = "response"),
    Model_5 = residuals(model_5_t3, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Ranking

```{r}
#| label: t3-blups

#BLUPs
blup1_t3 <- predict.asreml(model_1_t3, classify = "geno.cod", sed = TRUE)
blup2_t3 <- predict.asreml(model_2_t3, classify = "geno.cod", sed = TRUE)
blup3_t3 <- predict.asreml(model_3_t3, classify = "geno.cod", sed = TRUE)
blup4_t3 <- predict.asreml(model_4_t3, classify = "geno.cod", sed = TRUE)
blup5_t3 <- predict.asreml(model_5_t3, classify = "geno.cod", sed = TRUE)

blup_t3 <- list(blup1_t3, blup2_t3, blup3_t3, blup4_t3, blup5_t3)
predic_value_t3 <- list()

for (i in 1:length(blup_t3)) {
  
  predic_value_t3[[i]] <- blup_t3[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t3 <- predict(model_1_t3, classify = "check.cod")$pvals
blues_checks2_t3 <- predict(model_2_t3, classify = "check.cod")$pvals
blues_checks3_t3 <- predict(model_3_t3, classify = "check.cod")$pvals
blues_checks4_t3 <- predict(model_4_t3, classify = "check.cod")$pvals
blues_checks5_t3 <- predict(model_5_t3, classify = "check.cod")$pvals

blue_t3 <- list(blues_checks1_t3, blues_checks2_t3, blues_checks3_t3, blues_checks4_t3, blues_checks5_t3)
bpredic_value_t3 <- list()

for (i in 1:length(blue_t3)) {
  
  bpredic_value_t3[[i]] <- blue_t3[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

### Broad-sense Heritability

```{r}
#| eval: FALSE
#| label: tbl-heritability-t3
#| tbl-cap: Broad-sense Heritability of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t3 <- 1-(mean(blup1_t3$sed[upper.tri(blup1_t3$sed)]^2)/(2*varcomp_m1_t3[6,1])); h2_mod1_t3
h2_mod2_t3 <- 1-(mean(blup2_t3$sed[upper.tri(blup2_t3$sed)]^2)/(2*varcomp_m2_t3[6,1])); h2_mod2_t3
h2_mod3_t3 <- 1-(mean(blup3_t3$sed[upper.tri(blup3_t3$sed)]^2)/(2*varcomp_m3_t3[4,1])); h2_mod3_t3
h2_mod4_t3 <- 1-(mean(blup4_t3$sed[upper.tri(blup4_t3$sed)]^2)/(2*varcomp_m4_t3[6,1])); h2_mod4_t3
h2_mod5_t3 <- 1-(mean(blup5_t3$sed[upper.tri(blup5_t3$sed)]^2)/(2*varcomp_m5[6,1])); h2_mod5_t3

h2_t3 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t3, h2_mod2_t3, h2_mod3_t3, h2_mod4_t3, h2_mod5_t3)
)

options(knitr.table.format = "html")

h2_t3 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

## Best model

# Serere Season 2022/1 (Trial 4 - OT1SER22)

```{r}
#| label: fig-t4-prod

t4 <- data |> filter (env == "OT1SER22")

# Modifying the geno.cod and check.cod columns
t4 <- t4 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
t4 |> 
  group_by(check.cod) |>
  summarise(n_total = n(),
            n_na = sum(is.na(rytha)),
            n_obs = sum(!is.na(rytha))) |>
  arrange(n_obs)

t4 <- t4 |> mutate(
  check.cod = as.character(check.cod),
  check.cod = ifelse(check.cod == "RAK808", "Candidato", check.cod),
  check.cod = as.factor(check.cod))
```

In order to test the spatial correction, in this season and country was
analyzed six models. The most basic model is the same described in
[Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325).

```{r}
#| label: models-t4
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t4 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t4, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t4 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 3 - AR1 in residuals
model_3_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

```

### Variance components

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t4 <- summary(model_1_t4)$varcomp
varcomp_m1_t4 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t4 <- summary(model_2_t4)$varcomp
varcomp_m2_t4 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t4 <- summary(model_3_t4)$varcomp
varcomp_m3_t4 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t4 <- summary(model_4_t4)$varcomp
varcomp_m4_t4 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t4 <- summary(model_5_t4)$varcomp
varcomp_m5_t4 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t4)
```

### Model 2

```{r}
wald.asreml(model_2_t4)
```

### Model 3

```{r}
wald.asreml(model_3_t4)
```

### Model 4

```{r}
wald.asreml(model_4_t4)
```

### Model 5

```{r}
wald.asreml(model_5_t4)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t4 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t4, maxinter=50)

red_model2_t4 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model3_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model4_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model5_t4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t4, red_model1_t4)

```

### Model 2

```{r}
lrt(model_2_t4, red_model2_t4)

```

### Model 3

```{r}
lrt(model_3_t4, red_model3_t4)
```

### Model 4

```{r}
lrt(model_4_t4, red_model4_t4)

```

### Model 5

```{r}
lrt(model_5_t4, red_model5_t4)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t4
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t4 = icREML(list(model_1_t4, model_2_t4, model_3_t4, model_4_t4, model_5_t4)); modsel_t4
modsel_t4 <- as.data.frame(modsel_t4)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t4))
```

### Model 2

```{r}
icREML(list(model_2_t4))
```

### Model 3

```{r}
icREML(list(model_3_t4))
```

### Model 4

```{r}
icREML(list(model_4_t4))
```

### Model 5

```{r}
icREML(list(model_5_t4))
```
:::

### Spatial Distribution of Residuals

```{r}
resid_long <- t4 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t4, type = "response"),
    Model_2 = residuals(model_2_t4, type = "response"),
    Model_3 = residuals(model_3_t4, type = "response"),
    Model_4 = residuals(model_4_t4, type = "response"),
    Model_5 = residuals(model_5_t4, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Ranking

```{r}
#| label: t4-blups

#BLUPs
blup1_t4 <- predict.asreml(model_1_t4, classify = "geno.cod", sed = TRUE)
blup2_t4 <- predict.asreml(model_2_t4, classify = "geno.cod", sed = TRUE)
blup3_t4 <- predict.asreml(model_3_t4, classify = "geno.cod", sed = TRUE)
blup4_t4 <- predict.asreml(model_4_t4, classify = "geno.cod", sed = TRUE)
blup5_t4 <- predict.asreml(model_5_t4, classify = "geno.cod", sed = TRUE)

blup_t4 <- list(blup1_t4, blup2_t4, blup3_t4, blup4_t4, blup5_t4)
predic_value_t4 <- list()

for (i in 1:length(blup_t4)) {
  
  predic_value_t4[[i]] <- blup_t4[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t4 <- predict(model_1_t4, classify = "check.cod")$pvals
blues_checks2_t4 <- predict(model_2_t4, classify = "check.cod")$pvals
blues_checks3_t4 <- predict(model_3_t4, classify = "check.cod")$pvals
blues_checks4_t4 <- predict(model_4_t4, classify = "check.cod")$pvals
blues_checks5_t4 <- predict(model_5_t4, classify = "check.cod")$pvals

blue_t4 <- list(blues_checks1_t4, blues_checks2_t4, blues_checks3_t4, blues_checks4_t4, blues_checks5_t4)
bpredic_value_t4 <- list()

for (i in 1:length(blue_t4)) {
  
  bpredic_value_t4[[i]] <- blue_t4[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

### Broad-sense Heritability

```{r}
#| eval: FALSE
#| label: tbl-heritability-t4
#| tbl-cap: Broad-sense Heritability of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t4 <- 1-(mean(blup1_t4$sed[upper.tri(blup1_t4$sed)]^2)/(2*varcomp_m1_t4[6,1])); h2_mod1_t4
h2_mod2_t4 <- 1-(mean(blup2_t4$sed[upper.tri(blup2_t4$sed)]^2)/(2*varcomp_m2_t4[6,1])); h2_mod2_t4
h2_mod3_t4 <- 1-(mean(blup3_t4$sed[upper.tri(blup3_t4$sed)]^2)/(2*varcomp_m3_t4[4,1])); h2_mod3_t4
h2_mod4_t4 <- 1-(mean(blup4_t4$sed[upper.tri(blup4_t4$sed)]^2)/(2*varcomp_m4_t4[6,1])); h2_mod4_t4
h2_mod5_t4 <- 1-(mean(blup5_t4$sed[upper.tri(blup5_t4$sed)]^2)/(2*varcomp_m5[6,1])); h2_mod5_t4

h2_t4 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t4, h2_mod2_t4, h2_mod3_t4, h2_mod4_t4, h2_mod5_t4)
)

options(knitr.table.format = "html")

h2_t4 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

## Best model
