---
title: "Spatial analysis of augmented designs in sweet potato"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

The dataset available here has six trials laid out in an augmented row-column design (more details about this design here). Analyse each trial individually with increasing order of spatial adjustment. Report the improvement of the modelling (if any) when spatial trends are dealt with. Your report should have differences in the variance components, broad-sense heritabilities, and rankings. For each trial, determine the best model using semivariograms and the AIC. Don't forget that models with different fixed effects must be compared using the adjusted version of the AIC available on the lmmtools package.

# Introduction (REFAZER!! PEGUEI DO RELATÓRIO ANTERIOR)

This report provides a data analysis of a multi-environment soybean
breeding in Zambia dataset, focusing on two agronomic traits **Grain
Yield** and **Plant Height**. Utilizing a Randomized Complete Block
Design (RCBD) framework, all of the 33 individual trials were evaluated
through mixed model methodology. Beyond estimation of variance
components, statistical hypothesis tests, and experimental quality were
audited using broad-sense heritability ($H^{2}$) and the coefficient of
variation ($CV%$). Furthermore, selected candidates were evaluated based
on their Best Linear Unbiased Predictions (BLUPs) and respective
reliability values. Finally, a cross-trial ranking comparison is
presented, providing a critical speculation on the underlying genetic
and environmental factors that govern the differences or similarities in
genotype performance across locations.

# Outliers Removal

First, we performed some analysis to remove outlier data that are not
agronomically consistent, and we standardized the codification of
missing information to NA.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

```{r}
#| label: trials-data

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml", "lmmtools")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data 
data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data$row_number <- as.factor(data$row_number)
data$col_number <- as.factor(data$col_number)
data$rowgroup <- as.factor(data$rowgroup)
data$colgroup <- as.factor(data$colgroup)
data$check <- as.factor(data$check)
data$geno <- as.factor(data$geno)
```


```{r}
# Trial identification
unique(data$env)
```

# Trial 1 - OT1NAM22

variance components, broad-sense heritabilities, and rankings

It is 50 rows and each row has 30 columns.

```{r}
t1 <- data |> filter (env == "OT1NAM22")

unique(data$row_number)
unique(data$col_number)
unique(data$rowgroup)
unique(data$colgroup)
unique(data$check.cod)
length(unique(data$geno.cod))
unique(data$rc.group)

# Conta os genótipos únicos por grupo de linha-coluna
tapply(data$geno, data$rc.group, function(x) length(unique(x)))

```
The experiment has 3 super columns and 5 super rows, which means the blocks are . They are testing 1178 genotypes and 8 checks.

The model used in the article Piepho & Williams (2016) can be represented

$$ y = \mu + rowgroup + colgroup + rowgroup.colgroup + rowgroup.row + colgroup.col + genotype + \epsilon$$
posso fazer tudo aleatorio e só as checks como fixo

```{r}
#| label: models

# Piepho & Williams (2016) model as a starter

model_1 <- asreml(fixed = rytha ~ at(check, "Check"),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:row_number + colgroup:col_number + at(check, "Candidate"):geno,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1)
plot(varioGram(model_1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

# Correcting col and row number

model_2 <-  asreml(fixed = rytha ~ at(check, "Check"),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:row_number + colgroup:col_number + at(check, "Candidate"):geno,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1)

plot(varioGram(model_2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))
# 

model_3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + at(check, "Check"),
             random = ~ rowgroup:colgroup + rowgroup:row_number + colgroup:col_number + at(check, "Candidate"):geno + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1)

plot(varioGram(model_3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```
```{r}
var_comp_model_1 <- summary(model_1)$varcomp

blups_model_1 <- predict.asreml(model_1, classify = "at(check, 'Candidate'):geno", sed = TRUE)
blups_model_1_pvals <- blups_model_1$pvals
blups_model_1_sed <- blups_model_1$sed

#blups
blups_model_1_pvals |> 
  arrange(desc(predicted.value))

#blues
blues_checks <- predict(model_1, classify = "check:geno")$pvals

#Reliability
reb_model_1 <- unique(1 - (blups_model_1$pvals$std.error^2/var_comp_model_1[6,1]))


ggplot(data = t1, aes(x = col_number, y = row_number, fill = rytha)) + 
  geom_tile() + 
  theme_minimal() + 
  theme(text = element_text(size = 20)) + 
  labs(x = "Column", y = "Row") +
  scale_fill_viridis_c(option = "magma")

```

```{r}
# Best model

```
```{r}
# Heritability

calcular_heritabilidade_cullis <- function(modelo) {
  # 1. Extrair a variância genética dos candidatos
  vc <- summary(modelo)$varcomp
  # Encontra a linha correspondente ao componente genético do candidato
  sigma2_g <- vc["at(check, Candidate):geno!geno.var", "component"] 
  
  # 2. Obter a matriz de predições e os erros padrões associados (PEV)
  pred <- predict(modelo, classify = "at(check, 'Candidate'):geno", sed = TRUE)
  
  # 3. Calcular a média quadrada das diferenças de erro padrão (Mean PEV)
  mean_sed <- mean(pred$sed^2)
  
  # 4. Fórmula de Cullis
  h2_cullis <- 1 - (mean_sed / (2 * sigma2_g))
  return(h2_cullis)
}

# Calcule para os três modelos e reporte no relatório:
calcular_heritabilidade_cullis(M1)
calcular_heritabilidade_cullis(M2)
calcular_heritabilidade_cullis(M3)
```

```{r}
# Variance components

# 1. Comparar as tabelas de componentes de variância de forma direta
summary(M1)$varcomp
summary(M2)$varcomp
summary(M3)$varcomp

# 2. Extrair e ranquear os BLUPs de cada modelo para ver a mudança de posições
get_ranking <- function(modelo) {
  predict(modelo, classify = "at(check, 'Candidate'):geno")$pvals %>%
    filter(check == "Candidate") %>%
    mutate(predicted.value = as.numeric(as.character(predicted.value))) %>%
    filter(!is.na(predicted.value)) %>%
    arrange(desc(predicted.value)) %>%
    mutate(Posicao = row_number()) %>%
    select(Posicao, geno, predicted.value)
}

rank1 <- get_ranking(M1)
rank2 <- get_ranking(M2)
rank3 <- get_ranking(M3)

# Dica: Use uma correlação de Spearman para reportar o impacto no ranking
cor(rank1$predicted.value, rank3$predicted.value, method = "spearman")
```



Testing whether the checks separated  

```{r}
wald.asreml(model_1)
```
The checks are significantly different, at least two of them.

### Ranking

como eu vou saber se tenho rows iguais, como descrito no artigo?
## Best model

# Trial 2 - OT1RWE21

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 3 - OT1RWE22

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 4 - OT1SER22

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 5 - OT2NAM21

variance components, broad-sense heritabilities, and rankings

## Best model

# Trial 6 - OT2SER21

variance components, broad-sense heritabilities, and rankings

## Best model
