---
title: "Spatial analysis of augmented designs in sweet potato"
author: 
  - name: Bruna Moreno, Daniele Bononi, João Vitor Calvo, Renan Fantine
    affiliation:
      - name: Department of Genetics, ESALQ-USP
        city: Piracicaba, São Paulo, Brazil
format: 
  html:
    toc: true
    toc-location: left
    code-overflow: wrap
    code-fold: true
    fig-align: center
    theme: 
      light: journal
      dark: darkly
    embed-resources: true
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
        }
        </style>
crossref: {}
editor: 
  markdown: 
    wrap: 72
---

# Exercise

The dataset available here has six trials laid out in an augmented
row-column design (more details about this design here). Analyse each
trial individually with increasing order of spatial adjustment. Report
the improvement of the modelling (if any) when spatial trends are dealt
with. Your report should have differences in the variance components,
broad-sense heritabilities, and rankings. For each trial, determine the
best model using semivariograms and the AIC. Don't forget that models
with different fixed effects must be compared using the adjusted version
of the AIC available on the lmmtools package.

# Introduction

This report provides a data analysis of a spatial correction of a
multi-environment sweet potato experiment in Africa. Each one of the six
environments were analyzed with multiple models testing different
approaches to spatial correction. It was estimated variance components,
broad-sense heritabilities (Cullis, 2006), and rankings based on BLUP
predictions.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.show = "hold",
  warning = FALSE,
  message = FALSE,
  results = "hold")
```

The trials were conducted across Uganda in three locations (Rwebitaba -
RWE, Namulonge - NAM and Serere - SER) during the 2021 and 2022 growing
seasons. To account for spatial variation, each trial was analyzed
individually to implement single-trial spatial corrections.

This dataset is derived from a multi-environment sweet potato trial. The
experimental layout follows an augmented row-column design, where each
environment comprises a grid of 50 rows and 30 columns. These rows are
grouped into 5 super-rows, and the columns into 3 super-columns. Each
resulting combination (hereafter referred to as a miniblock) consists of
10 rows and 10 columns, encompassing a total of 8 checks and 1,178 test
genotypes.

The model used in this report was the same described in the article of
[Piepho & Williams (2016)](https://doi.org/10.2134/agronj2016.06.0325),
which can be represented as the model below:

$$ y = \mu + rowgroup + colgroup + rowgroup.colgroup + rowgroup.row + colgroup.col + genotype + check + \epsilon$$

Where, $\mu$ is the intercept,

$rowgroup$ is the effect of the super-row,

$colgroup$ is the effect of the super-column,

$rowgroup.colgroup$ is the effect of the intersection of super-row and
super-column (the miniblock),

$rowgroup.row$ the random effect of the row number nested in the
super-row,

$colgroup.col$ the random effect of the column number nested in the
super-column,

$genotype$ the random effect of the genotype,

$check$ the fixed effect of the check and

$\epsilon$ the random effect of the error.

In order to verify the spatial correction, each trial was analysed
individually, through five different models:

-   Model 1: Reference model (the same described in [Piepho & Williams
    (2016)](https://doi.org/10.2134/agronj2016.06.0325)), considering
    row and column groups as random

-   Model 2: Same as 'Model 1', but using row and column groups as fixed

-   Model 3: 'Model 1' + residuals with AR1 structure

-   Model 4: 'Model 1' + residuals with AR1 structure + nugget effect

-   Model 5: 'Model 1' + residuals with AR1 structure + nugget effect +
    linear trend across row and columns

After that, for each trial/model combination, check effects (fixed
effects) were tested using Wald tests, whereas candidate genotype
effects (random effects) were evaluated using likelihood ratio tests
(LRT). The semi-variograms, spatial distribution of residuals, and AIC
values were then assessed to select the most appropriate model for each
trial.

```{r}
#| label: trials-data
#| warning: false

# Loading packages
packages <- c("ggplot2", "patchwork", "tidyr", "dplyr", "purrr", "knitr", "kableExtra", "grid","ggcorrplot", "asreml", "lmmtools")
invisible(lapply(packages, function(p){
  suppressMessages(suppressPackageStartupMessages(
  suppressWarnings(library(p, character.only = TRUE))))}))

# Loading phenotypic data 
data = read.csv("https://raw.githubusercontent.com/ufv-molecular-breeding-lab/schaves_sweetpotato_two_stage/refs/heads/main/Data/Phenotype.csv", sep = ',')

data$row_number <- as.factor(data$row_number)
data$col_number <- as.factor(data$col_number)
data$rowgroup <- as.factor(data$rowgroup)
data$colgroup <- as.factor(data$colgroup)
data$check <- as.factor(data$check)
data$geno <- as.factor(data$geno)
data$check.cod = as.factor(data$check.cod)
data$geno.cod  = as.factor(data$geno.cod)
```

```{r}
#| label: blup_comparison
blup_comparison <- function(model1, model2, m1_name = "Model 1", m2_name = "Best Model") {
  
  blup_m1 <- model1[,c("geno.cod", "predicted.value")]
  colnames(blup_m1) <- c('geno.cod','blup.m1')
  blup_m2 <- model2[,c("geno.cod", "predicted.value")]
  colnames(blup_m2) <- c('geno.cod','blup.m2')
  comparison <- merge(blup_m1,blup_m2,by = "geno.cod")
  ngen <- nrow(comparison) # number of genotypes
  
  # Rankings
  comparison <- comparison |> 
    mutate(rank.m1 = rank(-blup.m1, ties.method = "average"),
      rank.m2 = rank(-blup.m2, ties.method = "average"),
      blup_diff = abs(blup.m1 - blup.m2),
      rank_diff = rank.m1 - rank.m2)
  
  rank_cor <- cor(comparison$blup.m1, comparison$blup.m2, method = "spearman") # Spearman correlation (Pearson correlation between ranks)
  pearson_cor <- cor(comparison$blup.m1, comparison$blup.m2, method = "pearson") # Pearson correlation between blups
  
  # Summary metrics
  metrics <- data.frame(
    n_genotypes = nrow(comparison),
    pearson_correlation = pearson_cor,
    spearman_correlation = rank_cor,
    mean_blup_diff = mean(comparison$blup_diff),
    sd_blup_diff = sd(comparison$blup_diff),
    min_blup_diff = min(comparison$blup_diff),
    max_blup_diff = max(comparison$blup_diff),
    sd_rank_diff = sd(comparison$rank_diff),
    min_rank_diff = min(abs(comparison$rank_diff)),
    max_rank_diff = max(abs(comparison$rank_diff))
    )
  
  # Rankings plot
  ranking_plot <- plotly::ggplotly(
    ggplot(comparison,aes(x = rank.m1,y = rank.m2)) +
    geom_point(aes(color = geno.cod),size = 2) +
    theme(axis.line = element_line(colour = "black"),
      legend.position = "none",
      panel.background = element_blank(),
      text = element_text(size = 14)) +
    labs(x = paste(m1_name, "- Ranking (M1)"),
      y = paste(m2_name, "- Ranking (M2)"),
      title = paste("Spearman correlation =", round(rank_cor, 4))))

  # BLUPs scatterplot
  blup_plot <- ggplot(comparison, aes(x = blup.m1, y = blup.m2)) +
    geom_point(size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    theme_bw() +
    labs(x = paste(m1_name, "- BLUP"),
      y = paste(m2_name, "- BLUP"),
      title = paste("Pearson correlation =", round(pearson_cor,4)))
  
  return(list(comparison_table = comparison,
    metrics = metrics,
    ranking_plot = ranking_plot,
    blup_plot = blup_plot))
  
    }
```

# Namulonge Season 2022/1 (Trial 1 - OT1NAM22)

```{r}
#| label: fig-t1-prod

t1 <- data |> filter (env == "OT1NAM22")

# Modifying the geno.cod and check.cod columns
t1 <- t1 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t1 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#             n_obs = sum(!is.na(rytha))) |>
#   arrange(n_obs)
```

```{r}
#| label: models-t1
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t1 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Model 3 - AR1 in residuals
model_3_t1 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t1 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t1 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

```

### Variance components {#sec-VC1}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t1 <- summary(model_1_t1)$varcomp
varcomp_m1_t1 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t1 <- summary(model_2_t1)$varcomp
varcomp_m2_t1 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t1 <- summary(model_3_t1)$varcomp
varcomp_m3_t1 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t1 <- summary(model_4_t1)$varcomp
varcomp_m4_t1 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t1 <- summary(model_5_t1)$varcomp
varcomp_m5_t1 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t1)
```

### Model 2

```{r}
wald.asreml(model_2_t1)
```

### Model 3

```{r}
wald.asreml(model_3_t1)
```

### Model 4

```{r}
wald.asreml(model_4_t1)
```

### Model 5

```{r}
wald.asreml(model_5_t1)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t1 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t1, maxinter=50)

red_model2_t1 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model3_t1 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model4_t1 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)

red_model5_t1 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t1, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t1, red_model1_t1)

```

### Model 2

```{r}
lrt(model_2_t1, red_model2_t1)

```

### Model 3

```{r}
lrt(model_3_t1, red_model3_t1)
```

### Model 4

```{r}
lrt(model_4_t1, red_model4_t1)

```

### Model 5

```{r}
lrt(model_5_t1, red_model5_t1)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t1
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t1, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC1}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t1 = icREML(list(model_1_t1, model_2_t1, model_3_t1, model_4_t1, model_5_t1)); modsel_t1
modsel_t1 <- as.data.frame(modsel_t1)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t1))
```

### Model 2

```{r}
icREML(list(model_2_t1))
```

### Model 3

```{r}
icREML(list(model_3_t1))
```

### Model 4

```{r}
icREML(list(model_4_t1))
```

### Model 5

```{r}
icREML(list(model_5_t1))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t1
#| fig-cap: "Spatial Distribution of residuals (Trial 1)"
resid_long <- t1 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t1, type = "response"),
    Model_2 = residuals(model_2_t1, type = "response"),
    Model_3 = residuals(model_3_t1, type = "response"),
    Model_4 = residuals(model_4_t1, type = "response"),
    Model_5 = residuals(model_5_t1, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t1-blups
#| results: hide

#BLUPs
blup1_t1 <- predict.asreml(model_1_t1, classify = "geno.cod", sed = TRUE)
blup2_t1 <- predict.asreml(model_2_t1, classify = "geno.cod", sed = TRUE)
blup3_t1 <- predict.asreml(model_3_t1, classify = "geno.cod", sed = TRUE)
blup4_t1 <- predict.asreml(model_4_t1, classify = "geno.cod", sed = TRUE)
blup5_t1 <- predict.asreml(model_5_t1, classify = "geno.cod", sed = TRUE)

blup_t1 <- list(blup1_t1, blup2_t1, blup3_t1, blup4_t1, blup5_t1)
predic_value_t1 <- list()

for (i in 1:length(blup_t1)) {
  
  predic_value_t1[[i]] <- blup_t1[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t1 <- predict(model_1_t1, classify = "check.cod")$pvals
blues_checks2_t1 <- predict(model_2_t1, classify = "check.cod")$pvals
blues_checks3_t1 <- predict(model_3_t1, classify = "check.cod")$pvals
blues_checks4_t1 <- predict(model_4_t1, classify = "check.cod")$pvals
blues_checks5_t1 <- predict(model_5_t1, classify = "check.cod")$pvals

blue_t1 <- list(blues_checks1_t1, blues_checks2_t1, blues_checks3_t1, blues_checks4_t1, blues_checks5_t1)
bpredic_value_t1 <- list()

for (i in 1:length(blue_t1)) {
  
  bpredic_value_t1[[i]] <- blue_t1[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}
```

```{r}
#| label: fig-h2-t1
#| fig-cap: "BLUPs ranking correlation (Trial 1)"
#| 
# Selecting baseline model and best model
t1_comp <- blup_comparison(blup1_t1$pvals, blup4_t1$pvals, m1_name = "Model 1", m2_name = "Best Model")

t1_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t1
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t1 <- 1-(mean(blup1_t1$sed[upper.tri(blup1_t1$sed)]^2)/(2*varcomp_m1_t1[6,1]))
h2_mod2_t1 <- 1-(mean(blup2_t1$sed[upper.tri(blup2_t1$sed)]^2)/(2*varcomp_m2_t1[3,1]))
h2_mod3_t1 <- 1-(mean(blup3_t1$sed[upper.tri(blup3_t1$sed)]^2)/(2*varcomp_m3_t1[6,1]))
h2_mod4_t1 <- 1-(mean(blup4_t1$sed[upper.tri(blup4_t1$sed)]^2)/(2*varcomp_m4_t1[6,1]))
h2_mod5_t1 <- 1-(mean(blup5_t1$sed[upper.tri(blup5_t1$sed)]^2)/(2*varcomp_m5_t1[6,1]))

h2_t1 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t1, h2_mod2_t1, h2_mod3_t1, h2_mod4_t1, h2_mod5_t1)
)

options(knitr.table.format = "html")

h2_t1 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Rwebitaba Season 2021/1 (Trial 2 - OT1RWE21)

```{r}
#| label: fig-t2-prod

t2 <- data |> filter (env == "OT1RWE21")

# Modifying the geno.cod and check.cod columns
t2 <- t2 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t2 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#             n_obs = sum(!is.na(rytha))) |>
#   arrange(n_obs)
```

```{r}
#| label: models-t2
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t2 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t2, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t2 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t2, maxint=50)

# Model 3 - AR1 in residuals
model_3_t2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t2 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

```

### Variance components {#sec-VC2}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t2 <- summary(model_1_t2)$varcomp
varcomp_m1_t2 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t2 <- summary(model_2_t2)$varcomp
varcomp_m2_t2 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t2 <- summary(model_3_t2)$varcomp
varcomp_m3_t2 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t2 <- summary(model_4_t2)$varcomp
varcomp_m4_t2 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t2 <- summary(model_5_t2)$varcomp
varcomp_m5_t2 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t2)
```

### Model 2

```{r}
wald.asreml(model_2_t2)
```

### Model 3

```{r}
wald.asreml(model_3_t2)
```

### Model 4

```{r}
wald.asreml(model_4_t2)
```

### Model 5

```{r}
wald.asreml(model_5_t2)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t2 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t2, maxinter=50)

red_model2_t2 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t2, maxint=50)

red_model3_t2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

red_model4_t2 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)

red_model5_t2 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t2, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t2, red_model1_t2)

```

### Model 2

```{r}
lrt(model_2_t2, red_model2_t2)

```

### Model 3

```{r}
lrt(model_3_t2, red_model3_t2)
```

### Model 4

```{r}
lrt(model_4_t2, red_model4_t2)

```

### Model 5

```{r}
lrt(model_5_t2, red_model5_t2)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t2
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t2, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC2}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t2 = icREML(list(model_1_t2, model_2_t2, model_3_t2, model_4_t2, model_5_t2)); modsel_t2
modsel_t2 <- as.data.frame(modsel_t2)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t2))
```

### Model 2

```{r}
icREML(list(model_2_t2))
```

### Model 3

```{r}
icREML(list(model_3_t2))
```

### Model 4

```{r}
icREML(list(model_4_t2))
```

### Model 5

```{r}
icREML(list(model_5_t2))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t2
#| fig-cap: "Spatial Distribution of residuals (Trial 2)"
resid_long <- t2 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t2, type = "response"),
    Model_2 = residuals(model_2_t2, type = "response"),
    Model_3 = residuals(model_3_t2, type = "response"),
    Model_4 = residuals(model_4_t2, type = "response"),
    Model_5 = residuals(model_5_t2, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t2-blups
#| results: hide

#BLUPs
blup1_t2 <- predict.asreml(model_1_t2, classify = "geno.cod", sed = TRUE)
blup2_t2 <- predict.asreml(model_2_t2, classify = "geno.cod", sed = TRUE)
blup3_t2 <- predict.asreml(model_3_t2, classify = "geno.cod", sed = TRUE)
blup4_t2 <- predict.asreml(model_4_t2, classify = "geno.cod", sed = TRUE)
blup5_t2 <- predict.asreml(model_5_t2, classify = "geno.cod", sed = TRUE)

blup_t2 <- list(blup1_t2, blup2_t2, blup3_t2, blup4_t2, blup5_t2)
predic_value_t2 <- list()

for (i in 1:length(blup_t2)) {
  
  predic_value_t2[[i]] <- blup_t2[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t2 <- predict(model_1_t2, classify = "check.cod")$pvals
blues_checks2_t2 <- predict(model_2_t2, classify = "check.cod")$pvals
blues_checks3_t2 <- predict(model_3_t2, classify = "check.cod")$pvals
blues_checks4_t2 <- predict(model_4_t2, classify = "check.cod")$pvals
blues_checks5_t2 <- predict(model_5_t2, classify = "check.cod")$pvals

blue_t2 <- list(blues_checks1_t2, blues_checks2_t2, blues_checks3_t2, blues_checks4_t2, blues_checks5_t2)
bpredic_value_t2 <- list()

for (i in 1:length(blue_t2)) {
  
  bpredic_value_t2[[i]] <- blue_t2[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

```{r}
#| label: fig-h2-t2
#| fig-cap: "BLUPs ranking correlation (Trial 2)"

# Selecting baseline model and best model
t2_comp <- blup_comparison(blup1_t2$pvals, blup2_t2$pvals, m1_name = "Model 1", m2_name = "Best Model")
t2_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t2
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t2 <- 1-(mean(blup1_t2$sed[upper.tri(blup1_t2$sed)]^2)/(2*varcomp_m1_t2[6,1]))
h2_mod2_t2 <- 1-(mean(blup2_t2$sed[upper.tri(blup2_t2$sed)]^2)/(2*varcomp_m2_t2[3,1]))
h2_mod3_t2 <- 1-(mean(blup3_t2$sed[upper.tri(blup3_t2$sed)]^2)/(2*varcomp_m3_t2[6,1]))
h2_mod4_t2 <- 1-(mean(blup4_t2$sed[upper.tri(blup4_t2$sed)]^2)/(2*varcomp_m4_t2[6,1]))
h2_mod5_t2 <- 1-(mean(blup5_t2$sed[upper.tri(blup5_t2$sed)]^2)/(2*varcomp_m5_t2[6,1]))

h2_t2 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t2, h2_mod2_t2, h2_mod3_t2, h2_mod4_t2, h2_mod5_t2)
)

options(knitr.table.format = "html")

h2_t2 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Rwebitaba Season 2022/1 (Trial 3 - OT1RWE22)

```{r}
#| label: fig-t3-prod

t3 <- data |> filter (env == "OT1RWE21")

# Modifying the geno.cod and check.cod columns
t3 <- t3 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t3 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#             n_obs = sum(!is.na(rytha))) |>
#   arrange(n_obs)

t3 <- t3 |> mutate(
  check.cod = as.character(check.cod),
  check.cod = ifelse(check.cod == "RAK808", "Candidato", check.cod),
  check.cod = as.factor(check.cod))

#unique(t3$row_number)
#unique(t3$col_number)
#unique(t3$rowgroup)
#unique(t3$colgroup)
#unique(t3$check.cod)
#length(unique(t3$geno.cod))
#unique(t3$rc.group)

```

```{r}
#| label: models-t3
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t3 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t3, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 3 - AR1 in residuals
model_3_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t3 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

```

### Variance components {#sec-VC3}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t3 <- summary(model_1_t3)$varcomp
varcomp_m1_t3 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t3 <- summary(model_2_t3)$varcomp
varcomp_m2_t3 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t3 <- summary(model_3_t3)$varcomp
varcomp_m3_t3 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t3 <- summary(model_4_t3)$varcomp
varcomp_m4_t3 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t3 <- summary(model_5_t3)$varcomp
varcomp_m5_t3 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t3)
```

### Model 2

```{r}
wald.asreml(model_2_t3)
```

### Model 3

```{r}
wald.asreml(model_3_t3)
```

### Model 4

```{r}
wald.asreml(model_4_t3)
```

### Model 5

```{r}
wald.asreml(model_5_t3)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t3 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t3, maxinter=50)

red_model2_t3 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model3_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model4_t3 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)

red_model5_t3 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t3, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t3, red_model1_t3)

```

### Model 2

```{r}
lrt(model_2_t3, red_model2_t3)

```

### Model 3

```{r}
lrt(model_3_t3, red_model3_t3)
```

### Model 4

```{r}
lrt(model_4_t3, red_model4_t3)

```

### Model 5

```{r}
lrt(model_5_t3, red_model5_t3)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t3
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t3, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC3}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t3 = icREML(list(model_1_t3, model_2_t3, model_3_t3, model_4_t3, model_5_t3)); modsel_t3
modsel_t3 <- as.data.frame(modsel_t3)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t3))
```

### Model 2

```{r}
icREML(list(model_2_t3))
```

### Model 3

```{r}
icREML(list(model_3_t3))
```

### Model 4

```{r}
icREML(list(model_4_t3))
```

### Model 5

```{r}
icREML(list(model_5_t3))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t3
#| fig-cap: "Spatial Distribution of residuals (Trial 3)"
resid_long <- t3 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t3, type = "response"),
    Model_2 = residuals(model_2_t3, type = "response"),
    Model_3 = residuals(model_3_t3, type = "response"),
    Model_4 = residuals(model_4_t3, type = "response"),
    Model_5 = residuals(model_5_t3, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t3-blups
#| results: hide

#BLUPs
blup1_t3 <- predict.asreml(model_1_t3, classify = "geno.cod", sed = TRUE)
blup2_t3 <- predict.asreml(model_2_t3, classify = "geno.cod", sed = TRUE)
blup3_t3 <- predict.asreml(model_3_t3, classify = "geno.cod", sed = TRUE)
blup4_t3 <- predict.asreml(model_4_t3, classify = "geno.cod", sed = TRUE)
blup5_t3 <- predict.asreml(model_5_t3, classify = "geno.cod", sed = TRUE)

blup_t3 <- list(blup1_t3, blup2_t3, blup3_t3, blup4_t3, blup5_t3)
predic_value_t3 <- list()

for (i in 1:length(blup_t3)) {
  
  predic_value_t3[[i]] <- blup_t3[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t3 <- predict(model_1_t3, classify = "check.cod")$pvals
blues_checks2_t3 <- predict(model_2_t3, classify = "check.cod")$pvals
blues_checks3_t3 <- predict(model_3_t3, classify = "check.cod")$pvals
blues_checks4_t3 <- predict(model_4_t3, classify = "check.cod")$pvals
blues_checks5_t3 <- predict(model_5_t3, classify = "check.cod")$pvals

blue_t3 <- list(blues_checks1_t3, blues_checks2_t3, blues_checks3_t3, blues_checks4_t3, blues_checks5_t3)
bpredic_value_t3 <- list()

for (i in 1:length(blue_t3)) {
  
  bpredic_value_t3[[i]] <- blue_t3[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

```{r}
#| label: fig-h2-t3
#| fig-cap: "BLUPs ranking correlation (Trial 3)"

# Selecting baseline model and best model
t3_comp <- blup_comparison(blup1_t3$pvals, blup4_t3$pvals, m1_name = "Model 1", m2_name = "Best Model")
t3_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t3
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t3 <- 1-(mean(blup1_t3$sed[upper.tri(blup1_t3$sed)]^2)/(2*varcomp_m1_t3[6,1]))
h2_mod2_t3 <- 1-(mean(blup2_t3$sed[upper.tri(blup2_t3$sed)]^2)/(2*varcomp_m2_t3[3,1]))
h2_mod3_t3 <- 1-(mean(blup3_t3$sed[upper.tri(blup3_t3$sed)]^2)/(2*varcomp_m3_t3[6,1]))
h2_mod4_t3 <- 1-(mean(blup4_t3$sed[upper.tri(blup4_t3$sed)]^2)/(2*varcomp_m4_t3[6,1]))
h2_mod5_t3 <- 1-(mean(blup5_t3$sed[upper.tri(blup5_t3$sed)]^2)/(2*varcomp_m5_t3[6,1]))

h2_t3 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t3, h2_mod2_t3, h2_mod3_t3, h2_mod4_t3, h2_mod5_t3)
)

options(knitr.table.format = "html")

h2_t3 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Serere Season 2022/1 (Trial 4 - OT1SER22)

```{r}
#| label: fig-t4-prod

t4 <- data |> filter (env == "OT1SER22")

# Modifying the geno.cod and check.cod columns
t4 <- t4 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t4 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#             n_obs = sum(!is.na(rytha))) |>
#   arrange(n_obs)

t4 <- t4 |> mutate(
  check.cod = as.character(check.cod),
  check.cod = ifelse(check.cod == "RAK808", "Candidato", check.cod),
  check.cod = as.factor(check.cod))
```

```{r}
#| label: models-t4
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t4 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t4, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t4 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 3 - AR1 in residuals
model_3_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

```

### Variance components {#sec-VC4}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t4 <- summary(model_1_t4)$varcomp
varcomp_m1_t4 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t4 <- summary(model_2_t4)$varcomp
varcomp_m2_t4 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t4 <- summary(model_3_t4)$varcomp
varcomp_m3_t4 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t4 <- summary(model_4_t4)$varcomp
varcomp_m4_t4 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t4 <- summary(model_5_t4)$varcomp
varcomp_m5_t4 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t4)
```

### Model 2

```{r}
wald.asreml(model_2_t4)
```

### Model 3

```{r}
wald.asreml(model_3_t4)
```

### Model 4

```{r}
wald.asreml(model_4_t4)
```

### Model 5

```{r}
wald.asreml(model_5_t4)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t4 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t4, maxinter=50)

red_model2_t4 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model3_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model4_t4 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)

red_model5_t4 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t4, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t4, red_model1_t4)

```

### Model 2

```{r}
lrt(model_2_t4, red_model2_t4)

```

### Model 3

```{r}
lrt(model_3_t4, red_model3_t4)
```

### Model 4

```{r}
lrt(model_4_t4, red_model4_t4)

```

### Model 5

```{r}
lrt(model_5_t4, red_model5_t4)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t4
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t4, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC4}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t4 = icREML(list(model_1_t4, model_2_t4, model_3_t4, model_4_t4, model_5_t4)); modsel_t4
modsel_t4 <- as.data.frame(modsel_t4)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t4))
```

### Model 2

```{r}
icREML(list(model_2_t4))
```

### Model 3

```{r}
icREML(list(model_3_t4))
```

### Model 4

```{r}
icREML(list(model_4_t4))
```

### Model 5

```{r}
icREML(list(model_5_t4))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t4
#| fig-cap: "Spatial Distribution of residuals (Trial 4)"
resid_long <- t4 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t4, type = "response"),
    Model_2 = residuals(model_2_t4, type = "response"),
    Model_3 = residuals(model_3_t4, type = "response"),
    Model_4 = residuals(model_4_t4, type = "response"),
    Model_5 = residuals(model_5_t4, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t4-blups
#| results: hide

#BLUPs
blup1_t4 <- predict.asreml(model_1_t4, classify = "geno.cod", sed = TRUE)
blup2_t4 <- predict.asreml(model_2_t4, classify = "geno.cod", sed = TRUE)
blup3_t4 <- predict.asreml(model_3_t4, classify = "geno.cod", sed = TRUE)
blup4_t4 <- predict.asreml(model_4_t4, classify = "geno.cod", sed = TRUE)
blup5_t4 <- predict.asreml(model_5_t4, classify = "geno.cod", sed = TRUE)

blup_t4 <- list(blup1_t4, blup2_t4, blup3_t4, blup4_t4, blup5_t4)
predic_value_t4 <- list()

for (i in 1:length(blup_t4)) {
  
  predic_value_t4[[i]] <- blup_t4[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t4 <- predict(model_1_t4, classify = "check.cod")$pvals
blues_checks2_t4 <- predict(model_2_t4, classify = "check.cod")$pvals
blues_checks3_t4 <- predict(model_3_t4, classify = "check.cod")$pvals
blues_checks4_t4 <- predict(model_4_t4, classify = "check.cod")$pvals
blues_checks5_t4 <- predict(model_5_t4, classify = "check.cod")$pvals

blue_t4 <- list(blues_checks1_t4, blues_checks2_t4, blues_checks3_t4, blues_checks4_t4, blues_checks5_t4)
bpredic_value_t4 <- list()

for (i in 1:length(blue_t4)) {
  
  bpredic_value_t4[[i]] <- blue_t4[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

```{r}
#| label: fig-h2-t4
#| fig-cap: "BLUPs ranking correlation (Trial 4)"

# Selecting baseline model and best model
t4_comp <- blup_comparison(blup1_t4$pvals, blup4_t4$pvals, m1_name = "Model 1", m2_name = "Best Model")
t4_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t4
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t4 <- 1-(mean(blup1_t4$sed[upper.tri(blup1_t4$sed)]^2)/(2*varcomp_m1_t4[6,1]))
h2_mod2_t4 <- 1-(mean(blup2_t4$sed[upper.tri(blup2_t4$sed)]^2)/(2*varcomp_m2_t4[3,1]))
h2_mod3_t4 <- 1-(mean(blup3_t4$sed[upper.tri(blup3_t4$sed)]^2)/(2*varcomp_m3_t4[6,1]))
h2_mod4_t4 <- 1-(mean(blup4_t4$sed[upper.tri(blup4_t4$sed)]^2)/(2*varcomp_m4_t4[6,1]))
h2_mod5_t4 <- 1-(mean(blup5_t4$sed[upper.tri(blup5_t4$sed)]^2)/(2*varcomp_m5_t4[6,1]))

h2_t4 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t4, h2_mod2_t4, h2_mod3_t4, h2_mod4_t4, h2_mod5_t4)
)

options(knitr.table.format = "html")

h2_t4 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Namulonge Season 2021/2 (Trial 5 - OT2NAM21)

```{r}
#| label: fig-t5-prod

t5 <- data |> filter (env == "OT2NAM21")

# Modifying the geno.cod and check.cod columns
t5 <- t5 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t5 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#            n_obs = sum(!is.na(rytha))) |>
#  arrange(n_obs)
```

```{r}
#| label: models-t5
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t5 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t5, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t5 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t5, maxint=50)

# Model 3 - AR1 in residuals
model_3_t5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t5 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)

```

### Variance components {#sec-VC5}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t5 <- summary(model_1_t5)$varcomp
varcomp_m1_t5 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t5 <- summary(model_2_t5)$varcomp
varcomp_m2_t5 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t5 <- summary(model_3_t5)$varcomp
varcomp_m3_t5 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t5 <- summary(model_4_t5)$varcomp
varcomp_m4_t5 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t5 <- summary(model_5_t5)$varcomp
varcomp_m5_t5 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t5)
```

### Model 2

```{r}
wald.asreml(model_2_t5)
```

### Model 3

```{r}
wald.asreml(model_3_t5)
```

### Model 4

```{r}
wald.asreml(model_4_t5)
```

### Model 5

```{r}
wald.asreml(model_5_t5)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t5 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t5, maxinter=50)

red_model2_t5 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t5, maxint=50)

red_model3_t5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)

red_model4_t5 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)

red_model5_t5 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t5, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t5, red_model1_t5)

```

### Model 2

```{r}
lrt(model_2_t5, red_model2_t5)

```

### Model 3

```{r}
lrt(model_3_t5, red_model3_t5)
```

### Model 4

```{r}
lrt(model_4_t5, red_model4_t5)

```

### Model 5

```{r}
lrt(model_5_t5, red_model5_t5)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t5
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t5, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC5}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t5 = icREML(list(model_1_t5, model_2_t5, model_3_t5, model_4_t5, model_5_t5)); modsel_t5
modsel_t5 <- as.data.frame(modsel_t5)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t5))
```

### Model 2

```{r}
icREML(list(model_2_t5))
```

### Model 3

```{r}
icREML(list(model_3_t5))
```

### Model 4

```{r}
icREML(list(model_4_t5))
```

### Model 5

```{r}
icREML(list(model_5_t5))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t5
#| fig-cap: "Spatial Distribution of residuals (Trial 5)"
resid_long <- t5 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t5, type = "response"),
    Model_2 = residuals(model_2_t5, type = "response"),
    Model_3 = residuals(model_3_t5, type = "response"),
    Model_4 = residuals(model_4_t5, type = "response"),
    Model_5 = residuals(model_5_t5, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t5-blups
#| results: hide

#BLUPs
blup1_t5 <- predict.asreml(model_1_t5, classify = "geno.cod", sed = TRUE)
blup2_t5 <- predict.asreml(model_2_t5, classify = "geno.cod", sed = TRUE)
blup3_t5 <- predict.asreml(model_3_t5, classify = "geno.cod", sed = TRUE)
blup4_t5 <- predict.asreml(model_4_t5, classify = "geno.cod", sed = TRUE)
blup5_t5 <- predict.asreml(model_5_t5, classify = "geno.cod", sed = TRUE)

blup_t5 <- list(blup1_t5, blup2_t5, blup3_t5, blup4_t5, blup5_t5)
predic_value_t5 <- list()

for (i in 1:length(blup_t5)) {
  
  predic_value_t5[[i]] <- blup_t5[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t5 <- predict(model_1_t5, classify = "check.cod")$pvals
blues_checks2_t5 <- predict(model_2_t5, classify = "check.cod")$pvals
blues_checks3_t5 <- predict(model_3_t5, classify = "check.cod")$pvals
blues_checks4_t5 <- predict(model_4_t5, classify = "check.cod")$pvals
blues_checks5_t5 <- predict(model_5_t5, classify = "check.cod")$pvals

blue_t5 <- list(blues_checks1_t5, blues_checks2_t5, blues_checks3_t5, blues_checks4_t5, blues_checks5_t5)
bpredic_value_t5 <- list()

for (i in 1:length(blue_t5)) {
  
  bpredic_value_t5[[i]] <- blue_t5[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

```{r}
#| label: fig-h2-t5
#| fig-cap: "BLUPs ranking correlation (Trial 5)"

# Selecting baseline model and best model
t5_comp <- blup_comparison(blup1_t5$pvals, blup5_t5$pvals, m1_name = "Model 1", m2_name = "Best Model")
t5_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t5
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t5 <- 1-(mean(blup1_t5$sed[upper.tri(blup1_t5$sed)]^2)/(2*varcomp_m1_t5[6,1]))
h2_mod2_t5 <- 1-(mean(blup2_t5$sed[upper.tri(blup2_t5$sed)]^2)/(2*varcomp_m2_t5[3,1]))
h2_mod3_t5 <- 1-(mean(blup3_t5$sed[upper.tri(blup3_t5$sed)]^2)/(2*varcomp_m3_t5[6,1]))
h2_mod4_t5 <- 1-(mean(blup4_t5$sed[upper.tri(blup4_t5$sed)]^2)/(2*varcomp_m4_t5[6,1]))
h2_mod5_t5 <- 1-(mean(blup5_t5$sed[upper.tri(blup5_t5$sed)]^2)/(2*varcomp_m5_t5[6,1]))

h2_t5 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t5, h2_mod2_t5, h2_mod3_t5, h2_mod4_t5, h2_mod5_t5)
)

options(knitr.table.format = "html")

h2_t5 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Serere Season 2021/2 (Trial 6 - OT2SER21)

```{r}
#| label: fig-t6-prod

t6 <- data |> filter (env == "OT2SER21")

# Modifying the geno.cod and check.cod columns
t6 <- t6 |> 
  mutate(
    check.cod = ifelse(is.na(check.cod), "Candidato", as.character(check.cod)),
    geno.cod  = ifelse(is.na(geno.cod), "Testemunha", as.character(geno.cod))
  ) |> 
  mutate(
    check.cod = as.factor(check.cod),
    geno.cod  = as.factor(geno.cod),
    rowgroup  = as.factor(rowgroup),
    colgroup  = as.factor(colgroup),
    rn        = as.factor(rn),
    cn        = as.factor(cn))

# Identifying checks with no valid observations
# t6 |> 
#   group_by(check.cod) |>
#   summarise(n_total = n(),
#             n_na = sum(is.na(rytha)),
#             n_obs = sum(!is.na(rytha))) |>
#   arrange(n_obs)

t6 <- t6 |> mutate(
  check.cod = as.character(check.cod),
  check.cod = ifelse(check.cod %in% c("RAK808", "New_Kawogo"), "Candidato", check.cod),
  check.cod = as.factor(check.cod))
```

```{r}
#| label: models-t6
#| output: false

# Model 1 - Piepho & Williams (2016) model as a starter
model_1_t6 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t6, maxinter=50)

# Model 2 - Row and column groups as fixed
model_2_t6 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t6, maxint=50)

# Model 3 - AR1 in residuals
model_3_t6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)

# Model 4 - AR1 in residuals + nugget effect
model_4_t6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)

# Model 5 - AR1 in residuals + nugget effect + row/column as covariables
model_5_t6 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + geno.cod + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)

```

### Variance components {#sec-VC6}

::: panel-tabset
### Model 1

```{r}
varcomp_m1_t6 <- summary(model_1_t6)$varcomp
varcomp_m1_t6 |> 
  kbl(caption = "Variance components for Model 1.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

### Model 2

```{r}
varcomp_m2_t6 <- summary(model_2_t6)$varcomp
varcomp_m2_t6 |> 
  kbl(caption = "Variance components for Model 2.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 3

```{r}
varcomp_m3_t6 <- summary(model_3_t6)$varcomp
varcomp_m3_t6 |> 
  kbl(caption = "Variance components for Model 3.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 4

```{r}
varcomp_m4_t6 <- summary(model_4_t6)$varcomp
varcomp_m4_t6 |> 
  kbl(caption = "Variance components for Model 4.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Model 5

```{r}
varcomp_m5_t6 <- summary(model_5_t6)$varcomp
varcomp_m5_t6 |> 
  kbl(caption = "Variance components for Model 5.", digits = 4) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```
:::

### Wald test

::: panel-tabset
### Model 1

```{r}
#| warning: false

#Fixed effects
wald.asreml(model_1_t6)
```

### Model 2

```{r}
wald.asreml(model_2_t6)
```

### Model 3

```{r}
wald.asreml(model_3_t6)
```

### Model 4

```{r}
wald.asreml(model_4_t6)
```

### Model 5

```{r}
wald.asreml(model_5_t6)
```
:::

### LRT

```{r}
#| output: false

#Random effects
red_model1_t6 <- asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~idv(units),
             na.action = na.method("include"),
             data = t6, maxinter=50)

red_model2_t6 <-  asreml(fixed = rytha ~ rowgroup + colgroup + rowgroup:colgroup + check.cod,
             random = ~ rowgroup:rn + colgroup:cn,
             residual = ~ idv(units),
             na.action = na.method("include"),
             data = t6, maxint=50)

red_model3_t6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)

red_model4_t6 <-  asreml(fixed = rytha ~ check.cod,
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)

red_model5_t6 <-  asreml(fixed = rytha ~ check.cod +  lin(row_number) + lin(col_number),
             random = ~ rowgroup + colgroup + rowgroup:colgroup + rowgroup:rn + colgroup:cn + units,
             residual = ~ ar1v(row_number):ar1(col_number),
             na.action = na.method("include"),
             data = t6, maxint=50)
```

::: panel-tabset
### Model 1

```{r}
lrt(model_1_t6, red_model1_t6)

```

### Model 2

```{r}
lrt(model_2_t6, red_model2_t6)

```

### Model 3

```{r}
lrt(model_3_t6, red_model3_t6)
```

### Model 4

```{r}
lrt(model_4_t6, red_model4_t6)

```

### Model 5

```{r}
lrt(model_5_t6, red_model5_t6)

```
:::

### Semivariograms

```{r}
#| label: fig-variograms-t6
#| warning: false
#| fig-cap: 
#|   - "Semivariogram for Model 1 (Baseline Model)"
#|   - "Semivariogram for Model 2 (Row and column groups as fixed)"
#|   - "Semivariogram for Model 3 (AR1 in residuals)"
#|   - "Semivariogram for Model 4 (AR1 in residuals + nugget effect)"
#|   - "Semivariogram for Model 5 (AR1 in residuals + nugget effect + linear trend)"

plot(varioGram(model_1_t6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_2_t6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_3_t6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_4_t6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

plot(varioGram(model_5_t6, type = "default", spatial = "trend",
formula = ~NULL, composite = TRUE, model = c("empirical"),
metric = c("euclidean"), angle = 0, angle.tol = 180,
nlag = 20, maxdist = 0.5, xlag = NA, lag.tol = 0.5, grid = TRUE))

```

### AIC {#sec-AIC6}

```{r}
#| output: false
#| warning: false

#AIC comparison

modsel_t6 = icREML(list(model_1_t6, model_2_t6, model_3_t6, model_4_t6, model_5_t6)); modsel_t6
modsel_t6 <- as.data.frame(modsel_t6)

```

::: panel-tabset
### Model 1

```{r}
icREML(list(model_1_t6))
```

### Model 2

```{r}
icREML(list(model_2_t6))
```

### Model 3

```{r}
icREML(list(model_3_t6))
```

### Model 4

```{r}
icREML(list(model_4_t6))
```

### Model 5

```{r}
icREML(list(model_5_t6))
```
:::

### Spatial Distribution of Residuals

```{r}
#| label: fig-resplot-t6
#| fig-cap: "Spatial Distribution of residuals (Trial 6)"
resid_long <- t6 %>%
  mutate(
    row_number = as.numeric(as.character(row_number)),
    col_number = as.numeric(as.character(col_number)),
    Model_1 = residuals(model_1_t6, type = "response"),
    Model_2 = residuals(model_2_t6, type = "response"),
    Model_3 = residuals(model_3_t6, type = "response"),
    Model_4 = residuals(model_4_t6, type = "response"),
    Model_5 = residuals(model_5_t6, type = "response"))  |> 
  pivot_longer(
    cols = starts_with("Model_"),
    names_to = "Model",
    values_to = "Residual")

lim <- max(abs(resid_long$Residual), na.rm = TRUE)

ggplot(
  resid_long,
  aes(
    x = col_number,
    y = row_number,
    fill = Residual
  )
) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
  low = "#2166AC",
  mid = "#F7F7F7",
  high = "#B35806",
  midpoint = 0,
  limits = c(-lim, lim)) +
  coord_fixed() +
  scale_y_reverse() +
  facet_wrap(~Model) +
  theme_minimal() +
  labs(
    title = "Spatial distribution of residuals",
    x = "Column",
    y = "Row"
  )
```

### Best Model and Ranking

```{r}
#| label: t6-blups
#| results: hide

#BLUPs
blup1_t6 <- predict.asreml(model_1_t6, classify = "geno.cod", sed = TRUE)
blup2_t6 <- predict.asreml(model_2_t6, classify = "geno.cod", sed = TRUE)
blup3_t6 <- predict.asreml(model_3_t6, classify = "geno.cod", sed = TRUE)
blup4_t6 <- predict.asreml(model_4_t6, classify = "geno.cod", sed = TRUE)
blup5_t6 <- predict.asreml(model_5_t6, classify = "geno.cod", sed = TRUE)

blup_t6 <- list(blup1_t6, blup2_t6, blup3_t6, blup4_t6, blup5_t6)
predic_value_t6 <- list()

for (i in 1:length(blup_t6)) {
  
  predic_value_t6[[i]] <- blup_t6[[i]]$pvals |> 
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))

}

#blues - average of all checks together
blues_checks1_t6 <- predict(model_1_t6, classify = "check.cod")$pvals
blues_checks2_t6 <- predict(model_2_t6, classify = "check.cod")$pvals
blues_checks3_t6 <- predict(model_3_t6, classify = "check.cod")$pvals
blues_checks4_t6 <- predict(model_4_t6, classify = "check.cod")$pvals
blues_checks5_t6 <- predict(model_5_t6, classify = "check.cod")$pvals

blue_t6 <- list(blues_checks1_t6, blues_checks2_t6, blues_checks3_t6, blues_checks4_t6, blues_checks5_t6)
bpredic_value_t6 <- list()

for (i in 1:length(blue_t6)) {
  
  bpredic_value_t6[[i]] <- blue_t6[[i]] |> 
        dplyr::filter(check.cod != "Candidato") |>
    dplyr::select(predicted.value, dplyr::everything()) |> 
    dplyr::arrange(desc(predicted.value))  
}

```

```{r}
#| label: fig-h2-t6
#| fig-cap: "BLUPs ranking correlation (Trial 6)"

# Selecting baseline model and best model
t6_comp <- blup_comparison(blup1_t6$pvals, blup3_t6$pvals, m1_name = "Model 1", m2_name = "Best Model")
t6_comp$ranking_plot
```

### Broad-sense Heritability

```{r}
#| label: tbl-heritability-t6
#| tbl-cap: Broad-sense Heritability (Cullis, 2006) of the different models to spatial correction

# Heritabilities - Generalized heritability
h2_mod1_t6 <- 1-(mean(blup1_t6$sed[upper.tri(blup1_t6$sed)]^2)/(2*varcomp_m1_t6[6,1]))
h2_mod2_t6 <- 1-(mean(blup2_t6$sed[upper.tri(blup2_t6$sed)]^2)/(2*varcomp_m2_t6[3,1]))
h2_mod3_t6 <- 1-(mean(blup3_t6$sed[upper.tri(blup3_t6$sed)]^2)/(2*varcomp_m3_t6[6,1]))
h2_mod4_t6 <- 1-(mean(blup4_t6$sed[upper.tri(blup4_t6$sed)]^2)/(2*varcomp_m4_t6[6,1]))
h2_mod5_t6 <- 1-(mean(blup5_t6$sed[upper.tri(blup5_t6$sed)]^2)/(2*varcomp_m5_t6[6,1]))

h2_t6 <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  H2 = c(h2_mod1_t6, h2_mod2_t6, h2_mod3_t6, h2_mod4_t6, h2_mod5_t6))

options(knitr.table.format = "html")

h2_t6 |> 
  kbl(caption = " ", digits = 2) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
```

# Discussion

Across all trial–model combinations analyzed, both the check and
candidate genotype effects were significant. For the check effects
(fixed), Wald test results indicated that at least two check means
differed significantly. Regarding the candidate genotype effects
(random), LRT tests indicated that the variance component associated
with this effect was significantly greater than zero in all cases,
confirming that genetic variance among the genotypes evaluated exists in
the trials.

Now, we will discuss about the spatial corrections within trials.

### Trial 1

When comparing the adjusted AIC values among the five models, the lowest
value was observed for Model 4 (7978.3), compared with 8095.3 for Model
1 (no spatial correction) @sec-AIC1 .

These results indicate that, in this trial, the inclusion of an AR1
residual correlation structure improved the model relative to the
uncorrected model, however, the semivariogram still displayed a lot of
scattered peaks @fig-variograms-t1-3 . With the addition of a nugget
effect (Model 4), the semivariogram surface became smoother and the
range of the z-axis decreased, indicating an improvement in the spatial
correction

The spatial distribution of residuals across plots did not reveal an
evident systematic pattern and the spatial correction structures reduced
the magnitude of the residuals relative to the uncorrected model
@fig-resplot-t1 , which was also reflected in an increase in the
estimated heritability, since residual variance decreased
@tbl-heritability-t1

When comparing the genotype BLUP rankings between the uncorrected model
and the best-fitting model (Model 4, selected based on AIC and
semivariogram inspection), the Spearman rank correlation was 0.9105,
indicating the presence of reordering of genotypes, particularly among
genotypes of intermediate rank @fig-h2-t1.

As shown in the corresponding figure, genotypes at the extremes of the
ranking distribution were the least affected by re-ranking (the diagonal
of the plot) .

### Trial 2

For the second trial, the lowest adjusted AIC was also obtained for
Model 4 (8684.8), however, the reduction relative to the uncorrected
model (8695.9) was minimal. Models 3 and 5 even showed an increase in
AIC, indicating that the additional parameters included in those models
did not sufficiently account for the residual variation @sec-AIC2 .

Although Model 4 achieved the lowest AIC in this trial, its
semivariogram revealed a more pronounced residual pattern in the central
area of the experiment @fig-variograms-t2-4

As shown in @fig-variograms-t2-5 and @fig-resplot-t2 , the more heavily
parameterized spatial models reduced the overall magnitude of the
residuals but, in doing so, made an underlying residual spatial pattern
more apparent, with larger residual values concentrated in the central
region of the trial. This pattern could potentially be addressed through
the use of other residual covariance structures.

Given that the variance components @sec-VC2 , heritability estimates
@tbl-heritability-t2 and BLUP rankings @fig-h2-t2 were highly similar
between Models 1 and 4, these two models can be considered to perform
comparably in this trial.

### Trial 3

For the third trial, the lowest adjusted AIC was obtained for Model 4
(8682.9), compared with 8694.1 for the uncorrected Model 1. Although the
magnitude of the AIC reduction was larger than that observed in Trial 2,
the overall pattern was similar @sec-AIC3 .

Inspection of the semivariograms @fig-variograms-t3-3 and residual plots
@fig-resplot-t3 also revealed a more pronounced error pattern in the
central area of the trial. Because this trial was conducted at the same
location as Trial 2, this pattern may be attributable to an uncontrolled
factor related to soil fertility, water availability or other management
factor, that is not accounted for by the current model. Alternative
spatial correction structures might further improve model fit.

As in Trial 2, the variance components (genetic and residual) @sec-VC3,
heritability estimates @tbl-heritability-t3, and ranking correlations
@fig-h2-t3 were highly similar between Models 1 and 4, indicating
comparable performance between these models.

### Trial 4

In this trial, Model 4 produced the largest AIC reduction among all
models evaluated (6910.6 vs. 6972.3 for the uncorrected model). As in
Trial 1, the inclusion of the AR1 residual structure improved the AIC
relative to the uncorrected model @sec-AIC4, however, the semivariogram
still showed numerous scattered peaks @fig-variograms-t4-3 .

With the addition of the nugget effect (Model 4), the semivariogram
surface became smoother and the range of the z-axis decreased, again
indicating an improvement in the spatial correction @fig-variograms-t4-4
.

As shown in the corresponding figure, Model 4 reduced the magnitude of
the residuals without generating an evident spatial dispersion pattern.
This reduction in residual variance also led to higher heritability
estimates @tbl-heritability-t4 .

Considering Model 4 the best-performing model in this trial, some
changes in genotype rankings were observed relative to the uncorrected
model; however, genotypes with the most extreme performance retained
their rankings, particularly KML942, which was also the top-ranked
genotype in Trial 1 @fig-h2-t4 .

### Trial 5

For this trial, Model 5 produced the lowest adjusted AIC among all
models tested (5445.0 vs. 5483.7 for the uncorrected model) @sec-AIC5 .
The semivariogram of Model 5 showed a reduction in the range of z-axis
values, indicating an improved spatial fit comparing to Model 3
@fig-variograms-t5-5 .

Analysis of the variance components showed that the spatial correction
reduced the magnitude of the residual variance component @sec-VC5 . In
Model 5, the estimated spatial correlation parameters for rows and
columns, $\rho_{row} = 0.76$ and $\rho_{col} = -0.26$, indicate a
stronger correlation among residuals along rows than along columns. With
this spatial correction, other variance components (such as the colgroup
and rowgroup effects) were estimated as zero, probably because the
spatial structure was able to capture the variation previously
attributed to these terms.

Heritability estimates also increased with the spatial correction,
indicating that Model 5 may be the most appropriate model for this trial
@tbl-heritability-t5 .

### Trial 6

In this trial, the lowest adjusted AIC was obtained for Model 3 (3565.0
vs. 3574.2 for the uncorrected model) @sec-AIC6 .

Note that Model 3 only includes an AR1 residual correlation structure,
and the addition of a nugget effect (Model 4) did not improve the model
fit, as the variance component corresponding to the nugget effect was
estimated as null @sec-VC6 .

The semivariograms of the different models also showed a very similar
pattern, possibly because the AR1 structure had already captured most of
the spatially structured variation present in the residuals
@fig-variograms-t6-4 .

The estimated correlation parameters for rows and columns in Model 3,
$\rho_{row} = 0.19$ and $\rho_{col} = 0.04$ indicate a very weak spatial
correlation, comparing to the values observed in the previous trials
@sec-VC6 .

This result, together with the similarity in heritability estimates
@tbl-heritability-t6 and residual plots between models @fig-resplot-t6 ,
indicates that the spatial corrections tested did not substantially
change the magnitude of the residuals. Other covariance structures could
be explored in future analyses, however, considering only the structures
tested here and selecting the model with the lowest AIC (Model 3), no
major changes in genotype ranking were observed relative to the
uncorrected model @fig-h2-t6 . Overall, model performance was very
similar across models in this trial.
